---
title: "Bayes Theorem for Distributions"
output: html_notebook
---

```{r load_pkgs}
library(dplyr)
```

Suppose that we have 600 coins of the following types:

* 300 fair coins (equal chance of Heads and Tails 1:1)

* 200 slightly unfair coins (heads:tails with odds 2:1)

* 100 very unfair coins (heads:tails with odds 5:1)

If we take a coin at random from the 600 coins, flip it 10 times and observe 8 heads, what is the probability that the coin we've flipped is a fair (1:1) coin?

$$\begin{array}{lcl}Pr\Big[\text{fair coin} \space{|}\space \text{got 8 heads in 10 tosses}\Big] &=& \frac{\text{number of the 300 fair (1:1) coins on which we'd expect to get 8 heads in 10 tosses if we flipped all 300 (ten flips on each coin)}}{\text{total number of coins on which we'd expect to get 8 heads in 10 tosses (across all coin types) if we tossed all 600 (ten flips on each coin)}} \\
&=& \frac{\text{(probability of 8 heads in 10 tosses for a fair coin)}\times 300}{\text{(probability of 8 heads in 10 tosses for a fair coin)}\times 300 + \text{(probability of 8 heads in 10 tosses for an unfair 1:2 coin)}\times 200 + \text{(probability of 8 heads in 10 tosses for a very unfair 1:5 coin)}\times 100} \\
&=& \frac{\binom{10}{8}(0.5)^8(0.5)^2 \space\times\space 300}{\binom{10}{8}(0.5)^8(0.5)^2 \space\times\space 300 + \binom{10}{8}(\frac{2}{3})^8(\frac{1}{3})^2 \space\times\space 200 + \binom{10}{8}(\frac{5}{6})^8(\frac{1}{6})^2 \space\times\space 100} \hspace{20mm} \text{using the binomial distribution} \\
&=& 0.1622136 \\
\end{array}$$

```{r showtheresult}
# calculation in R
choose(10,8)*(0.5)^8*(0.5)^2*300 / ( choose(10,8)*(0.5)^8*(0.5)^2*300+choose(10,8)*(2/3)^8*(1/3)^2*200+choose(10,8)*(5/6)^8*(1/6)^2*100 ) 
```

```{r showtheresult_alt}
# calculation in R (alternate)
dbinom(x=8,size=10,prob=0.5)*300 / ( dbinom(x=8,size=10,prob=0.5)*300 + dbinom(x=8,size=10,prob=2/3)*200 + dbinom(x=8,size=10,prob=5/6)*100 )
```

We can verify this by simulation:

```{r verify_q1_by_simulation}
system.time(
run_sims <-   
  sapply( 1:1e5L,
          function(x){
          (
             rbind(
             # flip 30 000 fair 1:1 coins:
             cbind( coin_type = 1, 
                    num_heads = sapply( 1:300,
                                        function(x){ sum( sample(c(1,0),size=10,prob=c(1,1), replace=TRUE) ) }
                                      )
                  )
             ,
             # flip 20 000 unfair 2:1 coins
             cbind( coin_type = 2, 
                    num_heads = sapply( 1:200,
                                        function(x){ sum( sample(c(1,0),size=10,prob=c(2,1), replace=TRUE) ) }
                                      )
                  )
             ,
             # flip 10 000 very unfair 5:1 coins
             cbind( coin_type = 3, 
                    num_heads = sapply( 1:100,
                                        function(x){ sum( sample(c(1,0),size=10,prob=c(5,1), replace=TRUE) ) }
                                      )
                  )
             ) 
          ) %>% 
             # only keep coins which tossed 8 heads, and store their coin_type 
             .[.[,2]==8,1] %>% 
             sample(., size=1)      # sample at random one of the coins
        }
)
)

table(run_sims)

table(run_sims)["1"] / length(run_sims)
```

We can see that this is the same result as if we had directly applied Bayes rule:

$$Pr\Big[A \space | B\Big] \quad = \quad \frac{Pr\Big[B \space | \space A \space \Big] \space \space Pr\Big[A\space \Big]}{\underset{i}{\sum} Pr\Big[B \space | \space A_i \space \Big] \space \space Pr\Big[A_i \space \Big] } \hspace{15mm}\text{where the denominator is summing over all possible values of variable } A$$

We can frame our coin example in terms of distributions instead: 

For each individual coin, the distribution of 

$$X \quad = \quad \text{number of heads observed in 10 tosses of a coin}$$

is 

$$Pr\Big[X=x \space | \space \text{coin type}\Big] \quad = \quad 
\begin{cases}
\binom{10}{x}(0.5)^x(0.5)^{10-x} & \text{ if coin is a fair coin (1:1)}\\
\\
\binom{10}{x}(\frac{2}{3})^x(\frac{1}{3})^{10-x} & \text{ if coin is an unfair coin (2:1)}\\
\\
\binom{10}{x}(\frac{5}{6})^x(\frac{1}{6})^{10-x} & \text{ if coin is a very unfair coin (5:1)}\\
\end{cases}
\\$$

If we consider the coin_type as a parameter $\beta$ of the distribution of $X$, taking on a value in $\{1,2,3\}$:

$$\beta \quad = \quad 
\begin{cases}
1 & \text{if coin is a fair coin (1:1)}\\
2 & \text{if coin is an unfair coin (2:1)}\\
3 & \text{if coin is a very unfair coin (5:1)}\\
\end{cases}$$


then we can write the distribution of $X = \text{number of heads observed in 10 tosses}$ as

$$Pr\Big[X=x \space | \space \beta \Big] \quad = \quad \mathbf{I}_1(\beta)\space\binom{10}{x}(0.5)^x(0.5)^{10-x} + \mathbf{I}_2(\beta)\space\binom{10}{x}(\frac{2}{3})^x(\frac{1}{3})^{10-x} + \mathbf{I}_3\binom{10}{x}(\frac{5}{6})^x(\frac{1}{6})^{10-x}  $$

where $\mathbf{I}_b(\beta)$ is the indicator function

$$\mathbf{I}_b(\beta) \quad = \quad 
\begin{cases} 1 & \text{if } \beta = b \\
              0 & \text{if } \beta \neq b \\     
\end{cases}$$



