---
title: 'Confidence Interval for a Proportion: Bayesian vs. Frequentist'
author: 'Joseph Bolton'
output:
  html_document:
    df_print: paged
---

This post is about statistical methods for comparing proportions between two groups. For example, the proportion of people surviving in a treatment group vs. surviving in a control group, or the proportion of customers subjected to targeted advertising who shopped vs. the proportion of customers who shopped without being targeted. 

Here are the packages that I am going to use:

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)           # for general data manipulation
library(HDInterval)          # for calculating Highest Density Regions for posterior distributions
library(DescTools)           # for the function BinomDiffCI()
library(patchwork)           # for arranging ggplots
library(binom)               # for generating confidence intervals for a 1-sample proportion
```

I admit that this post is heavier on Bayesian content. This is because the Frequentist methods are conceptually simpler in construction, although not necessarily in interpretation. 

# Random Variables
A [random variable](http://www.stat.yale.edu/Courses/1997-98/101/ranvar.htm) is a quantity whose value is governed by a random system or process. Some values are more likely to occur than others, and this information is encoded in a [probability distribution](https://en.wikipedia.org/wiki/Probability_distribution).    

# Quantifying Confidence in a Sample Proportion (an example)
Does drug A or drug B work better? 

Suppose that if we were to extract a random sick person from the general population and give them drug A, that the probability that this person would recover from their sickness is $\pi_A$. 
$\pi_A$ is the true proportion of sick people treated with drug A who recover from their sickness.

If we were to extract a random sick person from the general population and give them drug B, then the probability that we'd see that person recover from their sickness is $\pi_B$. 

$\pi_B$, then, is the true proportion of sick people treated with drug B that overcome their sickness.

Suppose that we don't known the true value of $\pi_A$, nor the true value of $\pi_B$ (how could anyone ever know something like this?). 

In the interest of learning something about $\pi_A$ and $\pi_B$, we do the following:

1. We give 100 random sick people drug A. We observe that 90 of them recover. This data gives us a point estimate $\hspace{3mm} \hat\pi_A=0.9 \hspace{3mm}$ of $\pi_A$.

2. We give 100 random sick people drug B, and 95 of them recover. This gives us a point estimate $\hspace{3mm} \hat\pi_B=0.95 \hspace{3mm}$ of $\pi_B$.

Apart from the absolute effects of drug A and drug B, we are also interested in their relative effects (i.e. we want to compare drug A to drug B). $\hat\pi_A-\hat\pi_B=-0.05$ gives us a point estimate of the true difference $\hspace{2mm}(\pi_A-\pi_B)\hspace{2mm}$ between the proportion of people who survive when given drug A vs. the proportion of people who survive if given drug B.  

How can we objectively quantify our uncertainty in these estimates of $\pi_A$, $\pi_B$ and $(\pi_A-\pi_B)$? 

# Confidence in a Single Sample Proportion: Bayesian Approach
The basic logic behind the Bayesian approach is this:

* $\pi_A$ is the true proportion of sick people on drug A who recover in the population. Not knowing this true value of $\pi_A$, we consider it a random variable. We quantify our uncertainty in possible values of this variable using a probability distribution (i.e. the shape of the distribution describes which values we think are more likely, and which values we think are less likely, to be the true value). 

* If we draw a random sick person from the population and give them drug A, then the probability that we would see that person recover is $\pi_A$. If we take a random sample of 100 sick people and give them drug A, then $X$ of them will recover. $X$, then, is also a random variable. The distribution of $X$ is clearly highly dependent on the value of random variable $\pi_A$. Our experiment, in which we gave 100 people drug A and 90 people recovered, is a single realisation $x=90$ of the random variable $X$ (a random draw from the probability distribution of $X$).

* The distribution of the random variables $\hspace{2mm}\{X,\pi_A\}\hspace{2mm}$ is described by a joint/bivariate probability function, representing the fact that some combinations of $X$ and $\pi_A$ are more likely than others.  

* Using results from conditional probability ([Bayes Theorem](https://www.youtube.com/watch?v=HZGCoVF3YvM)), we can calculate the [conditional distribution](https://en.wikipedia.org/wiki/Conditional_probability_distribution) of $\pi_A$ given that the outcome ($x$) of the random variable $X$ has been observed:

$$\overset{posterior}{\overbrace{f\Big(\pi_A \space\Bigl|X=x\Big)}} 
\quad=\quad 
\displaystyle\frac{\overset{likelihood}{\overbrace{Pr\Big[X=x\space\Bigl|\pi_A\Big]}}\cdot \overset{prior}{\overbrace{f(\pi_A)}}}{\underset{\pi_A}{\displaystyle\int} Pr\Big[X=x\space\Bigl|\pi_A\Big]\cdot f(\pi_A) \hspace{2mm} d\pi_A} $$

This means that if we are comfortable specifying a marginal distribution (*prior*) $f(\pi_A)$ for $\pi_A$ (distribution of $\pi_A$ ignoring $X$) and specifying the form of the conditional distribution $Pr\Big[X=x\Bigl|\pi_A\Big]$ of $X$, then we get "**the conditional distribution of $\pi_A$ given $X$**" (the distribution of parameter of interest $\pi_A$, having taken into account the evidence provided by our data $X$).

This result is consistent with the fact that observing $X$ gives us additional information about $\pi_A$, which should cause us to update our believes about the possible values that $\pi_A$ could take. 

Assuming a form for the conditional distribution of our data $X\Bigl|\pi_A$ (the *likelihood*) is easy: $\hspace{3mm}$ we assume that if we draw a sick person at random from the population and give them drug A, then the probability that we will see that person recover is $\pi_A$. This gives the following distribution for $X\Bigl|\pi_A$ in our random sample of 100 people:

$$\begin{array}{lcl} X \space \Bigl| \space \pi_A &\sim& \text{Binomial}\Big(n=100, \quad prob=\pi_A\Big) \\
Pr\Big[X=x \space\Bigl|\space\pi_A \Big] &=& \displaystyle\binom{100}{x} \cdot (\pi_A)^{x} \cdot (1-\pi_A)^{100-x} \\
\end{array}$$

See https://en.wikipedia.org/wiki/Binomial_distribution for further details on this distribution.

This is not to say that a particular randomly drawn person given drug A will recover with probability of $\pi_A$ (the particular person drawn could be strong as an ox), but rather that the proportion of randomly drawn out people given drug A that recover is $\pi_A$. 

The choice of the marginal (*prior*) distribution $f(\pi_A)$ is more subjective, and a point of some contention. This distribution is used to quantify our degree of belief in different possible values of $\pi_A$ before having observed our data $X$.

For our data, suppose that we choose the following **prior distribution** for $\pi_A$:

```{r echo=FALSE}
x_seq <- seq( from=0, to=1, by=0.01 )
plot(    x = x_seq, 
         y = dbeta(x=x_seq, shape1=5, shape2=1), 
      type ="l", 
      main = bquote("Beta distribution with parameters"~alpha~"= 5, "~beta~"= 1"),
      xlab = bquote(pi[A]),
      ylab = bquote("f ("*pi[A]*")"),
      ylim = c(0,5)
    )
axis( side=1, at=seq(0,1,0.1) )
abline( v = seq(0,1,0.1), col="grey" )
```

Notice that we are specifying that we believe that higher values of $\pi_A$ are much more likely than lower values - we don't believe that either drug A or the disease is likely to decimate the population. The choice of prior should in practice be justified by evidence. [Empirical Bayesian methods](http://varianceexplained.org/r/empirical_bayes_baseball/) use data to estimate a prior distribution.  

The choice of a **beta** distribution for the **prior distribution** here is for mathematical convenience (see https://en.wikipedia.org/wiki/Conjugate_prior for details).

Combining this choice of prior $f(\pi_A)$ with our binomial likelihood $Pr\Big[X=x \space\Bigl|\space\pi_A \Big]$ gives us our **posterior distribution** of $\pi_A$:

$$\begin{array}{lcl} f\Big(\pi_A \space\Bigl|X=90\Big) 
&=& \text{Beta}\Big(\alpha=5+90, \quad \beta=1+100-90\Big) \\
&=& \text{Beta}\Big(\alpha=95, \quad \beta=11\Big) \\
\end{array}$$

For details on this calculation, see https://stats.stackexchange.com/questions/70661/how-does-the-beta-prior-affect-the-posterior-under-a-binomial-likelihood.

Here are plots of these **prior** and **posterior** distributions:

```{r echo=FALSE, fig.width=6, fig.height=6}
par( mfrow=c(2,1) )
x_seq <- seq(from=0,to=1,by=0.01)
plot(    x = x_seq, 
         y = dbeta(x=x_seq, shape1=5, shape2=1), 
      type ="l", 
      main = bquote("Prior (Beta) Distribution with parameters"~alpha~"= 5, "~beta~"= 1"),
      xlab = bquote(pi[A]),
      ylab = bquote("f ("*pi[A]*")"),
      ylim = c(0,5),
       las = 1
    )
axis( side=1, at=seq(0,1,0.1) )
abline( v = seq(0,1,0.1), col="grey" )
abline( v = 0.9, lwd=2.5, col=2 )
legend( "topleft",
        legend = "Observed proportion of people recovering",
        col = 2,
        lty = 1,
        lwd = 2.5
      )

## POSTERIOR ##
plot(    x = x_seq, 
         y = dbeta(x=x_seq, shape1=95, shape2=11), 
      type ="l", 
      main = bquote("Posterior (Beta) Distribution with parameters"~alpha~"= 95, "~beta~"= 11"),
      xlab = bquote(pi[A]),
      ylab = bquote("f ("*pi[A]*"|X)"),
       las = 1
    )
axis( side=1, at=seq(0,1,0.1) )
abline( v = seq(0,1,0.1), col="grey" )
abline( v = 0.9, lwd=2.5, col=2 )
legend( "topleft",
        legend = "Observed proportion of people recovering",
        col = 2,
        lty = 1,
        lwd = 2.5
      )
```

Here is another way to think about this:

1. Suppose that we draw out a million values of $\pi_A$ from the prior distribution $\text{beta}\Big(5,1\Big)$:

```{r}
pi_A_sample <- rbeta( n=1e6L, shape1=5, shape2=1 )
hist(      x = pi_A_sample, 
      breaks = 70,
        main = bquote( "1,000,000 simulated values of"~pi~"from prior distribution Beta(5,1)"),
        xlab = bquote(pi)
    )
```

2. Now, suppose that for each value $\quad \pi_A^{(i)}: \hspace{4mm} i=1,...,1\hspace{1mm}000\hspace{1mm}000 \quad$ of the million $\pi_A$ values, we simulate a value $X_i$ from the distribution $X_i \sim \text{Binomial}\Big(n=100, \space prob=\pi_A^{(i)}\Big)$:

```{r fig.width=8, fig.height=4}
binomial_data_sample <- 
  rbinom(    n = length(pi_A_sample), 
          size = 100, 
          prob = pi_A_sample 
        )

barplot(         table(binomial_data_sample),
          xlab = "X",
          ylab = "Frequency",
          main = "Distribution of Number of People Recovered in 1,000,000 binomial samples"
        )
```

3. Now, we throw away all of the values $\pi_A^{(i)}$ which led to $X_i=90$ (this is the vast majority of them). We now consider the distribution of only those $\pi_A^{(i)}$ values which led to samples $X_i=90$ (this happened in `r sum(binomial_data_sample==90)` out of the million binomial values):

```{r echo=FALSE}
posterior_pi_A_sample <- pi_A_sample[ binomial_data_sample == 90 ]

hist(      x = posterior_pi_A_sample, 
      breaks = 50,
        freq = FALSE,
        xlab = bquote(pi),
        main = bquote( "Distribution of "~pi[A]~"over samples with exactly 90 recoveries" )
    )

curve( dbeta( x, shape1=95, shape2=11),
       add = TRUE,
       col = 2
      )

legend(          "topleft",
        legend = "Theoretical Posterior Distribution",
           lty = 1,
           col = 2
      )
```

This is the **posterior distribution**.

So, the **posterior disribution** is the distribution of the $\pi_A$ values drawn from prior distribution $f(\pi_A)$ which gave rise to a dataset identical to our observed data $X$.  

Going back to our original question: $\hspace{3mm}$ we can use this posterior probability distribution (density) of $\pi_A$ to generate a $100(1-\alpha)\%$ *credibility interval* for $\pi_A$ i.e. an interval containing $\pi_A$ with probability $(1-\alpha)$.

We calculate a 90% **Highest Density Interval** (HDI) using the R package **HDInterval**. The 90% **HDI** is the narrowest credibility interval (smallest range of $\pi_A$ values) containing 90% of the density of the posterior distribution:  

```{r}
calculate_hpd_0.9 <- 
  HDInterval::hdi( qbeta,
                   credMass = 0.9,
                     shape1 = 95,
                     shape2 = 11
                 )

c( calculate_hpd_0.9 )
```

Check that the interval truly contains 90% of the density (a sense check):

```{r}
pbeta( q = calculate_hpd_0.9["upper"], 95, 11 ) - pbeta( q = calculate_hpd_0.9["lower"], 95, 11 )   
```

Here is a plot of the 90% HDI: 

```{r echo=FALSE}
# plot the HPD interval:
coord_x <- seq( 0, 1, by=0.0001 )
coord_y <- dbeta( x = coord_x, shape1=95, shape2=11 )
coord_y <- ifelse( coord_x>=calculate_hpd_0.9[1] & coord_x<=calculate_hpd_0.9[2],
                   coord_y,
                   0
                 )
curve( dbeta(x, 95, 11), 
       xlim=c(0.6, 1), 
       xlab=expression(pi[A]), 
       ylab=bquote("f ("*pi[A]*"|X)"),
       main = bquote( "90% Credibility Interval (HDI) for"~pi[A] )
     )
polygon( coord_x, coord_y, col='skyblue')
axis(   side = 1,
          at = calculate_hpd_0.9,
      labels = c("", ""),
         lwd = 3 
    )
curve( dbeta(x, 95, 11), 
       xlim=c(0.6, 1), 
       xlab=expression(pi[A]), 
       ylab=bquote("f ("*pi[A]*")"),
       main = bquote( "90% Credibility Interval (HDI) for"~pi[A] ),
       col = "red",
       lwd = 2,
       add = TRUE
     )
legend( "topleft",
        lty = 1,
        lwd = 2,
        col = 2,
        legend = bquote( "Posterior Distribution of "*pi )
      )
```

Using the same prior distribution for $\pi_B$, we can in the same way get a 90% credibility interval for $\pi_B$:

```{r echo=FALSE}
prior <- c()
prior["alpha"] <- 5
prior["beta"] <- 1

n <- 100

x <- c()
x["drug A"] <- 90
x["drug B"] <- 95

drugA_posterior <- c()
drugA_posterior["alpha"] <- prior["alpha"] + x["drug A"]
drugA_posterior["beta"] <- prior["beta"] + n - x["drug A"]

drugB_posterior <- c()
drugB_posterior["alpha"] <- prior["alpha"] + x["drug B"]
drugB_posterior["beta"] <- prior["beta"] + n - x["drug B"]
 
drug_A_confint_hdi_0.9 <- 
  HDInterval::hdi( qbeta,
                   credMass = 0.9,
                     shape1 = drugA_posterior["alpha"],
                     shape2 = drugA_posterior["beta"]
                 )

drug_B_confint_hdi_0.9 <- 
  HDInterval::hdi( qbeta,
                   credMass = 0.9,
                     shape1 = drugB_posterior["alpha"],
                     shape2 = drugB_posterior["beta"]
                 )

data_for_posterior_pi_A_ggplot <- 
  tibble( x = seq(0, 1, 0.001),
          y = dbeta(      x = seq(0, 1, 0.001), 
                     shape1 = drugA_posterior["alpha"],
                     shape2 = drugA_posterior["beta"]
                   )
        ) 
posterior_pi_A_ggplot <- 
  ggplot( data = data_for_posterior_pi_A_ggplot,
          aes( x = x,
               y = y
             )
        ) +
    geom_line() +
    geom_area( data = subset(  data_for_posterior_pi_A_ggplot, 
                               x>=drug_A_confint_hdi_0.9["lower"] &
                               x<=drug_A_confint_hdi_0.9["upper"]  
                             ),
               fill = "red",
               alpha = 0.5
             ) +
    coord_cartesian( xlim = c(0.75,1) ) +
    labs( title = bquote("Posterior Distribution of "*pi[A]*" given Observed Data "*X[A]*"=90"),
              x = bquote(pi[A]),
              y = bquote("f ("*pi[A]*"|"*X[A]*"=90)")
          )

data_for_posterior_pi_B_ggplot <- 
  tibble( x = seq(0, 1, 0.001),
          y = dbeta(      x = seq(0, 1, 0.001), 
                     shape1 = drugB_posterior["alpha"],
                     shape2 = drugB_posterior["beta"]
                   )
        ) 
posterior_pi_B_ggplot <- 
  ggplot( data = data_for_posterior_pi_B_ggplot,
          aes( x = x,
               y = y
             )
        ) +
    geom_line() +
    geom_area( data = subset(  data_for_posterior_pi_B_ggplot, 
                               x>=drug_B_confint_hdi_0.9["lower"] &
                               x<=drug_B_confint_hdi_0.9["upper"]  
                             ),
               fill = "blue",
               alpha = 0.5
             ) +
    coord_cartesian( xlim = c(0.75,1) ) +
    labs( title = bquote("Posterior Distribution of "*pi[B]*" given Observed Data "*X[B]*"=95"),
              x = bquote(pi[B]),
              y = bquote("f ("*pi[B]*"|"*X[B]*"=95)")
          )

errobar_plot <- 
  tibble( pi = c("A","B"),
          point_estimate = c(0.9, 0.95),
          lower = c( drug_A_confint_hdi_0.9["lower"],
                     drug_B_confint_hdi_0.9["lower"]
                   ),
          upper = c( drug_A_confint_hdi_0.9["upper"],
                     drug_B_confint_hdi_0.9["upper"]
                   )
        ) %>% 
    ggplot( data = .,
            aes( x = relevel( factor(pi), "B" ),
                 y = point_estimate,
                 colour = pi
               )
          ) + 
    geom_point( size=2, colour=c("red","blue") ) +
    geom_errorbar( aes(ymin=lower, ymax=upper), colour=c("red","blue") ) +
    coord_flip() +
    labs( y = bquote(pi[j]),
          x = "Drug",
          title = bquote("90% Bayesian Credibility Intervals for "*pi[A]~"and"~pi[B])
        ) +
    theme( legend.position = "none" ) +
    ylim( c(0.75,1) )

posterior_pi_A_ggplot / errobar_plot / posterior_pi_B_ggplot 


```

If we don't want to specify a prior distribution for $\pi_A$, we could just consider the relative likelihoods that our observed data was generated under different possible values of $\pi_A$:

```{r echo=FALSE}
tibble( pi_A = seq(0,1,0.01) ) %>% 
  mutate( unscaled_prob_X = map_dbl( .x = pi_A,
                                     .f = function(p){ dbinom(x=90, size=100, prob=p) } 
                                   )
  ) %>% 
  mutate( density_prob_X = unscaled_prob_X/sum(unscaled_prob_X) ) %>% 
  # summarise( sum(density_prob_X) )
  ggplot( data = .,
          aes( x = pi_A,
               y = density_prob_X
             )
        ) +
  geom_line() +
  labs( title = 
          bquote("Likelihood (probability) of Observed Data ("*X[A]*"=90) for Different Values of "*pi[A] ),
            x = bquote(pi[A]),
            y = "Density"
  )
```

We could then proceed to use this distribution as the **posterior distribution** (for calculating credibility intervals etc.). 

We would obtain an identical **posterior distribution** to this under the Bayesian approach if we specified the prior distribution to be $f(\pi_A)\propto 1$. This is known as an [uninformative prior](https://en.wikipedia.org/wiki/Prior_probability#Uninformative_priors).

# Confidence in a Single Sample Proportion: Frequentist Approach
The frequentist approach to this problem is to choose a method for creating confidence intervals that we know will, if we use the same method many times, capture the true value of $\pi_A$ within the interval 100($1-\alpha$)% of the time. $\alpha$ is a parameter that we choose: smaller values of $\alpha$ result in intervals that are correct more often but give wider intervals.

Foe example, if we draw a whole lot of random samples, each sample containing 100 people, from the population of sick people, give them drug A, and calculate the following confidence interval for each of the samples:

$$\Bigg[\underset{\quad lower}{\underbrace{\displaystyle\frac{1}{1+\displaystyle\frac{n-x+1}{x}F_{2(n-x+1), 2x, \frac{\alpha}{2}}}}} \quad;\quad \underset{upper}{\underbrace{\displaystyle\frac{\displaystyle\frac{x+1}{n-x}F_{2(x+1),2(n-x),\frac{\alpha}{2}}}{1 + \displaystyle\frac{x+1}{n-x}F_{2(x+1),2(n-x),\frac{\alpha}{2}}}}} \quad\Bigg]$$

$$\begin{array}{lcl}
n &=& \text{number of samples (people)} \\
  &=& 100 \\
x &=& \text{number of people recovering} \\
\end{array}$$

..then this interval $\Big[lower \space ; upper \Big]$ will contain the true value of $\pi_A$ in **at least** $100(1-\alpha)\%$ of the samples. 

See [exact binomial confidence intervals](https://www.researchgate.net/publication/255630317_Exact_Binomial_Confidence_Interval_for_Proportions) for more details. 

So, using this method, we know that our confidence intervals will **not contain** the true value of $\pi_A$ in some small number of samples ($100(\alpha)\%$ of them). We can decrease this value of $\alpha$, to be safe, but this will give us wider confidence intervals. 

There are many different types of frequentist confidence intervals for a proportion ($\pi_A$), each with different properties (interval length, time to compute, suitability to different sample sizes and true proportions $\pi_A$, location of interval relative to true proportion $\pi_A$ etc.). See [Two‐sided confidence intervals for the single proportion: comparison of seven methods](https://onlinelibrary.wiley.com/doi/abs/10.1002/%28SICI%291097-0258%2819980430%2917%3A8%3C857%3A%3AAID-SIM777%3E3.0.CO%3B2-E) and [Binomial confidence intervals and contingency tests](https://www.ucl.ac.uk/english-usage/staff/sean/resources/binomialpoisson.pdf) for more details.

Here is an example of this method: 

We simulate 100,000 random samples, where each sample contains 100 sick people. For purposes of illustration, we specify that the true proportion of sick people on drug A who recover is $\pi_A=0.85$. We calculate the exact confidence interval discussed above, at a 90% significance level ($\alpha=0.1$), on each sample. We observe the fraction of samples in which the confidence interval contained the true value of $\pi_A$ ($\pi_A=0.85$).

The confidence interval method used is the one discussed at the beginning of this section (*Pearson-Klopper exact method*). 

```{r}
# function to generate exact binomial confidence intervals for a single proportion:
# this is the same as as the function binom::binom.confint( ..., method="exact" )
get_exact_binomial_confint <- function( n, x, alpha ){
  
    if( n == x ){ upper_limit <- 1 } else{
      upper_limit <- (      ((x+1)/(n-x)) * qf(df1=2*(x+1), df2=2*(n-x), p=1-alpha/2)   ) /
                     (  1 + ((x+1)/(n-x)) * qf(df1=2*(x+1), df2=2*(n-x), p=1-alpha/2)   )  
    }
  
    if( x == 0 ){ lower_limit <- 0 } else{
      lower_limit <- 1 / (   1 + ((n-x+1)/x) * qf( df1=2*(n-x+1), df2=2*x, p=1-alpha/2 )  )
    }
  
  return( 
          list( lower_confint = lower_limit,
                upper_confint = upper_limit
              ) 
        )
}

true_pi <- 0.85

# simulate the 100,000 random samples, each of size 100 
sim_x_values <- rbinom(    n = 1e5L,
                        size = 100,
                        prob = true_pi
                      )

# plot the distribution of number of people recovered in each sample:
barplot( table(sim_x_values),
         xlab = "Number Recovered (X)",
         ylab = "Frequency",
         main = "Distribution of Number of People Drug A who Recovered, in 100,000 Simulations"
        )

# calculate the exact confidence intervals for each of the samples:
calc_confint <- 
  tibble( n_recovered = sim_x_values ) %>% 
    mutate( point_estimate = sim_x_values / 100 ) %>% 
    mutate( get_confint = map( .x = n_recovered,
                               .f = function(x){ 
                                 get_exact_binomial_confint(     n = 100, 
                                                                 x = x,
                                                             alpha = 0.1
                                 )  
                                               }
                              )
    ) %>% 
    unnest_wider( col = get_confint ) %>% 
    mutate( contains_true_pi = 0.85 >= lower_confint & 
                                  0.85 <= upper_confint
          )

```

Here are the first 100 random samples plotted with their 90% confidence intervals:

```{r echo=FALSE}
ggplot( data = calc_confint %>% slice(1:100) %>% mutate(rownum = row_number()),
        aes(      x = rownum,
                  y = point_estimate,
             colour = contains_true_pi 
           )  
      ) +
  geom_point( size=1.5 ) +
  geom_linerange( aes(ymin=lower_confint, ymax=upper_confint) ) +
  geom_hline( yintercept = 0.85 ) +
  theme( legend.position = "none" ) +
  labs( x = "sample",
        y = bquote(hat(pi)),
        title = "90% Confidence Intervals for the First 100 Samples"
      )
```

The proportion of the 100,000 simulated samples for which the confidence interval contained the true value $\pi_A=0.85$: 

```{r}
sum(calc_confint$contains_true_pi) / nrow(calc_confint)
```

$\pi_A$ is being captured in the interval more than 90% of the time, which is what we wanted.

Going back to our original example, this gives us the following 90% confidence intervals for $\pi_A$ and $\pi_B$ (recall that our observed data was 90 recoveries out of 100 patients on drug A and 95 recoveries out of 100 patients treated with drug B):

```{r echo=FALSE}
cat( "      -- Drug A -- \n" )
get_exact_binomial_confint(     n = 100,
                                x = 90,
                            alpha = 0.1
                          ) %>% 
  unlist() %>% 
  round(., digits=3)
cat("\n")
cat("      -- Drug B -- \n")
get_exact_binomial_confint(     n = 100,
                                x = 95,
                            alpha = 0.1
                          ) %>% 
  unlist() %>% 
  round(., digits=3)

freqentist_exact_intervals_for_pi_A <- 
  get_exact_binomial_confint(     n = 100, 
                                  x = 90,
                              alpha = 0.1
                            )

freqentist_exact_intervals_for_pi_B <- 
  get_exact_binomial_confint(     n = 100, 
                                  x = 95,
                              alpha = 0.1
                            )

tibble( pi = c("A","B"),
        point_estimate = c(0.9, 0.95),
        lower = c(freqentist_exact_intervals_for_pi_A$lower_confint,
                  freqentist_exact_intervals_for_pi_B$lower_confint),
        upper = c(freqentist_exact_intervals_for_pi_A$upper_confint,
                  freqentist_exact_intervals_for_pi_B$upper_confint)  
      ) %>% 
  ggplot( data = .,
          aes( x = pi,
               y = point_estimate
             )
        ) + 
  geom_point( size=2 ) +
  geom_errorbar( aes( ymin=lower, ymax=upper) ) +
  coord_flip() +
  labs( y = bquote(pi[j]),
        x = "Drug",
        title = bquote("90% (Frequentist) Exact Binomial Confidence Intervals for "*pi[A]~"and"~pi[B])
      ) +
  scale_y_continuous( breaks=seq(0.83, 0.99, 0.01) )
```

Methods for many different sorts of confidence intervals for a single population proportion are available in the R package [binom](https://cran.r-project.org/web/packages/binom/index.html).

Here is a comparison of the Bayesian and Frequentist confidence intervals for $\pi_A$ and $\pi_B$:

```{r echo=FALSE}

bind_rows( 
  
  tibble( pi = c("A","B"),
        point_estimate = c(0.9, 0.95),
        lower = c(freqentist_exact_intervals_for_pi_A$lower_confint,
                  freqentist_exact_intervals_for_pi_B$lower_confint),
        upper = c(freqentist_exact_intervals_for_pi_A$upper_confint,
                  freqentist_exact_intervals_for_pi_B$upper_confint)  
      ) %>% 
    mutate( model = "Frequentist" )
  ,
  
  tibble( pi = c("A","B"),
        point_estimate = c(0.9, 0.95),
        lower = c( drug_A_confint_hdi_0.9["lower"],
                   drug_B_confint_hdi_0.9["lower"]
                 ),
        upper = c( drug_A_confint_hdi_0.9["upper"],
                   drug_B_confint_hdi_0.9["upper"]
                 )
      ) %>% 
    mutate( model = "Bayesian" )
  
  ) %>% 
  ggplot( data = .,
          aes(      x = pi,
                    y = point_estimate,
               colour = model 
             )
        ) + 
  geom_point( size=2 ) +
  geom_errorbar( aes( ymin=lower, ymax=upper) ) +
  coord_flip() +
  facet_grid( model ~ . ) +
  theme( legend.position = "none" ) +
  scale_y_continuous( breaks=seq(0.83, 0.99, 0.01) ) +
  labs( title = bquote("Confidence/Credibility Intervals for"~pi[A]~"and"~pi[B]),
            x = "Drug",
            y = bquote(pi[j])
        )
```

# Intervals for the Difference Between 2 Proportions $\pi_A-\pi_B$
Instead of considering the proportions $\pi_A$ and $\pi_B$ separately, we can conduct inference on the difference between the two $\quad(\pi_A-\pi_B)$. Alternatively, one could also use the ratio $\hspace{3mm}\displaystyle\frac{\pi_A}{\pi_B}\hspace{3mm}$ as a comparison between $\pi_A$ and $\pi_B$, but we'll consider the difference $\hspace{3mm}(\pi_A-\pi_B)\hspace{3mm}$ here. 

# Bayesian Intervals for the Difference Between 2 Proportions $\pi_A-\pi_B$

From the previous Bayesian section, we made the following choices/assumptions:

$$\begin{array}{lcl}
\pi_A \text{ prior: } 
&\space&  f(\pi_A) 
&=&  \text{Beta}\Big(\alpha_A, \space \beta_A\Big) \\
\pi_B \text{ prior: } 
&\space&  f(\pi_B) 
&=&  \text{Beta}\Big(\alpha_B, \space \beta_B\Big) \\
\text{Likelihood A: } 
&\space&  X_A 
\space \Bigl|\space\pi_A 
&\sim& \text{Binomial}\Big(n_A, \space \pi_A \Big) \\
\text{Likelihood B: } 
&\space&  X_B 
\space \Bigl|\space\pi_B 
&\sim& \text{Binomial}\Big(n_B, \space \pi_B \Big) \\
\end{array}$$

Once realisations $x_A$ of $X_A$ and $x_B$ of $X_B$ were observed (i.e. we observed data), then the posterior distributions of $\pi_A$ and $\pi_B$ were given by:

$$\begin{array}{lcl}
\pi_A \text{ posterior: } 
&\space&  f \Big(\pi_A \space\Bigl| \space X_A=x_A \Big) 
&=&  \text{Beta}\Big(\alpha_A + x_A, \space \beta_A - n_A + x_A \Big) \\
\pi_B \text{ posterior: } 
&\space&  f \Big(\pi_B \space\Bigl| \space X_B=x_B \Big) 
&=&  \text{Beta}\Big(\alpha_B + x_B, \space \beta_B - n_B + x_B \Big) \\
\end{array}$$

Continuing with these same priors and likelihoods, suppose that we are interested in the posterior distribution of the random variable $Y=\pi_A-\pi_B$. The mathematical form of the posterior distribution $f(Y|X_A,X_B)$ of $Y$ is complicated and difficult to work with, but it is straightforward to draw random values (simulate) from this distribution:

* To draw a random value $y_i$ from the distribution $f(Y|X_A,X_B)$:
    
    1. Draw a random value $\pi_A^{(i)}$ from the posterior distribution $f(\pi_A\Bigl|X_A=x_A)$
    
    2. Draw a random value $\pi_B^{(i)}$ from the posterior distribution $f(\pi_B\Bigl|X_B=x_B)$
    
    3. Calculate $y_i=\pi_A^{(i)}-\pi_B^{(i)}$

We can use this method to simulate a very large random sample from the posterior distribution $f(Y)$, then use this random sample for inference. For example, we can use empirical quantiles of this large sample as estimates of the theoretical quantiles of the distribution $f(Y|X_A,X_B)$ - the larger the sample, the more accurate these estimates will be.  
 
Using the distribution $\text{Beta}(5,1)$ for both priors $\space f(\pi_A) \space$ and $\space f(\pi_B) \space$ implies the following prior distribution for the difference $\hspace{3mm} Y=\pi_A-\pi_B$:  

```{r echo=FALSE}
# prior for pi[A]:
x_seq <- seq(0,1,0.01)
prior_A_plot <- 
  tibble( x = x_seq,
          y = dbeta(x_seq, shape1=5, shape2=1 )
        ) %>%  
  ggplot(   data = .,
          aes(     x = x,
                   y = y,
               group = 1
             )  
        ) +
  geom_line() +
  labs( title = bquote("Prior Distribution of "*pi[A]),
        subtitle= bquote( "alpha = "*5*", "*beta*" = 1"),
            x = bquote(pi[A]),
            y = bquote("f ("*pi[A]*")")
      ) +
  scale_x_continuous( breaks=seq(0,1,0.1) )

# prior for pi[B]:
prior_B_plot <- 
  tibble( x = x_seq,
          y = dbeta(x_seq, shape1=5, shape2=1 )
        ) %>%  
  ggplot(   data = .,
          aes(     x = x,
                   y = y,
               group = 1
             )  
        ) +
  geom_line() +
  labs( title = bquote("Prior Distribution of "*pi[B]),
        subtitle= bquote( "alpha = "*5*", "*beta*" = 1"),
            x = bquote(pi[B]),
            y = bquote("f ("*pi[B]*")")
      ) +
  scale_x_continuous( breaks=seq(0,1,0.1) )

# prior difference distribution:
prior_diffplot <- 
  tibble( pi_A = rbeta(n=1e6L, shape1=5, shape2=1),
          pi_B = rbeta(n=1e6L, shape1=5, shape2=1),
          diff = pi_A - pi_B
        ) %>%
  ggplot( data = .,
          aes( x = diff )
        ) +
  # geom_histogram( aes(y=..density..), bins=100 ) +
  geom_density() +
  labs( title = bquote("Prior Distribution of ("*pi[A]-pi[B]*")" ),
        x = bquote(pi[A]-pi[B]),
        y = bquote("f ("*pi[A]-pi[B]*")" )
      ) +
  scale_x_continuous( breaks=seq(-1,1,0.1) )
  
( prior_A_plot + prior_B_plot ) / prior_diffplot
```

Recall that our observed data was:

 Drug Administered | Number of People Receiving Drug  | Number of People Recovering  
-------------------|----------------------------------|------------------------------
A                  |          100                     |   90
B                  |          100                     |   95

These priors and observed data give the following posterior distribution $f(Y|X_A,X_B)$:

```{r echo=FALSE}
nsims = 1e6L

#drugA_posterior
#drugB_posterior

plot_data <- 
  tibble( pi_A = rbeta(n=nsims, shape1=95, shape2=11),
          pi_B = rbeta(n=nsims, shape1=100, shape2=6),
             y = pi_A - pi_B
        )

posterior_pi_A_plot <- 
  ggplot( data = plot_data,
          aes( x = pi_A )
        ) +
  geom_histogram( aes(y=..density..), bins=100 ) +
  geom_density( colour="red" ) +
  labs( title = bquote("Posterior Distribution of "*pi[A]),
            x = bquote(pi[A])
      ) +
  scale_x_continuous( breaks=seq(0.7,1,0.05) ) +
  coord_cartesian( xlim = c(0.7, 1) )
  

posterior_pi_B_plot <- 
  ggplot( data = plot_data,
          aes( x = pi_B )
        ) +
  geom_histogram( aes(y=..density..), bins=100 ) +
  geom_density( colour="red" ) +
    labs( title = bquote("Posterior Distribution of "*pi[B]),
            x = bquote(pi[B])
      ) +
  scale_x_continuous( breaks=seq(0.7,1,0.05) ) +
  coord_cartesian( xlim = c(0.7, 1) )
  
posterior_pi_diff_plot <- 
  ggplot( data = plot_data,
          aes( x = y )
        ) +
  geom_histogram( aes(y=..density..), bins=100 ) +
  geom_density( colour="red" ) +
  labs( title = bquote("Posterior Distribution of "*pi[A]-pi[B]),
            x = bquote(pi[A]-pi[B])
        ) +
  geom_vline( xintercept = 0 ) +
  scale_x_continuous( breaks = round(seq(-0.3, 0.2, 0.05), digits=2) )

( posterior_pi_A_plot + posterior_pi_B_plot ) / posterior_pi_diff_plot
```

Here is a simple way to understand this Bayesian calculation:

Suppose that we draw 5 million random values from each of the priors:

```{r echo=FALSE, fig.width=9, fig.height=4}
prior_A <- rbeta( n = 5e6L,
                  shape1 = 5,
                  shape2 = 1
                )

prior_B <- rbeta( n = 5e6L,
                  shape1 = 5,
                  shape2 = 1
                )


( ggplot( data = tibble(pi_a=prior_A), aes(x=pi_a) ) + 
    geom_histogram(bins=100) +
    labs( title = bquote("Random sample from prior f("*pi[A]*")=Beta(5,1)"),
              x = bquote(pi[A])
        ) 
) +
( ggplot( data = tibble(pi_b=prior_B), aes(x=pi_b) ) + 
    geom_histogram(bins=100) +
    labs( title = bquote("Random sample from prior f("*pi[B]*")=Beta(5,1)"),
              x = bquote(pi[B])
        ) 
)

```

For each of the 5 million simulated values of $\pi_A^{(i)}$, we draw a random value $X_A^{(i)}$ from the distribution $\hspace{2mm}X_A^{(i)}\sim \text{Binomial}\Big(n=100, \space prob=\pi_A^{(i)}\Big)$.

Similarly, for each of the 5 million simulated values of $\pi_B^{(i)}$, we draw a random value $X_B^{(i)}$ from the distribution $\hspace{2mm}X_B^{(i)}\sim \text{Binomial}\Big(n=100, \space prob=\pi_B^{(i)}\Big)$.
  
```{r echo=FALSE}
sim_X_A <- rbinom( n=5e6L, size=100, prob=prior_A )
sim_X_B <- rbinom( n=5e6L, size=100, prob=prior_B )
```  
  
Now, we keep only the simulated samples $i$ where we saw that both $X_A^{(i)}=90$ and $X_B^{(i)}=95$. We throw all the rest of the simulated samples away. This happened in `r sum(sim_X_A==90 & sim_X_B==95)` of the one million samples.

Over only these `r sum(sim_X_A==90 & sim_X_B==95)` samples which resulted in $X_A=90$ and $X_B=95$, we look at the distribution of $\pi_A-\pi_B \hspace{2mm}$
(this is the posterior distribution of $\pi_A-\pi_B$):

```{r echo=FALSE}
posterior_dbn <- 
  tibble( pi_A = prior_A[sim_X_A==90 & sim_X_B==95],
           X_A = sim_X_A[sim_X_A==90 & sim_X_B==95],
          pi_B = prior_B[sim_X_A==90 & sim_X_B==95],
           X_B = sim_X_B[sim_X_A==90 & sim_X_B==95]
        ) %>% 
  mutate( pi_diff = pi_A - pi_B )

ggplot( data = posterior_dbn,
        aes( x = pi_diff )
) +
  geom_histogram( aes(y = ..density..), bins=100 ) +
  geom_density( data = plot_data, aes(x=y), colour="red", size=1 ) +
  labs( x = bquote(pi[A]-pi[B]) )
```

Again, we can create 90% credibility intervals (HDI) using this posterior distribution:

```{r}
HDInterval::hdi(   object = plot_data$y,
                 credMass = 0.9  
               ) %>% 
  c()
```

```{r echo=FALSE}
get_intervals_pi_diff <- 
  HDInterval::hdi(   object = plot_data$y,
                 credMass = 0.9  
               ) %>% 
  c()

extract_plot_data <-  
  ggplot_build(posterior_pi_diff_plot)$data[[2]]

posterior_pi_diff_plot +
  geom_area( data = subset( extract_plot_data, 
                            x >= get_intervals_pi_diff["lower"] & x <= get_intervals_pi_diff["upper"]
                          ),
             aes( x=x, y=y ),
             fill = "blue"
           ) +
  geom_vline( xintercept = 0 ) +
  labs( subtitle = bquote("With 90% Credibility Interval for"~pi[A]-pi[B]) ) + 
  geom_vline( xintercept = get_intervals_pi_diff, colour="blue", linetype = "dashed" )

```

As with the single-sample case, if we do not wish to specify priors for $\pi_A$ and $\pi_B$, we could simply consider the likelihoods of our observed data under different possible values of $\pi_A$ and $\pi_B$:

```{r}
# create a grid of different pi_A and pi_B values:
# (I'm approximating the continuous surface with a discrete grid of points)
expand_grid( pi_A = seq(0,1,0.01),
             pi_B = seq(0,1,0.01)
           ) %>% 
  
  # calculate the probability of our observed data under each {pi_A, pi_B} combination:
  mutate( unscaled_prob = 
            map2_dbl( .x = pi_A, 
                      .y = pi_B,
                      .f = function(a,b)
                             { dbinom(x=90, size=100, prob=a) *
                               dbinom(x=95, size=100, prob=b)
                             }
                    )    
        ) %>% 
  
  # work out the difference (pi_A - pi_B) for each {pi_A, pi_B} combination:
  mutate( pi_diff = round( pi_A-pi_B,
                           digits=2
                          ) 
        ) %>% 
  
  # work out the probability of our observed data under each (pi_A - pi_B): 
  group_by( pi_diff ) %>% 
  summarise( unscaled_prob = sum(unscaled_prob) ) %>%
  mutate( prob = unscaled_prob / sum(unscaled_prob) ) %>% 
  #summarise( sum(prob) )
  
  # make the plot:
  ggplot( data = .,
          aes( x = pi_diff, 
               y = prob
             )  
        ) +
  geom_line( colour = "blue" ) +
  geom_area( fill = "blue", alpha=0.5 ) +
  labs( x = bquote(pi[A]-pi[B]),
        y = "Density",
        title = 
          bquote( "Likelihood of Observed Data under Alternative ("*pi[A]-pi[B]*") Values" )
      ) +
  coord_cartesian( xlim = c(-0.5, 0.5) ) +
  geom_vline( xintercept = 0 ) +
  scale_x_continuous( breaks = seq(-0.5, 0.5, 0.1) )

```

We'd achieve the same result by using the Bayesian method, but with uninformative priors $f(\pi_A)\propto 1$ and $f(\pi_B)\propto 1$. 

# Frequentist Intervals for the Difference Between 2 Proportions $\pi_A-\pi_B$

A frequentist method for quantifying uncertainty in the difference between the two proportions $\hspace{2mm} Y=\pi_A-\pi_B\hspace{2mm}$ is to create *confidence intervals* for our observed data sample, which are guaranteed to contain the true difference $Y$ in a certain proportion of random samples.  

I choose (for purposes of illustration) $\hspace{2mm}\pi_A=0.7\hspace{2mm}$ as the true value of $\pi_A$ and $\hspace{2mm}\pi_B=0.64\hspace{2mm}$ as the true value of $\pi_B$. This makes the true difference $\hspace{2mm}\delta=0.7-0.64=0.06$. 

I generate 1,000,000 random samples $\hspace{2mm}\{X_A^{(i)},X_B^{(i)}\}\hspace{2mm}$ (giving us 1,000,000 estimates $d_i$ of $\delta$), where $\hspace{2mm} d_i=\underset{\hat\pi_A}{\underbrace{\displaystyle\frac{X_A^{(i)}}{100}}}-\underset{\hat\pi_B}{\underbrace{\displaystyle\frac{X_B^{(i)}}{100}}}\hspace{2mm}$ for sample $i$, and $\hspace{2mm}X_A^{(i)}\sim\text{Binomial}\Big(n=100, \space prob=\pi_A\Big)\hspace{2mm}$ and $\hspace{2mm}X_B^{(i)}\sim\text{Binomial}\Big(n=100, \space prob=\pi_B\Big)\hspace{2mm}$.  

Using the **Agresti-Caffo** adjusted Wald confidence interval method (implemented using the [DescTools::BinomDiffCI()](https://rdrr.io/cran/DescTools/man/BinomDiffCI.html) function) with a chosen confidence level of 90%: the proportion of these these 1,000,000 million samples in which the confidence intervals generated contained the true difference $\delta=0.06$ is:

```{r}
nsim <- 1e6L

tibble( x_A = rbinom(    n = nsim,
                      size = 100,
                      prob = 0.7
                    ),
        x_B = rbinom(    n = nsim,
                      size = 100,
                      prob = 0.64
                    ),
        diff = x_B - x_A
      ) %>% 
  mutate( binom_ci = map2( .x = x_A,
                           .y = x_B,
                           .f = function(x,y){
                                  BinomDiffCI( x1=x, x2=y, n1=100, n2=100, 
                                               conf.level=0.9, method="ac" )
                                             }
                         )
        ) %>% 
  mutate( prop_diff = map_dbl( .x=binom_ci, .f=~ .[[1]] ),
              lower = map_dbl( .x=binom_ci, .f=~ .[[2]] ),
              upper = map_dbl( .x=binom_ci, .f=~ .[[3]] )
  ) %>% 
  select( -binom_ci ) %>% 
  mutate( interval_contains_true_diff = 0.06>=lower & 0.06<=upper ) %>% 
  summarise( sum(interval_contains_true_diff) / nsim ) 

```

What is important is that this is 90% (or more) of the random samples. 

Using this same method on our original data...

 Drug Administered | Number of People Receiving Drug  | Number of People Recovering  
-------------------|----------------------------------|------------------------------
A                  |          100                     |   90
B                  |          100                     |   95

...gives the 90% confidence interval:

```{r}
DescTools::BinomDiffCI(         
          x1 = 90, 
          x2 = 95, 
          n1 = 100,
          n2 = 100, 
  conf.level = 0.9, 
      method = "ac"
  )
```

```{r echo=FALSE}

confint_pi_A <- binom::binom.confint( x=90, n=100, method="exact" ) %>% mutate(drug = "A")
confint_pi_B <- binom::binom.confint( x=95, n=100, method="exact" ) %>% mutate(drug = "B")

pi_ci_ggplot <- 
  ggplot( data = bind_rows( confint_pi_A, confint_pi_B ),
          aes( x = drug,
               y = mean, 
               colour = drug
             )
  ) + 
  geom_point( size=2 ) +
  geom_errorbar( aes(ymin=lower, ymax=upper) ) +
  coord_flip() +
  labs( title = bquote("90% Exact Confidence Intervals for "*pi[A]*" and "*pi[B]),
            y = bquote(pi[j]),
            x = "Drug"
      ) +
  theme( legend.position = "none" ) +
  scale_y_continuous( breaks = seq(0.8, 1, 0.025) )
  
pi_ci_ggplot

pi_diffplot_ci_ggplot <- 
  DescTools::BinomDiffCI(         
            x1 = 90, 
            x2 = 95, 
            n1 = 100,
            n2 = 100, 
    conf.level = 0.9, 
        method = "ac"
    ) %>% 
    as_tibble()

```

# Resources

1. [NCSS Statistical Software Chapter 514: Two Proportions](https://ncss-wpengine.netdna-ssl.com/wp-content/themes/ncss/pdf/Procedures/NCSS/Two_Proportions.pdf)

2. [Confidence and Credibility Intervals for the Difference of Two Proportions](http://www.scielo.org.co/scielo.php?script=sci_abstract&pid=S0120-17512010000100005)
l,.
3. [Exact Binomial Confidence Interval for Proportions](https://www.researchgate.net/publication/255630317_Exact_Binomial_Confidence_Interval_for_Proportions)

4. https://stats.stackexchange.com/questions/70661/how-does-the-beta-prior-affect-the-posterior-under-a-binomial-likelihood.

5. [Confidence Intervals for Binomial Proportions in R using the **DescTools** package](https://www.rdocumentation.org/packages/DescTools/versions/0.99.34/topics/BinomDiffCI)

6. http://varianceexplained.org/r/empirical_bayes_baseball/

7. [Binomial confidence intervals and contingency tests: mathematical fundamentals and the evaluation of alternative methods](https://www.ucl.ac.uk/english-usage/staff/sean/resources/binomialpoisson.pdf)

8. [Bayesian Approaches to Clinical Trials and Health-Care Evaluation](https://onlinelibrary.wiley.com/doi/book/10.1002/0470092602)

