---
title: "B-splines"
author: "joe"
date: "7/12/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

# Basis Splines 

This information is based on the Vignette __A Primer on Regression Splines__ by Jeffrey Racine, available at https://cran.r-project.org/web/packages/crs/vignettes/spline_primer.pdf

B-spline stands for 'Basis Spline', which is a generalisation of a Bezier curve. We concern ourselves with regression splines, as opposed to smoothing splines

# Quadratic Bezier Curve
A univariate quadratic bezier curve is a function $B(x)$ of $x$ with domain $x\in[0,1]$:

$$\begin{array}{lcl} B(x) &=& \beta_0(1-x)^2 + \beta_1 2(1-x)x + \beta_2x^2 \\
&=& \beta_0 B_0(x) + \beta_1 B_1(x) + \beta_2 B_2(x)   \\
&=& \sum_{i=0}^2 \beta_i B_i(x) \\
\end{array}$$

So, we are created a curve by from a weighted sum of different polynomial functions of $x$.  

```{r define_bezier}

# define a grid of x-values:
x <- seq(0,1,0.001)

# define the univariate quadratic bezier function  
B_x <- function(x,beta){ beta$beta0*(1-x)^2 + beta$beta1*2*(1-x)*x + beta$beta2*x^2 }

```

```{r examples}
par( mfrow=c(2,2), pty = "s")

plot( x = x,
      y = (1-x)^2,
      pch = ".",
      ylab = expression( (1-x)^2 ),
      main = expression( (1-x)^2 ),
      xlab = "x"
    )
plot( x = x,
      y = 2*(1-x)*x,
      pch = ".",
      ylab = expression( 2 %.% (1-x) %.% x ),
      main = expression( 2 %.% (1-x) %.% x ),
      xlab = "x"
    )
plot( x = x,
      y = x^2,
      pch = ".",
      ylab = expression( x^2 ),
      main = expression( x^2 ),
      xlab = "x"
    )

plot( x = x, 
      y = B_x( x = x, beta = list( beta0=1, beta1=-1, beta2=2 ) ),
      main = expression( 1 %.% (1-x)^2 +
                         -1 %.% 2(1-x)%.% x +
                         2 %.% x^2 ),
      ylab = "B(x)",
      pch = "."
    )
```

We could fit a curve like this to univariate data using least squares estimation. Here are examples on four random datasets: 

```{r randat,echo=FALSE}
# create random data
par( mfrow = c(2,2) )
x <- runif(10,0,1)
y <- sin(x*8) + rnorm(10)

B0 <- (1-x)^2
B1 <- 2*(1-x)*x
B2 <- x^2

X <- cbind(B0,B1,B2)
OLS_coef <- solve( t(X)%*%X ) %*% t(X) %*% y
# this is the same as lm( y ~ B0 + B1 + B2 - 1 )

x_grid <- seq(0,1,0.001)
fitted_vals <- cbind( (1-x_grid)^2, 2*(1-x_grid)*x_grid, x_grid^2) %*% OLS_coef
plot( y ~ x, 
      main = paste0( round(OLS_coef[1], 2) , " * (1-x)^2 + ",
                     round(OLS_coef[2], 2), " * 2(1-x)x + ",  
                     round(OLS_coef[3], 2), " * x^2" ),
      xlim=c(0,1) 
    )
points( x=x_grid, y=fitted_vals, pch=".", main = "test")

x <- runif(10,0,1)
y <- sin(x*8) + rnorm(10)

B0 <- (1-x)^2
B1 <- 2*(1-x)*x
B2 <- x^2

X <- cbind(B0,B1,B2)
OLS_coef <- solve( t(X)%*%X ) %*% t(X) %*% y
# this is the same as lm( y ~ B0 + B1 + B2 - 1 )

x_grid <- seq(0,1,0.001)
fitted_vals <- cbind( (1-x_grid)^2, 2*(1-x_grid)*x_grid, x_grid^2) %*% OLS_coef
plot( y ~ x, 
      main = paste0( round(OLS_coef[1], 2) , " * (1-x)^2 + ",
                     round(OLS_coef[2], 2), " * 2(1-x)x + ",  
                     round(OLS_coef[3], 2), " * x^2" ),
      xlim=c(0,1) 
    )
points( x=x_grid, y=fitted_vals, pch=".", main = "test")

x <- runif(10,0,1)
y <- sin(x*8) + rnorm(10)

B0 <- (1-x)^2
B1 <- 2*(1-x)*x
B2 <- x^2

X <- cbind(B0,B1,B2)
OLS_coef <- solve( t(X)%*%X ) %*% t(X) %*% y
# this is the same as lm( y ~ B0 + B1 + B2 - 1 )

x_grid <- seq(0,1,0.001)
fitted_vals <- cbind( (1-x_grid)^2, 2*(1-x_grid)*x_grid, x_grid^2) %*% OLS_coef
plot( y ~ x, 
      main = paste0( round(OLS_coef[1], 2) , " * (1-x)^2 + ",
                     round(OLS_coef[2], 2), " * 2(1-x)x + ",  
                     round(OLS_coef[3], 2), " * x^2" ),
      xlim=c(0,1) 
    )
points( x=x_grid, y=fitted_vals, pch=".", main = "test")

x <- runif(10,0,1)
y <- sin(x*8) + rnorm(10)

B0 <- (1-x)^2
B1 <- 2*(1-x)*x
B2 <- x^2

X <- cbind(B0,B1,B2)
OLS_coef <- solve( t(X)%*%X ) %*% t(X) %*% y
# this is the same as lm( y ~ B0 + B1 + B2 - 1 )

x_grid <- seq(0,1,0.001)
fitted_vals <- cbind( (1-x_grid)^2, 2*(1-x_grid)*x_grid, x_grid^2) %*% OLS_coef
plot( y ~ x, 
      main = paste0( round(OLS_coef[1], 2) , " * (1-x)^2 + ",
                     round(OLS_coef[2], 2), " * 2(1-x)x + ",  
                     round(OLS_coef[3], 2), " * x^2" ),
      xlim=c(0,1) 
    )
points( x=x_grid, y=fitted_vals, pch=".", main = "test")
```