---
title: "Calculated Features to Capture Customer Order Behaviour in Time"
author: "Joseph Bolton"
output:
  html_document:
    df_print: paged
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(patchwork)
library(broom)
```

```{r}
gen_random_timeofday <- function(){
  hour <- sample( 8:16, size=1 ) %>% as.character()
  minute <- sample( 0:59, size=1 ) %>% as.character()
  hour <- str_pad( string=hour, width=2, side="left", pad="0")
  minute <- str_pad( string=minute, width=2, side="left", pad="0")
  return( paste0(hour, ":", minute) )
}

gen_random_timeofday()
gen_random_timeofday()
gen_random_timeofday()
```

```{r}
# gen_customer_order_viz <- function( order_date_vec,
#                                     xlim = c( lubridate::date("2018-01-01"),
#                                               lubridate::date("2019-12-31")
#                                             ),
#                                     cluster_proximity = 7   # all orders within this distance of one another
#                                                             # will be treated as 1 order
#                                   ){
#   
#     order_date_vec <- lubridate::date( order_date_vec )
#     start_date <- xlim[1]
#     end_date <- xlim[2]
#     
#     plot_data <- tibble( date = order_date_vec,
#                            y = 0,
#                          grp = "customer"    # add this because geom_line compains otherwise
#                        ) %>% 
#             mutate( gaplength = date - lag(date) )
#     
#     # 
#     
#     sd_gaplength <- sd(plot_data$gaplength, na.rm=TRUE)
#     
#     return( 
#             ggplot( data = plot_data,
#                     aes(     x = date,
#                              y = y,
#                          group = grp
#                         )
#                   ) + 
#             geom_point(colour="red", size=3) + 
#             geom_line() +
#             geom_vline( aes(xintercept = date), colour = "red" ) +
#             geom_text( aes(label=week_number), hjust=2, angle=90 ) +
#             labs( subtitle = paste0( "SD_gaplength: ", round(sd_gaplength),
#                                      "\n",
#                                      "n_orders_past_7days: ", plot_data %>% filter( date >= (end_date-7) ) %>% nrow(),
#                                      "\n",
#                                      "n_orders_past_14days: ", plot_data %>% filter( date >= (end_date-14) ) %>% nrow(),
#                                      "\n",
#                                      "n_orders_past_30days: ", plot_data %>% filter( date >= (end_date-30) ) %>% nrow()
#                                     )
#                 ) +
#             xlim( xlim ) 
#             
#     )
#   
#   
# }
# 
# gen_customer_order_viz( order_date_vec = c("2018-01-01", "2018-01-05", "2018-01-31", "2018-03-01") )
```

```{r}
gen_customer_order_viz2 <- function( order_datetime_vec,
                                     xlim = c( lubridate::as_datetime("2018-01-01 00:00:00"),
                                               lubridate::as_datetime("2019-12-31 23:59:59")
                                             ),
                                     ylim = as.POSIXct( strptime(c("08:00",
                                                                   "17:00"), format = "%H:%M")
                                                      )
                                   ){
  
    order_datetime_vec <- lubridate::as_datetime( order_datetime_vec )
    start_date_time <- xlim[1]
    end_date_time <- xlim[2]
    
    plot_data <- tibble(      date = order_datetime_vec,
                         timeofday = as.POSIXct( 
                                          strftime(order_datetime_vec, format="%H:%M"),
                                          format = "%H:%M"
                                          ) - 2*60*60,
                               grp = "customer"    # add this because geom_line compains otherwise
                        ) %>% 
             mutate( gaplength = difftime( date,lag(date), unit="days" ) ) %>% 
             mutate( dayofweek = lubridate::wday(date) ) %>% 
             mutate( period = case_when( lubridate::hour(date) < 11 ~ "8:00-10:59",
                                         lubridate::hour(date) %in% c(11,12,13) ~ "11:00-13:59",
                                         lubridate::hour(date) > 13 ~ "14:00-16:59",
                                       ) 
                   ) %>% 
             mutate( period = factor(period, levels=c("8:00-10:59", "11:00-13:59", "14:00-16:59") ))
             
    
    return( 
             ggplot( data = plot_data,
                     aes(     x = date,
                              y = timeofday,
                          group = grp,
                          label = dayofweek,
                          shape = period
                        )
                   ) + 
             geom_vline( aes(xintercept = date), colour = "red" ) +
             geom_hline( yintercept = c(as.POSIXct( strptime(c("08:00",
                                                                 "11:00",
                                                                 "14:00",
                                                                 "17:00"), format = "%H:%M")
                                                      )),
                         alpha = 0.3
                        ) +
             geom_point( colour="black", size=5) +
             geom_text( colour = "white" ) +
             xlim( xlim ) + 
             scale_y_datetime( limits = ylim,
                               breaks = c(as.POSIXct( strptime(c("08:00",
                                                                 "09:00",
                                                                 "10:00",
                                                                 "11:00",
                                                                 "12:00",
                                                                 "13:00",
                                                                 "14:00",
                                                                 "15:00",
                                                                 "16:00",
                                                                 "17:00"), format = "%H:%M")
                                                      )),
                               labels = c("08:00",
                                                                 "09:00",
                                                                 "10:00",
                                                                 "11:00",
                                                                 "12:00",
                                                                 "13:00",
                                                                 "14:00",
                                                                 "15:00",
                                                                 "16:00",
                                                                 "17:00")
                             )
     )
  
  
}

gen_customer_order_viz2( order_datetime_vec = c("2018-01-01 09:15:36", "2018-03-05 13:49:02", "2018-03-19 10:01:02", "2018-07-01 16:12:58") )
```

# Recency
When was the last time that the customer was active? 
```{r echo=FALSE}
(
  gen_customer_order_viz2( order_datetime_vec = c("2018-01-01 12:15:00", "2019-01-01 13:30:02") ) /
  gen_customer_order_viz2( order_datetime_vec = c("2018-01-01 12:15:00", "2019-11-01 13:30:02") )
)
```

Features describing *recency*:

1. Number of days since last transaction

2. Number of days since customer had $x$ close transactions (e.g. how many days ago did customer have 3 transactions, where those transactions were a maximum of 1 month apart). This is a stricter definition of an *active* customer, requiring a number of transactions in close proximity in time.  

# Consistency
How consistent is a customer in their transactions, in terms of amount of time between each transaction?  

```{r echo=FALSE}
(
  gen_customer_order_viz2( order_datetime_vec = c("2018-01-01 12:15:00", "2018-04-01 12:15:00","2018-07-01 12:15:00", "2018-10-01 12:15:00", "2019-01-01 12:15:00","2019-04-01 12:15:00", "2019-07-01 12:15:00","2019-10-01 12:15:00") ) /
  gen_customer_order_viz2( order_datetime_vec = c("2018-01-01 12:15:00", "2018-02-01 12:15:00","2018-03-01 12:15:00", "2018-11-01 12:15:00", "2018-12-01 12:15:00","2019-12-01 12:15:00", "2019-07-01 12:15:00","2019-08-01 12:15:00") ) /
    gen_customer_order_viz2( order_datetime_vec = c("2018-01-01 12:15:00", "2018-02-01 12:15:00","2018-07-01 12:15:00", "2018-08-01 12:15:00", "2019-01-01 12:15:00","2019-02-01 12:15:00", "2019-07-01 12:15:00","2019-08-01 12:15:00") )
)
```

Meausures of consistency: 

1. Standard Deviation of Gap Length (gap length=number of days between 2 transactions).

2. Standard Deviation of gap length, where data is first binned into active weeks (week containing at least one transaction) and inactive weeks (week containing no transactions). *Gap length* is then defined as number of inactive weeks between any 2 given active weeks. Could also bin transactions into *active* and *inactive* *months*. 

# Trajectory
```{r echo=FALSE}
(
  gen_customer_order_viz2( order_datetime_vec = c("2018-01-01 09:15:00", "2018-07-01 13:30:02", "2019-01-01 14:22:02", "2019-07-01 11:35:58", "2019-07-15 11:35:58", "2019-08-01 11:35:58", "2019-08-15 11:35:58", "2019-10-01 11:35:58", "2019-12-01 11:35:58") ) /
  gen_customer_order_viz2( order_datetime_vec = c("2018-01-01 09:15:00", "2018-07-01 13:30:02", "2019-01-01 14:22:02", "2019-07-01 11:35:58", "2019-08-01 11:35:58", "2019-09-01 11:35:58", "2019-10-01 11:35:58", "2019-11-01 11:35:58", "2019-12-01 11:35:58") )  /
  gen_customer_order_viz2( order_datetime_vec = c("2018-01-01 09:15:00", "2018-07-01 13:30:02", "2019-01-01 14:22:02", "2019-07-01 11:35:58", "2019-10-01 11:35:58", "2019-11-01 11:35:58", "2019-11-15 11:35:58", "2019-12-01 11:35:58", "2019-12x-15 11:35:58") )
)
```

Measures of trajectory:

1. *Trajectory in last $x$ weeks*: Count number of transactions per week in the last $x$ months. Fit an OLS line to the relationship **n_transactions ~ time**. The slope of this line is a measure of the trajectory of the last $x$ months.     

2. *Trajectory in last $x$ weeks*: Count number of transactions per week in the last $x$ months. For each week, calculate the change in number of transactions (absolute or %) for this week compared to every subsequent week. Define the trajectory as the average of these changes. 

# Frequency
```{r echo=FALSE}
(
  gen_customer_order_viz2( order_datetime_vec = c("2018-01-01 12:15:00", "2018-04-01 12:15:00","2018-07-01 12:15:00", "2018-10-01 12:15:00", "2019-01-01 12:15:00","2019-04-01 12:15:00", "2019-07-01 12:15:00","2019-10-01 12:15:00", "2019-12-31 12:15:00") ) /
  gen_customer_order_viz2( order_datetime_vec = c("2018-01-01 12:15:00","2018-07-01 12:15:00", "2019-01-01 12:15:00", "2019-07-01 12:15:00", "2019-12-31 12:15:00") )
)
```

Measures of frequency:

1. Number of transactions in last $x$ weeks.  

# Day-of-Week Preference
```{r echo=FALSE}
(
  gen_customer_order_viz2( order_datetime_vec = c("2018-01-04 12:15:00", "2018-04-05 12:15:00","2018-07-05 12:15:00", "2018-10-04 12:15:00", "2019-01-03 12:15:00","2019-04-04 12:15:00", "2019-07-04 12:15:00","2019-10-03 12:15:00", "2019-12-26 12:15:00") ) /
  gen_customer_order_viz2( order_datetime_vec = c("2018-01-01 12:15:00", "2018-04-01 12:15:00","2018-07-04 12:15:00", "2018-10-04 12:15:00", "2019-01-06 12:15:00","2019-04-01 12:15:00", "2019-07-06 12:15:00","2019-10-01 12:15:00", "2019-12-26 12:15:00") )
)
```

Measures of *Day-of-Week Preference*: 

1. Count number of transactions per day of week in the last $x$ weeks. Turn these into a proportional spread of day-of-week across all of these transactions. Gini coefficient of these values measures spread over days (i.e. ...).

2. Report favourite day-of-week. Could scale this by number of customers active on each day-of-week in the period of interest. 

# Time-of-Day Preference
```{r echo=FALSE}
(
  gen_customer_order_viz2( order_datetime_vec = 
            c(paste0( "2018-01-01 ", gen_random_timeofday(), ":00"),
              paste0( "2018-04-01 ", gen_random_timeofday(), ":00"),
              paste0( "2018-07-04 ", gen_random_timeofday(), ":00"),
              paste0( "2018-10-04 ", gen_random_timeofday(), ":00"),
              paste0( "2019-01-06 ", gen_random_timeofday(), ":00"),
              paste0( "2019-04-01 ", gen_random_timeofday(), ":00"),
              paste0( "2019-07-06 ", gen_random_timeofday(), ":00"),
              paste0( "2019-10-01 ", gen_random_timeofday(), ":00"),
              paste0( "2019-12-26 ", gen_random_timeofday(), ":00")
             )
    )/
  gen_customer_order_viz2( order_datetime_vec = c("2018-01-01 12:15:00", "2018-04-01 12:15:00","2018-07-04 12:15:00", "2018-10-04 12:15:00", "2019-01-06 12:15:00","2019-04-01 12:15:00", "2019-07-06 12:15:00","2019-10-01 12:15:00", "2019-12-26 12:15:00") )
)
```

# Change in behaviour regime
Compare change in various metrics over 2 consecutive periods. 


# Old Code

```{r}
orders <-
  bind_rows( 
    tibble( person = 1,
            order_datetime = c("2018-01-01 09:15:00", "2018-07-01 13:30:02", "2019-01-01 14:22:02", "2019-07-01 11:35:58", "2019-07-15 11:35:58", "2019-08-01 11:35:58", "2019-08-15 11:35:58", "2019-10-01 11:35:58", "2019-12-01 11:35:58")   
          ),
    tibble( person = 2,
            order_datetime = c("2018-01-01 09:15:00", "2018-07-01 13:30:02", "2019-01-01 14:22:02", "2019-07-01 11:35:58", "2019-08-01 11:35:58", "2019-09-01 11:35:58", "2019-10-01 11:35:58", "2019-11-01 11:35:58", "2019-12-01 11:35:58")
          ),
    tibble( person = 3,
            order_datetime = c("2018-01-01 09:15:00", "2018-07-01 13:30:02", "2019-01-01 14:22:02", "2019-07-01 11:35:58", "2019-10-01 11:35:58", "2019-11-01 11:35:58", "2019-11-15 11:35:58", "2019-12-01 11:35:58", "2019-12x-15 11:35:58")
          )
  ) %>% 
  mutate( order_datetime = lubridate::as_datetime(order_datetime) ) %>% 
  mutate( time_since_period_end = difftime( "2020-01-01 00:00:00",    # end of observation period
                                            order_datetime,
                                            unit = "days"
                                            ) 
                                  
        ) 

#   trajectory_last_6_months <- 
#     orders %>% 
#       filter( time_since_period_end < 180 ) %>% 
#       mutate( yearmon = lubridate::date( 
#                               paste0( lubridate::year(order_datetime),
#                                       "-",
#                                       str_pad( string = lubridate::month(order_datetime),
#                                                 width = 2,
#                                                  side = "left",
#                                                   pad = "0"
#                                              ),
#                                       "-01"
#                                       )
#                               )
#       ) %>% 
#       group_by( person, yearmon ) %>% 
#       summarise( n_orders = n() ) %>% 
#       ungroup() %>% 
#       arrange( person, yearmon ) %>% 
#       group_by( person ) %>% 
#       mutate( time = yearmon - min(yearmon) ) %>%
#       nest() %>% 
#       mutate(fit_lm = map(data, ~ lm(n_orders ~ time, data = .) %>% 
#                             tidy() %>%
#                             select(term, estimate) %>% 
#                             spread(term, estimate)
#                          )
#             ) %>% 
#     unnest(fit_lm) %>% 
#     rename( slope = time ) %>%
#     unnest( data ) 
#   
# ggplot( trajectory_last_6_months,
#         aes( x = time, 
#              y = n_orders
#            )
#       ) +
#   geom_point() +
#   geom_abline( aes(intercept = `(Intercept)`, slope = slope), colour="red" ) +
#   labs( aes(title = slope) ) %>% 
#   facet_grid( person ~ . ) 
  
(
  gen_customer_order_viz2( order_datetime_vec = orders %>% filter(person==1) %>% pull(order_datetime) ) /
  gen_customer_order_viz2( order_datetime_vec = orders %>% filter(person==2) %>% pull(order_datetime) )  /
  gen_customer_order_viz2( order_datetime_vec = orders %>% filter(person==3) %>% pull(order_datetime) )
)
```





```{r fig.width=10, fig.height=15}
# show 10 random series with between 2 and 5 points in each:


# for( j in 1:5 ){
#   
#   plot_list <- list()
#   
#   for( i in 1:10 ){
#   
#     n_orders <- sample(3:10, size=1)
#     order_dates <- sample( 1:180, size=n_orders )
#     order_dates <- order_dates[ cumsum(order_dates) < 729 ]
#     order_dates <- lubridate::date("2018-01-01") + cumsum(order_dates)
#     plot_list[[i]] <- gen_customer_order_viz( order_dates )
#     }
# 
#   plot( patchwork::wrap_plots( plot_list, ncol = 2) )
# 
# }

```


