---
title: "Cold Start/Leads Recommender Model Concept"
author: "Joseph Bolton"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    theme: darkly
    highlight: espresso
    code_folding: hide
---

<style type="text/css">
@import url("https://fonts.googleapis.com/css?family=Inconsolata|Roboto+Mono|Ubuntu+Mono|Cutive+Mono");
pre{
display: inline-block;
font-size: 7px;
text-align: center;
font-family: Ubuntu Mono, monospace !important;
text-rendering: optimizeSpeed;
}
</style>
<pre style="font-size: 7px; font-family: &quot;Ubuntu Mono&quot;; background-color: rgb(255, 255, 255); color: rgb(0, 0, 0); line-height: 7px;">........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
..........................................................................................................................71117.........................................................................................................................................................................................................
.........................................................................................................................722227.........................................................................................................................................................................................................
........................................................................................................................72222227.......7................................................................................................................................................................................................
......................................................................................................................71222222227.......7...............................................................................................................................................................................................
....................................................................................................................7122244222227.......................................................................................................................................................................................................
..................................................................................................................7122224422222227..........7...........................................................................................................................................................................................
.................................................................................................................74422244222222217......................................................................................................................................................................................................
...............................................................................................................7234224422222222222......................................................................................................................................................................................................
.............7...............................................................................................723334222442222222222.......................................................77.............................................................................................................................................
............747...........................................................................................7724444422444242222422227...................................................771227............................................................................................................................................
..........716057........................................................................................714222222244444442224422221................................................7712222222177........................................................................................................................................
..........750884......................................................................................7722222222444422422222432222227............................................7122222222222222..71427................................................................................................................................
..........72688277..........................................................................77144177122222444224444222222222432222227...7.............................77.......71224444222222222212444427...............................................................................................................................
..........773807..........................................................................7233342222222443333334442222222223442222227...............................722222171712244444422222433344444222221.............................................................................................................................
..........2588017.........................................................................13422244442243333333334424333333433322222217............................71224442222244444422222444444444334222227.........7...................................................................................................................
.........725880327.......................................................................744224444444333333333333333333333333322222222777....7................7.71222444222434344442222244244444444442227..........7777.................................................................................................................
.........7748888527.....................................................................7422244444333333333333333333333333333334422221....77..................712222244422444222222222224244444444444217.........77777117...............................................................................................................
.......77258865837.......................7...........................................771422222443333333333333333333333333333333533334427.........................71122244244222222222222444442222444231................77777............................................................................................................
......73880088888017....................77......................................777122244222224333333333333333333333333333333333333333333177........................71444442222222222442444442442244347...................7717777.......................................................................................................
.......71655880211241.................7244477.................................712224444444444333353333333333333333333333333333333333333333333417..................77.....7744222224442444444444433327.7277..............................................................................................................................
...7772598880360517..................7768888617.............................772333333333333333333333333333333333333333333333333333333333333333337..........................74333333233333333333337177...7777777..7..77..................................................................................................................
...722222588888856542177...........740888888047...........................723333333333333333333333333333333333333333333333333333333333333333333337........77.................72217..74333333333317....7..77777........7.................................................................................................................
.....7230530888883441117............726888880637...................7771233333333333333333333333333333333333333333343333333333333333333333333333333277777.............77....7............7712442777............777117777.....................77.....................................7....................................................
..77450888888650034177..........77730888888831117..............777233333333333333333333333333443333335333333333342224433433553333333333333333333333333333417...77.............7777........................7..........777.77..................7777.............................77...17.....................................77............
72560888888880636886367.....7713608888356888654777..........7723333333433333333333333335333325333333333342224422222222222243333333333333333333333333333333337...777.........7....77....777.....777....................77777.....77.................7..........................717..37.........77..........................74............
14344120055368888880632217..7771453688888888941777.......714333322444212444242244433333424323043333222222222221111111111122223333333333333333333333333333333337....7......77177171177711711777771177............7777711177.....7....77................777....................77777497.77.....7777.........................755...........
1111114805088635888888888517.77459888888853658888627.777233334422222221222221222222224222112892442221221111111111111111111122224434433333333333333333333333333327....777777117111111111111111117117177777......77777177717777777..........77.............................77..7714600512177771117777......................4001......77.77
1112236888888805536047777777.71388888888888050888889542223334121115911111111111111122211111268553222111111111111111111111111111222223333333333333333333333333333377777771111111111111111111111111111111117717777111111111111171117117777771177.....7.........7....777...7117177126000655211112111711177777..7.....77.....750177....77177
124250888065888888888096342177771248880508888895342422212222221450886277111111111111111112588031111111111111111111111111111111111222244233333333333333333333333434217111111111111111111111111111111111111111111111112111111111111111111111711711177177.7....717.722777771111232555000322211255211111111717771777.777....7400031777111111
1122230033508888888804111111111745053336888888034417111112111121488888563211111111111111111588851111111111111111111111111111111111112222242233333333333334322222222111111112211111111111111111111112421111111111111144211111221111111221111111111111111421111112465171111112442243660053212556511112211111143211111111773366031111111111
1111458888888886333339933427124508800888688888665421111111111146865880222111111111111111112506947111111111111111111111111111111111111111222224333333344222222221111111111125221111111111121111111125632111111111112256411112432111112532111111112432113654212214555421111111244395350032774556511123521222466411111117740055062111111111
1136088888953588800665655223088888888888543353111711111111112445088880321111111121111111126889355211122111111111142111111111111111111111111222242434422111111122111111111245311111111112452111112435653211111111124556632124565211123653211222123563223555322215566542232225556600000065211366521246632432566321111124350000566521111111
1224406335008888888880334217112443688888888888888054211112235424888564442211112352111111122060851112453111111111265211111111111111111112211111112222211111111245211111112456641112211122565211111456666321111111145556532125665421123565422565443655355666332423566655555334444450065600006535553556632163566522212444422560653332222233
4366088888888888888056690653450888888888888888033506411112334300888050031121123665211122468860863123565221111112566421111111221111111125421111111121111111112356521111112366632114542245565422222356666532223342456666553435666533355566655656566566556666444635666656634444445000063505333443655555633006366653556656666060555555555655
5653445888885588888054424224588888006508888888344443377223566543065888865334435666444442598888803456665532442223565542221112554211112456541111111234211111112566542122124566654212565435665563356665666655655555566666665556666656666666666666666666666666535064566666666600000533606663444466666666663400666666556665345560634436655566
2445608888535688888888888886088854350888888888888888055566663368650508944421456566666565665685080665666565565655666535554223665211112566652111112366421222114566654245443666666545566666666666666666666666666666666666556666666666666666666666666666666663456064366666444444439000000055656666666666655605466666666666666500536066666666
6533308888888888333339888888888888888888633558888888544422424445888656508005666666666642445088544466666666666666666656655336665421123666664224424666542365443666655555566666666666666666666666666666666666666666666666666666666666666666666666666666666665445095443666669996305533600505444445666666660660096666666644445006000645666666
4555588880558888888000350888888888888888800555308888888800041455568888803435366666666660006688055566666666666666666666655666665544436656665356635666555665555666566655666666666666666666666666666666666666666666666666666666666666666666666666666666666544445005500666544443500000000960000096666665445005545666666653365333606444356666
8888853333508888888888053442433088888888888888888886556560880333688808888805344444666654430888953366666666666666666666666666666665666666666556666666666665556666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666535566500544364444336005500644360533344366666535500633666665443350006346053236666
0888888888888888888888888888800888888808888888888888093225053688888863568633456666333444408880583446666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666344450005300335000600000000060006555334445666353606096655654444450560000542143666
4444444445888888888888888888533333333330880368888888888888888806536888888000535443660608856888888954456666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666344222445600005453350336000033550005443600000006344605654445666000000350000544444566
0008888888888888888888608888880088888888888888888854335333356088888805688533332333666598808880088800666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666605355600590065536000006344500000555333444500000063456695555605350000350634444334441
88888888888888635668889343333333350888888888888888888880000536806588805358888800554688888508888886443666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666533333333366544445636055005243390000000000005333360000544444350000633333553055000344456000277..
4433333308888880544335888800000885988888888888888805555555689305358888888888888880533568888888056888834446666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666666653333333333333334111117777.................71455000550000053335000000000900000000000006900060006665533334222334555566224432717.7
00000888888888888888888888888888888888888888888888000000000000888888533333588888880008888888888058888051777777771222223555666666666666666666666666666666666666666666666666666666666552222222777777.............................................771444444335000644443446656532242421235333277111144171717111777127777777777777717777777..
588888888888888888888888888888888888888853333333355558888888888888888806663398633333356668888988886217....................77777777244442466666666666666666666666666666666664277777..............................................................777244465334242111351777.777777.777777777.......77...........77.............77777777....
8888888888888888888888888888888888888888888888888888888885308888888888888895963335688888888885358888881.....................................7711144456666666666654427777...............................................................................7..77.777..77....7...............................................................
8888888888888888888888888888888888888888888888888033333335088888888888888888888880088533335888888888881...............................................77777777..........................................................................................................................................................................
88888888888888888888888888888888888888888888888888880088888888888888888888888864333355080888808888888817................................................................................................................................................................................................................................
88888888888888888888888888888888888888888888888888888888888888888888888888888888888855536888888056686334254.............................................................................................................................................................................................................................
5888888888880688865688888888888888688888888888888888888888888888888888888888655555555655533088088888888517..............................................................................................................................................................................................................................
7453588888865553334066536888895366666808965088806606665088888888888888888888955555568888888888565358632217..............................................................................................................................................................................................................................
6543311171744177777777.7352721444323353433560653556908890880660880665888888888888888888888888888888888884...............................................................................................................................................................................................................................
...777..7777..7777..7.777.7.77777143431142217211711777124566088066888096688888880088888888888888888886332117............................................................................................................................................................................................................................
..7..777....777.......7771124177777777777777777.77777211245965545534256806659653213360606566668888880663427.............................................................................................................................................................................................................................
..777................77..77777.77777777777777777777777777777777777214321712177721135422712455533532211777...............................................................................................................................................................................................................................
.........................77.7777777777777777777777777777777777777777777777777777777777777111777777......................................................................................................................................................................................................................................
.................................77777777777777777777777777777777777777.7777777777777777777777777777....................................................................................................................................................................................................................................
................................................77777777777777..777..................7..................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
........................................................................................................................................................................................................................................................................................................................................
..................................................................................................................................................................................................................................................................................................................................7.....
.............................................................................................................................................................................................................................................................................................................................77.77......
.............................................................................................................................................................................................................................................................................................................................7..7.......
.............................................................................................................................................................................................................................................................................................................................7.77.......
........................................................................................................................................................................................................................................................................................................................................</pre>

```{r setup, include=FALSE}
library(tidyverse)
library(ranger)
```

I describe here a model-based item-recommender model that I have conceived, designed for the following context:

* Personalized recommendations are required for users who we have data about, but who have not yet bought an item 

* Predictive item data is not readily available

The concept is this: to create a recommender using only user features and user purchase history: 

* Create rich item features data by aggregating user features over each item  

The framework is also flexible in that any other desired/available item-level features can be added to the model (e.g. item price, or contextual features such as season). 

example of how a continuous user feature can be converted into an item feature:

```{r}
n_users <- 80
simdata <- 
  tibble( 
          user_ID = 1:n_users, 
          item_ID = sample( c("itemA","itemB","itemC"), size=n_users, replace=TRUE ),
          user_age =  ceiling( rnorm(n=n_users, mean=35, sd=10) )
        ) %>% 
  mutate( user_age = case_when( 
                                item_ID=="itemA" ~ user_age*1.3,
                                item_ID=="itemC" ~ user_age*0.8,
                                TRUE ~ user_age
                              ) 
        ) %>% 
  mutate( age_quintile = paste0( "ageQuintile_", ntile(user_age, n=5) ) ) 

n_users_bought_each_item <- 
  simdata %>% 
    group_by( item_ID ) %>% 
    summarise( n_users_bought_this_item = n() )

# Quantile Features:
simdata %>% 
    group_by( item_ID, age_quintile ) %>% 
    summarise( n_users_this_age_quintile_bought_this_item = n() ) %>% 
    left_join( n_users_bought_each_item ) %>% 
    mutate( proportion_of_users = n_users_this_age_quintile_bought_this_item / n_users_bought_this_item ) %>% 
    arrange( age_quintile ) %>% 
    pivot_wider( 
                 id_cols = item_ID,
                 names_from = age_quintile,
                 values_from = proportion_of_users
              )
  
# Mean Features:
simdata %>% 
  group_by( item_ID ) %>% 
  summarise( mean_buyer_age = mean(user_age) ) %>% 
  mutate( 
              global_mean_buyer_age = 
                simdata %>% summarise( global_mean_buyer_age = mean(user_age) ) %>% pull(global_mean_buyer_age)
         ) %>% 
  mutate( relative_age_this_item = mean_buyer_age/global_mean_buyer_age-1 ) %>% 
  select( item_ID, mean_buyer_age, relative_age_this_item, global_mean_buyer_age )

```

example of how a binary user feature can be converted into an item feature:

```{r}

```


```{r }
sim_n_users <- 50
sim_user_data <- 
  tibble(
    user_ID = 1:sim_n_users,
    age = ceiling( rnorm(n=sim_n_users, mean=35, sd=10) ),
    gender = sample( c("m","f"), size=sim_n_users, replace=TRUE),
    location = sample( c("A_town", "B_town", "C_town", "D_town"), size=sim_n_users, prob=sample(1:4), replace=TRUE  ) 
  ) %>% 
  mutate( # quick dirty manual 1-hot encoding:
          gender_male = if_else(gender=="m",1,0),
          gender_female = if_else(gender=="f",1,0),
          from_Atown = if_else(location=="A_town",1,0),
          from_Btown = if_else(location=="B_town",1,0),
          from_Ctown = if_else(location=="C_town",1,0),
          from_Dtown = if_else(location=="D_town",1,0)
        ) %>% 
  select( -gender, -location )

sim_user_data
```

simulate items with item attributes:

```{r}
sim_n_items <- 50
sim_items_attribute_table <-
  tibble(
          item_ID = 1:sim_n_items,
          base_item_popularity = runif( min=0, max=0.5, n=sim_n_items ),
          item_ageband_0_24 = sample( 0:1, size=sim_n_items, replace=TRUE ),
          item_ageband_25_39 = sample( 0:1, size=sim_n_items, replace=TRUE ),
          item_ageband_40_64 = sample( 0:1, size=sim_n_items, replace=TRUE ),
          item_ageband_65plus = sample( 0:1, size=sim_n_items, replace=TRUE ),
          item_male = sample( 0:1, size=sim_n_items, replace=TRUE ),
          item_female = sample( 0:1, size=sim_n_items, replace=TRUE ),
          item_townA = sample( 0:1, size=sim_n_items, replace=TRUE ),
          item_townB = sample( 0:1, size=sim_n_items, replace=TRUE ),
          item_townC = sample( 0:1, size=sim_n_items, replace=TRUE ),
          item_townD = sample( 0:1, size=sim_n_items, replace=TRUE )
  )

sim_items_attribute_table  
```

Show each user 20 random items. Calculate their buy probability for each. They select 1 to buy:

```{r}
orders_history <- tibble( user_ID=sim_user_data$user_ID, item_ID=NA, y=1 )

item_buy_prob_ftn <- function( specify_user_ID, specify_item_ID ){
  # calculates the relative probability that [user_ID] will buy [item_ID] 
  user_attributes <- 
    sim_user_data %>% 
      filter( user_ID == specify_user_ID ) 
  
  item_attributes <-
      sim_items_attribute_table %>% 
        filter( item_ID == specify_item_ID )
  
  buy_prob <- 
    item_attributes$base_item_popularity + 
    0.1 * item_attributes$item_ageband_0_24 * (user_attributes$age < 25) +
    0.1 * item_attributes$item_ageband_25_39 * (user_attributes$age>=25 & user_attributes$age<40) +
    0.1 * item_attributes$item_ageband_40_64 * (user_attributes$age>=40 & user_attributes$age<65) +
    0.1 * item_attributes$item_ageband_65plus * (user_attributes$age>=65) +
    0.1 * item_attributes$item_male * user_attributes$gender_male +
    0.1 * item_attributes$item_female * user_attributes$gender_female +
    0.1 * item_attributes$item_townA * user_attributes$from_Atown +
    0.1 * item_attributes$item_townB * user_attributes$from_Btown +
    0.1 * item_attributes$item_townC * user_attributes$from_Ctown +
    0.1 * item_attributes$item_townD * user_attributes$from_Dtown
  
  return( buy_prob )
}


counter <- 0
for( user_ID_i in sim_user_data$user_ID ){
  
  random_item_selection <- sample( sim_items_attribute_table$item_ID, size=20, replace=FALSE )
  random_item_selection_buy_probs_user_i <- 
    sapply( 
            random_item_selection,
            function(itemID){ item_buy_prob_ftn( specify_user_ID=user_ID_i, specify_item_ID=itemID ) }
          )
  orders_history$item_ID[ orders_history$user_ID==user_ID_i ] <- 
    sample(
               x = random_item_selection,
            size = 1,
            prob = random_item_selection_buy_probs_user_i / sum(random_item_selection_buy_probs_user_i) 
    )
  
  percent_complete <- 100 * counter / nrow(sim_user_data)
  if( percent_complete%%1 == 0 ){
      print( paste0(percent_complete, "%") )
  }
  counter <- counter+1
}
  
orders_history
```

add negative examples to the orders data:
```{r}
negative_examples <- 
  tibble(
    user_ID = orders_history$user_ID,
          y = 0
  ) %>% 
  mutate( item_ID = sample( sim_items_attribute_table$item_ID, size=n(), replace=TRUE ) ) %>% 
  # remove negative examples that the user actually did buy:
  left_join( orders_history %>% select(user_ID, item_ID) %>% mutate(remove=1) ) %>% 
  filter( is.na(remove) ) %>% 
  select( user_ID, item_ID, y )

train_data <- 
  bind_rows( orders_history, negative_examples ) %>% 
  arrange( user_ID )
```

create a test dataset: 

* each customer given 2 random items (only items seen during training)

```{r}
test_data <- tibble( user_ID=-1, item_ID=-1, y=-1) %>% slice(0)

test_data <- 
  tibble(
    user_ID = rep( sim_user_data$user_ID, 2 ),
    item_ID = sample( train_data$item_ID, size=nrow(sim_user_data)*2, replace=TRUE )
  ) %>% 
  mutate( buy_prob = map2_dbl( 
                      .x = user_ID,
                      .y = item_ID,
                      .f = item_buy_prob_ftn
                    )
  )

test_data %>% 
  arrange( user_ID )
```

compute item features by aggregating attributes of users who bought those items:

```{r}
item_meanBased_features_traindata <- 
  train_data %>%
    filter( y==1 ) %>% 
    left_join( sim_user_data ) %>% 
    select( -user_ID, -y ) %>%   
    rename_all( function(x) paste0( "itemMean_", x) ) %>% 
    rename( item_ID = itemMean_item_ID ) %>% 
    group_by( item_ID ) %>% 
    summarise_all( mean )

# calculate mean user attributes over all items
mean_item_population <- 
  train_data %>%
      filter( y==1 ) %>% 
      left_join( sim_user_data ) %>% 
      select( -user_ID, -y, -item_ID ) %>%   
      rename_all( function(x) paste0( "itemMean_", x) ) %>%   
      summarise_all( mean )

item_meanBased_features_traindata <- 
  item_meanBased_features_traindata %>% 
    mutate( 
            relative_itemMean_age = 
                itemMean_age / mean_item_population$itemMean_age -1,
            relative_itemMean_gender_male = 
                itemMean_gender_male / mean_item_population$itemMean_gender_male -1,
            relative_itemMean_gender_female = 
                itemMean_gender_female / mean_item_population$itemMean_gender_female -1,
            relative_itemMean_from_Atown = 
                itemMean_from_Atown / mean_item_population$itemMean_from_Atown -1,
            relative_itemMean_from_Btown = 
                itemMean_from_Btown / mean_item_population$itemMean_from_Btown -1,
            relative_itemMean_from_Ctown = 
                itemMean_from_Ctown / mean_item_population$itemMean_from_Ctown -1,
            relative_itemMean_from_Dtown = 
                itemMean_from_Dtown / mean_item_population$itemMean_from_Dtown -1
          ) %>% 
    select( 
            item_ID, 
            itemMean_age, relative_itemMean_age,
            itemMean_gender_male, relative_itemMean_gender_male, 
            itemMean_gender_female, relative_itemMean_gender_female,          
            itemMean_from_Atown, relative_itemMean_from_Atown,          
            itemMean_from_Btown, relative_itemMean_from_Btown,  
            itemMean_from_Ctown, relative_itemMean_from_Ctown,          
            itemMean_from_Dtown, relative_itemMean_from_Dtown  
          )

# add item popularity measure:
item_meanBased_features_traindata <- 
  item_meanBased_features_traindata %>% 
    left_join( 
                train_data %>% 
                  group_by( item_ID ) %>% 
                  summarise( item_popularity_count = n() ) %>% 
                  ungroup() %>% 
                  mutate( item_popularity_score = 
                            100 * item_popularity_count / sum(item_popularity_count)
                          ) %>% 
                  select( item_ID, item_popularity_score)
              ) %>% 
    select( item_ID, item_popularity_score, everything()  )
  

# check if the features have picked up the (unknown to model) true item attributes:
#item_meanBased_features_traindata %>% 
#  left_join( sim_items_attribute_table ) %>% 
#  mutate_all( round, digits=2 ) %>% #write.csv("clipboard")
#  View()
    
```

```{r}
train_data_complete <- 
  train_data %>% 
    # add user features:  
    left_join( sim_user_data ) %>% 
    # add item features:
    left_join( item_meanBased_features_traindata )
```

train random forest on the training data:
```{r}
fit_randomforest <- 
  ranger(
    y ~ 
        # user attributes:
        age + 
        gender_male + 
        gender_female +
        from_Atown +
        from_Btown + 
        from_Ctown + 
        from_Dtown +
        # item attributes:
        item_popularity_score + 
        itemMean_age + 
        itemMean_gender_male + 
        itemMean_gender_female + 
        itemMean_from_Atown + 
        itemMean_from_Btown + 
        itemMean_from_Ctown + 
        itemMean_from_Dtown +
        relative_itemMean_age + 
        relative_itemMean_gender_male + 
        relative_itemMean_gender_female + 
        relative_itemMean_from_Atown + 
        relative_itemMean_from_Btown + 
        relative_itemMean_from_Ctown + 
        relative_itemMean_from_Dtown,
    data = train_data_complete %>% mutate(y=factor(y)),
    classification = TRUE,
    probability = TRUE
  )
```

test set performance:

```{r}
test_data_complete <- 
  test_data %>%
    # add user features:  
    left_join( sim_user_data ) %>% 
    # add item features:
    left_join( item_meanBased_features_traindata )

model_predictions <- 
  predict(
          fit_randomforest, 
          data = test_data_complete,
          type = "response",
          verbose = TRUE
  )

test_data_complete$ypred <- model_predictions$predictions[,"1"]

cor( 
      test_data_complete$buy_prob,  #true buy probability
      test_data_complete$ypred      # estimates buy probability
)
```

Allow model to recommend 5 items to each user, and calculate total buy probability for each user across those top 5 items:

```{r}
choose_model_topN_item_recommendations_per_user <- 
  expand_grid(
    user_ID = sample( sim_user_data$user_ID),
    item_ID = unique(orders_history$item_ID)
  ) %>% 
    # add user features:
    left_join( sim_user_data ) %>% 
    # add item features:
    left_join( item_meanBased_features_traindata )

choose_model_topN_item_recommendations_per_user$item_score <- 
  predict(
          fit_randomforest, 
          data = choose_model_topN_item_recommendations_per_user,
          type = "response",
          verbose = TRUE
    )$predictions[,"1"]

choose_random_topN_item_recommendations_per_user <- 
  choose_model_topN_item_recommendations_per_user %>% 
    mutate( rand01 = runif( n() ) ) %>% 
    group_by( user_ID ) %>% 
    arrange( desc(rand01) ) %>% 
    slice_head( n=5 ) %>% 
    # calculate true buy probability per item:
    mutate( true_buy_prob = map2_dbl( 
                        .x = user_ID,
                        .y = item_ID,
                        .f = item_buy_prob_ftn
                      )
    )

choose_most_popular_topN_item_recommendations_per_user <- 
  choose_model_topN_item_recommendations_per_user %>% 
    group_by( user_ID ) %>% 
    arrange( desc(item_popularity_score) ) %>% 
    slice_head( n=5 ) %>% 
    # calculate true buy probability per item:
    mutate( true_buy_prob = map2_dbl( 
                        .x = user_ID,
                        .y = item_ID,
                        .f = item_buy_prob_ftn
                      )
    )

# take top 5 items per user:
choose_model_topN_item_recommendations_per_user <- 
  choose_model_topN_item_recommendations_per_user %>% 
    group_by( user_ID ) %>% 
    arrange( desc(item_score) ) %>% 
    slice_head( n=5 ) %>% 
    # calculate true buy probability per item:
    mutate( true_buy_prob = map2_dbl( 
                        .x = user_ID,
                        .y = item_ID,
                        .f = item_buy_prob_ftn
                      )
    )
  

result_calc <- 
  tibble(
    selection_method = c("random", "model"),
    mean_total_prob_per_user = 
      c(
          choose_random_topN_item_recommendations_per_user %>% 
            group_by(user_ID) %>% 
            summarise( total_true_buyprob_this_user = sum(true_buy_prob) ) %>% 
            summarise( mean_total_true_byprob_per_user = mean(total_true_buyprob_this_user) ) %>% 
            pull( mean_total_true_byprob_per_user ),
          choose_model_topN_item_recommendations_per_user %>% 
            group_by(user_ID) %>% 
            summarise( total_true_buyprob_this_user = sum(true_buy_prob) ) %>% 
            summarise( mean_total_true_byprob_per_user = mean(total_true_buyprob_this_user) ) %>% 
            pull( mean_total_true_byprob_per_user )        
       )
  )


result_calc

print( 
  paste0(
          "i.e. 100 * (", 
          round(result_calc$mean_total_prob_per_user[2],2),
          "/",  
          round(result_calc$mean_total_prob_per_user[1],2),
          "-1) = ",
          round(result_calc$mean_total_prob_per_user[2]/result_calc$mean_total_prob_per_user[1]-1,2)
       )
)
```

"mean_total_prob_per_user" is the average number of items consumed per user.

We can see that the model is working

However, we don't have the true probabilities when working with real data. Can we estimate this quantity using the data that we have? 

Have a look at top recommendations for a random user:

```{r}
random_user <- 
  tibble( 
        age = sample( 10:100, size=1 ),
        gender = sample( c("m","f"), size=1),
        location = sample( c("A_town", "B_town", "C_town", "D_town"), size=1 )
    )

print( paste0(" age:       ", random_user$age) )
print( paste0(" gender:    ", random_user$gender) )
print( paste0(" location:  ", random_user$location) )

score_all_products_this_user <- 
    sim_items_attribute_table %>% 
      mutate(
        # user features:
        age = random_user$age, 
        gender_male = if_else( random_user$gender=="m", 1, 0 ),
        gender_female = if_else( random_user$gender=="f", 1, 0 ),
        from_Atown = if_else( random_user$location=="A_town", 1, 0 ),
        from_Btown = if_else( random_user$location=="B_town", 1, 0 ),
        from_Ctown = if_else( random_user$location=="C_town", 1, 0 ),
        from_Dtown = if_else( random_user$location=="D_town", 1, 0 )
      ) %>% 
      left_join( item_meanBased_features_traindata )

score_all_products_this_user$item_model_score <- 
    predict(
          fit_randomforest, 
          data = score_all_products_this_user,
          type = "response",
          verbose = TRUE
    )$predictions[,"1"]

score_all_products_this_user <- 
  score_all_products_this_user %>% 
    mutate( 
            true_buyprob = 
              base_item_popularity + 
              0.1 * item_ageband_0_24 * (age < 25) +
              0.1 * item_ageband_25_39 * (age>=25 & age<40) +
              0.1 * item_ageband_40_64 * (age>=40 & age<65) +
              0.1 * item_ageband_65plus * (age>=65) +
              0.1 * item_male * gender_male +
              0.1 * item_female * gender_female +
              0.1 * item_townA * from_Atown +
              0.1 * item_townB * from_Btown +
              0.1 * item_townC * from_Ctown +
              0.1 * item_townD * from_Dtown
          ) %>% 
    select( 
            item_model_score,
            item_ID, 
            base_item_popularity,
            item_ageband_0_24,
            item_ageband_25_39,
            item_ageband_40_64,
            item_ageband_65plus,
            item_male,
            item_female,
            item_townA,
            item_townB,
            item_townC,
            item_townD
          ) %>% 
  arrange( desc(item_model_score) )

score_all_products_this_user


```





