---
title: "Constrained Linear Optimisation in R"
author: "Joseph Bolton"
date: "08 February 2019"
output: 
  html_document:
    theme: darkly
    highlight: espresso  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plotly)
library(dplyr)
```

The **constrOptim()** function is used to minimise an objective function based on linear inequality constraints. 

# Example 
Suppose that we want to maximise the function

$$f(\theta) \quad = \quad f(x,y) \quad = \quad x^2 + y^2$$ 

subject to the constraints

$$\begin{array}{lcl} 
   \frac{8}{5}x + y &\leq& 15 \\
   x &\geq& 0 \\
   y &\geq& 0 \\
\end{array}$$

Here is an interactive visualisation of the objective function and 3 constraints (i.e. the allowable $x,y$ region):

```{r show_the_function}

plot_ly( data = expand.grid( x = seq(-15, 15, 1),
                             y = seq(-15, 20, 1)
                           ) %>% 
                   mutate(f_xy = x^2+y^2), 
         x = ~x, 
         y = ~y, 
         z = ~f_xy
        ) %>% 
  add_trace(type="mesh3d", opacity=0.5) %>% 
  add_markers( data = expand.grid( x = seq(0,50,0.1),
                                   y = seq(0,50,0.1)
                                   ) %>%
                 mutate( constraint = 8/5*x + y,
                         f_xy = x^2 + y^2
                       ) %>%
                 filter( constraint <= 15 ),
                 x = ~x,
                 y = ~y,
                 z = 0,
                 size = 0.1
              ) 
  
     
```

**constrOptim()** requires the constraints in the form

$$\text{ui} \space \theta \quad \geq \quad \text{ci}$$

our constraints are 

$$\begin{array}{lcl} 
   \frac{8}{5}x + y &\leq& 15 \\
   x &\geq& 0 \\
   y &\geq& 0 \\
\end{array}$$

which in the notation $\text{ui} \space \theta \geq \text{ci}$ is:

$$\begin{bmatrix} -\frac{8}{5} & -1 \\ 1 & 0 \\ 0 & 1\end{bmatrix}\begin{bmatrix}x \\ y \end{bmatrix} \quad \geq \quad \begin{bmatrix} -15 \\ 0 \\ 0\end{bmatrix}$$

```{r specify_in_R}
my_ui <- matrix( c( -8/5, -1,
                       1,  0,
                       0,  1
                   ),
                 byrow=TRUE,
                 ncol=2
               )
my_ci <- matrix( c(-15,0,0), ncol=1)                 
                 
my_ui
my_ci
                
```

We specify the function to maximise:
```{r ftn_to_max}
ftn_to_max <- function(theta){ theta[1]^2 + theta[2]^2 }
```

**constrOptim()** by default minimises functions, so we maximise $f(\theta)$ by using **constrOptim()** to minimise $g(\theta)=-f(\theta)$:

$$g(\theta) \quad = \quad -f(\theta) \quad = \quad -x^2 - y^2$$

```{r ftn_to_min}
ftn_to_min <- function(theta){ -ftn_to_max(theta) }
```

**constrOptim()** requires the gradient function (function returning all partial first derivatives) of the objective function. This function in our case is:

$$\nabla g(\theta) \quad = \quad \Big(\frac{\partial g}{ \partial x} \space,\space \frac{\partial g}{ \partial y} \Big) \quad = \quad \Big( -2x \space, -2y \space \Big)$$

We specify the gradient function in R:

```{r give_grad_ftn}
grad_ftn <- function(theta){ c(-2*theta[1], -2*theta[2]) }
```

and for our grande finale:
```{r grand_finish}
do_the_optimisation <- 
    constrOptim( theta = c(1,1),    # starting parameter guesses
                 f = ftn_to_min,    # function to minimise
                 ui = my_ui,
                 ci = my_ci,
                 grad = grad_ftn
    )

do_the_optimisation
```

So the values maximising $f(x,y)$ are $x=$ `r round(do_the_optimisation$par[1],2)` and $y=$ `r round(do_the_optimisation$par[2],2)`

We can see that the calculation is correct by plotting this optimal value:
```{r last_show}

plot_ly( data = expand.grid( x = seq(-15, 15, 1),
                             y = seq(-15, 20, 1)
                           ) %>% 
                   mutate(f_xy = x^2+y^2), 
         x = ~x, 
         y = ~y, 
         z = ~f_xy
        ) %>% 
  add_trace(type="mesh3d", opacity=0.5) %>% 
  add_markers( data = expand.grid( x = seq(0,50,0.1),
                                   y = seq(0,50,0.1)
                                   ) %>%
                 mutate( constraint = 8/5*x + y,
                         f_xy = x^2 + y^2
                       ) %>%
                 filter( constraint <= 15 ),
                 x = ~x,
                 y = ~y,
                 z = 0,
                 size = 0.1
              ) %>% 
  add_markers( data = data.frame( x = do_the_optimisation$par[1],
                                  y = do_the_optimisation$par[2],
                                  f_xy = -do_the_optimisation$value
                                  ),
               x = ~x,
               y = ~y,
               z = ~f_xy
            )

```