---
title: "Content-based filtering"
author: "Joe"
output: 
  html_notebook:
    code: hide
---

still to do:

* does normalising the item vectors (rows) make any difference to cosine distance?

```{r}
library(tidyverse)
library(plotly)
library(knitr)
library(reshape2)
```

thoughts:

* need to down-weight product attributes that are very common 

* Do I need to deal with products that have more attributes?

* note that this method will very likely want to give customers products that they've just bought before.

```{r}
product_data <- 
  tibble( product_id = paste("product", 1:4, sep="_"), 
          product_desc = c( "red, green, green, green",
                           " red, blue, green",
                           " blue, blue, blue",
                           "red, red, blue"
                           )
        ) %>% 
  mutate(   tf_red = c( 1, 1, 0, 2),
           tf_blue = c( 0, 1, 2, 1 ),
          tf_green = c( 3, 1, 0, 0 ),
        ) %>% 
  mutate( red_idf = log( 4 / 3 ),
          blue_idf = log( 4 / 3 ),
          green_idf = log( 4 / 2 )
        ) %>% 
  mutate( red_tf_idf = tf_red * red_idf,
          blue_tf_idf = tf_blue * blue_idf,
          green_tf_idf = tf_green * green_idf
        )

product_data %>% kable()

```

plot the products in product space:
```{r fig.width=10, fig.height=10}

product1_vector <- product_data[1,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist()

product2_vector <- product_data[2,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist()

product3_vector <- product_data[3,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist()

product4_vector <- product_data[4,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist()

plot_ly( data = product_data,
            x = ~red_tf_idf,
            y = ~blue_tf_idf,
            z = ~green_tf_idf
       ) %>% 
  add_markers() %>% 
  add_trace( type = "scatter3d",
             x = c( 0, product1_vector[1] ), 
             y = c( 0, product1_vector[2] ),
             z = c( 0, product1_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            ) %>% 
    add_trace( type = "scatter3d",
             x = c( 0, product2_vector[1] ), 
             y = c( 0, product2_vector[2] ),
             z = c( 0, product2_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            ) %>% 
    add_trace( type = "scatter3d",
             x = c( 0, product3_vector[1] ), 
             y = c( 0, product3_vector[2] ),
             z = c( 0, product3_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            )  %>% 
    add_trace( type = "scatter3d",
               x = c( 0, product4_vector[1] ), 
               y = c( 0, product4_vector[2] ),
               z = c( 0, product4_vector[3] ),
               mode = "lines",
               line = list(width = 4, color="blue")
          )  %>% 
  layout(scene = list( xaxis = list(title = 'red'),
                       yaxis = list(title = 'blue'),
                       zaxis = list(title = 'green')
                    )
       )
```

Here is the product table for reference:
```{r}
product_data %>% 
  select( product_id, red_tf_idf, blue_tf_idf, green_tf_idf )
```

Suppose that customer $A$ buys four of product_3 and two of product_4. We average the 5 product vectors that customer $A$ has bought, giving customer $A$ an item profile of: 

```{r}
customerA_profile <- 
( 4*product3_vector + 2*product4_vector ) / 6 

names( customerA_profile ) <- c( "red", "blue", "green" )

customerA_profile
```

We can now plot this customer in the product space:
```{r fig.width=10, fig.height=10}
plot_ly( data = product_data,
            x = ~red_tf_idf,
            y = ~blue_tf_idf,
            z = ~green_tf_idf
       ) %>% 
  add_markers() %>% 
  add_trace( type = "scatter3d",
             x = c( 0, product1_vector[1] ), 
             y = c( 0, product1_vector[2] ),
             z = c( 0, product1_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            ) %>% 
    add_trace( type = "scatter3d",
             x = c( 0, product2_vector[1] ), 
             y = c( 0, product2_vector[2] ),
             z = c( 0, product2_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            ) %>% 
    add_trace( type = "scatter3d",
             x = c( 0, product3_vector[1] ), 
             y = c( 0, product3_vector[2] ),
             z = c( 0, product3_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            )  %>% 
    add_trace( type = "scatter3d",
             x = c( 0, product4_vector[1] ), 
             y = c( 0, product4_vector[2] ),
             z = c( 0, product4_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            )  %>% 
  layout(scene = list( xaxis = list(title = 'red'),
                       yaxis = list(title = 'blue'),
                       zaxis = list(title = 'green')
                    )
       ) %>% 
  add_trace( type = "scatter3d",
             x = c( 0, customerA_profile[1] ), 
             y = c( 0, customerA_profile[2] ),
             z = c( 0, customerA_profile[3] ),
             mode = "lines",
             line = list(width = 4, color="red")
            ) %>% 
    add_trace( type = "scatter3d",
             x = c( 0, customerA_profile[1] ), 
             y = c( 0, customerA_profile[2] ),
             z = c( 0, customerA_profile[3] ),
             mode = "points"
            )
```

Here is a reminder of what customer $A$ bought:
```{r}
product_data %>% 
  select( product_id, product_desc, red_tf_idf, blue_tf_idf, green_tf_idf ) %>% 
  bind_cols( ., tibble(n_bought=c(0, 0, 4, 2) ) ) %>% 
  select( product_id, product_desc, n_bought, everything() )
```

The angle $\theta$ between any 2 product vectors indicates their similarity. $cos(\theta)$ will therefore range as follows:

$\theta$   |   $cos(\theta)$
-----------|------------------
$0^\circ$  | 1
$30^\circ$ | 0.866
$45^\circ$ | 0.707
$60^\circ$ | 0.5
$90^\circ$ | 0

The cosine of the angle between 2 vectors $\overset{\rightarrow{}}{u}$ and $\overset{\rightarrow{}}{u}$ can be calculated using

$$cos\Big(\theta_{uv}\Big) \quad=\quad \displaystyle\frac{u \cdot v}{|u| \space |v|}$$

Here are the angles between the product vectors in our example:

```{r}
cosine_distmat <- matrix( nrow = 4, ncol = 4 )

for( i in 1:4 ){
  for( j in 1:4 ){
    
    product_i <- product_data[i,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist() 
    product_j <- product_data[j,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist() 
    
    cos_theta <- ( product_i %*% product_j ) / 
                 ( sqrt( sum(product_i^2) ) * sqrt( sum(product_j^2) ) )
    
    cosine_distmat[i,j] <- cos_theta
  }
}

rownames(cosine_distmat) <- colnames(cosine_distmat) <- paste("product", 1:4, sep="")

cosine_distmat %>% round(2)
```

Here are the cosine distances between the customer and each product:
```{r}
cbind( product = 1:4,
       cos_dist_to_customerA = 
       c( 
          ( product1_vector %*% customerA_profile ) / 
            ( sqrt( sum(product1_vector^2) ) * sqrt( sum(customerA_profile^2) ) )
          ,
          ( product2_vector %*% customerA_profile ) / 
            ( sqrt( sum(product2_vector^2) ) * sqrt( sum(customerA_profile^2) ) )
          ,
          ( product3_vector %*% customerA_profile ) / 
            ( sqrt( sum(product3_vector^2) ) * sqrt( sum(customerA_profile^2) ) )
          ,
          ( product4_vector %*% customerA_profile ) / 
            ( sqrt( sum(product4_vector^2) ) * sqrt( sum(customerA_profile^2) ) )
       )
     )
```

So, if we needed to offer this customer a new product (choosing from either product_1 or product_2 which they haven't bought yet), we'd offer them product_2.

# Higher-dimensional example
```{r}

generate_random_prod_ftn <- function(){
  
  product_vector <- 
        tibble( formal=0, casual=0, sports=0,
                shoes=0, pants=0, shirt=0, tie=0, bag=0, socks=0,
                black=0, red=0, green=0, blue=0, white=0,
                stripes=0, bold=0, leopard_print=0, spots=0, shiny=0
              )
  
  choose_use <- sample( 1:3, size=1 )
  if( choose_use == 1 ){ product_vector$formal <- 1 } else 
  if( choose_use == 2 ){ product_vector$casual <- 1 } else
                       { product_vector$sports <- 1 }
  
  choose_garment <- sample( 1:6, size=1 )
  if( choose_garment == 1 ){ product_vector$shoes <- 1 } else 
  if( choose_garment == 2 ){ product_vector$pants <- 1 } else
  if( choose_garment == 3 ){ product_vector$shirt <- 1 } else
  if( choose_garment == 4 ){ product_vector$tie <- 1 } else
  if( choose_garment == 5 ){ product_vector$bag <- 1 } else
                           { product_vector$socks <- 1 }
  
  choose_colour <- sample( 1:5, size=1 )
  if( choose_colour == 1 ){ product_vector$black <- 1 } else 
  if( choose_colour == 2 ){ product_vector$red <- 1 } else
  if( choose_colour == 3 ){ product_vector$green <- 1 } else
  if( choose_colour == 4 ){ product_vector$blue <- 1 } else
  { product_vector$white <- 1 }
  
  choose_style <- sample( 1:5, size=1 )
  if( choose_style == 1 ){ product_vector$stripes <- 1 } else 
  if( choose_style == 2 ){ product_vector$bold <- 1 } else
  if( choose_style == 3 ){ product_vector$leopard_print <- 1 } else
  if( choose_style == 4 ){ product_vector$spots <- 1 } else
  { product_vector$shiny <- 1 }

  return( product_vector )
}

set.seed(1925)
full_product_list <- 
  bind_rows(
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn()
  ) %>% 
  mutate( product_ID = row_number() 
) %>% 
  select( product_ID, everything() )
 
kable( full_product_list )
```

get item/product profiles (tf-idf) for full product list:
```{r}
product_colsums <- full_product_list %>% select(-product_ID) %>% colSums()
product_IDFs <- log( nrow(full_product_list) / product_colsums ) 

cbind( count = t( t(product_colsums) ), 
       IDF = t( t(product_IDFs) ) 
     )

product_IDFs_mat <- matrix( rep(product_IDFs, nrow(full_product_list)),
                            byrow = TRUE,
                            nrow = nrow(full_product_list)  
                          )
  
product_profiles <- full_product_list * cbind( 1, product_IDFs_mat )
```

Customer A buys the following:
```{r}
# customer A likes formal and sports stuff
# and he like shoes and socks a lot
# and he like black and white a lot
# and he likes shiny leopard print things

customerA_product_probs <- 
  full_product_list %>% 
    mutate( prob_this_product = 
              formal*10 + casual*1 + sports*10 +
              shoes*10 + pants*1 + shirt*1 + tie*1 + bag*4 + socks*10 +
              black*10 + red*2 + green*2 + blue*2 + white*10 +
              stripes*1 + bold*3 + leopard_print*10 + spots*2 + shiny*10
          ) %>% 
  pull( prob_this_product )

set.seed(6969420)
customerA_products_chosen <- 
  full_product_list[sample(1:nrow(full_product_list),prob=customerA_product_probs,size=5, replace=TRUE),]

customerA_products_chosen

customer_profile_vector <- 
  customerA_products_chosen %>% 
    select( product_ID ) %>% 
    left_join( product_profiles ) %>% 
    select( -product_ID ) %>% 
    summarise_all( mean ) %>% 
    unlist()

cosine_distances <- rep( NA, nrow(full_product_list) )
names(cosine_distances) <- paste("product", 1:nrow(full_product_list), sep="")

for( prod_i in 1:nrow(full_product_list) ){
  
  product_vector <- product_profiles[prod_i,-1] %>% unlist()
  
  cosine_distances[prod_i] <- 
                 ( product_vector %*% customer_profile_vector ) / 
                 ( sqrt( sum(product_vector^2) ) * sqrt( sum(customer_profile_vector^2) ) )
}
  
```

So, to recap, the customer bought:

```{r}
customerA_products_chosen
```

and his top matched products are:
( the ones he's already bought are obviously on top )

```{r}
bind_cols( full_product_list,
           tibble( cosine_dist = cosine_distances )
) %>% 
  arrange( desc(cosine_dist) ) %>% 
  kable()
```

# TF-IDF
```{r}
bind_rows( tibble( n = 10, df = 0:10 ) %>% mutate(idf = log( (1+n)/(1+df) ) +1 )
           ,
           tibble( n = 100, df = 0:100 ) %>% mutate(idf = log( (1+n)/(1+df) ) +1 )
           ,
           tibble( n = 1000, df = 0:1000 ) %>% mutate(idf = log( (1+n)/(1+df) ) +1 )
           ,
           tibble( n = 1e4L, df = 0:1e4L ) %>% mutate(idf = log( (1+n)/(1+df) ) +1 )
           ,
           tibble( n = 1e5L, df = 0:1e5L ) %>% mutate(idf = log( (1+n)/(1+df) ) +1 )
         ) %>% 
  ggplot( data = .,
          aes( x = df,
               y = idf
              )  
        ) +
  geom_line() +
  facet_wrap( ~n, scales = "free" ) +
  labs( title = "IDF values for different document sizes")
```

# other useful text tools
```{r}
library(tidyverse)
library(tidytext)
library(janeaustenr)
library(stringr)
library(SnowballC) # for the function wordStem() for  stemming words

text <- c("Because I could not stop for Death -",
          "He kindly stopped for me -",
          "The Carriage held but just Ourselves -",
          "and Immortality"
        )

text_df <- tibble(line = 1:4, text = text)

text_df

# tokenize our text dataframe:
# a 'token' is a meaningful unit of text (e.g. a word)
text_df %>% 
  unnest_tokens( output = "word",     # name of output column
                 input = "text",      # name of column containing the text to be tokenized 
                 to_lower = TRUE,     # convert text to lowercase
                 strip_punct = TRUE  # remove punctuation (.,-/ etc.)
               )

  
raw_books <- austen_books()

raw_books %>% slice(500:510)

original_books <- 
  raw_books %>% 
    group_by(book) %>%   
      # within each book:
      mutate( # add a unique line number to each line:
              linenumber = row_number(),    
              
              # add a chapter column by getting the chapter number out of the raw text
              chapter = cumsum(str_detect(text, regex("^chapter [\\divxlc]",
                                                     ignore_case = TRUE
                                                     )
                                          )
                               )
              ) %>%
      ungroup()

# look at a random few lines from different books:
original_books %>% sample_n(10)  

# put into tidy text format:
tidy_books <- 
  original_books %>%
    unnest_tokens(input="text", output="word")

tidy_books %>%  head(20)

# get a list of stop-words (words not useful in analysis, like 'the', 'of', 'to' etc.)
data(stop_words)

nrow(stop_words)
head(stop_words, 20)

tidy_books_no_stopwords <-
  tidy_books %>% 
    anti_join(stop_words)
  
tidy_books %>% head(20)
tidy_books_no_stopwords %>% head(20)

# count most common words across all books:
tidy_books_no_stopwords %>% 
  count( word, sort=TRUE ) %>% 
  head(20)

?str_extract
# We use str_extract() here because the UTF-8 encoded texts from Project 
# Gutenberg have some examples of words with underscores around them to indicate 
# emphasis (like italics). The tokenizer treated these as words, but we don't want 
# to count "_any_" separately from "any" as we saw in our initial data 
# exploration before choosing to use str_extract()
# go search for this text in https://www.tidytextmining.com/tidytext.html

tibble( word = c("word", "_word_", "%w%o%r%d%") ) %>% 
  mutate( word_clean1 = str_extract(word, "[a-z']+"),
          word_clean2 = str_remove_all( word, "\\W+" )
        )



# checking out tf-idf:
example_dat <- tibble( document = paste("document", 1:5, sep="_"),
                       document_text = c( "black and big black boots and black",
                                          "black pants formal and with buttons",
                                          "BIG black and red pants with big buttons",
                                          "black hat with red and green stripes",
                                          "blue formal waistcoat"
                                        ) 
                       )

get_tf.idf <- 
example_dat %>% 
  unnest_tokens( input="document_text", output="word" ) %>% 
  group_by( document, word ) %>% 
  summarise( n_times_in_doc = n() ) %>% 
  ungroup() %>% 
  group_by( word ) %>% 
  mutate( n_docs_containing_word = n_distinct(document) ) %>% 
  group_by( document ) %>% 
  mutate( n_words_in_doc = sum(n_times_in_doc) ) %>% 
  ungroup() %>% 
  mutate( joe_tf = n_times_in_doc / n_words_in_doc ) %>% 
  ungroup() %>% 
  mutate( `n_docs/n_docs_containing_word` = 5 / n_docs_containing_word ) %>% 
  mutate( joe_idf = log(`n_docs/n_docs_containing_word`) ) %>% 
  mutate( joe_tf_idf = joe_tf * joe_idf ) %>% 
  bind_tf_idf( tbl =.,
               term = word,
               document = document,
               n = n_times_in_doc
  ) %>% 
  arrange( document, desc(tf_idf) ) 
# 
# # turn our tidy dataframe into a document-term matrix:
# example_dat %>% 
#   unnest_tokens( input="document_text", output="word" ) %>% 
#   group_by( document, word ) %>% 
#   summarise( n_times_in_doc = n() ) %>% 
#   ungroup() %>% 
#   cast_dtm( term = word,
#             document = document
#           )

# checking out stemming:
tibble( unstemmed_word = c("running", "run", "runs", "runned", "runner", "runners", "ran") ) %>% 
  mutate( stemmed_word = map_chr( .x = unstemmed_word,
                                .f = function(word){ wordStem( word, language="en" ) } 
                                )
  )


# check out document term matrix:
```
