---
title: "Content-based filtering"
author: "Joe"
output: 
  html_notebook:
    code: hide
---

```{r}
library(tidyverse)
library(plotly)
library(knitr)
library(reshape2)
```

thoughts:

* need to down-weight product attributes that are very common 

* Do I need to deal with products that have more attributes?

* note that this method will very likely want to give customers products that they've just bought before.

```{r}
product_data <- 
  tibble( product_id = paste("product", 1:4, sep="_"), 
          product_desc = c( "red, green, green, green",
                           " red, blue, green",
                           " blue, blue, blue",
                           "red, red, blue"
                           )
        ) %>% 
  mutate(   tf_red = c( 1, 1, 0, 2),
           tf_blue = c( 0, 1, 2, 1 ),
          tf_green = c( 3, 1, 0, 0 ),
        ) %>% 
  mutate( red_idf = log( 4 / 3 ),
          blue_idf = log( 4 / 3 ),
          green_idf = log( 4 / 2 )
        ) %>% 
  mutate( red_tf_idf = tf_red * red_idf,
          blue_tf_idf = tf_blue * blue_idf,
          green_tf_idf = tf_green * green_idf
        )

product_data %>% kable()

```

plot the products in product space:
```{r fig.width=9, fig.height=6}

product1_vector <- product_data[1,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist()

product2_vector <- product_data[2,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist()

product3_vector <- product_data[3,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist()

product4_vector <- product_data[4,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist()

plot_ly( data = product_data,
            x = ~red_tf_idf,
            y = ~blue_tf_idf,
            z = ~green_tf_idf
       ) %>% 
  add_markers() %>% 
  add_trace( type = "scatter3d",
             x = c( 0, product1_vector[1] ), 
             y = c( 0, product1_vector[2] ),
             z = c( 0, product1_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            ) %>% 
    add_trace( type = "scatter3d",
             x = c( 0, product2_vector[1] ), 
             y = c( 0, product2_vector[2] ),
             z = c( 0, product2_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            ) %>% 
    add_trace( type = "scatter3d",
             x = c( 0, product3_vector[1] ), 
             y = c( 0, product3_vector[2] ),
             z = c( 0, product3_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            )  %>% 
    add_trace( type = "scatter3d",
               x = c( 0, product4_vector[1] ), 
               y = c( 0, product4_vector[2] ),
               z = c( 0, product4_vector[3] ),
               mode = "lines",
               line = list(width = 4, color="blue")
          )  %>% 
  layout(scene = list( xaxis = list(title = 'red'),
                       yaxis = list(title = 'blue'),
                       zaxis = list(title = 'green')
                    )
       )
```

Here is the product table for reference:
```{r}
product_data %>% 
  select( product_id, red_tf_idf, blue_tf_idf, green_tf_idf )
```

Suppose that customer $A$ buys four of product_2 and two of product_1. We average the 5 product vectors that customer $A$ has bought, giving customer $A$ an item profile of: 

```{r}
customerA_profile <- 
( 4*product3_vector + 2*product4_vector ) / 6 

names( customerA_profile ) <- c( "red", "blue", "green" )

customerA_profile
```

We can now plot this customer in the product space:
```{r fig.width=9, fig.height=6}
plot_ly( data = product_data,
            x = ~red_tf_idf,
            y = ~blue_tf_idf,
            z = ~green_tf_idf
       ) %>% 
  add_markers() %>% 
  add_trace( type = "scatter3d",
             x = c( 0, product1_vector[1] ), 
             y = c( 0, product1_vector[2] ),
             z = c( 0, product1_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            ) %>% 
    add_trace( type = "scatter3d",
             x = c( 0, product2_vector[1] ), 
             y = c( 0, product2_vector[2] ),
             z = c( 0, product2_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            ) %>% 
    add_trace( type = "scatter3d",
             x = c( 0, product3_vector[1] ), 
             y = c( 0, product3_vector[2] ),
             z = c( 0, product3_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            )  %>% 
    add_trace( type = "scatter3d",
             x = c( 0, product4_vector[1] ), 
             y = c( 0, product4_vector[2] ),
             z = c( 0, product4_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            )  %>% 
  layout(scene = list( xaxis = list(title = 'red'),
                       yaxis = list(title = 'blue'),
                       zaxis = list(title = 'green')
                    )
       ) %>% 
  add_trace( type = "scatter3d",
             x = c( 0, customerA_profile[1] ), 
             y = c( 0, customerA_profile[2] ),
             z = c( 0, customerA_profile[3] ),
             mode = "lines",
             line = list(width = 4, color="red")
            ) %>% 
    add_trace( type = "scatter3d",
             x = c( 0, customerA_profile[1] ), 
             y = c( 0, customerA_profile[2] ),
             z = c( 0, customerA_profile[3] ),
             mode = "points"
            )
```

Here is a reminder of what customer $A$ bought:
```{r}
product_data %>% 
  select( product_id, product_desc, red_tf_idf, blue_tf_idf, green_tf_idf ) %>% 
  bind_cols( ., tibble(n_bought=c(0, 0, 4, 2) ) ) %>% 
  select( product_id, product_desc, n_bought, everything() )
```

The angle $\theta$ between any 2 product vectors indicates their similarity. $cos(\theta)$ will therefore range as follows:

$\theta$   |   $cos(\theta)$
-----------|------------------
$0^\circ$  | 1
$30^\circ$ | 0.866
$45^\circ$ | 0.707
$60^\circ$ | 0.5
$90^\circ$ | 0

The cosine of the angle between 2 vectors $\overset{\rightarrow{}}{u}$ and $\overset{\rightarrow{}}{u}$ can be calculated using

$$cos\Big(\theta_{uv}\Big) \quad=\quad \displaystyle\frac{u \cdot v}{|u| \space |v|}$$

Here are the angles between the product vectors in our example:

```{r}
cosine_distmat <- matrix( nrow = 4, ncol = 4 )

for( i in 1:4 ){
  for( j in 1:4 ){
    
    product_i <- product_data[i,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist() 
    product_j <- product_data[j,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist() 
    
    cos_theta <- ( product_i %*% product_j ) / 
                 ( sqrt( sum(product_i^2) ) * sqrt( sum(product_j^2) ) )
    
    cosine_distmat[i,j] <- cos_theta
  }
}

rownames(cosine_distmat) <- colnames(cosine_distmat) <- paste("product", 1:4, sep="")

cosine_distmat %>% round(2)
```

Here are the cosine distances between the customer and each product:
```{r}
cbind( product = 1:4,
       cos_dist_to_customerA = 
       c( 
          ( product1_vector %*% customerA_profile ) / 
            ( sqrt( sum(product1_vector^2) ) * sqrt( sum(customerA_profile^2) ) )
          ,
          ( product2_vector %*% customerA_profile ) / 
            ( sqrt( sum(product2_vector^2) ) * sqrt( sum(customerA_profile^2) ) )
          ,
          ( product3_vector %*% customerA_profile ) / 
            ( sqrt( sum(product3_vector^2) ) * sqrt( sum(customerA_profile^2) ) )
          ,
          ( product4_vector %*% customerA_profile ) / 
            ( sqrt( sum(product4_vector^2) ) * sqrt( sum(customerA_profile^2) ) )
       )
     )
```

So, if we needed to offer this customer a new product (choosing from either product_1 or product_2 which they haven't bought yet), we'd offer them product_2.

# Higher-dimensional example
```{r}

generate_random_prod_ftn <- function(){
  
  product_vector <- 
        tibble( formal=0, casual=0, sports=0,
                shoes=0, pants=0, shirt=0, tie=0, bag=0, socks=0,
                black=0, red=0, green=0, blue=0, white=0,
                stripes=0, bold=0, leopard_print=0, spots=0, shiny=0
              )
  
  choose_use <- sample( 1:3, size=1 )
  if( choose_use == 1 ){ product_vector$formal <- 1 } else 
  if( choose_use == 2 ){ product_vector$casual <- 1 } else
                       { product_vector$sports <- 1 }
  
  choose_garment <- sample( 1:6, size=1 )
  if( choose_garment == 1 ){ product_vector$shoes <- 1 } else 
  if( choose_garment == 2 ){ product_vector$pants <- 1 } else
  if( choose_garment == 3 ){ product_vector$shirt <- 1 } else
  if( choose_garment == 4 ){ product_vector$tie <- 1 } else
  if( choose_garment == 5 ){ product_vector$bag <- 1 } else
                           { product_vector$socks <- 1 }
  
  choose_colour <- sample( 1:5, size=1 )
  if( choose_colour == 1 ){ product_vector$black <- 1 } else 
  if( choose_colour == 2 ){ product_vector$red <- 1 } else
  if( choose_colour == 3 ){ product_vector$green <- 1 } else
  if( choose_colour == 4 ){ product_vector$blue <- 1 } else
  { product_vector$white <- 1 }
  
  choose_style <- sample( 1:5, size=1 )
  if( choose_style == 1 ){ product_vector$stripes <- 1 } else 
  if( choose_style == 2 ){ product_vector$bold <- 1 } else
  if( choose_style == 3 ){ product_vector$leopard_print <- 1 } else
  if( choose_style == 4 ){ product_vector$spots <- 1 } else
  { product_vector$shiny <- 1 }

  return( product_vector )
}

set.seed(1925)
full_product_list <- 
  bind_rows(
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn()
  ) %>% 
  mutate( product_ID = row_number() 
) %>% 
  select( product_ID, everything() )
 
kable( full_product_list )
```

get item/product profiles (tf-idf) for full product list:
```{r}
product_colsums <- full_product_list %>% select(-product_ID) %>% colSums()
product_IDFs <- log( nrow(full_product_list) / product_colsums ) 

product_IDFs

product_IDFs_mat <- matrix( rep(product_IDFs, nrow(full_product_list)),
                            byrow = TRUE,
                            nrow = nrow(full_product_list)  
                          )
  
product_profiles <- full_product_list * cbind( 1, product_IDFs_mat )
```

Customer A buys the following:
```{r}
# customer A likes formal and sports stuff
# and he like shoes and socks a lot
# and he like black and white a lot
# and he likes shiny leopard print things

customerA_product_probs <- 
  full_product_list %>% 
    mutate( prob_this_product = 
              formal*10 + casual*1 + sports*10 +
              shoes*10 + pants*1 + shirt*1 + tie*1 + bag*4 + socks*10 +
              black*10 + red*2 + green*2 + blue*2 + white*10 +
              stripes*1 + bold*3 + leopard_print*10 + spots*2 + shiny*10
          ) %>% 
  pull( prob_this_product )

set.seed(6969420)
customerA_products_chosen <- 
  full_product_list[sample(1:nrow(full_product_list),prob=customerA_product_probs,size=5, replace=TRUE),]

customerA_products_chosen

customer_profile_vector <- 
  customerA_products_chosen %>% 
    select( product_ID ) %>% 
    left_join( product_profiles ) %>% 
    select( -product_ID ) %>% 
    summarise_all( mean ) %>% 
    unlist()

cosine_distances <- rep( NA, nrow(full_product_list) )
names(cosine_distances) <- paste("product", 1:nrow(full_product_list), sep="")

for( prod_i in 1:nrow(full_product_list) ){
  
  product_vector <- product_profiles[prod_i,-1] %>% unlist()
  
  cosine_distances[prod_i] <- 
                 ( product_vector %*% customer_profile_vector ) / 
                 ( sqrt( sum(product_vector^2) ) * sqrt( sum(customer_profile_vector^2) ) )
}
  
```

So, to recap, the customer bought:

```{r}
customerA_products_chosen
```

and his top matched products are:
( the ones he's already bought are obviously on top )

```{r}
bind_cols( full_product_list,
           tibble( cosine_dist = cosine_distances )
) %>% 
  arrange( desc(cosine_dist) ) %>% 
  kable()
```



