---
title: "cuped_cupac_experiment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(MASS)
library(tidyverse)
library(ranger)
```

interesting to add:

* reduce variance using a regression model $y = \beta_0 + \text{treatment effect} + \beta_1X_1 + \beta_2X_2+...$ 

* reduce variance with a more complex regression model 

* add ranger CUPAC

* is propensity score matching working correctly?

* simulate y by partitioning the feature space

* simulate y by nearest neighbour method

# CUPED (*Controlled-experiment Using Pre-Experiment Data*)

https://www.exp-platform.com/Documents/2013-02-CUPED-ImprovingSensitivityOfControlledExperiments.pdf

$\hat{Y}_{cv}=\overline{Y}-\theta\overline{X} + \theta E\Big[X\Big]$ is an unbiased estimator of $E\Big[Y\Big]$. However, it requires us to know $E\Big[X\Big]$.

The variance of $\hat{Y}_{cv}$ is minimized where $\theta=\displaystyle\frac{COV(Y,X)}{VAR(X)}$. Using this $\theta$, the variance of the estimator is $VAR\Big[\hat{Y}_{cv}\Big]=VAR[\overline{Y}](1-\rho^2)$, where $\rho$ is the correlation between $Y$ and $X$.  

Given that $E\Big[X^{(t)}\Big]=E\Big[X^{(c)}\Big]$, then $\Delta_{cv}=\hat{Y}_{cv}^{(t)}-\hat{Y}_{cv}^{(c)}$ is an unbiased estimator of $\delta=E\Big[\Delta\Big]=E\Big[\overline{Y}^{(t)}-\overline{Y}^{(c)}\Big]$. The variance of this estimator is $VAR\Big[\Delta_{cv}\Big]=VAR\Big[\Delta\Big](1-\rho^2)$, where $\rho$ is the correlation between $Y$ and $X$. In order for the estimate to be unbiased, the same $\theta$ must be used for both control and treatment - it can be estimated on both groups pooled together. 

# CUPAC (*Control Using Predictions As Covariates*)

https://doordash.engineering/2020/06/08/improving-experimental-power-through-control-using-predictions-as-covariate-cupac/

**CUPAC** is **CUPED**, where the control variable $X$ is a prediction $\hat{Y}$ of $Y$ produced by a predictive model trained on pre-experiment (i.e. pre-treatment) data.  

```{r}
run_1_experiment <- function( n_control_group,
                              n_target_group,
                              n_training_group,
                              population_sim_method = c( 
                                                         "multivariate_normal",
                                                         "linear_model_with_interactions",
                                                         "tree"        # not implemented yet
                                                       ),
                              effect_size
                            ){

   results_this_experiment <- c( true_difference = effect_size )
  
   # simulate populations --------------------------------------------------------------------------------
   if( population_sim_method == "multivariate_normal" ){
     
       control_mu <- c( y  = 5,
                        x1 = 2,
                        x2 = -1,
                        x3 = 0
                      )
       
       target_mu <- c( y  = control_mu["y"] + effect_size,
                       x1 = 2,
                       x2 = -1,
                       x3 = 0
                     )
       names(target_mu) <- c("y","x1","x2","x3")
       
       shared_sigma <- matrix( c(    # y     x1     x2     x3 
                                      1  ,   0.2,   0.5,   0.3, # y 
                                      0.2,   1  ,   0.7,   0.1, # x1 
                                      0.5,   0.7,   1  ,   0.6, # x2 
                                      0.3,   0.1,   0.6,   1     # x3 
                                 ),
                               byrow = TRUE,
                               nrow = 4,
                               ncol = 4
                              )
       colnames(shared_sigma) <- c("y","x1","x2","x3")
       rownames(shared_sigma) <- colnames(shared_sigma)
       
       control_group <- 
         MASS::mvrnorm( n     = n_control_group,
                        mu    = control_mu,
                        Sigma = shared_sigma
                       ) %>% 
         as_tibble()
       
       target_group <- 
         MASS::mvrnorm( n     = n_target_group,
                        mu    = target_mu,
                        Sigma = shared_sigma
                       ) %>% 
         as_tibble()
      
       training_data <- 
         MASS::mvrnorm( n     = n_training_group,
                        mu    = control_mu,
                        Sigma = shared_sigma
                       ) %>% 
         as_tibble()
   } else if( population_sim_method == "linear_model_with_interactions" ){
    
      x1_effect <- sample(0:10, size=1) 
      x2_effect <- sample(0:10, size=1)
      x3_effect <- sample(0:10, size=1)
      x1_x2_effect <- sample(0:10, size=1)
      x1_x3_effect <- sample(0:10, size=1)
      x2_x3_effect <- sample(0:10, size=1)
      x1_x2_x3_effect <- sample(0:10, size=1)
      
      control_group <- tibble( x1 = runif( n=n_control_group, min=-5, max=5 ),
                               x2 = runif( n=n_control_group, min=-5, max=5 ),
                               x3 = runif( n=n_control_group, min=-5, max=5 )
                             ) 
      control_group$y <- 
            x1_effect*control_group$x1 + 
            x2_effect*control_group$x2 + 
            x3_effect*control_group$x3 +
            x1_x2_effect*control_group$x1*control_group$x2 +
            x1_x3_effect*control_group$x1*control_group$x3 +
            x2_x3_effect*control_group$x2*control_group$x3 +
            x1_x2_x3_effect*control_group$x1*control_group$x2*control_group$x3 +
            rnorm( nrow(control_group), mean=0, sd=1 )
      control_group <- control_group %>% dplyr::select(y, x1, x2, x3)
      target_group <- tibble( x1 = runif( n=n_target_group, min=-5, max=5 ),
                               x2 = runif( n=n_target_group, min=-5, max=5 ),
                               x3 = runif( n=n_target_group, min=-5, max=5 )
                             ) 
      target_group$y <- 
            effect_size +
            x1_effect*target_group$x1 + 
            x2_effect*target_group$x2 + 
            x3_effect*target_group$x3 +
            x1_x2_effect*target_group$x1*target_group$x2 +
            x1_x3_effect*target_group$x1*target_group$x3 +
            x2_x3_effect*target_group$x2*target_group$x3 +
            x1_x2_x3_effect*target_group$x1*target_group$x2*target_group$x3 +
            rnorm( nrow(target_group), mean=0, sd=1 )  
      target_group <- target_group %>% dplyr::select(y, x1, x2, x3)
      training_data <- tibble( x1 = runif( n=n_training_group, min=-5, max=5 ),
                               x2 = runif( n=n_training_group, min=-5, max=5 ),
                               x3 = runif( n=n_training_group, min=-5, max=5 )
                             ) 
      training_data$y <- 
            x1_effect*training_data$x1 + 
            x2_effect*training_data$x2 + 
            x3_effect*training_data$x3 +
            x1_x2_effect*training_data$x1*training_data$x2 +
            x1_x3_effect*training_data$x1*training_data$x3 +
            x2_x3_effect*training_data$x2*training_data$x3 +
            x1_x2_x3_effect*training_data$x1*training_data$x2*training_data$x3 +
            rnorm( nrow(training_data), mean=0, sd=1 )     
      training_data <- training_data %>% dplyr::select(y, x1, x2, x3) 
  }
   
  # calculate estimators of E[Y|target] - E[Y|control] --------------------------------------------------
   
  # calculate simple difference mean(target y) - mean(control y):
  results_this_experiment["simple_difference"] <- mean(target_group$y) - mean(control_group$y)
   
  # using control variate X2, with unknown distribution (i.e. must estimate COV[Y,X2] and VAR[X2]): 
  pooled_y <- c( control_group$y, target_group$y ) 
  pooled_x2 <- c( control_group$x2, target_group$x2 ) 
  theta <- cov( pooled_y, pooled_x2 ) /
           var( pooled_x2 )
  results_this_experiment["cv_x2_UNknown_dbn"] <- 
             mean(target_group$y) -
             mean(control_group$y) -    
             ( theta * mean(target_group$x2) ) + 
             ( theta * mean(control_group$x2) ) 
  
  # create matched control group:
  # (sample with replacement)
  combined_target_control <- 
    bind_rows( target_group %>% dplyr::select(x1, x2, x3) %>% dplyr::mutate(test_control="test")
               ,
               control_group %>% dplyr::select(x1, x2, x3) %>% dplyr::mutate(test_control="control")  
            )
  distmat <- dist(      x = combined_target_control[, c("x1","x2","x3")],
                   method = "euclidean"
                 ) %>% 
              as.matrix()
  target_indices <- which(combined_target_control$test_control=="test")
  control_indices <- which(combined_target_control$test_control=="control")
  distmat <- distmat[ target_indices, control_indices ] # rows as only target, columns as only control 
  closest_control_indices <- apply(      X = distmat,
                                    MARGIN = 1,                       # apply function to each row
                                       FUN = function(row){ which.min(row) }
                                    )
  matched_control_group <- control_group[closest_control_indices, ]
  results_this_experiment["matched_control"] <- mean(target_group$y) - mean(matched_control_group$y)
  
  # CUPAC linear model
  training_data_for_model <- data.frame( y  = training_data$y,
                                         x1 = training_data$x1,
                                         x2 = training_data$x2,
                                         x3 = training_data$x3
                                       )
  fit_lm <- lm( y ~ x1 + x2 + x3 + .*., data=training_data_for_model  )      # including all degree-2 interactions too
  yhat_control <- predict(  object = fit_lm,
                           newdata = control_group
                         )
  yhat_target <- predict(  object = fit_lm,
                          newdata = target_group
                        )
  pooled_y <- c( control_group$y, target_group$y ) 
  pooled_yhat <- c( yhat_control, yhat_target ) 
  # estimate theta:
  theta <- cov( pooled_y, pooled_yhat ) /
           var( pooled_yhat )
  results_this_experiment["cupac_lm"] <- 
             mean(target_group$y) -
             mean(control_group$y) -    
             ( theta * mean(yhat_target) ) + 
             ( theta * mean(yhat_control) ) 
  
  # stratify y based on quantiles of X:
  combined_target_control <- 
      bind_rows( target_group %>% dplyr::mutate(test_control="test")
                 ,
                 control_group %>% dplyr::mutate(test_control="control")  
              ) %>% 
      mutate( x1_median = median(x1),
              x2_median = median(x2),
              x3_median = median(x3)
            ) %>% 
      mutate( stratum = paste0( as.numeric(x1>x1_median),
                                as.numeric(x2>x2_median),
                                as.numeric(x3>x3_median)
                              )
            ) 
  within_strata_calc <- 
    combined_target_control %>% 
        group_by( stratum ) %>% 
        summarise( n_test_in_stratum = sum( if_else(test_control=="test",1,0) ),
                   n_control_in_stratum = sum( if_else(test_control=="control",1,0) ),
                   sum_test_y_in_stratum = sum( if_else(test_control=="test",y,0) ),
                   sum_control_y_in_stratum = sum( if_else(test_control=="control",y,0) ),
                   .groups = "drop"     # drop grouping after summarise
                 ) %>% 
        filter( n_test_in_stratum > 0 & n_control_in_stratum > 0 ) %>%     # discard strata without both test & control
        mutate( # estimate treatment effect on y in each stratum:
                y_diff_est_in_stratum = sum_test_y_in_stratum/n_test_in_stratum - 
                                        sum_control_y_in_stratum/n_control_in_stratum
              ) 
  within_strata_calc$stratum_weight <- 
    ( within_strata_calc$n_test_in_stratum + within_strata_calc$n_control_in_stratum ) /
    ( sum(within_strata_calc$n_test_in_stratum) + sum(within_strata_calc$n_control_in_stratum) )
  
  results_this_experiment["y_stratified"] <- sum( within_strata_calc$y_diff_est_in_stratum * within_strata_calc$stratum_weight )
  
  # propensity score matching
  combined_target_control <- 
      bind_rows( target_group %>% dplyr::mutate(test_control="test")
                 ,
                 control_group %>% dplyr::mutate(test_control="control")  
              ) 
  propensity_model_logreg <- glm( formula = test_control ~ x1+x2+x3,
                                     data = combined_target_control %>% mutate( test_control = if_else(test_control=="test",1,0) ),
                                   family = binomial(link="logit") 
                                )
  target_group_propensity_scores <- predict(  object = propensity_model_logreg,
                                             newdata = target_group,
                                                type = "link"                # return scores on the link scale 
                                    )
  holdout_group_propensity_scores <- predict(  object = propensity_model_logreg,
                                              newdata = control_group,
                                                 type = "link"                # return scores on the link scale 
                                    )
  y_diff_vec <- rep( as.numeric(NA), n_target_group )
  for( target_customer_i in 1:n_target_group ){
      this_target_customer_propensity_score <- target_group_propensity_scores[ target_customer_i ]
      this_target_customer_y <- target_group$y[target_customer_i]
      closest_holdout_propensity_score <- which.min( abs(this_target_customer_propensity_score-holdout_group_propensity_scores) )
      propensity_matched_control_customer_y <- control_group$y[closest_holdout_propensity_score] 
      y_diff_vec[target_customer_i] <- this_target_customer_y - propensity_matched_control_customer_y     
  }
  results_this_experiment["y_matched_propensity_score"] <- mean(y_diff_vec)
  
  # reduce variance with linear regression model:
  combined_target_control <- 
    bind_rows( target_group %>% dplyr::mutate(test_control="test")
               ,
               control_group %>% dplyr::mutate(test_control="control")  
             ) %>% 
    mutate( test_control = factor(test_control, levels=c("control","test")) )
  fit_lm <- lm( formula = y ~ x1 + x2 + x3 + test_control,
                   data = combined_target_control    
              )
  results_this_experiment["reduce_variance_by_lm"] <- coef(fit_lm)["test_controltest"] 
  
  # repeated CUPED (one covariate at a time):
  # (note: could only include covariates X passing a chosen correlation threshold with Y)
  for( X_j in 1:3 ){
    
    
  }
  
  return( results_this_experiment )                   
}
```

run multiple experiments, and look at the distribution of the results:

```{r}

# user inputs ------------------------------------------
n_experiments_to_run <- 1000
n_control_group <- 50
n_target_group <- 200
n_training_group <- 200
population_simulation_method <- "multivariate_normal" 
true_effect_size <- 2
# ------------------------------------------------------

store_experiment_results_matrix <- matrix( 
                                           rep( as.numeric(NA), 8 * n_experiments_to_run ),
                                           ncol = 8
                                        )
colnames( store_experiment_results_matrix ) <- 
  c( "true_difference", 
     "simple_difference",
     "cv_x2_UNknown_dbn",
     "matched_control",
     "cupac_lm",
     "y_stratified",
     "y_matched_propensity_score",
     "reduce_variance_by_lm"
    )
rownames( store_experiment_results_matrix ) <- 
  paste( "experiment", 1:n_experiments_to_run ) 

for( i in 1:n_experiments_to_run ){
  
  store_experiment_results_matrix[i,] <-  
        run_1_experiment(       
                                n_control_group = n_control_group,
                                 n_target_group = n_target_group,
                               n_training_group = n_training_group,
                          population_sim_method = population_simulation_method,
                                    effect_size = true_effect_size
                        )  
}

head(store_experiment_results_matrix)
```

```{r fig.width=12, fig.height=5}

data_for_plot <- 
  tibble(            estimator = "simple_difference",
          estimated_difference = store_experiment_results_matrix[,"simple_difference"]  
        )
for( col_j in 3:ncol(store_experiment_results_matrix) ){
  
  estimator_name <- colnames(store_experiment_results_matrix)[col_j]
  data_for_plot <- bind_rows(
                              data_for_plot
                              ,
                              tibble(            estimator = estimator_name,
                                      estimated_difference = store_experiment_results_matrix[,estimator_name]
                                     )
                   )
}
 
ggplot( data = data_for_plot,
        aes(  
                    x = estimated_difference,
             group    = estimator,
             fill     = estimator,
             colour   = estimator,
             linetype = estimator
           )  
      ) + 
  geom_density( alpha=0.1 ) +
  labs( title = "Multivariate Normal y" )
  
```

do the same thing with a different y=1 simulation style:

```{r}
# user inputs ------------------------------------------
n_experiments_to_run <- 1000
n_control_group <- 50
n_target_group <- 200
n_training_group <- 200
population_simulation_method <- "linear_model_with_interactions" 
true_effect_size <- 2
# ------------------------------------------------------

store_experiment_results_matrix <- matrix( 
                                           rep( as.numeric(NA), 8 * n_experiments_to_run ),
                                           ncol = 8
                                        )
colnames( store_experiment_results_matrix ) <- 
  c( "true_difference", 
     "simple_difference",
     "cv_x2_UNknown_dbn",
     "matched_control",
     "cupac_lm",
     "y_stratified",
     "y_matched_propensity_score",
     "reduce_variance_by_lm"
    )
rownames( store_experiment_results_matrix ) <- 
  paste( "experiment", 1:n_experiments_to_run ) 

for( i in 1:n_experiments_to_run ){
  
  store_experiment_results_matrix[i,] <-  
        run_1_experiment(       
                                n_control_group = n_control_group,
                                 n_target_group = n_target_group,
                               n_training_group = n_training_group,
                          population_sim_method = population_simulation_method,
                                    effect_size = true_effect_size
                        )  
}

head(store_experiment_results_matrix)

```

```{r fig.width=10, fig.height=5}
data_for_plot <- 
  tibble(            estimator = "simple_difference",
          estimated_difference = store_experiment_results_matrix[,"simple_difference"]  
        )
for( col_j in 3:ncol(store_experiment_results_matrix) ){
  
  estimator_name <- colnames(store_experiment_results_matrix)[col_j]
  data_for_plot <- bind_rows(
                              data_for_plot
                              ,
                              tibble(            estimator = estimator_name,
                                      estimated_difference = store_experiment_results_matrix[,estimator_name]
                                     )
                   )
}
 
ggplot( data = data_for_plot,
        aes(        x = estimated_difference,
               group  = estimator,
               fill   = estimator,
               colour = estimator,
             linetype = estimator
           )  
      ) + 
  geom_density( alpha=0.1 ) +
  labs( title = "y Simulated by Linear Model" )
```


