---
title: "Track_records_visualisation"
author: "Joseph Bolton"
date: "17 February 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(ggcorrplot)
library(knitr)
library(GGally)             # for ggpairs()
library(tidyr)
library(formattable)
```

Here is the dataset, printed in full:

```{r data_load}
track_records_dataset <-
  read.csv( "C:/Users/Joe Bolton/Desktop/sync_to_google_drive/UCT honours work/Multivariate Analysis/data/men.csv") %>% 
  select(-Country) %>% 
  rename( country = countrylabel) %>% 
  mutate( country = as.factor(country) )
  
table_with_formatting <- 
    formattable( track_records_dataset, 
                 list(  m100 = color_tile("transparent", "pink"),
                        m200 = color_tile("transparent", "green"),
                        m400 = color_tile("transparent", "blue"),
                        m800 = color_tile("transparent", "orange"),
                        m1500 = color_tile("transparent", "red"),
                        m5000 = color_tile("transparent", "yellow"),
                        m5000 = color_tile("transparent", "purple"),
                        m10000 = color_tile("transparent", "dark grey"),
                        marathon = color_tile("transparent", "violet")
                     )
                )
    
table_with_formatting
```
    
# Scatterplot Matrix
```{r scaterplotmaytrickz, fig.width=9, fig.height=9}
GGally::ggpairs( track_records_dataset %>% select(-country),
                 diag = list(continuous = wrap("barDiag", bins=10) ),
                 upper = list(continuous = "density",
                              mapping = ggplot2::aes_string(fill = "..level..")
                             ),
                 lower = list(continuous = wrap("points", alpha = 0.3, size=1) )
                )

```

All of the variables (events) look very positively correlated, meaning that high performance in one event is a good predictor of high performance in any other event. 

The univariate distribution of each event record is more Gaussian for shorter events, and increasingly long-tailed on the right as events become longer (the long right tail is populated by very badly-performing countries).

# Correlation plot
```{r ggcorplotyo, fig.with=9}
corr_matrix <- 
  track_records_dataset %>% 
    select(-country) %>% 
    cor() %>% 
    round(., digits=2)

kable(corr_matrix)

ggcorrplot( corr = corr_matrix,               # correlation matrix to visualise
            type = "lower",         # only fill bottom diagonal
            ggtheme = theme_bw,     # simple theme for plot
            title = "Visualisation of Correlations Between Variables", 
            show.legend = TRUE,
            show.diag = TRUE,
            legend.title = "Correlation",
            hc.order = TRUE,         # order the coefficients       
            lab = TRUE,              # show correlation coefficients on the plot 
            lab_size = 3,            # size of displayed coefficients
            method="circle",   
            colors = c("tomato2", "white", "springgreen3")
          )
```

The variables are all positively correlated, meaning that countries with high performance in any one event are likely to be high-performers in any other event (and likewise for poorly performing countries).

The strongest correlations are between events with similar distances (e.g. short distance performance is strongly correlated with other short-distance event performance, and likewise long-distance event performance is strongly correlated with performance in other long-distance events).

# Best event by country

Now, event performances are reported by country, where each column of the data has been standardised (centred to have mean 0 and scaled to have standard deviation 1). This plot highlights whether each country is performing above or below the mean performance in each event. 

```{r labeltheoptions, fig.width=12, fig.height=10}
track_records_dataset %>%
  mutate_if( is.numeric, scale) %>% 
  gather( key="event", value="record", -country ) %>% 
  mutate( event = factor(event, 
                         levels = c("m100",
                                    "m200",
                                    "m400",
                                    "m800",
                                    "m1500",
                                    "m5000",
                                    "m10000",
                                    "marathon")
                         )
        ) %>% 
ggplot( data = .,
        aes( x = event, 
             y = record,
             fill = event
          )  
     ) +
 geom_bar( stat="identity", position="dodge") + 
 theme( axis.text.x = element_text(angle = 90, hjust = 1),
                 panel.background = element_rect(fill = "black",
                                colour = "black",
                                size = 0.5, linetype = "solid"),
                 panel.grid = element_blank()
      ) +
 facet_wrap( ~country ) +
 labs( title = "Event records by Country",
       subtitle = "Variables (events) centred to have mean 0 and scaled to have standard deviation 1"  
     )

```

Cook Islands performs by far the worst across all events, followed closely in poor performance by Samoa. 

U.S.A. is clearly the best in the short-distance events. 

Kenya looks like the strongest performer in all medium to long-distance events. 

India and Sweden perform almost exactly on the average (mean) in all events. 

Most of the Western European countries seem to perform better than average across all events: France, Germany, Great Britain, Italy, Netherlands, Switzerland, Portugal. 


# Compare country performance by event
```{r comparisons, fig.width=12, fig.height=12, warning=FALSE}

track_records_dataset %>%
  mutate_if( is.numeric, scale, scale=TRUE, center=FALSE ) %>% 
  gather( key="event", value="record", -country ) %>%
  ggplot( data = .,
          aes( x = country,
               y = record,
               colour = record,
               shape = event
             )
        ) +
  geom_point(size=3) +
  scale_colour_gradientn( colours = rainbow(4) ) + 
  theme( axis.text.x = element_text(angle=45, hjust=1, vjust=1),
         panel.background = element_rect(fill = "black",
                                colour = "black",
                                size = 0.5, linetype = "solid"),
         panel.grid = element_blank()
         ) +
  scale_shape_manual( values = c("@","#","$","%","&","?","X","O") ) +
  facet_grid( event~., scale="free_y") +
  geom_vline( xintercept = 0:100, colour = "white", alpha=0.3 ) +
  labs( title = "Country Performance by Event",
        subtitle = "each variable standardised to have standard deviation of 1"
  )
  
```

# Event Rankings by Country

```{r rankung, fig.width=12, fig.height=10}
track_records_dataset %>% 
  mutate_if(is.numeric, dense_rank ) %>% 
  gather( key="event", value="rank", -country ) %>% 
  mutate( event = factor(event, 
                       levels = c("m100",
                                  "m200",
                                  "m400",
                                  "m800",
                                  "m1500",
                                  "m5000",
                                  "m10000",
                                  "marathon")
                       )
        ) %>%
ggplot( data = .,
        aes( x = event,
             y = rank,
             fill = event
          )  
     ) +
 geom_bar( stat="identity", position="dodge") + 
 theme( axis.text.x = element_text(angle = 90, hjust = 1),
        panel.background = element_rect(fill = "black",
                  colour = "black",
                  size = 0.5, linetype = "solid"),
        panel.grid = element_blank()
      ) +
 facet_wrap( ~country ) +
 labs( title = "Rank of each Record, by Country" )
  
```
