---
title: "Time Series & Dynamic Regression. Fourier terms"
author: "Joe"
output: html_notebook
---

This information is taken from

https://kevinkotze.github.io/ts-5-decompose/

and

https://robjhyndman.com/nyc2018/3-1-Dynamic-Regression.pdf
https://robjhyndman.com/uwafiles/10-Dynamic-Regression.pdf

# Backshift notation
$$B y_t \quad=\quad y_{t-1}$$

$$\begin{array}{lcl} 
B^3 y_t 
&=& 
B B By_t \\
&=&
B B y_{t-1} \\
&=&
B y_{t-2} \\
&=&
y_{t-3} \\
\end{array}$$

a 12-period (seasonal) difference is (e.g. subtract observed same month last year if data is monthly):
$$\begin{array}{lcl}
(1-B^{12})y_t 
&=&
y_t - y_{t-12}
\end{array}$$

a first difference:
$$\begin{array}{lcl} 
y_t' &=& 
(1-B)y_t \\
&=& 
y_t - y_{t-1} \\
\end{array}$$

a $2^{nd}$-order difference:
$$\begin{array}{lcl}
y_t'' 
&=&
y_t' - y_{t-1}' \\
&=&
(y_t-y_{t-1}) - (y_{t-1}-y_{t-2}) \\
&=&
y_t - 2y_{t-1} + y_{t-2} \\
&=&
(1-B)^2y_t \\
&=&
(1-2B+B^2)y_t \\
&=&
y_t - 2y_{t-1} + y_{t-2} \\
\end{array}$$

a $\text{d}^{th}$-order difference is:
$$(1-B)^dy_t$$

seasonal difference (4 periods) followed by first difference
$$\begin{array}{lcl} 
(1-B)(1-B^4)y_t 
&=& 
(1-B)(y_t - y_{t-4}) \\
&=& 
y_t - y_{t-1} - y_{t-4} + y_{t-5} \\
&=&
(1 - B - B^4 + B^5)y_t
\end{array}$$

# (non-seasonal) Arima process
$y_t'$ is a differenced series (could be differenced more than once - $d$ times) 
$$\begin{array}{lcl}
y_t' 
&=&
\mu + \underbrace{ \phi_1 y'_{t-1} + \phi_2 y'_{t-2} + ... + \phi_p y'_{t-p}} + \underbrace{\phi_1\epsilon_{t-1}+\phi_2\epsilon_{t-2} + ... + \phi_q\epsilon_{t-q}} 
+ \epsilon_t   \\
(1-\phi_1B-\phi_2B^2-...-\phi_p B^p)(1-B)^dy_t &=& \mu + (1+\theta_1B+\theta_2B^2+...+\theta_qB^q)\epsilon_t \\
\end{array}$$

# Seasonal Arima process
Need to do this

# Regression with ARIMA errors
The (time-series) regression model (with $k$ explanatory variables) is

$$y_t \quad=\quad \beta_0 + \beta_1 x_{1t} + \beta_2 x_{2t} + ... + \beta_k x_{kt} + \epsilon_t$$

The normal OLS regression makes the assumption that the errors $\epsilon_t$ are independent (a *white noise* process). We wish to allow the $\epsilon_t$ to be autocorrelated. 

The linear regression model with an $\text{ARIMA}(p,d,q)$ error structure is:

$$\begin{array}{lcl}
y_t &=& \beta_0 + \beta_1 x_{1t} + \beta_2 x_{2t} + ... + \beta_k x_{kt} + \eta_t \\
\eta_t' 
&=&
\underbrace{ \phi_1 \eta'_{t-1} + \phi_2 \eta'_{t-2} + ... + \phi_p \eta'_{t-p}} + \underbrace{\phi_1\epsilon_{t-1}+\phi_2\epsilon_{t-2} + ... + \phi_q\epsilon_{t-q}} 
+ \epsilon_t   \\
(1-\phi_1B-\phi_2B^2-...-\phi_p B^p)(1-B)^d\eta_t 
&=& 
(1+\theta_1B+\theta_2B^2+...+\theta_qB^q)\epsilon_t \\
&\space& \text{where } \epsilon_t \text{ is a white noise process} \\
\end{array}$$

We don't want to assume that the series $\epsilon_t$ is *white noise* (i.e. no autocorrelation).

The regression model with ARIMA errors is

$$\begin{array}{lcl}
y_t &=& 
\beta_0 + \beta_1 x_{1t} + \beta_2 x_{2t} + ... + \beta_k x_{kt} + \eta_t \\
\\
\text{where} \quad \phi(B)(1-B)^d \eta_t &=& \theta(B) \epsilon_t 
\quad\quad \eta_t \text{ is an (non-seasonal) ARIMA process} 
\\
\end{array}$$

Before fitting the regression model with ARIMA errors, all variables (both features $X$ and response $y$) need to be differenced until they are stationary. The same order of differencing may be applied to all variables in order to aid interpretability. 






# to do: 
simulate data with autocorrelation and look at what happens to coefficient estimates and confidence intervals. 

also try fitting a model with the wrong autocorrelation structure. 






# other stuff

```{r setup}
library(tidyverse)
```

An observed time series can be considered a weighted sum of a number of different cyclical underlying series (with different frequencies). 

consider the function

$$\begin{array}{lcl} 
f\Big(t\Big) 
&=& 
a \space \cdot cos\Big(2\pi \omega t + \phi\Big) \\
&=& 
\underset{\displaystyle U_1}{\underbrace{a \cdot cos\phi}} \times cos\Big(2\pi\omega t\Big)
\quad+\quad
\underset{\displaystyle U_2}{\underbrace{-a \cdot sin\phi}} \times sin\Big(2\pi\omega t\Big)
\\
\end{array}$$

$\omega$ is cycles per unit time (frequency index)

$a = \sqrt{U_1^2+U_2^2}$ is the height/amplitude of the function

$\phi = tan^{-1}\Big(-\frac{U_1}{U_2}\Big)$ is the phase of the function (the starting point of the cosine function)

```{r}
t <- seq(1, 5, length.out=1000)

parameter_sets_to_try <- 
  tibble( omega = c(1, 1, 1,  0.5, 1, 2,  1,1,1),
              a = c(1, 2, 5,  2, 2, 2,    2,2,2),
            phi = c(0, 0, 0,  0, 0, 0,    0,pi,2*pi) 
        )

par(mfrow = c(3,3) )

for( i in 1:nrow(parameter_sets_to_try) ){
  
  omega <- parameter_sets_to_try$omega[i]
  a <- parameter_sets_to_try$a[i]
  phi <- parameter_sets_to_try$phi[i]
  
  plot( x = t,
        y = a * cos( 2*pi*omega*t + phi),
        main = bquote(omega~"="~.(omega)~"     a = "~.(a)~"     "*phi~"="~.(phi)),
        xlab = "t", ylab = "f(t)",
        ylim = c(-5, 5),
        type = "l"
  )
}
```

# Regression with fourier terms

A regression model with only fourier terms (and an intercept) is:

$$\begin{array}{lcl} y_t 
&=& 
\beta_0 + \bigg[ \displaystyle{\sum}_{k=1}^K \quad \alpha_k  \underset{sin\Big(\frac{2\pi k t}{m}\Big)}{\underbrace{s_k(t)}} + \beta_k  \underset{cos\Big(\frac{2\pi k t}{m}\Big)}{\underbrace{c_k(t)}} \Bigg] + \epsilon_t
\end{array}$$
