---
title: "Embedding Categorical Variables in Python"
output: html_document
---

```{r setup, include=FALSE}
library(reticulate)
reticulate::use_python("C:/Program Files (x86)/Microsoft Visual Studio/Shared/Python37_64/")
```

```{python}
import pandas as pd
import numpy as np
```

```{r}
library(tidyverse)
```

This article is flipping great:

https://machinelearningmastery.com/how-to-prepare-categorical-data-for-deep-learning-in-python/

```{python}
bc_data = pd.read_csv("https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer/breast-cancer.data",
                      header = None
                     )

bc_data_df = bc_data.copy()
bc_data_df.columns = ['class',       # 0 
                      'age',           # 1
                      'menopause',       # 2
                      'tumour-size',     # 3
                      'inv-nodes',     # 4
                      'node-caps',       # 5
                      'deg-malig',     # 6
                      'breast',       # 7
                      'breast-quad',    # 8
                      'irradiat (y)'        # 9
                     ]
bc_data = bc_data.values
```

```{r}
py$bc_data_df
```


Here is how R sees the data:
```{r}
py$bc_data[1:5,]
```

```{python}
feature_names = ['class',       # 0 
                 'age',           # 1
                 'menopause',       # 2
                 'tumour-size',     # 3
                 'inv-nodes',     # 4
                 'node-caps',       # 5
                 'deg-malig',     # 6
                 'breast',       # 7
                 'breast-quad',    # 8
                 'irradiat'        # 9
                ]
encoding_strategy = ['(1-hot)',       # 0 
                     '(ordinal)',     # 1
                     '(1-hot)',       # 2
                     '(ordinal)',     # 3
                     '(ordinal)',     # 4
                     '(embed 6D)',       # 5
                     '(ordinal)',     # 6
                     '(1-hot)',       # 7
                     '(embed 3D)',    # 8
                     'response (y) variable! (binary)'        # 9
                ]                
                
for i in range(len(bc_data[0])):
  print( feature_names[i] )
  print( encoding_strategy[i] )
  print( set( [x[i] for x in bc_data] ) )
  print('\n')
```

```{python}
y = bc_data[:,-1]
#y = y.reshape( (len(y), 1) )
y[0:10]
```


```{python}
X = bc_data[:,:-1]
X = X.astype(str)     # format all fields as string
X
```

here is how R sees the X data (each row a sample/observation, each column a feature): 
```{r}
py$X[1:5,]
```

train/test split:
```{python}
from sklearn.model_selection import train_test_split
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.33, random_state=1)
```

```{python}
X_train.shape
X_test.shape
y_train.shape
y_test.shape
```

```{python}
y_train[0:5]
```

make the response variable (y) numeric using LabelEncoder():
```{python}
from sklearn.preprocessing import LabelEncoder

lab_encoder = LabelEncoder()
lab_encoder.fit( y_train )
y_train_labenc = lab_encoder.transform( y_train )
y_test_labenc = lab_encoder.transform( y_test )

y_train[0:10]               # before using LabelEncoder()
y_train_labenc[0:10]        # after using LabelEncoder() 
```

```{python}
X_train[0]
X_train[1]
```

separate out the features which we've decided to treat as ordinal: 
```{python}
# pull out the columns (features) that we want to treat as ordinal:
X_train_ordinal = [ [ X_train[row][col] for col in [1,3,4,6] ] for row in range(len(X_train)) ]
X_test_ordinal = [ [ X_test[row][col] for col in [1,3,4,6] ] for row in range(len(X_test)) ]
```

```{r}
py$X_train_ordinal[1:4]
```

```{python}
from sklearn.preprocessing import OrdinalEncoder

ord_encoder = OrdinalEncoder(
                  categories = [ ['20-29', '30-39', '40-49', '50-59', '60-69', '70-79'],
                                 ['0-4','5-9','10-14','15-19','20-24','25-29','30-34','35-39','40-44','45-49','50-54'],
                                 ['0-2', '3-5', '6-8', '9-11', '12-14', '15-17', '24-26'],
                                 ['1','2','3']
                              ]

)

# train the encoder on the training data:
ord_encoder.fit( X_train_ordinal )

# apply the ordinal encoding to the training data: 
X_train_ordinal_encoded = ord_encoder.transform( X_train_ordinal )

# apply the ordinal encoding to the test data:
X_test_ordinal_encoded = ord_encoder.transform( X_test_ordinal )
```

Let's see the ordinal coding for a random observation (using row 95):
```{python}
[ X_train[95][i] for i in [1,3,4,6] ]   # the features that we chose for ordinal encoding, for observation 69 
X_train_ordinal_encoded[95]   # the ordinal encoded features (after encoding) for observation 69
```
separate out the features which we've decided to represent with 1-hot encoding: 

```{python}
# pull out the columns (features) that we want to treat as ordinal:
X_train_1hot = [ [ X_train[row][col] for col in [0, 2, 7] ] for row in range(len(X_train)) ]
X_test_1hot = [ [ X_test[row][col] for col in [0, 2, 7] ] for row in range(len(X_test)) ]
```

```{python}
from sklearn.preprocessing import OneHotEncoder  
one_hot_encoder = OneHotEncoder(  # you don't have to specify the category ordering for the encoding, but you can:
                     categories = [ ['no-recurrence-events', 'recurrence-events'],
                                    ['premeno', 'ge40', 'lt40'],
                                    ['right', 'left']
                                  ]
                                )

# train 1hot-encoder on X training data:
one_hot_encoder.fit( X_train_1hot )

X_train_1hot_encoded = one_hot_encoder.transform( X_train_1hot )
```

Let's have a look at a specific sample/observation. 

Here are the features of observation 97 from the original raw training data:
```{python}
[ f'{feature_names[i]}:  {X_train[97][i]}' for i in range(len(X_train[0])) ]
```
and here are the features of observation 97 which we decided to 1-hot encode:
```{python}
[ X_train[97][i] for i in [0,2,7] ]
```
here is the one-hot encoding of these variables:
```{python}
X_train_1hot_encoded[97].todense()
```

The first 2 values in the 1-hot vector encode the variable **class**, which has 2 possible labels:

* [1,0] is the encoding for **class**='no-recurrence-events'

* [0,1] is the 1-hot encoding for **class**='recurrence-events'

The 3rd, 4th, 5th values in the 1-hot vector encode the variable **menopause**, which has 3 possible labels:

* [1,0,0] is the 1-hot encoding for **menopause**='premeno'

* [0,1,0] is the 1-hot encoding for **menopause**='ge40'

* [0,0,1] is the 1-hot encoding for **menopause**='lt40'

The final 2 values (6th and 7th) in the 1-hot vector encode the variable ****, which has 2 possible labels:

* [1,0] is the 1-hot encoding for **breast**='right'

* [0,1] is the 1-hot encoding for **breast**='left'

Here are the feature (**node-caps** and **breast-quad**) that we've decided to represent with latent embeddings:

```{python}
X_train_embed = [ [ X_train[row][col] for col in [5,8] ] for row in range(len(X_train)) ]
X_test_embed = [ [ X_test[row][col] for col in [5,8] ] for row in range(len(X_test)) ]
X_train_embed[0:5]   # print out this variable for the first 5 observations
```

```{r}
py$X_train_embed[1:5] %>% reduce(rbind)
```

```{python}
X_train_enc = []
X_test_enc = []

for j in range(len(X_train_embed[0])):
  lab_enc = LabelEncoder()
  X_train_embed_col_j = [ X_train_embed[row][j] for row in range(len(X_train_embed)) ] 
  X_test_embed_col_j = [ X_test_embed[row][j] for row in range(len(X_test_embed)) ] 
  lab_enc.fit( X_train_embed_col_j )
  train_enc_col_j = lab_enc.transform( X_train_embed_col_j ) 
  test_enc_col_j = lab_enc.transform( X_test_embed_col_j ) 
  X_train_enc.append( train_enc_col_j )
  X_test_enc.append( test_enc_col_j )

```


```{python}
X_train_enc
```

```{python}
from keras.layers import Input, Embedding

store_input_layers = []
store_embed_layers = []

# feature "deg-malig" - (embed in 6-D) --------------------------------------------------------------------
n_labels = len( np.unique(X_train_enc[0]) )    # store number of unique labels/levels for this variable
input_layer = Input( shape=(1,) )
embed_layer = Embedding(n_labels, 6)(input_layer)
store_input_layers.append( input_layer )
store_embed_layers.append( embed_layer )

# feature "breast quad" - (embed in 3-D) --------------------------------------------------------------------
n_labels = len( np.unique(X_train_enc[1]) )    # store number of unique labels/levels for this variable
input_layer = Input( shape=(1,) )
embed_layer = Embedding(n_labels, 3)(input_layer)
store_input_layers.append( input_layer )
store_embed_layers.append( embed_layer )

```

```{python}
store_embed_layers
```

combine all of the input layers:
```{python}
from keras.layers.merge import concatenate

#merge_input_layers = concatenate( 
#                                  [ concatenate(store_embed_layers),
#                                    X_train_ordinal_encoded
#                                  ]
#)
```

```{python}
X_train_ordinal_encoded
```


```{python}

```








































