---
title: "feature_embedding"
output:
  pdf_document: default
---

```{r setup}
# R
library(reticulate)
reticulate::use_python("C:/Program Files (x86)/Microsoft Visual Studio/Shared/Python37_64/")
```

```{python load_python_packages}
# python
import numpy as np
import random
import pandas as pd
```

```{python}
# python
user_list = [ 'user' + str(i+1) for i in range(10) ]
user_list
```

https://stackoverflow.com/questions/42762849/keras-embedding-layers-how-do-they-work

one-hot encode each user: 

```{python}
# python
def item_1hot(username):
  vec = [0]*len(user_list)
  vec[ user_list.index(username) ] = 1
  return vec
  
user_list_1hot = [ item_1hot(x) for x in user_list ]
user_list_1hot = np.array(user_list_1hot)

for s in user_list_1hot:
    print(*s)
```
So, for example, **user3** has the 1-hot encoding:

```{python}
# python
user_list_1hot[2]
```

So, using the 1-hot encoding, the users are represented by vectors in a 10-dimensional space.

**Embedding** means representing the users in a smaller-dimensional space. This process necessarily means a loss of information but this is offset by the gains from reducing the size of the input data (memory/space, computation time etc.).   

```{python}
# python
n_dims = 2   # desired number of dimensions of new space
n_users = len(user_list_1hot)   # number of users 

embedding_matrix = np.random.rand(n_users, n_dims).round(2)

print( embedding_matrix )
```

```{python}
# python
user_embeddings = user_list_1hot.dot( embedding_matrix  ) 
print(user_embeddings)
```

```{r}
# R
user_embedding_matrix_R <- py$user_embeddings
plot( x = user_embedding_matrix_R[,1],
       y = user_embedding_matrix_R[,2],
       xlab = "embedding dimension 1",
       ylab = "embedding dimension 2",
       pch = 16
     )
 text( x = user_embedding_matrix_R[,1],
       y = user_embedding_matrix_R[,2],
       labels = paste("user", 1:nrow(user_embedding_matrix_R), sep=""),
       col = 1:nrow(user_embedding_matrix_R)
     )
```

```{python}
user_ages = [ random.randint(0,90) for x in range(len(user_list) ) ]
user_ages
```

```{python}
user_list_1hot
```

```{python}
# python

# https://www.datacamp.com/community/tutorials/deep-learning-python

from keras.models import Sequential
from keras.layers import Embedding, Dense, Flatten

model = Sequential()    # initialise feed-forward neural network model

# add embedding layer:
model.add( 
           Embedding( input_dim = 11,
                      output_dim = 2,       # desired dimension of latent/embedded vector (output of the embedding)
                      embeddings_initializer = "uniform", # how starting values in embedding matrix are chosen 
                      embeddings_regularizer = None, # if there is regularization on the embedding matrix       
                      embeddings_constraint = None,  # constraints on embedding matrix
                      mask_zero = False,    # used when input vectors are different lengths
                      input_length = 10    # required if we want to connect to Dense layers later in the network
                    )
         )
         
# need to do this apparently:
model.add( Flatten() )

# (dense) output layer:
model.add( 
           Dense( 1, 
                  activation = 'linear'
                )
        )
        
model.compile( optimizer = 'adam',
                    loss = 'MSE'
             )        
        
model.fit( x = user_list_1hot,
           y = user_ages,
           epochs = 50,
           verbose = 0
         )        

```
* https://keras.io/api/layers/core_layers/embedding/



