---
title: "Simple Frequentist vs. Bayesian Hypothesis Testing Example"
author: "Joseph Bolton"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    theme: darkly
    highlight: espresso
---

```{r setup}
library(tidyverse)
```

*Bayesian Hypothesis Testing: An Alternative to Null Hypothesis Significance Testing (NHST) in Psychology and Social Sciences* by Alonso Ortega and Gorka Navarrete.

Suppose we observe 62 people, as follows: 

```{r echo=FALSE}
mydata <- matrix( c(23, 24,
                     8,  7
                    ), 
                  byrow = TRUE,
                  ncol = 2
                )
rownames(mydata) = c("Yes","No")
colnames(mydata) = c("Yes","No")

mydata <- as.table(mydata)
names( dimnames(mydata) ) <- c("Exposure","Disease")
```

```{r}

mydata
addmargins( mydata )
prop.table( mydata, margin=1 )
```

Is the proportion of people getting the disease truly different between those who got the drug (exposed) and those who didn't get the drug (unexposed)?  

# Frequentist Approach 
```{r}
chisq.test( mydata, correct=FALSE ) 
```

```{r echo=FALSE, eval=FALSE}

# Suppose that Pr[Disease|Exp] = 0.3
#              Pr[Disease|Not_Exp] = 0.6  
possibilities_table <- 
  tibble( a = 0:31 ) %>% 
    mutate( b = 47 - a,
            c = 31 - a,
            d = 15-c
          ) %>%
    filter( a>=0 & b>=0 & c>=0 & d>=0 ) %>%      # remove negative counts
    # check that the row and column totals are correct:
    mutate( exposureYes_total = a+b,
             exposureNo_total = c+d,
             DiseaseYes_total = a+c,
              DiseaseNo_total = b+d
          ) %>% s
    mutate( prob = 0.3^a * 0.7^b * 0.6^c * 0.4^d ) 
  
possibilities_table %>% 
  sample_n(    size = 1e4L,
            replace = TRUE,
             weight = prob  
          ) %>% 
  # count( a,b,c,d, prob ) %>% arrange( desc(n) )
  mutate( chisq_teststat = pmap_dbl( .l = list(a,b,c,d),
                                     .f = function(a,b,c,d){
                                        
                                            chisq.test( matrix( c(a,c,b,d), ncol=2 ) )$statistic
                                         
                                       }  
                                    ) 
        ) %>% 
  
  pull(chisq_teststat) %>% 
  hist( ., freq=FALSE, breaks=50 )
  
curve( dchisq(x, df=1), add=TRUE, col=2 )

# n_sim <- 1e4L
# store_test_statistics <- rep(NA, n_sim)
# for( i in 1:n_sim ){
#   
#   # create hypothetical data:
#   exposure_grp_n_diseased <- rbinom(n=1, size=23+24, prob=exposure_grp_p)
#   exposure_grp_n_NOT_diseased <- 23+24-exposure_grp_n_diseased 
#   non_exposure_grp_n_diseased <- rbinom(n=1, size=8+7, prob=non_exposure_grp_p)
#   non_exposure_grp_n_NOT_diseased <- 23+24-non_exposure_grp_n_diseased
#   simdata <- matrix( c(    exposure_grp_n_diseased, exposure_grp_n_NOT_diseased,
#                        non_exposure_grp_n_diseased, non_exposure_grp_n_NOT_diseased
#                        ), 
#                      byrow = TRUE,
#                       ncol = 2
#                    )
#   X_test_stat <- chisq.test(simdata)
#   store_test_statistics[i] <- X_test_stat$statistic
# }
# 
# hist(store_test_statistics, breaks=50, freq=FALSE)

```

# Bayesian Approach
Here is the data again: 
```{r}
mydata
addmargins(mydata)
prop.table(mydata, margin=1)
```

$$\begin{array}{lcl} BF_{H_a;H_0} &=& \displaystyle\frac{\text{Posterior Odds } \frac{H_a}{H_0}}{\text{Prior Odds } \frac{H_a}{H_0}} \\
&=& \displaystyle\frac{\frac{Pr\Big[H_a\Bigl|Data\Big]}{Pr\Big[H_0\Bigl|Data\Big]}}{\frac{Pr\Big[H_a\Big]}{Pr\Big[H_0\Big]}} \\
&=& \displaystyle\frac{\frac{\Big( Pr\Big[Data\Bigl|H_a\Big]\cdot Pr\Big[H_a\Big] \Big) / \Big(Pr\Big[Data\Bigl|H_a\Big]\cdot Pr\Big[H_a\Big]+Pr\Big[Data\Bigl|H_0\Big]\cdot Pr\Big[H_0\Big]\Big)}{\Big( Pr\Big[Data\Bigl|H_0\Big]\cdot Pr\Big[H_0\Big] \Big) / \Big(Pr\Big[Data\Bigl|H_a\Big]\cdot Pr\Big[H_a\Big]+Pr\Big[Data\Bigl|H_0\Big]\cdot Pr\Big[H_0\Big]\Big)}}{\frac{Pr\Big[H_a\Big]}{Pr\Big[H_0\Big]}} \\
&=& \displaystyle\frac{Pr\Big[Data\Bigl|H_a\Big]}{Pr\Big[Data\Bigl|H_0\Big]}\\
\end{array}$$




# References

https://lingpipe-blog.com/2009/10/13/bayesian-counterpart-to-fisher-exact-test-on-contingency-tables/