---
title: "Intuition for the CRVTW uplift model"
author: "Joseph Bolton"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    theme: darkly
    highlight: espresso
---

<style type="text/css">
@import url("https://fonts.googleapis.com/css?family=Inconsolata|Roboto+Mono|Ubuntu+Mono|Cutive+Mono");
pre{
display: inline-block;
font-size: 7px;
text-align: center;
font-family: Ubuntu Mono, monospace !important;
text-rendering: optimizeSpeed;
}
</style>
<pre style="font-size: 10px; font-family: &quot;Ubuntu Mono&quot;; background-color: rgb(0,0,0); color: rgb(250,250,250); line-height: 10px;">......................................................................
......................................................................
..............................$$$.....................................
...................................$$$$...............................
.................................$$$$$$$..............................
.................................$$$$$$$$.............................
..................................$$$$$$..............................
.........................$$$$$........................................
........................$$$$$$$.......................................
.........................$$$$$$.....$$$...............................
.......................................$$$............................
......................................$$$$$...........................
.................................$$$..$$$$$...........................
......................................................................
............................$$$$$....$................................
............................$$$$......................................
......................................................................
..........................$$$$$$$$$$$$$$$$............................
........................$$................$$..........................
........................$$.................$$.........................
.........................$$..............$$$..........................
..........................$...........$..$$...........................
..........................$...........$..$$...........................
..........................$..............$$...........................
..........................$..............$$...........................
..........................$..............$$...........................
..........................$.....$$$$.....$$...........................
..........................$.....$$$$$....$$...........................
..........................$......$$......$$...........................
..........................$..............$$...........................
..........................$..............$$...........................
.........................$$...............$$..........................
........................$............$.....$$.........................
......................$$...........$$$$$....$$........................
.....................$$.............$$$$......$$......................
....................$$.........................$$.....................
..................$$............................$$....................
.................$$..$$$$$$$$$$..$$$$$$$.....$$$..$$..................
................$$..$$$$$$$$$$$.$$$$$$$$......$$$..$$.................
..............$$..$$$$$$$$$$$$$$$$$$$$$$......$$$$...$................
.............$$..$$$$$$$$$$$$$$$...$$$$$$...$$$$$$$$..$$..............
............$...$$$$$$$$$$$$$$$$$.$$$$$$$$$$$$$$$$$$$..$$.............
..........$$..$$$$$$$$$$$$$$$$$$$$$$$....$$$$$$$$$$$$$...$$...........
.........$$..$$$$$$$$$$$$$$$$$$...$$$....$$$$$$$$$$$$$$$..$$..........
........$$..$$$$$$$$$$$$$$$$$$$..$$$$$$$$$$$$$$$$$$$$$$$$..$$.........
......$$...$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$...$$.......
.....$$..$$$$$$$$$$$$$$$$$$$$$$$$..$$$$$$$$$$$$$$$$$$$$$$$$$..$$......
....$$..$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$..$$.....
....$..$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$..$.....
....$..$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$..$.....
.....$.........................................................$$.....
......$$$...................................................$$$.......
..........$.$..........................................$$$$...........
......................................................................</pre>

This page is a very brief intuitive introduction to the **Continuous Response
Variable Transformation with Weightings** (CRVTW) uplift model discussed by the paper:

"*Response Transformation and Profit Decomposition for Revenue Uplift Modeling*" 

[Robin M. Gubela, Stefan Lessmann, Szymon Jaroszewicz. 2019]

```{r setup, echo=FALSE}
library(splines)
```

Consider the following data, in which we are interested in estimating how an administered treatment has affected outcome $y$ for units with different values of feature $x$. 

```{r echo=FALSE, fig.width=8, fig.height=8}
set.seed(9713)
sample_size_n <- 1000
y <- rexp( n=sample_size_n, rate=0.05 )
x <- runif( n=sample_size_n, min=0, max=100 )
treat_grp <- sample( 0:1, size=sample_size_n, replace=TRUE) 
#treat_effect <- 
  
# plot( y ~ x, col=treat_grp+2, pch=16 )
# legend( "topright",
#         legend = c("control","treatment"),
#         col = 2:3,
#         pch = 16
#       )
# # true uplift:
# lines( x = seq(0,100,0.1),
#        y = dnorm( seq(0,100,0.1), mean=50, sd=10 )*1000
#      )

par(bg="black", col="white", col.axis="white", col.lab="white", col.main="white")

true_uplift_effect <- treat_grp*1000*dnorm(x, mean=50, sd=12) + rnorm( n=length(y), mean=0, sd=20 )  

y <- y + true_uplift_effect     # add the true uplift effect

y <- abs(y)       # make sure that spend can not be negative

plot( y ~ x, col=treat_grp+2, pch=16 )
legend( "topright",
        legend = c("control group (untreated)","treated group"),
        col = 2:3,
        pch = 16,
        box.col = "white"
      )
```

There appears to be a treatment effect somewhere between $40<x<60$.

We create the following transformed outcome variable:  

$$z_i \quad=\quad \begin{cases} \hspace{3.5mm}\Bigg(\displaystyle\frac{n}{n_{treated}}\Bigg) \cdot y_i \quad \text{ if unit } i \text{ received treatment} \\ - \Bigg(\displaystyle\frac{n}{n_{control}}\Bigg) \cdot y_i \quad \text{ if unit } i \text{ did not receive treatment (control group)} \end{cases}$$

Here is how this transformed outcome variable $z$ looks:

```{r fig.height=8, fig.width=8, echo=FALSE}
par( mfrow = c(2,1), mar=c(5,5,5,5),
     bg="black", col="white", col.axis="white", col.lab="white", col.main="white"
     )

factor_treat1 <- length(treat_grp) / sum(treat_grp)
factor_treat2 <- length(treat_grp) / sum(treat_grp==0)

z_crvtw <- ifelse( treat_grp==1, factor_treat1, -factor_treat2 ) * y

plot( y ~ x, col=treat_grp+2, pch=16, main="original data" )
legend( "topright",
        legend = c("control group","treated group"),
        col = 2:3,
        pch = 16,
        box.col = "white"
      )
plot( z_crvtw ~ x, col=treat_grp+2, pch=16, main="transformed outcome [z]" )
abline( h = 0 )
legend( "topright",
        legend = c("control group","treated group"),
        col = 2:3,
        pch = 16,
        box.col = "white"
      )
```

We fit a natural spline regression model to the transformed outcome $z$:

```{r echo=FALSE, fig.height=8, fig.width=8}
fit_lm_z <- lm( z ~ ns(x, df=6), 
                data = data.frame( z = z_crvtw, x = x )
              ) 
  
x_seq <- seq(0, 100, by=0.1 )

par( mfrow = c(2,1), mar=c(5,5,5,5),
     bg="black", col="white", col.axis="white", col.lab="white", col.main="white"
     )

plot( y ~ x, col=treat_grp+2, pch=16, main="original data" )
legend( "topright",
        legend = c("control group","treated group"),
        col = 2:3,
        pch = 16,
        box.col = "white"
      )
plot( z_crvtw ~ x, col=treat_grp+2, pch=16, main="transformed outcome [z]" )
abline( h = 0 )
lines( x = x_seq,
       y = predict( fit_lm_z, newdata = data.frame(x=x_seq) ),
       col = 4,
       lwd = 2
     )
legend( "topright",
        legend = c("natural spline regression fit"),
        col = 4,
        lty = 1,
        box.col = "white"
      )
```

This fitted model gives an estimate of the Conditional Average Treatment Effect (*CATE*) $\space\tau(x)$ which, in a random sampling context, is: 

$$\tau(x) \quad=\quad E\Big[Y_i\space\Bigl|X_i=x, \space\text{Treatment}=Yes\Big] \quad\mathbf{-}\quad E\Big[Y_i\space\Bigl|X_i=x, \space\text{Treatment}=No\Big]$$

It does this since: 

$$E\Big[Z_i\Bigl|X_i=x\Big] \quad=\quad E\Big[Y_i\space\Bigl|X_i=x, \space\text{Treatment}=Yes\Big] \quad\mathbf{-}\quad E\Big[Y_i\space\Bigl|X_i=x, \space\text{Treatment}=No\Big]$$

Here is a comparison of the estimated uplift effect with the true uplift effect (I only know the true uplift effect because I simulated it): 

```{r fig.height=10, fig.width=8, echo=FALSE}

par( mfrow = c(2,1), mar=c(5,5,5,5),
     bg="black", col="white", col.axis="white", col.lab="white", col.main="white"
     )

plot( y ~ x, col=treat_grp+2, pch=16, main = "original data" )
legend( "topright",
        legend = c("control group","treated group"),
        col = 2:3,
        pch = 16,
        box.col="white"
      )

plot( z_crvtw ~ x, col=treat_grp+2, pch=16, main="transformed outcome Z (with estimated uplift signal)"
    )
abline( h = 0 )
lines( x = x_seq,
       y = predict( fit_lm_z, newdata = data.frame(x=x_seq) ),
       col = 4,
       lwd = 2
     )
legend( "topright",
        legend = c("natural spline regression fit"),
        col = 4,
        lty = 1,
        box.col="white"
      )

plot( y ~ x, col=treat_grp+2, pch=16, main = "original data (with estimated uplift signal)" )
legend( "topright",
        legend = c("natural spline regression fit"),
        col = 4,
        lty = 1,
        box.col="white"
      )
lines( x = x_seq,
       y = predict( fit_lm_z, newdata = data.frame(x=x_seq) ),
       col = 4,
       lwd = 2
     )

plot( y ~ x, col=treat_grp+2, pch=16,
        main = "true uplift signal"
      )
legend( "topright",
        legend = c("true uplift signal"),
        col = "white",
        lty = 1,
        box.col="white",
        lwd = 2
      )
# true uplift:
lines( x = seq(0,100,0.1),
       y = dnorm( seq(0,100,0.1), mean=50, sd=12 )*1000
     )
abline( h=0 )
```