---
title: "Likelihood Ratio Tests: Developing Intuition via Simulation"
output: html_notebook
---

```{r setup}
library(tidyverse)
```

Suppose that we have a univariate random variable $Y$ with the distribution $f(y|\theta)$, where $\theta$ is a vector containing the parameters of the distribution function $f()$.  

If we consider the function $f()$ to be a function of the parameters $\theta$ rather than of the data $y$, then we call $f()$ the *likelihood function* $\mathcal{L}(\theta|y)$

The *maximum likelihood estimators* of the parameters $\theta$ are the values of $\theta$ which maximise the value of $\mathcal{L}(\theta|y)$. Out of all possible parameter values, these are the parameters which are the *most likely* to have generated our particular observed data $y$. 

The *Likelihood-Ratio test statistic* $-2\cdot ln(LR)$ is 

$$\begin{array}{lcl} -2\cdot ln(LR) &=& -2\cdot ln\Bigg(\displaystyle\frac{\mathcal{L}(\hat\theta_r|y)}{\mathcal{L}(\hat\theta|y)}\Bigg) \\
&=& -2\Bigg[ ln\Big(\mathcal{L}(\hat\theta_r|y)\Big) - ln\Big(\mathcal{L}(\hat\theta|y)\Big)\Bigg]  \\
\end{array}$$

where $\hat\theta$ is the maximum-likelihood estimator of $\theta$ (i.e. the value of $\hat\theta$ maximising $\mathcal{L}(\hat\theta|y)$) and $\hat\theta_r$   

$$y_i \quad=\quad \begin{cases} 1 \quad \text{ if diseased patient has died} \\
0 \quad \text{ if diseased patient lived} \\
\end{cases}$$

Each $y_i$ is an independent random draw from a Bernoulli distribution $Y_i\sim Bernoulli\Big(p_i\Big)$

The true data-generating process is:

$$\begin{array}{lcl}
Pr\Big[Y_i=1\Big] 
&=& E\Big[Y_i\Big] \\
&=& 0.2 \cdot \Big(1 \cdot \mathbb{I}_{\{\}}\Big)
\end{array}$$

```{r}
simdata_ftn <- function(n){
  
  patient_id <- 1:n
  baseline_prob <- rep(0.1, n)
  gender <- sample( c("male","female"), size=n, replace=TRUE )
  age <- round( abs( rnorm( n=n, mean=30, sd=20 ) ) +1 )
  bmi <- sample( 5:35, size=n, replace=TRUE )
  
  disease_prob <- 
    baseline_prob *
      if_else( gender=="male", 1.5, 1 ) *
      (1+age/50)
  
  disease_outcome <- sapply( disease_prob, 
                             function(p){ sample( 0:1, size=1, prob=c(1-p,p) ) }  
                           )
  
  tibble( patient_id = patient_id,
          baseline_prob = baseline_prob,
          gender = gender,
          age = age,
          bmi = bmi,
          disease_prob = disease_prob,
          disease_outcome = disease_outcome
        ) %>% 
    return(.)
    
}
```

```{r}
# true_coefs <- list()
# true_coefs$intercept <- -5
# true_coefs$gender <- 0.5
# true_coefs$age <- 0.12
# true_coefs$bmi <- 0

simdata_nicedata_ftn <- function(n, true_coefs){
  
  patient_id <- 1:n
  gender <- sample( c("male","female"), size=n, replace=TRUE )
  age <- round( abs( rnorm( n=n, mean=30, sd=20 ) ) +1 )
  bmi <- sample( 5:35, size=n, replace=TRUE )
  
  eta <- rep(true_coefs$intercept, n) +
         true_coefs$gender*(gender=="male") +
         true_coefs$age*age
    
  disease_prob <- 1 / (1+exp(-eta) ) 

    # hist(disease_prob)
  # plot( age ~ disease_prob )
  # plot( factor(gender)~disease_prob )
  # 
  
  disease_outcome <- sapply( disease_prob, 
                             function(p){ sample( 0:1, size=1, prob=c(1-p,p) ) }  
                           )
  
  tibble( patient_id = patient_id,
          gender = gender,
          age = age,
          bmi = bmi,
          disease_prob = disease_prob,
          disease_outcome = disease_outcome
        ) %>% 
    return(.)
    
}
```

test out the data-simulating function:
```{r}
simdata_ftn(n=100)
```

```{r}
n_sims <- 100

store_beta_coef_estimates <- matrix( rep(NA,n_sims*4),ncol=4 )
colnames(store_beta_coef_estimates) <- c("[intercept]","gender","age","bmi")
store_beta_coef_SE_estimates <- store_beta_coef_estimates 

store_logLR_stat <- rep(NA, n_sims) 

true_coefs <- list()
true_coefs$intercept <- -5
true_coefs$gender <- 0.5
true_coefs$age <- 0.12
true_coefs$bmi <- 0

for( i in 1:n_sims ){
  
  generate_data <- simdata_nicedata_ftn(n=200, true_coefs=true_coefs)
  
  fit_glm_with_bmi <- glm( disease_outcome ~ gender + age + bmi,
                                   family = binomial(link="logit"),
                                   data = generate_data   
                                 )
  fit_glm_wout_bmi <- glm( disease_outcome ~ gender + age,
                           family = binomial(link="logit"),
                           data = generate_data   
                         )
  
  store_beta_coef_estimates[i, ] <- coef(fit_glm_with_bmi)
  store_beta_coef_SE_estimates[i, ] <- summary(fit_glm_with_bmi)[["coefficients"]][,"Std. Error"]

  y <- generate_data$disease_outcome
  # glm with bmi:
  pi_hat <- predict(fit_glm_with_bmi, type="response") 
  loglik_glm_with_bmi <- sum( y * log(pi_hat) + (1-y)*log(1-pi_hat) )  # same as logLik(fit_glm_with_bmi)  
  # glm without bmi:
  pi_hat <- predict(fit_glm_wout_bmi, type="response") 
  loglik_glm_wout_bmi <- sum( y * log(pi_hat) + (1-y)*log(1-pi_hat) )  # same as logLik(fit_glm_wout_bmi)  

  store_logLR_stat[i] <- -2 * (loglik_glm_wout_bmi - loglik_glm_with_bmi) 
}


```

The true distribution of the *Likelihood-Ratio test statistic* $-2\cdot ln(LR)$ over all of the simulations is:
```{r}
hist(store_logLR_stat, breaks=50, freq=FALSE)
curve( dchisq(x, df=1), col=2, add=TRUE )
```

The (true) distributions of the beta coefficients are:
```{r}

for( col_j in 1:ncol(store_beta_coef_estimates) ){
  
  hist( store_beta_coef_estimates[,col_j], 
          freq = FALSE, 
        breaks = 50,
          main = colnames(store_beta_coef_estimates)[col_j]
      )
  curve( dnorm(    x = x, 
                 mean = mean(store_beta_coef_estimates[,col_j]),
                   sd = sd(store_beta_coef_estimates[,col_j])
               ),
          add = TRUE,
          col = 2
        )
}

```

Suppose, though, that the data follows some other true distribution $g(y|\theta)$ as follows:

$$g(y|\theta) \quad=\quad dothisjoe$$
```{r eval=FALSE}

n_sims <- 100

store_beta_coef_estimates <- matrix( rep(NA,n_sims*4),ncol=4 )
colnames(store_beta_coef_estimates) <- c("[intercept]","gender","age","bmi")
store_beta_coef_SE_estimates <- store_beta_coef_estimates 

store_logLR_stat <- rep(NA, n_sims) 

for( i in 1:n_sims ){
  
  generate_data <- simdata_ftn(n=200)
  
  fit_glm_with_bmi <- glm( disease_outcome ~ gender + age + bmi,
                                   family = binomial(link="logit"),
                                   data = generate_data   
                                 )
  fit_glm_wout_bmi <- glm( disease_outcome ~ gender + age,
                           family = binomial(link="logit"),
                           data = generate_data   
                         )
  
  store_beta_coef_estimates[i, ] <- coef(fit_glm_with_bmi)
  store_beta_coef_SE_estimates[i, ] <- summary(fit_glm_with_bmi)[["coefficients"]][,"Std. Error"]

  y <- generate_data$disease_outcome
  # glm with bmi:
  pi_hat <- predict(fit_glm_with_bmi, type="response") 
  loglik_glm_with_bmi <- sum( y * log(pi_hat) + (1-y)*log(1-pi_hat) )  # same as logLik(fit_glm_with_bmi)  
  # glm without bmi:
  pi_hat <- predict(fit_glm_wout_bmi, type="response") 
  loglik_glm_wout_bmi <- sum( y * log(pi_hat) + (1-y)*log(1-pi_hat) )  # same as logLik(fit_glm_wout_bmi)  

  store_logLR_stat[i] <- -2 * (loglik_glm_wout_bmi - loglik_glm_with_bmi) 
}
```

# References

1. http://nitro.biosci.arizona.edu/courses/EEB519A-2007/pdfs/MLE.pdf
