---
title: "R Notebook"
author: "Joseph Bolton"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_notebook
---

Original paper: https://arxiv.org/abs/1602.04938

A very good introductory lecture on LIME: https://www.youtube.com/watch?v=CY3t11vuuOM

LIME interface in R: https://cran.r-project.org/web/packages/lime/vignettes/Understanding_lime.html

Using H2o in R: https://www.h2o.ai/wp-content/uploads/2018/01/RBooklet.pdf

```{r setup}
library(tidyverse)
library(lime)
#library(randomForest)
library(h2o)
```

**LIME** (**L**ocal **I**nterpretable **M**odel-agnostic **E**xplanations) is an algorithm that aims to make the predictions of complicated or 'black box' predictive models more transparent by creating simplified model explanation plots. 

**LIME** does this by assuming that *locally* (i.e. within a group of observations that are very similar to one another), the predictions of a complicated model can be approximated well by a much simpler model (a model whose output a human can easily interpret and understand). 

The **LIME** framework has been designed to be *model agnostic*, meaning that the algorithm works for the predictions of any classification or regression model (even ones that haven't been invented yet). Although the original paper introduces **LIME** in the context of classification modelling, the R implementation of **LIME** can be applied to both classification and regression frameworks. 

# The LIME algorithm
For a given observation $\{\overset{\rightarrow{}}{\mathbf{x}}_i, y_i\}$, our complicated/'black box' model provides us with a prediction $\hat{y}_i = f(\overset{\rightarrow{}}{\mathbf{x}}_i)$ of $y_i$. 

We wish to *explain* to some degree how the complicated/'black box' model came to this prediction $\hat{y}_i$.

The **LIME** algorithm is:

1. Permute the observation $\overset{\rightarrow{}}{\mathbf{x}}_i$ $n$ times. This creates $n$ observations 'similar' to $\overset{\rightarrow{}}{\mathbf{x}}_i$. How an observation is *permuted* is explained later.  

2. For each of the $n$ permuted observations $\overset{\rightarrow{}}{\mathbf{x}}_i^{(j)}$, get a predicted value $\hat{y}_i^{(j)}=f\big(\overset{\rightarrow{}}{\mathbf{x}}_i^{(j)}\big)$ for this observation from the complicated/'black box' model.

3. For each of the $n$ permuted observations, calculate the distance $\delta_{ij}$ between the original observation $\overset{\rightarrow{}}{\mathbf{x}}_i$ and the permuted observation $\overset{\rightarrow{}}{\mathbf{x}}_i^{(j)}$ using a chosen distance metric (e.g. *euclidean distance*).

4. Convert the distance measure $\delta_{ij}$ into a similarity score. The similarity metric chosen will depend on the type of data.

5. Select the $m$ 'most predictive' features from the data. 

6. Fit a simple model to the permuted data: the simple model tries to predict the complex model predictions $\hat{y}_i^{(j)}$ using the $m$ chosen features from $\mathbf{x}_i^{(j)}$.

7. Explain the prediction $\hat{y}_i^{(j)}$ of the complex/'black box' model by interpretting the fit of the simple model.


# **LIME** example on a classification model
From the help documentation for the **LIME** package in R:
"Out of the box, lime supports the following model objects:
* train from caret

* WrappedModel from mlr

* xgb.Booster from xgboost

* H2OModel from h2o

* keras.engine.training.Model from keras

* lda from MASS (used for low-dependency examples)"

```{r}
# load the [biopsy] dataset from the MASS package
data( "biopsy", package="MASS" )

# count number of rows with an NA value in them:
apply( X = biopsy, 
       MARGIN = 1, 
       FUN = anyNA
     ) %>% 
  sum()   

biopsy[ apply(biopsy,1,anyNA), ]

# remove the ID column:
biopsy <- biopsy %>% select(-ID)

biopsy <- na.omit(biopsy)       # remove rows with any NA values in them

# clean up the column names:
names(biopsy) <- c( 'clump_thickness',
                    'uniformity_of_cell_size',
                    'uniformity_of_cell_shape',
                    'marginal_adhesion',
                    'single_epithelial_cell_size',
                    'bare_nuclei',
                    'bland_chromatin',
                    'normal_nucleoli',
                    'mitoses',
                    'class'
                   )

# make the table into tibble() format:
biopsy <- as_tibble(biopsy) %>% select( class, everything() )

# have a look at the data:
biopsy

# pull 10 observations out as a test set:
test_indices <- sample( seq_len(nrow(biopsy)), size=10 )
biopsy_train <- biopsy[-test_indices,]
biopsy_test <- biopsy[test_indices,]
```

Let's fit a random forest model (with default hyperparameters) to predict the class of each tumour:
```{r}
# fit_rf <- 
#   randomForest::randomForest( class ~ .,      # explain class using all of the predictors
#                                data = biopsy_train
#                             )
# 
# fit_rf
```

Predictions on the holdout/test set (10 observations):
```{r}
# bind_cols( tibble( true_label = biopsy_test$class ) 
#            ,
#            predict(fit_rf, newdata=biopsy_test, type="prob") %>% 
#              as_tibble() %>% 
#              rename( prob_benign = benign,
#                      prob_malignant = malignant  
#                    )
#           )
# 
# test_set_predictions <- predict(fit_rf, newdata=biopsy_test, type="response")
# 
# tibble( true_label = biopsy_test$class,
#         prediction = test_set_predictions
#       ) %>%
#   table()
```

fit a random forest model using H2o:
```{r}
h2o.init( nthreads=-1 )     # start an H2o instance
h2o.clusterInfo()

# Clean slate - just in case the cluster was already running:
h2o.removeAll() 

# convert our training dataset testing dataset into format useable by H2o:
biopsy_train_h2o <- as.h2o(biopsy_train)
biopsy_test_h2o <- as.h2o(biopsy_test)

fit_biopsy_random_forest <- 
  h2o.randomForest( training_frame = biopsy_train_h2o,
                    y = 1,
                    x = 2:10, 
                    model_id = "biopsy_random_forest",
                    ntrees = 200
                  )

fit_biopsy_random_forest

```


Run the LIME model explainer:  
```{r}
set.seed(69420)
lime_explainer <- 
  lime::lime(              x = biopsy_train,         # the data used to train the model
                       model = fit_biopsy_random_forest,    # the trained 'black box' model  
              bin_continuous = TRUE,        # bin continuous variables in the simple explanatory model
                      n_bins = 4,           # number of bins when binning the continuous variables
               quantile_bins = FALSE,       # define bin bounds using quantiles (else all bins same width) 
                 use_density = NULL         # use kernel density estimation for permuting continuous vars?
            )

```

Here is what the **LIME** object contains:
```{r}
lime_explainer
```

Now, we use the object *lime_explainer* created above using the **lime()** function to explain the set of 10 predictions that we set aside as a test set: 
```{r}
explanation <- lime::explain(         x = biopsy_test,     # the test set data
                              explainer = lime_explainer,  # the explainer object made in the previous chunk
                               n_labels = 1,               # number of response labels explain 
                             n_features = 4,               # number of features to use in simple model
                         n_permutations = 30,       # number of times to permute each observation 
                         feature_select = "auto",   # feature selection method in simple model
                               dist_fun = "gower",  # how to measure distance between original observation
                                                    #     and its permutations  
                           kernel_width = NULL,     # not used if using GOWER distance metric
                              gower_pow = 1,        # parameter of Gower distance metric (this is the default value)
                     single_explanation = FALSE     # "pool all text in x into a single explanation"
                     
                            )
```

```{r}
explanation
```

```{r}
lime::plot_explanations(explanation)
```

```{r}
h2o.shutdown(prompt = FALSE)
```