---
title: "Logistic Regression"
author: "Joe"
date: "15 October 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(knitr)      # for kable()
```

Suppose that we have observed 10,000 individuals in the course of a study. We want to see if we can use their

* age

* number of previous convictions, prior to study.

* employment status (EMPLOYED or UNEMPLOYED)

* gender (MALE of FEMALE)

to predict whether each individual committed a crime in the study period or not. We also want to try to measure/infer how each of these features (variables) affect the probability of someone committing a crime in general. 

Let's simulate this data: 

```{r simulate_data}
set.seed(8008569)

sample_size <- 1000

test_dat <- 
  tibble( age = ceiling( runif(sample_size, 18, 25) ),
              previous_convictions = sample( 0:4, 
                                             size=sample_size, 
                                             replace=TRUE,
                                             prob = c(0.6, 0.2, 0.1, 0.05, 0.05)),
              employed = sample( c(1,0), size=sample_size, replace=TRUE ),
              gender = sample( c(1,0), size=sample_size, replace=TRUE )
            )

coef_effects <- list( base = 0.05, 
                      age = -0.001,
                      previous_convictions = 0.02,
                      employed = -0.15,
                      gender = 0.05
                    ) 

test_dat_add_prob_col <- 
  test_dat %>%
      mutate( crime_prob = 0.25 + 
                           age * coef_effects$age +
                           previous_convictions * coef_effects$previous_convictions +
                           employed * coef_effects$employed +
                           gender * coef_effects$gender
            ) %>% 
      mutate( runif01 = runif( n(), 0, 1) ) %>% 
      mutate( committed_crime = ifelse(crime_prob > runif01, 1, 0) )

``` 

So, what we have are various variables describing 10,000 observed individuals, as well as whether they committed a crime in the study period or not. Here are the first 20 rows of the data:

```{r head_data}
test_dat_add_prob_col %>%
  select( age, previous_convictions, employed, gender, committed_crime ) %>% 
  head(20)
```

We assume the following structure for the data:

The response measurements _committed crime_, which take on either the value 0 or 1, we notate as $Z_1, Z_2, Z_3, ... , Z_{10,000}$. So if our first individual committed a crime in the study period we notate this $Z_1 = 1$. If the first observed individual did not commit a crime in the study period we would have $Z_1 = 0$. So, everyone in our study has a label $z_j$, indicating whether they committed a crime in the study period or not. 

We impose the following structure/assumption on our data: we assume that everyone with the same observed features (same age, number of previous convictions, employment status and gender) has the same fixed probability of committing a crime. We denote this fixed probability $\pi_i$, meaning that if we group everyone with the same characteristics into the same group and label this group as group $i$, then everyone in group $i$ has probability $\pi_i$ of committing a crime in the study period. For example, if we have $\pi_3=0.3$, then we'd expect something like 30% of group 3 to have committed a crime in the study period. This is saying that each individual $Z_j$ is drawn from a Bernoulli distribution.

If we group together people with the same features/characteristics like this, then we create a new random variable $Y_i$. For any given group $i$, $Y_i$ is the number of people committing a crime during the study period in that group (group $i$). For example, in our data, we could label as group 1 individuals with the following features:

* age 19

* 2 previous convictions

* unemployed
 
* female

In our 10 000 people, there are `r test_dat_add_prob_col %>% filter(age==19 & previous_convictions==2 & employed == 0 & gender==0) %>% nrow()` people matching these characteristics. Of them, `r test_dat_add_prob_col %>% filter(age==19 & previous_convictions==2 & employed == 0 & gender==0 & committed_crime==1) %>% nrow()` committed a crime in the study period, meaning $Y_1 =$ `r test_dat_add_prob_col %>% filter(age==19 & previous_convictions==2 & employed == 0 & gender==0 & committed_crime==1) %>% nrow()`. So, we will have as many observations $Y_i$ as we have unique combinations of our features. 

In logistic regression, we model $\pi_i$ - the probability of committing a crime for an individual in group $i$ - as some function of the features. For example, we assume that each of age, number of previous convictions, employment status and gender will contribute in some fixed way to determining what $\pi_i$ is.  

Because we have assumed that each $Z_j$ is a random draw from a Bernoulli distribution, it follows that each $Y_i$ is a random draw from a Binomial distribution with parameters $n_i$ (size of group $i$) and $\pi_i$. 

Here is a summary of our groups (there are `r test_dat_add_prob_col %>% 
     distinct(age, previous_convictions, employed, gender) %>% nrow()` unique feature combinations):  

```{r print_groups}
# make a summary of each group, with observed and theoretical proportions: 
test_dat_add_prob_col %>% 
     group_by( age, previous_convictions, employed, gender ) %>% 
     summarise( Y_i = sum(committed_crime == 1),
                n_i = n(),
                did_not_commit = sum(committed_crime == 0)
     ) %>% 
     ungroup() %>% 
     mutate( proportion_crime_commit = Y_i / n_i ) %>% 
  arrange( desc(proportion_crime_commit) ) %>%
  
  # add the theoretical probabilities back on for comparison:
  left_join( test_dat_add_prob_col %>%
               distinct( age, previous_convictions, employed, gender, crime_prob )
           ) %>% 
  kable()

```


Our logistic regression model is 

$$\begin{array}{lcl} \pi_i &=& sigmoid \Big( \beta_0 + \beta_1 x^{\text{(age)}}_i + \beta_2 x^{\text{(prev. convict)}}_i + \beta_3 x^{\text{(employed)}}_i + \beta_4 x^{\text{(gender)}}_i \Big) \\
&=& sigmoid \Big( \eta \Big) \\
&=& \frac{1}{1+e^{-\eta}}\\ 
\end{array}$$

So, for each feature group $i$, we model the probability of committing a crime in this group $i$ as a linear combination of the features of this group put through a (non-linear) sigmoid transformation. The sigmoid transformation ensures that $\pi_i$ is constrained to $[0,1]$, as opposed to just a linear transformation which can take on any value in $(-\infty,\infty)$. The sigmoid function looks like this:

```{r show_sigmoid}
x <- seq(-10,10,0.001)
plot( x = x,
      y = 1 / (1+exp(-x)),
      type = "l",
      xlab = expression(x),
      ylab = expression(f(x)),
      yaxt = "n"
      )

axis( side = 2, at = seq(0,1,0.1), las=2)
```

We estimate the coefficients $\beta_0, \beta_1, \beta_2, \beta_3, \beta_4$ of our model using maximum likelihood estimation.  

For any chosen group $i$, $Y_i$ is the number of people committing a crime in the study period in that group. $Y_i$ has a binomial distribution

$$Y_i \quad \sim \quad \text{binomial} \bigg( n_i, \space \pi_i \bigg) $$

..where $n_i$ is the total number of people in group $i$, and $\pi_i$ is the probability of any given individual in group $i$ committing a crime in the study period. 

This allows us, for any group $i$, to calculate the probability 

$$Pr\Big[Y_i=y_i \Big] \quad = \quad \binom{n_i}{y_i} \space \pi_i^{y_i} \space (1-\pi_i)^{n_i-y_i}$$

For example, if group 15 were to have 20 people in it, with probability of any given member committing a crime in the study period of 0.4, then the probability that 18 members of group 15 would commit a crime in the study period is

$$Pr\bigg[Y_{15}=20\bigg] \quad = \quad \binom{20}{18} \space (0.4)^{18} \space (0.6)^{2} \quad = \quad 0.000004700412$$

If we assume that the groups are independent (i.e. that the probability of committing a crime in one group does not affect the probability of committing a crime in a different group), then we can create a likelihood function:

$$\begin{array}{lcl}\mathcal{L} &=& \Pi_{i=1}^n \space Pr\bigg[Y_i = y_i\bigg] \\
&=& \Pi_{i=1}^n \space \binom{n_i}{y_i} \space \pi_i^{y_i} \space (1-\pi_i)^{n_i-y_i} \\ \end{array}$$

...where for each group, $y_i$ is the number of crimes committed in the study period that we have observed from our sample data.

Where the group probabilities $\pi_i$ are already decided and fixed, we can put in a set of sample data $y_i$, and this function will tell us how likely that sample data is to be observed under the binomial structure that we've imposed. Where our sample data $y_i$ are fixed and we allow $\pi_i$ to vary, maximising this function allows us to find the probabilities $\pi_i$ that make our observed data the data which was the most likely to have occurred given our binomial model.  

We can simplify this likelihood function, which will make it easier to maximise. The log of it is:

$$log(\mathcal{L}) \quad = \quad \mathcal{l} \quad = \quad \sum_{i=1}^n \Bigg[ \space log \binom{n_i}{y_i} + y_i \space log \bigg(\frac{\pi_i}{1-\pi_i}\bigg) + n_i \space log \bigg(1-\pi_i\bigg) \Bigg]$$

Recall that we are modelling each $\pi_i$ as $\pi_i = sigmoid \Big( \beta_0 + \beta_1 x^{\text{(age)}}_i + \beta_2 x^{\text{(prev. convict)}}_i + \beta_3 x^{\text{(employed)}}_i + \beta_4 x^{\text{(gender)}}_i \Big)$ 

So, in $\mathcal{l}$, we will replace each $\pi_i$ with this expression.

We can also drop the $log \binom{n_i}{y_i}$ term, because there are no parameters in here that are not fixed, so this is just a constant for each group that has no effect on the maximisation.

First, we define the likelihood function in R:

```{r fit_the_shit_myself}

feature_groups <- 
  test_dat_add_prob_col %>% 
     group_by( age, previous_convictions, employed, gender ) %>% 
     summarise( n_crime_committed = sum(committed_crime == 1),
                n_no_crime = sum(committed_crime == 0),
                groupsize=n()
     ) %>% 
     ungroup() %>% 
     mutate( proportion_crime_commit = n_crime_committed / (n_crime_committed + n_no_crime)) %>% 
     arrange( desc(proportion_crime_commit) )

y_i <- feature_groups$n_crime_committed        # observed numbers of crime committed per group
n_i <- feature_groups$groupsize                # observed group sizes

sigmoid <- function(x) { 1 / (1+exp(-x) ) }

log_likelihood_ftn <- 
  function( coefs ){
  
  -sum(                       # we return the negative sum so that we can minimise this function
      y_i * log( 
                  # vector of pi_i:
                    sigmoid( 
                             coefs[1] + 
                             coefs[2]*feature_groups$age +   
                             coefs[3]*feature_groups$previous_convictions + 
                             coefs[4]*feature_groups$employed +
                             coefs[5]*feature_groups$gender  
                    ) /
                    
                    # vector of (1 - pi_i)                 
                    ( 1 - sigmoid( 
                             coefs[1] + 
                             coefs[2]*feature_groups$age +   
                             coefs[3]*feature_groups$previous_convictions + 
                             coefs[4]*feature_groups$employed +
                             coefs[5]*feature_groups$gender  
                                  )
                    )
      ) +
      
      n_i * log( 1 - sigmoid( 
                             coefs[1] + 
                             coefs[2]*feature_groups$age +   
                             coefs[3]*feature_groups$previous_convictions + 
                             coefs[4]*feature_groups$employed +
                             coefs[5]*feature_groups$gender  
                            )
                    )
  )
}

# log_likelihood_ftn( coefs = c( intercept = 0, 
#                                   age = 0, 
#                                   previous_convictions = 0,
#                                   employed = 0.04,
#                                   gender = -0.01
#                                 )
#                   )


```

Maximising this function using the optim() function in R gives us our coefficient estimates:

```{r quick_optim}
optim( par = c( intercept = 0, 
                age = 0, 
                previous_convictions = 0,
                employed = 0,
                gender = 0
              ),
       fn = log_likelihood_ftn,
       control = list(maxit=1e4L)
    )$par
```

and using the glm() package (IWLS):

```{r use_the_glm_function}

glm_fit <- glm( committed_crime ~ age + previous_convictions + employed + gender,
                  family = binomial(link="logit"),
                  data = test_dat_add_prob_col )
glm_coefs <- glm_fit$coefficients
glm_coefs
```

## interpretting the coefficients
Recall that our logistic regression model is 

$$\pi_i \quad = \quad sigmoid \Big( \beta_0 + \beta_1 x^{\text{(age)}}_i + \beta_2 x^{\text{(prev. convict)}}_i + \beta_3 x^{\text{(employed)}}_i + \beta_4 x^{\text{(gender)}}_i \Big) $$

Rearranging this equation gives us a form which is interesting from an interpretation point of view. This rearrangement is:

$$\frac{\pi_i}{1-\pi_i} \quad = \quad e^{ \Big( \beta_0 + \beta_1 x^{\text{(age)}}_i + \beta_2 x^{\text{(prev. convict)}}_i + \beta_3 x^{\text{(employed)}}_i + \beta_4 x^{\text{(gender)}}_i \Big)}$$

$\frac{\pi_i}{1-\pi_i}$ is an odds ratio. For example, if the probability of committing a crime in the study period in group $i$ is 0.2, then the odds ratio for this group is $\frac{0.2}{0.8}=0.25$, meaning that the odds of committing a crime are 0.25:1, which is 1:4. We can interpret this as saying that we expect 1 crime-committing person for every 4 people not committing a crime in this group, or committing a crime is a quarter as likely as not committing one in this group.  

```{r exp_coefs}
exp(glm_coefs)
```

```{r checking_these_out}

person1 <- 
        tibble( age=21,
                    previous_convictions=2,
                    employed=0,
                    gender=0
                   )
person2 <- 
        tibble( age=21,
                    previous_convictions=3,
                    employed=0,
                    gender=0
                   )

pi_person1 <- predict( glm_fit, newdata=person1, type = "response" )
pi_person2 <- predict( glm_fit, newdata=person2, type = "response" )

pi_person1
pi_person2

pi_person1 / (1-pi_person1)
pi_person2 / (1-pi_person2)

( pi_person2 / (1-pi_person2) )/
( pi_person1 / (1-pi_person1) )  


```

Of course, can include splines, interactions, polynomials, GAMs etc. in the linear combination
