---
title: "Logistic Regression from Scratch"
output: html_notebook
---

```{r setup}
library(tidyverse)
library(mlbench)      # for the "BreastCancer" dataset
library(Hmisc)      # for EDA
```

https://documentation.sas.com/?docsetId=statug&docsetTarget=statug_logistic_details25.htm&docsetVersion=15.1&locale=en

Venzon, D. J. and Moolgavkar, S. H. (1988), “A Method for Computing Profile-Likelihood Based Confidence Intervals,” Applied Statistics, 37, 87–94.

```{r}
data(BreastCancer)
breast_cancer_data_cleaned <- 
  BreastCancer %>% 
    rename( patient_id = Id,
            y_tumour_class = Class,
            x1_clump_thickness = Cl.thickness,
            x2_uniformity_of_cell_size = Cell.size,
            x3_uniformity_of_cell_shape = Cell.shape,
            x4_marginal_adhesion_single_epithelial_cell = Marg.adhesion,
            x5_single_epithelial_cell_size = Epith.c.size,
            x6_bare_nuclei = Bare.nuclei,
            x7_bland_chromatin = Bl.cromatin,
            x8_normal_nucleoli = Normal.nucleoli,
            x9_mitoses = Mitoses
          ) %>% 
  select( patient_id, y_tumour_class, everything() ) %>%
  mutate_at( vars(matches("x")), as.numeric ) %>% 
  tidyr::drop_na()       # drop rows which contain any NA values
  
Hmisc::describe(breast_cancer_data_cleaned)
```

```{r}
breast_cancer_data_cleaned
```

So, we want to build a model to predict $y_i$ (*tumour class* of patient *i*) using patient features $x_1,x_2,...,x_9$

Response/outcome variable $y$ is *binary* - it can only one of the values $y=1$ (malignant) and $y=0$ (benign).

We assume that every observation $y_i$ is a random draw from it's own Bernoulli distribution:

$$y_i \quad\sim\quad Bernoulli\Big(p_i\Big)$$

We want to model $\pi_i=E\Big[y_i\Big]$ as a non-linear function of the predictors $x_{i1},x_{i2},...,x_{i9}$

The *logistic regression* model, with the *logit* link function, is:

$$\underset{\displaystyle \pi_i}{\underbrace{E\Big[y_i\Big]}} \quad=\quad \displaystyle\frac{1}{1+e^{-\eta_i}} \hspace{20mm} \eta_i=\beta_0+\beta_1x_{i1}+\beta_2x_{i2}+...+\beta_9x_{i9}$$

In this binary case ($y_i\in\{0,1\}$), $E\Big[y_i\Big]=Pr\Big[y_i=1\Big]$.

The likelihood of observing our dataset under a given model (i.e. a fixed set of parameters $\theta=\{\beta_0,\beta_1,...,\beta_9\}$) is:

$$\begin{array}{lcl}\mathcal{L}\Big(y,\theta\Big) &=& \displaystyle\prod_{i=1}^{699} \Bigg( \Big(\displaystyle\frac{1}{1+e^{-\eta_i}} \Big)^{y_i}  \Big(1-\displaystyle\frac{1}{1+e^{-\eta_i}} \Big)^ {1-y_i} \Bigg) \\
&=& \displaystyle\prod_{i=1}^{699} \Bigg( \pi_i^{y_i} (1-\pi_i)^ {1-y_i} \Bigg) \\
\end{array}$$

We choose the coefficients $\beta_0, \beta_1, ..., \beta_9$ by choosing the coefficients which maximise the likelihood function $\mathcal{L}\Big(y\Big)$.

This is equivalent to maximising the log of the likelihood function $\mathcal{L}\Big(y,\theta\Big)$, which is less prone to loss of information through underflow:

$$log\Bigg[\mathcal{L}\Big(y,\theta\Big)\Bigg] \quad=\quad \displaystyle\sum_{i=1}^{699}\Bigg( y_i\cdot log(\pi_i) \space+\space (1-y_i)\cdot log(1-\pi_i)\Bigg)$$

```{r}
X <- breast_cancer_data_cleaned %>% 
        select(-patient_id,-y_tumour_class) %>% 
        mutate_all(as.numeric) %>% 
        as.matrix() %>% 
        cbind( intercept=1, .)

y <- if_else( breast_cancer_data_cleaned$y_tumour_class == "benign",
              0,
              1
            )

negative_of_log_likelihood_ftn <- function(theta){
  
     calc_eta <- c( X %*% as.matrix(theta,ncol=1) )
     calc_pi <- 1 / (1 + exp(-calc_eta) )
     
     return( 
             -sum( y * log(calc_pi) + (1-y)*log(1-calc_pi) )
     )
}

run_optim <- optim( par=rep(0,10), fn=negative_of_log_likelihood_ftn, method="L-BFGS-B" )
run_optim
```

```{r}
fit_full_glm <- 
  glm( y_tumour_class ~ x1_clump_thickness + 
                        x2_uniformity_of_cell_size + 
                        x3_uniformity_of_cell_shape + 
                        x4_marginal_adhesion_single_epithelial_cell + 
                        x5_single_epithelial_cell_size + 
                        x6_bare_nuclei + 
                        x7_bland_chromatin + 
                        x8_normal_nucleoli + 
                        x9_mitoses,
       data = breast_cancer_data_cleaned,
       family = binomial(link="logit")
 )

coef(fit_full_glm)
```

Let's see that we got the same coefficients as the *glm()* function in R:

```{r}
tibble(     variable_name = colnames(X),
          glm_coefficient = coef(fit_full_glm),
        optim_coefficient = run_optim$par 
      )
```

The *glm()* function uses the method *Iteratively Weighted Least Squares*, which may explain the small discrepancy between the 2 results. 

# Likelihood Ratio Test

https://www.stern.nyu.edu/rengle/LagrangeMultipliersHandbook_of_Econ__II___Engle.pdf

The likelihood ratio is the ratio of the probability of the observed data under 2 different models:

$$\lambda = \displaystyle\frac{\mathcal{L}\Big(y\Big)}{\mathcal{L}\Big(\Big)}$$

# Confidence Intervals for $\beta$ coefficients

The confidence intervals reported by the R *glm()* function are *profile confidence intervals*.

First, some contextual information:

We can test the hypothesis $H_0: \space \beta_j=\beta_0$ (i.e. parameter $\beta_j$ has the value $\beta_0$) using a *likelihood ratio* hypothesis test:



