---
title: "Assessing top-k Recommender Systems: MAP@k and nDCG@k"
output: html_document
---

It is very useful to be able to summarise the performance of a model using a single-number summary. For example, we can summarise the performance of a regression model with the [mean squared error](https://en.wikipedia.org/wiki/Mean_squared_error) and [coefficient of determination $R^2$](https://en.wikipedia.org/wiki/Coefficient_of_determination), or the performance of a classication model with [accuracy](https://en.wikipedia.org/wiki/Confusion_matrix), [AUC](https://en.wikipedia.org/wiki/Receiver_operating_characteristic#Area_under_the_curve) and [F1 score](https://en.wikipedia.org/wiki/F-score).  

A $top-k$ recommender system is an algorithm which, given a large list of potential items to recommend, is required to produce a reduced list of $k$ maximally relevant items to each user. 

When we're comparing $top-k$ recommender systems, it doesn't make sense to use metrics designed to assess other kinds of models (like classification or regression). So, specific metrics have been designed for this purpose. 




What irritated me when I investigated these metrics for the first time was the lack of concrete examples of their implementation, so that is what I have given here - the calculations clearly laid out. 

I'm happy that my calculations are correct - I am able to perfectly replicate the output of the  [XG-Boost](https://github.com/dmlc/xgboost) package on these metrics ($MAP@k$ and $nDCG@k$). 

It turns out that the definitions of these metrics **MAP@k** and **NDCG@k** differ somewhat by source, although they are all variations on the same theme (see, for example, [DCG](https://en.wikipedia.org/wiki/Discounted_cumulative_gain).  

$AP@k$ is defined, for a single user, as: 

$$\begin{array}{lcl}
AP@k &=& \displaystyle\frac{\displaystyle\sum_{i=1}^k P(i)\cdot rel(i)}{min\big\{k,n \big\}} \\
n &=& \text{total number of items in dataset relevant to this user} \\
P(i) &=& \text{cumulative precision at rank } i \\ 
&=& \displaystyle\frac{}{} \\
rel(k) &=& \begin{cases} 1 \text{ if item in rank } i \text{ is relevant to the user}\\ 0 \text{ if item in rank } i \text{ is NOT relevant to the user} \end{cases}
\end{array}$$

Some sources (e.g. the [XG-Boost](https://github.com/dmlc/xgboost) package and [wikipedia](https://en.wikipedia.org/wiki/Evaluation_measures_(information_retrieval)#Average_precision)) define the denominator of the $AP@k$ simply as $n$. I don't think that this makes sense in the $top-k$ recommendations context: In my opinion, the ranking model should not be penalised where there are more than $k$ relevant items for a given user in the database, since we are not allowing the ranker to recommend more than $k$ items.

$nDCG@k$ is defined (for a single user) as:

$$\begin{array}{lcl}
nDCG@k &=& \displaystyle\frac{DCG@k}{\text{ideal } DCG@k} \\
DCG@k &=& \displaystyle\sum_{i=1}^k \displaystyle\frac{2^{y_i}-1}{log_2(i+1)} \\
y_i &=& \text{relevance of (gain from) the item in rank } i \\
\text{ideal } DCG@k &=& \text{the highest achievable } DCG@k \text{ on the observed data}  \\
k &=& \text{number of items recommended to user} \\
\end{array}$$

So, $DCG@k$ (for a single user) is the sum of the relevance values of the highest ranked $k$ items, where the relevance values are discounted (shrunk) as they fall further down in the ranking. $nDCG@k$ is simply the relative size of the user's $DCG@k$ value compared to the highest possible $DCG@k$ value, which we could have achieved if we had an item-ranking system with perfect knowledge. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(kableExtra)
library(formattable)   # see https://haozhu233.github.io/kableExtra/awesome_table_in_html.html
```

```{r echo=FALSE}
full_itemset <- 
  tibble(
    item_id = 1:16,
    relevant_to_user_1 = c(0,0,0,1,1,0,0,1,1,1,0,0,0,1,0,1)
  )
```

```{r echo=FALSE}
ideal_DCG <- 
  full_itemset %>% 
    arrange(
              desc(relevant_to_user_1)
    ) %>% 
    mutate( rownumber = row_number() ) %>% 
    mutate( 
            `ideal_DCG@k` = cumsum( 
                                    (2^relevant_to_user_1 - 1) /
                                    log( rownumber + 1, base=2 )
                                  )
          ) %>% 
    select( rownumber, `ideal_DCG@k` )
```

```{r echo=FALSE}
full_itemset$model1_item_rank_for_user1 <- c(9,15,13,5,1,16,10,4,7,11,3,6,12,14,8,2)

model1_metric_calcs <- 
  full_itemset %>% 
    arrange( model1_item_rank_for_user1 ) %>% 
    mutate( 
            `CG@k` = cumsum(relevant_to_user_1)
            ,
            `DCG@k` = cumsum( 
                              (2^relevant_to_user_1 - 1) /
                              log( model1_item_rank_for_user1 + 1, base=2 )
                            )
          ) %>% 
    mutate( 
            delta_DCG = if_else( 
                                              model1_item_rank_for_user1==1, 
                                             `DCG@k`,
                                             `DCG@k` - lag(`DCG@k`)
                                            ) 
          ) %>% 
    left_join( 
                ideal_DCG,
                by = c("model1_item_rank_for_user1"="rownumber") 
             ) %>% 
    mutate( 
              `NDCG@k` = `DCG@k` / `ideal_DCG@k`
          ) %>% 
    mutate( 
            precision = `CG@k` / model1_item_rank_for_user1
            ,
            delta_precision = if_else( 
                                        model1_item_rank_for_user1==1, 
                                        precision,
                                        precision - lag(precision)
                                    )
            ,
            recall = `CG@k` / sum(full_itemset$relevant_to_user_1)
            ,
            delta_recall = if_else(
                                    model1_item_rank_for_user1==1, 
                                    recall,
                                    recall - lag(recall)
                                  )
            ,
            Average_Precision = cumsum( precision * delta_recall)   
            ,
            Average_Precision_altcalc = cumsum( precision * relevant_to_user_1 / sum(.$relevant_to_user_1) )     
          ) 

```

```{r echo=FALSE}
full_itemset$model2_item_rank_for_user1 <- c(5,10,14,15,6,11,13,3,16,12,4,7,2,9,1,8)

model2_metric_calcs <- 
  full_itemset %>% 
    arrange( model2_item_rank_for_user1 ) %>% 
    mutate( 
            `CG@k ` = cumsum(relevant_to_user_1)
            ,
            `DCG@k ` = cumsum( 
                              (2^relevant_to_user_1 - 1) /
                              log( model2_item_rank_for_user1 + 1, base=2 )
                            )
          ) %>% 
    mutate( 
            delta_DCG = if_else( 
                                  model2_item_rank_for_user1==1, 
                                  `DCG@k `,
                                  `DCG@k ` - lag(`DCG@k `)
                               ) 
          ) %>% 
    left_join( 
                ideal_DCG,
                by = c("model2_item_rank_for_user1"="rownumber") 
             ) %>% 
    mutate( 
              `NDCG@k ` = `DCG@k ` / `ideal_DCG@k`
          ) %>%   
    mutate( 
            precision = `CG@k ` / model2_item_rank_for_user1
            ,
            delta_precision = if_else( 
                                        model2_item_rank_for_user1==1, 
                                        precision,
                                        precision - lag(precision)
                                    )
            ,
            recall = `CG@k ` / sum(full_itemset$relevant_to_user_1)
            ,
            delta_recall = if_else(
                                    model2_item_rank_for_user1==1, 
                                    recall,
                                    recall - lag(recall)
                                  )
            ,
            Average_Precision = cumsum( precision * delta_recall)   
            ,
            Average_Precision_altcalc = cumsum( precision * relevant_to_user_1 / sum(.$relevant_to_user_1) )     
          ) 
```


```{r echo=FALSE}
bind_cols(
  model1_metric_calcs
  ,
  model2_metric_calcs %>% rename_all(function(x){ paste0(x," ")} )
) %>% 
  rename( 
          item_rank = model1_item_rank_for_user1
          ,
          `item_rank ` = `model2_item_rank_for_user1 `
        ) %>% 
  mutate_all( round, digits = 3 ) %>% 
  
  # set up conditional formatting bars
  # note that I use fun=function(x){x} in order to make bars proportonal to the values themselves, rather than scaling them
  mutate( 
           `CG@k` = formattable::color_bar("lightgreen")(.$`CG@k`),
           `CG@k  ` = formattable::color_bar("lightgreen")(.$`CG@k  `),

           `DCG@k` = formattable::color_bar("lightblue")(.$`DCG@k`),
           `DCG@k  ` = formattable::color_bar("lightblue")(.$`DCG@k  `),
           
           `NDCG@k` = formattable::color_bar("pink")(.$`NDCG@k`),
           `NDCG@k  ` = formattable::color_bar("pink")(.$`NDCG@k  `)
           
          #`precision ` = formattable::color_bar("lightgreen", fun=function(x){x})(.$`precision `),
          #recall = formattable::color_bar("lightblue", fun=function(x){x})(.$recall),
          #`recall ` = formattable::color_bar("lightblue", fun=function(x){x})(.$`recall `),
          #Average_Precision = formattable::color_bar("pink", fun=function(x){x})(.$Average_Precision),
          #`Average_Precision ` = formattable::color_bar("pink", fun=function(x){x})(.$`Average_Precision `),
        ) %>%

  select( 
          # model 1
          item_rank, 
          item_id,
          relevant_to_user_1,
          `CG@k`,
          `DCG@k`,
          `ideal_DCG@k`,
          `NDCG@k`,
          
          # model 2
          `item_rank `,
          `item_id `,
          `relevant_to_user_1 `,
          `CG@k  `,
          `DCG@k  `,
          `NDCG@k  `
        ) %>% 
  kbl( 
       caption = "NDCG@k",
       escape = FALSE,
       align = "l",
       booktabs = TRUE
     ) %>%
  kable_paper( "hover", full_width = FALSE ) %>%
  # column_spec( column = 1:10, width = "6cm") %>%
  kable_styling(
                  bootstrap_options = "striped",
                  full_width = FALSE,
                  position = "center"
               ) %>%
  add_header_above( c("Model 1" = 7, "Model 2" = 6) )
```

```{r echo=FALSE}
bind_cols(
  model1_metric_calcs
  ,
  model2_metric_calcs %>% rename_all(function(x){ paste0(x," ")} )
) %>% 
  rename( 
          item_rank = model1_item_rank_for_user1
          ,
          `item_rank ` = `model2_item_rank_for_user1 `
        ) %>% 
  mutate_all( round, digits = 3 ) %>% 
  
  # set up conditional formatting bars
  # note that I use fun=function(x){x} in order to make bars proportonal to the values themselves, rather than scaling them
  mutate( 
          precision = formattable::color_bar("lightgreen", fun=function(x){x})(.$precision),
          `precision ` = formattable::color_bar("lightgreen", fun=function(x){x})(.$`precision `),
          recall = formattable::color_bar("lightblue", fun=function(x){x})(.$recall),
          `recall ` = formattable::color_bar("lightblue", fun=function(x){x})(.$`recall `),
          Average_Precision = formattable::color_bar("pink", fun=function(x){x})(.$Average_Precision),
          `Average_Precision ` = formattable::color_bar("pink", fun=function(x){x})(.$`Average_Precision `),
        ) %>%
   
  select( 
          # model 1
          item_rank, 
          item_id,
          relevant_to_user_1,
          precision,
          recall, 
          Average_Precision,
          
          # model 2
          `item_rank `,
          `item_id `,
          `relevant_to_user_1 `,
          `precision `,
          `recall `,
          `Average_Precision `
        ) %>% 
  kbl( 
       caption = "AP@k",
       escape = FALSE,
       align = "l",
       booktabs = TRUE
     ) %>%
  kable_paper("hover", full_width = F) %>%
  # column_spec( column = 1:10, width = "6cm") %>%
  kable_styling(
                  bootstrap_options = "striped",
                  full_width = FALSE,
                  position = "center"
               ) %>%
  add_header_above( c("Model 1" = 6, "Model 2" = 6) )
```

```{r fig.width=13, echo=FALSE}
ggplot(
        data = bind_rows( 
                            model1_metric_calcs %>% 
                              mutate( model = "Model 1") %>% 
                              rename( model_item_rank_for_user1 = model1_item_rank_for_user1 )
                            ,
                            model2_metric_calcs %>% 
                              rename_all( .funs = function(x){ gsub(" ","", x) } ) %>% 
                              mutate( model = "Model 2" ) %>% 
                              rename( model_item_rank_for_user1 = model2_item_rank_for_user1 )
                        ),
        aes( x = recall,
             y = precision,
             label = model_item_rank_for_user1
           )
) +
  geom_point() + 
  geom_line() +
  geom_text( 
            colour = "red",
            hjust = -0.5
          ) + 
  facet_wrap( ~model )

```

