---
title: "Experimentation of Modelling of Customer Spend Cycle"
author: "Joseph Bolton"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

Some customers buy in cycles. For example, a customer might regularly replenish a particular cosmetics product. Is there a simple way to model this behaviour using only the dates of their transactions? 

I've documented my experiments here.   

All of my code I've put in the appendix at the end. 

For example, consider the following observed transaction times for a single customer:
(lines indicate transactions, numbers the time between the transactions)

```{r echo=FALSE}

plot_transactions_ftn <- function( spend_times ){
  
    # plot the simulated data:
  plot(   x = spend_times,
          y = rep(0, length(spend_times) ),
        cex = 2,
        pch = 16,
       xlim = c(0, max(spend_times)+20 ),
        col = "white",
       ylab = "",
       xlab = "time",
       main = "Observed Transaction Times"
      )
      abline( v = spend_times )

  # plot the time gaps between transactions:    
  text( x = spend_times[-length(spend_times)] + diff(spend_times)/2, 
        y = runif( min=-1, max=1, n=length(spend_times)-1 ),
        labels = (diff(spend_times)-1),
        col = 2
      )
  
}

sim_single_customer_spend_data <- 
  function(   spend_pattern = c(5, 30, 15, 50),   # length of this determines true cycle length
             n_transactions = 30,
            noise_intensity = 0.2,    # value in [0,1]
            percent_of_values_to_rmv = 0.1    # more added signal noise: also a value in [0,1] 
          ){
    
    spend_times <- c(0)
    
    counter <- 1
    
    sd_spend_pattern <- sd(spend_pattern) %>% round()
    
    for( i in 2:n_transactions ){
      
      spend_times[i] <- spend_times[i-1] + spend_pattern[counter]
      noise <- rnorm( n=1, mean=0, sd=sd_spend_pattern*noise_intensity ) %>% abs() %>% round()
      
      spend_times[i] <- spend_times[i] + noise
  
      if( counter == length(spend_pattern) ){
          counter <- 1
          } else{ counter <- counter+1 }
    
    }
    
    # remove values (added noise to simulated data):
    values_to_remove <- sample(    x = 1:n_transactions, 
                                size = round(percent_of_values_to_rmv*n_transactions)
                              )
    
    spend_times <- spend_times[ setdiff(1:n_transactions, values_to_remove) ]

  print(spend_times)  
    
  plot_transactions_ftn( spend_times=spend_times )
  
  return(spend_times)  
}
```

```{r echo=FALSE, fig.width=8, fig.height=6}
set.seed(1999) 

sim_transactions <- 
  sim_single_customer_spend_data(
      spend_pattern = c(5, 25, 10, 50),   # length of this determines true cycle length
     n_transactions = 20,
    noise_intensity = 0.2,    # value in [0,1]
    percent_of_values_to_rmv = 0    # more added signal noise: also a value in [0,1] 
  )
```

# Model 1: Iterative Period/Cycle Grid Method

My idea here is that if someone is showing cyclic behaviour, then there is some period length and cycle length for which this cyclic behaviour can be captured. 

Let's try this out on the simulated data:

```{r echo=FALSE}
assign_spend_times_to_periods_ftn <- function( spend_times, 
                                               time_units_per_period,
                                               plot = FALSE
                                             ){
  
  assign_periods <- 
    cut( spend_times,
         breaks = seq(0, max(spend_times)+time_units_per_period, by=time_units_per_period),
         include.lowest = TRUE
       )  
  
  spend_times_period_form <- rep( -1, length(levels(assign_periods)) )
  names(spend_times_period_form) <- levels(assign_periods)

  for( period_i in names(spend_times_period_form) ){
    spend_times_period_form[period_i] <- if_else( length(spend_times[assign_periods==period_i])>0, 1, 0 )
  }


  if( plot==TRUE ){
    
      cat("spend times: \n")
      cat(spend_times)
      cat("\n")
      cat("\n")
      tibble(     period = names(spend_times_period_form),
              transacted = spend_times_period_form 
            ) %>% 
    print()

    
    plot( x = spend_times,
          y = rep(0, length(spend_times) ),
          cex = 2,
          pch = "+",
         xlim = c(0, max(spend_times)+20 ),
          col = 2, # red
         main = paste0("Transactions by Period. period length = ", time_units_per_period ) 
        )
    abline( v = seq(0, max(spend_times)+time_units_per_period, by=time_units_per_period ) )
 
    text( x = which(spend_times_period_form>0) * time_units_per_period - time_units_per_period/2,
          y = 0.5,
          labels = "1",
          col = 2
         )
  }
  
  return(spend_times_period_form) 
}


```

If we break up this persons transactions into periods of length 15, then the transactions look like this:

```{r fig.width=8, fig.height=6, echo=FALSE}
spend_times_periods <- 
  assign_spend_times_to_periods_ftn( spend_times = sim_transactions,
                                     time_units_per_period = 15,
                                     plot = TRUE
                                   )
```

Let's see how these transactions fit into a cycle of 3 periods:

```{r echo=FALSE, fig.width=7, fig.height=7}

fit_n_cycle_ftn <- function( spend_times,
                             spend_times_period_form,
                             time_units_per_period,
                             cycle_length,
                             plot = FALSE,
                             n_periods_to_predict = 5
                           ){
  
  # take as many recent observations as possible, keeping a clean multiple of the cycle length:
  n_observations_to_remove <- length(spend_times_period_form)%%cycle_length 
  
  if( n_observations_to_remove > 0 ){
    cropped_spend_times <- spend_times_period_form[ (n_observations_to_remove+1):length(spend_times_period_form) ] 
    if( plot == TRUE ){ 
      paste0("NOTE: first ", n_observations_to_remove, " observation(s) removed to divide into cycles of length ", cycle_length) %>% print()
    }
  } else{ cropped_spend_times <- spend_times_period_form }
  
  # store the observations by cycle number:
  store_cycle_data <- vector(mode="list", length=cycle_length)
  names(store_cycle_data) <- paste( "cycle", 1:cycle_length, sep="_" )
  
  for( cyc_period_j in 1:cycle_length ){
    get_data <- cropped_spend_times[ seq(cyc_period_j, length(cropped_spend_times), by=cycle_length) ]
    names(get_data) <- NULL
    store_cycle_data[[cyc_period_j]] <- get_data
    }
  
  #print(cropped_spend_times)
  
  # calculate % transacted per cycle:
  percent_time_transacted_per_cycle <- lapply( store_cycle_data, mean )
  
  if( plot == TRUE ){
    
    # Period Plot:
    plot( x = 1,
          y = 1,
          ylim = c(0,1),
          xlim = c(0,length(cropped_spend_times)+1+n_periods_to_predict),
          col = "white",
          xlab = "period",
          main = paste0("cycle length ", cycle_length, ",  period length ", time_units_per_period)
        )
    abline( v = 0:length(cropped_spend_times), lty=1 )
    # draw lines for the prediction periods: 
    abline(   v = (length(cropped_spend_times)+1):(length(cropped_spend_times)+1+n_periods_to_predict), 
            col = "grey"
          )
    # number the periods:
    text(      x = 0:(length(cropped_spend_times)-1) + 0.5,
               y = 0.9,
          labels = 1:cycle_length,   
             col = 1:cycle_length
        )
    # number the prediction periods:
    text(      x = (length(cropped_spend_times)):(length(cropped_spend_times)+n_periods_to_predict) + 0.5,
               y = 0.95,
          labels = 1:cycle_length,   
             col = 1:cycle_length
        )
    
    # mark periods containing a transaction with an "x":
    text( x = which(cropped_spend_times==1)-0.5,
          y = 0.8,
          labels = "x"
        )
    
    # put in stars to show % transactions per cycle:
    stars_to_print <- percent_time_transacted_per_cycle
    for( i in 1:length(stars_to_print) ){
      stars_to_print[[i]] <- 
        paste( rep("*\n", round(percent_time_transacted_per_cycle[[i]]*10)), collapse="" )
    }  
    stars_to_print <- unlist(stars_to_print)
    
    # put stars on the plot:
    text(      x = 0:(length(cropped_spend_times)+n_periods_to_predict) + 0.5,
               y = 0.7,
          labels = stars_to_print,
             col = 1:cycle_length,
             pos = 1
        )
    
    # put the actual proportions on the plot:
    text(      x = 0:(length(cropped_spend_times)-1) + 0.5,
               y = 0.2,
          labels = paste0( round(unlist(percent_time_transacted_per_cycle)*100), " %" ),   
             col = 1:cycle_length,
             srt = 270    # rotate text
        )
    
    # put the predicted proportions on the plot:
    text(      x = (length(cropped_spend_times)):(length(cropped_spend_times)+n_periods_to_predict) + 0.5,
               y = 0.5,
          labels = paste0( round(unlist(percent_time_transacted_per_cycle)*100), " %" ),   
             col = 1:cycle_length,
             srt = 270    # rotate text
        )
    
    ## Original Data Plot ##
     plot(  x = spend_times,
            y = rep(0, length(spend_times) ),
          cex = 2,
          pch = "+",
         xlim = c( 0, max(spend_times) + time_units_per_period*n_periods_to_predict ),
          col = 2, # red
         main = paste0("Transactions by Period:    period length = ", time_units_per_period ) 
        )
     # draw in periods removed:
    if( n_observations_to_remove>0 ){ 
        abline(   v = seq(0, time_units_per_period*n_observations_to_remove, by=time_units_per_period ),
                lty = 2,    # dashed linetype
                col = "grey"
              )
    }
    # draw in periods used:
    abline(   v = seq( time_units_per_period*n_observations_to_remove, 
                       length(spend_times_period_form)*time_units_per_period,
                       by = time_units_per_period
                     )
          )
    # draw in periods to predict:
        abline(   v = seq( length(spend_times_period_form)*time_units_per_period + time_units_per_period, 
                           length(spend_times_period_form)*time_units_per_period + time_units_per_period + n_periods_to_predict*time_units_per_period,
                       by = time_units_per_period
                         ),
                  col = "grey"
              )
    
    # number the observed periods:
        text(      x = seq( time_units_per_period*n_observations_to_remove,
                            (length(spend_times_period_form)-1)*time_units_per_period,
                            by = time_units_per_period
                          ) + time_units_per_period/2,
               y = 0.6,
          labels = 1:cycle_length,   
             col = 1:cycle_length
        ) 
    # number the prediction periods: 
         text(      x = seq( length(spend_times_period_form)*time_units_per_period + time_units_per_period,
                            length(spend_times_period_form)*time_units_per_period + time_units_per_period + n_periods_to_predict*time_units_per_period,
                            by = time_units_per_period
                          ) - time_units_per_period/2,
               y = 0.3,
          labels = 1:cycle_length,   
             col = 1:cycle_length
        )
         
    # put stars on prediction periods:
         # put stars on the plot:
    text(      x = seq( length(spend_times_period_form)*time_units_per_period + time_units_per_period,
                            length(spend_times_period_form)*time_units_per_period + time_units_per_period + n_periods_to_predict*time_units_per_period,
                            by = time_units_per_period
                          ) - time_units_per_period/2,
               y = 0.1,
          labels = stars_to_print,
             col = 1:cycle_length,
             pos = 1
        )
    
    
  }
  
  calc_gini <- lapply( store_cycle_data, 
                       function(x){ (sum(x==1)/length(x)) * (1 - (sum(x==1)/length(x))) +
                                    (sum(x==0)/length(x)) * (1 - (sum(x==0)/length(x)))
                                   } 
                      )
  calc_variance <- lapply(store_cycle_data, var)
  calc_entropy <- lapply( store_cycle_data, 
                          function(x){ -(sum(x==1)/length(x)) * log((sum(x==1)/length(x))) +
                                       -(sum(x==0)/length(x)) * log((sum(x==0)/length(x)))
                                     } 
                        )
  
  
  # ## calculate 'transaction-weighted' metrics ##
  # transaction_weight <- sum(cropped_spend_times==0)/sum(cropped_spend_times==1)
  # store_cycle_data_transaction_inflated <- store_cycle_data
  # for( cycle_i in 1:length(store_cycle_data_transaction_inflated) ){
  #   n_0 <- sum(store_cycle_data[[cycle_i]]==0)
  #   n_1 <-   
  #   
  # }  
  # 
  # #calc_transaction_inflated_gini <- 
  # #calc_transaction_inflated_entropy <- 
  # #calc_transaction_inflated_variance <- 
  
  return( 
          list( transactions = store_cycle_data,    # actual cycle data
                percent_time_transacted_per_cycle = percent_time_transacted_per_cycle, 
                # node impurity measures: 
                cycle_variance = calc_variance,
                 mean_variance = mean( unlist(calc_variance) ), 
                    cycle_gini = calc_gini,  
                     mean_gini = mean( unlist(calc_gini) ),
                cycle_entropy = calc_entropy,
                 mean_entropy = mean( unlist(calc_entropy) )
              ) 
        )
}


```

```{r fig.width=8, fig.height=6, echo=FALSE}
fit_3_cycle <- fit_n_cycle_ftn(             spend_times = sim_transactions,
                                spend_times_period_form = spend_times_periods,
                                  time_units_per_period = 15,           
                                           cycle_length = 3,
                                                   plot = TRUE,
                                   n_periods_to_predict = 10
                              )

fit_3_cycle$transactions

cat("proportion of times transacted per cycle:\n") 
fit_3_cycle$percent_time_transacted_per_cycle
```

Answer: very badly.

This doesn't seem like a very good fit. 

We can measure the *purity* of each cycle (i.e. cycle contains mostly the same value) using a metric such as *variance*, *entropy* or *gini*:

```{r echo=FALSE}
cat("variance: \n")
fit_3_cycle$cycle_variance
cat("\n gini: \n")
fit_3_cycle$cycle_gini
cat("\n entropy: \n")
fit_3_cycle$cycle_entropy
```

We could use the average (mean) value of one of these metrics as a measure of the goodness-of-fit of a cycle and period-length that we're considering: 

```{r echo=FALSE}
tibble( metric = c("variance", "gini", "entropy"),
        avg_value = c( fit_3_cycle$mean_variance,
                       fit_3_cycle$mean_gini,
                       fit_3_cycle$mean_entropy
                     )  
      )
```

Let's try out a grid of different cycle lengths and period lengths to find a good fit, extracting the one with the highest *gini*:

```{r echo=FALSE}

find_best_model_ftn <- function(   period_lengths_to_try,
                                    cycle_lengths_to_try,
                                             spend_times,
                                                  metric = "mean_gini"
){
  
  expand_grid( period_length = period_lengths_to_try,
                cycle_length = cycle_lengths_to_try   
             ) %>% 
  mutate( avg_metric = 
                     map2_dbl( .x = period_length,
                               .y = cycle_length,
                               .f = function(period.len, cycle.len){
                                      fit_periods <- 
                                        assign_spend_times_to_periods_ftn( 
                                                       spend_times = spend_times,
                                             time_units_per_period = period.len,
                                                              plot = FALSE
                                             )
                                    
                                      fit_n_cycle_ftn( 
                                                    spend_times = spend_times,
                                          time_units_per_period = period.len,
                                        spend_times_period_form = fit_periods,
                                                   cycle_length = cycle.len,
                                                           plot = FALSE
                                                     ) %>% 
                                        .[[metric]] %>% 
                                        return()
                                    } 
                            ) 
         ) %>% 
    return()
  
}


```

The best model by *gini* metric is:

```{r fig.width=15, fig.height=6, echo=FALSE}
find_best_model_by_gini <- find_best_model_ftn(   period_lengths_to_try = 10:50,
                                           cycle_lengths_to_try = 2:6,
                                                    spend_times = sim_transactions,
                                                         metric = "mean_gini"
                                )

# find_best_model_by_gini %>% arrange(avg_metric)

best_model_by_gini <- find_best_model_by_gini %>% arrange(avg_metric) %>% slice(1)

fit_n_cycle_ftn(             spend_times = sim_transactions,
                 spend_times_period_form = 
                   assign_spend_times_to_periods_ftn( spend_times = sim_transactions,
                                            time_units_per_period = best_model_by_gini$period_length,
                                                             plot = FALSE
                                                    ),
                                  time_units_per_period = best_model_by_gini$period_length,
                                           cycle_length = best_model_by_gini$cycle_length,
                                                   plot = TRUE,
                                   n_periods_to_predict = 20
              )
```


# Model 2: Gap-clustering Model

This method extracts the lengths of all of the gaps between transactions, and then tries to group similar gap lengths together using clustering. Following this, the model looks for commonly occuring sequences of gap lengths (e.g. short gap, short gap, long gap).

The algorithm is:

1. Extract length of (amount of time between) every gap between transactions

2. Fit a few different k-means clustering models.

3. Choose the best fitting k-means clustering to the data using the average *silhouette* metric.

4. Predict what the most likely next gap length is going to be based on past sequences of gap lengths. 

Here is this approach applied to the example data we've been looking at so far:

```{r echo=FALSE}
gap_clustering_model_ftn <- function( spend_times,
                                      seq_length = 3,
                                      k_values_to_try
                                    ){
  
  gap_lengths <- diff(spend_times)
  
  k_means_results <- list()
  
  for( k in k_values_to_try ){
  
    paste0( "fitting k-means, k = ", k, "\n" ) %>% cat()
    
    data_for_kmeans <- data.frame( id=1:length(gap_lengths), gapsize=gap_lengths )
    
    run_kmeans <- kmeans(       x = data_for_kmeans[,-1],
                          centers = k,
                          nstart = 25
                        )
    get_avg_silhouette_val <- cluster::silhouette( run_kmeans$cluster, 
                                                   dist(data_for_kmeans, method="manhattan") 
                                                 ) %>% .[,"sil_width"] %>% mean() 
    
    paste0( "avg silhouette value for k=", k, ": ", get_avg_silhouette_val ,"\n" ) %>% cat()
    
    k_means_results[[paste0("k",k)]] <- list( clusters = cbind(data_for_kmeans, cluster=run_kmeans$cluster),
                                               avg_sil = get_avg_silhouette_val 
                                            )
  }
  
  best_model_index <- lapply( k_means_results,
                              function(x){ x$avg_sil }
                             ) %>% 
                  unlist() %>% 
                  which.max()
  
  best_k <- k_values_to_try[best_model_index] 
  
  paste0( "best k identified: k=", best_k, "(highest mean silhouette value) \n" ) %>% cat()
  
  best_k_means_result <- k_means_results[[best_model_index]]$clusters 
  
  cluster_sequence <- best_k_means_result$cluster
  
  extract_windows <- list()
  
  continue <- 1
  position <- 1
  while( continue == 1 ){
    if( position+seq_length-1 <= length(cluster_sequence) ){
      extract_windows[[position]] <- cluster_sequence[position:(position+seq_length-1)]
      position <- position+1
    } else{ continue <- 0 }    
  }
  
  extract_windows_tbl <- reduce( .x = extract_windows,
                                 .f = rbind
                               ) %>% 
                         as.data.frame()  
  colnames(extract_windows_tbl) <- paste( "val", 1:seq_length, sep="" )
  extract_windows_tbl <- as_tibble(extract_windows_tbl)
  count_seq_tbl <- 
      extract_windows_tbl %>% 
        group_by_all() %>% 
        summarise( n_times_appearing = n() ) %>%
        ungroup() %>% 
        mutate( `%` = n_times_appearing / sum(n_times_appearing) )
 
 avg_gapsize_in_each_cluster <-
   best_k_means_result %>% 
      group_by( cluster ) %>% 
      summarise( mean_gapsize = mean(gapsize) ) %>% 
      arrange( cluster )
   
 most_recent_state_sequence <- 
   cluster_sequence[ (length(cluster_sequence)-seq_length+2):length(cluster_sequence) ]
 
 # have we seen the most recent state sequence before? 
 cols_to_check <- 
   count_seq_tbl[,1:(seq_length-1)]
 rows_of_cols_to_check_matching <- c()
 for( row_i in 1:nrow(cols_to_check) ){
   
   if( sum( cols_to_check[row_i, ] != most_recent_state_sequence )==0 ){ 
       rows_of_cols_to_check_matching <- c(rows_of_cols_to_check_matching, row_i)
   }
 }
 
 count_seq_tbl
 
 possible_next_states <- 
   count_seq_tbl[rows_of_cols_to_check_matching,]
 
 names(possible_next_states)[seq_length] <- "potential_next_state"
 
 possible_next_states <- 
   possible_next_states %>% 
    group_by( potential_next_state ) %>% 
    summarise( n_times_appearing = sum(n_times_appearing) ) %>%
    left_join( avg_gapsize_in_each_cluster,
               by = c("potential_next_state"="cluster")  
             ) %>% 
    mutate( `%` = paste0( n_times_appearing,"/", sum(n_times_appearing), " = ", round( n_times_appearing/sum(n_times_appearing), digits=2 ) ) 
          )
 
 # plot the transactions data:
 plot(   x = spend_times,
          y = rep(0, length(spend_times) ),
        cex = 2,
        pch = 16,
       xlim = c(0, max(spend_times)+max(avg_gapsize_in_each_cluster$mean_gapsize) ),
        col = "white",
       ylab = "",
       xlab = "time",
       main = paste0( "Observed Transaction Times with Clustered Gaps \n",
                      "Number of Clusters = ", best_k,
                      ",   n preceding observations in prediction = ", seq_length-1
                    )
      )
  abline( v = spend_times )
  
  # write in the cluster numbers:
  text(      x = spend_times[-length(spend_times)] + gap_lengths/2,
             y = -0.9,
        labels = as.character(cluster_sequence),
           col = cluster_sequence
  )
  
  if( nrow(possible_next_states) > 0 ){ 
    
    # plot possible next states:  
    for( possible_next_state_i in 1:nrow(possible_next_states) ){ 
        
         segments( x0 = max(spend_times),
                   y0 = 1 - possible_next_state_i/5,
                   x1 = max(spend_times) + possible_next_states$mean_gapsize[possible_next_state_i],
                   y1 = 1 - possible_next_state_i/5,
                   col = possible_next_state_i,
                   lwd = 3
                 ) 
      
         text( x = max(spend_times),
               y = 1.05 - possible_next_state_i/5,
               labels = paste0( "cluster ", 
                                possible_next_states$potential_next_state[possible_next_state_i],
                                ":   ",
                                possible_next_states$`%`[possible_next_state_i]
                              ),
                   col = possible_next_states$potential_next_state[possible_next_state_i]
             )
    }
  } else{ 
    
      text( x = max(spend_times),
            y = 0,
            labels = "Cannot \n Predict \n Next \n Gapsize",
            col = 4
          )
  }
  
  # plot in possible next states, with  
  
  # print the sequence counts table:
  count_seq_tbl %>% 
    arrange( desc(`%`) ) %>% 
    print()
  
  return(count_seq_tbl)
  
}

```

```{r fig.width=8, fig.height=6}
run_gap_clustering <- gap_clustering_model_ftn(     spend_times = sim_transactions,
                                                     seq_length = 4,
                                                k_values_to_try = 2:5
                                              )
```

```{r fig.width=8, fig.height=6}
run_gap_clustering <- gap_clustering_model_ftn(     spend_times = sim_transactions,
                                                     seq_length = 3,
                                                k_values_to_try = 2:5
                                              )
```

# Model comparison example 2


# Model comparison example 3

# Model comparison example 4

# Appendix: all code
