---
title: "naive_bayes_collaborative_filtering"
author: "Joseph Bolton"
date: "26/07/2020"
output: html_document
---

```{r setup}
library(tidyverse)
library(Matrix)
```

Random ratings matrix: 

```{r}
n_items <- 6
n_customers <- 20
percent_sparse <- 0.5

# assign some random 1-3 ratings for various customer/item combinations:
rating_matrix <- matrix( sample(1:3, size=n_items*n_customers, replace=TRUE),
                         ncol = n_items,
                         nrow = n_customers
                       )
rating_matrix <- 
  rating_matrix * 
  sample( 0:1, size=prod(dim(rating_matrix)), replace=TRUE, prob=c(percent_sparse,1-percent_sparse) )

colnames(rating_matrix) <- paste0("item", 1:ncol(rating_matrix), sep="" )
rownames(rating_matrix) <- paste0("customer", 1:nrow(rating_matrix), sep="" )

# make sure that each customer has rated at least 1 item:
for( row_i in 1:nrow(rating_matrix) ){
  if( sum(rating_matrix[row_i,])==0 ){ 
      rating_matrix[row_i,sample(1:ncol(rating_matrix),size=1)] <- 
        sample( 1:3, size=1)
    }
}

# make sure that each item has been rated at least once:
for( col_j in 1:ncol(rating_matrix) ){
  if( sum(rating_matrix[,col_j])==0 ){ 
      rating_matrix[sample(1:nrow(rating_matrix),size=1),col_j] <- 
        sample( 1:3, size=1)
    }
}

rating_matrix <- Matrix( rating_matrix, sparse=TRUE )

rating_matrix
```

$$\begin{array}{lcl}
Pr\Big[r_{uj}=v_s \Bigl| \text{observed ratings of customer } u\Big] &=& \text{probability customer } u \text{ gives } \text{(unknown) rating } v_s \\
&\space&   \text{ to item } j, \text{ given the observed ratings of customer } u  \\                       
&\propto& Pr\Big[\text{observed ratings of customer } u\Bigl| r_{uj}=v_s\Big] \times Pr\Big[r_{uj}=v_s\Big] \hspace{10mm} ...\text{by Bayes rule} \\
\end{array}$$

$Pr\Big[r_{uj}=v_s\Big]$, the *prior probability that customer $u$ gave rating $v_s$ to item $j$*, is estimated as the fraction/proportion of all users who rated item $j$ that gave it a rating of $v_s$.   

For a single item, $Pr\Big[r_{uk}=v_t\Bigl|r_{uj}=v_s\Big]$ (the probability that user $u$ gives rating $v_t$ to item $k$ given that user $u$ gave rating $v_s$ to item $j$) is estimated by the fraction/proportion of the users giving rating $v_t$ to item $k$, who gave rating $v_s$ to item $j$.  

$Pr\Big[\text{observed ratings of customer } u\Bigl| r_{uj}=v_s\Big]$, the *likelihood of all of the observed ratings of customer $u$, given that they gave a rating of $v_s$ to item $j$*, is estimated as:

