---
title: "object_allocation_experiments"
output: html_notebook
---


```{r setup}
library(tidyverse)
library(Rglpk)
```

My idea:

1. Split the data into small subsets and solve the problem with integer linear programming in each subset
(2.) [optional] do step 1 lots of times and choose the best solution
3. Combine all of the subset optimal solutions to create an approximate global solution
4. Perform iterative swapping/random voucher assignment on the global solution, accepting each random action 
    when it keeps the solution feasible and increases the objective function. 

Here is a quick example of how the linear integer programming function Rglpk_solve_LP() function works in R:
(it is from their help function in R)

```{r}
## Simple mixed integer linear program.
## maximize:    3 x_1 + 1 x_2 + 3 x_3
## subject to: -1 x_1 + 2 x_2 +   x_3 <= 4
##                      4 x_2 - 3 x_3 <= 2
##                x_1 - 3 x_2 + 2 x_3 <= 3
##                x_1, x_3 are non-negative integers
##                x_2 is a non-negative real number

obj <- c(3, 1, 3)
mat <- matrix(c(-1, 0, 1, 2, 4, -3, 1, -3, 2), nrow = 3)
dir <- c("<=", "<=", "<=")
rhs <- c(4, 2, 3)
types <- c("I", "C", "I")
max <- TRUE

Rglpk_solve_LP(obj, mat, dir, rhs, types = types, max = max)
```
   
So, suppose that our profit matrix looks like this:   

```{r}
n_ppl <- 8
n_objects <- 4

lower_bounds <- c(2,2,2,2)
upper_bounds <- c(4,4,4,4)

sim_data <- matrix( round(runif(n_ppl*n_objects,0,100)), ncol=n_objects ) 
colnames(sim_data) <- paste( "obj", 1:n_objects )
rownames(sim_data) <- paste( "person", 1:n_ppl )

sim_data 
```

Each cell shows the profit obtained if that object is assigned to that person. 

The constraints specified mean that no object can be assigned less than 2 times, or more than 4 times. 

Solve this problem using Rglpk_solve_LP():

```{r}
obj <- c( t(sim_data) )
lhs_matrix <- matrix( rep( rep(0,length(sim_data)), nrow(sim_data) ), ncol=length(sim_data) )

for( i in 1:nrow(sim_data) ){      # iterate through all of the people
   lhs_matrix[i, (1:ncol(sim_data))+ncol(sim_data)*(i-1)] <- 1
 }

dir <- rep( "==", nrow(lhs_matrix) )
rhs <- rep( 1, nrow(lhs_matrix) )

# add lower bound and upper bound constraints:
lower_bound_lhs_mat <- matrix( rep(0, ncol(sim_data)*length(obj)), ncol=length(obj) ) 
for( obj_i in 1:nrow(lower_bound_lhs_mat) ){
  lower_bound_lhs_mat[obj_i, seq(obj_i, nrow(sim_data)*ncol(sim_data),ncol(sim_data))] <- 1
}
upper_bound_lhs_mat <- lower_bound_lhs_mat

lhs_matrix <- rbind( lhs_matrix, lower_bound_lhs_mat, upper_bound_lhs_mat )

dir <- c( dir, rep(">=", nrow(lower_bound_lhs_mat)) )   # add lower bound
dir <- c( dir, rep("<=", nrow(upper_bound_lhs_mat)) )   # add upper bound

rhs <- c( rhs, lower_bounds )   # add lower bounds
rhs <- c( rhs, upper_bounds )   # add upper bounds

types <- rep( "B", length(obj) )

solve_problem <- Rglpk_solve_LP( obj, 
                                 lhs_matrix, 
                                 dir, 
                                 rhs, 
                                 types = types, 
                                 max = TRUE
                               )
solve_problem$status
```


clean up the Rglpk_solve_LP() solution and have a look at it:

```{r}
clean_solution <- matrix( solve_problem$solution, byrow=TRUE, ncol=ncol(sim_data) )
colnames(clean_solution) <- colnames(sim_data)
rownames(clean_solution) <- rownames(sim_data)

clean_solution

paste0( "objective value at optimal solution:  ", solve_problem$optimum )
```

let's check by brute force that this actually is the optimal solution:
```{r}
brute_force_solution <-
  expand.grid( person1=1:ncol(sim_data),
               person2=1:ncol(sim_data),
               person3=1:ncol(sim_data),
               person4=1:ncol(sim_data),
               person5=1:ncol(sim_data),
               person6=1:ncol(sim_data),
               person7=1:ncol(sim_data),
               person8=1:ncol(sim_data) #,
               #person9=1:ncol(sim_data),
               #person10=1:ncol(sim_data),
               #person11=1:ncol(sim_data)
            )

store_objective_values <- rep( 0, nrow(brute_force_solution) )

brute_force_solution %>% head

for( i in 1:nrow(brute_force_solution) ){
 
  extract_solution <- unlist( brute_force_solution[i,] )
  extract_solution_coef <- rep(-99, length(extract_solution))
   
  for( j in 1:length(extract_solution) ){
    extract_solution_coef[j] <- sim_data[j, extract_solution[j] ]
  }
 
  # check for feasibility:
  count_objects <- lower_bounds*0
  for( k in 1:ncol(sim_data) ){
    count_objects[k] <- sum(extract_solution==k)
  }
  if( ( sum(count_objects >= lower_bounds) == ncol(sim_data) ) &
      ( sum(count_objects <= upper_bounds) == ncol(sim_data) )
    ){ store_objective_values[i] <- sum(extract_solution_coef) } else{  
       store_objective_values[i] <- 0
  }
}

# show the brute force solution:
brute_force_solution <- unlist( brute_force_solution[ which.max(store_objective_values), ] )
obj_vals <- rep( 0, nrow(sim_data) )
for( i in 1:nrow(sim_data) ){
  obj_vals[i] <- sim_data[ i, brute_force_solution[i] ]
}

brute_force_solution

# objective value (total profit) obtained under brute force solution:
sum( obj_vals )
```

and, to confirm, the LP solution was:
```{r}
solve_problem$optimum
sum( clean_solution * sim_data )
clean_solution
colSums(clean_solution)
```

# Let's see how big we can go with a pool of 20 offers
```{r}
n_ppl <- 5000
n_objects <- 20

lower_bounds <- rep( round(n_ppl*0.05), n_objects )
upper_bounds <- rep( round(n_ppl*0.25), n_objects )

sim_data <- matrix( round(runif(n_ppl*n_objects,0,100)), ncol=n_objects ) 
colnames(sim_data) <- paste( "obj", 1:n_objects )
rownames(sim_data) <- paste( "person", 1:n_ppl )

sim_data %>% head()
```

solve using linear integer programming:
```{r}
obj <- c( t(sim_data) )
lhs_matrix <- matrix( rep( rep(0,length(sim_data)), nrow(sim_data) ), ncol=length(sim_data) )

for( i in 1:nrow(sim_data) ){      # iterate through all of the people
   lhs_matrix[i, (1:ncol(sim_data))+ncol(sim_data)*(i-1)] <- 1
 }

dir <- rep( "==", nrow(lhs_matrix) )
rhs <- rep( 1, nrow(lhs_matrix) )

# add lower bound and upper bound constraints:
lower_bound_lhs_mat <- matrix( rep(0, ncol(sim_data)*length(obj)), ncol=length(obj) ) 
for( obj_i in 1:nrow(lower_bound_lhs_mat) ){
  lower_bound_lhs_mat[obj_i, seq(obj_i, nrow(sim_data)*ncol(sim_data),ncol(sim_data))] <- 1
}
upper_bound_lhs_mat <- lower_bound_lhs_mat

lhs_matrix <- rbind( lhs_matrix, lower_bound_lhs_mat, upper_bound_lhs_mat )

dir <- c( dir, rep(">=", nrow(lower_bound_lhs_mat)) )   # add lower bound
dir <- c( dir, rep("<=", nrow(upper_bound_lhs_mat)) )   # add upper bound

rhs <- c( rhs, lower_bounds )   # add lower bounds
rhs <- c( rhs, upper_bounds )   # add upper bounds

types <- rep( "B", length(obj) )

system.time( 
  solve_problem <- Rglpk_solve_LP( obj, 
                                   lhs_matrix, 
                                   dir, 
                                   rhs, 
                                   types = types, 
                                   max = TRUE,
                                   control = list(verbose=TRUE)
                                 )
)

solve_problem$status
```

clean up the solution and have a look:
```{r}
clean_solution <- matrix( solve_problem$solution, byrow=TRUE, ncol=ncol(sim_data) )
colnames(clean_solution) <- colnames(sim_data)
rownames(clean_solution) <- rownames(sim_data)

head( clean_solution )

paste0( "objective value at optimal solution:  ", solve_problem$optimum )

print( "Number of each object allocated:" )
colSums(clean_solution)
```

# Define my algorithm:

define LP solving function for my problem:

```{r}

# profit_matrix <- sim_data[1:1000,]  
# lower_object_bounds <- rep( round(nrow(profit_matrix)*0.01), n_objects )
# upper_object_bounds <- rep( round(nrow(profit_matrix)*0.2), n_objects )

LP_solver_ftn <- function( profit_matrix,
                           lower_object_bounds,    # min number of each object type that can be assigned
                           upper_object_bounds     # max number of each object type that can be assigned
                        ){

  # set up objective function   
  obj <- c( t(profit_matrix) )

  # set up constraint matrix (LHS)
  lhs_matrix <- matrix( rep( rep(0,length(profit_matrix)), nrow(profit_matrix) ), ncol=length(profit_matrix) )

  for( i in 1:nrow(profit_matrix) ){      # iterate through all of the people
     lhs_matrix[i, (1:ncol(profit_matrix))+ncol(profit_matrix)*(i-1)] <- 1
  }

  dir <- rep( "==", nrow(lhs_matrix) )
  rhs <- rep( 1, nrow(lhs_matrix) )

  # add lower bound and upper bound constraints to constraint matrix (LHS):
  lower_bound_lhs_mat <- matrix( rep(0, ncol(profit_matrix)*length(obj)), ncol=length(obj) ) 
  for( obj_i in 1:nrow(lower_bound_lhs_mat) ){
    lower_bound_lhs_mat[obj_i, seq(obj_i, nrow(profit_matrix)*ncol(profit_matrix),ncol(profit_matrix))] <- 1
  }
  upper_bound_lhs_mat <- lower_bound_lhs_mat

  lhs_matrix <- rbind( lhs_matrix, lower_bound_lhs_mat, upper_bound_lhs_mat )

  dir <- c( dir, rep(">=", nrow(lower_bound_lhs_mat)) )   # add lower bound
  dir <- c( dir, rep("<=", nrow(upper_bound_lhs_mat)) )   # add upper bound

  rhs <- c( rhs, lower_object_bounds )   # add lower bounds
  rhs <- c( rhs, upper_object_bounds )   # add upper bounds

  # set decision variable to binary type 
  types <- rep( "B", length(obj) )

  # run LP:
  solve_problem <- Rglpk_solve_LP( obj, 
                                   lhs_matrix, 
                                   dir, 
                                   rhs, 
                                   types = types, 
                                   max = TRUE
                                 )
  
  # solution with nice formatting:
  clean_solution <- matrix( solve_problem$solution, byrow=TRUE, ncol=ncol(profit_matrix) )
  colnames(clean_solution) <- colnames(profit_matrix)
  rownames(clean_solution) <- rownames(profit_matrix)

  
  return( 
          list( model_object = solve_problem,
                clean_solution = clean_solution,
                total_profit_this_group = sum( clean_solution * profit_matrix ),
                object_allocation_summary = colSums(clean_solution)
              )  
  )
}
```

make a proper huge dataset:
```{r}

n_ppl <- 123456
n_objects <- 20

sim_data <- matrix( round(runif(n_ppl*n_objects,0,100)), ncol=n_objects ) 
colnames(sim_data) <- paste( "obj", 1:n_objects )
rownames(sim_data) <- paste( "person", 1:n_ppl )

sim_data %>% head()
dim( sim_data )
```

define iterative LP algorithm:
```{r}

run_iterative_LP <- 
  function( profit_matrix,              # matrix of profit of each object for each person
            max_groupsize = 200,         # max size of each subgroup on which LP is performed   (approximate)
            lower_object_bounds_percent,         # min number of each object type that can be assigned
            upper_object_bounds_percent          # max number of each object type that can be assigned
          ){
 
    # assign people to groups:
    num_groups <- ceiling(  nrow(profit_matrix) / max_groupsize  )
    group_IDs <- sample( 1:num_groups, size=nrow(profit_matrix), replace = TRUE )
    
    run_LPs_inner <- list() 
      
    for( group_i in 1:num_groups ){
      print( paste0( "run group ", group_i ) )
      run_LPs_inner[[group_i]]  <- LP_solver_ftn( profit_matrix[ which(group_IDs==group_i), ],    # select only individuals from group_i
                                                lower_object_bounds = 
                                                  round( lower_object_bounds_percent * sum(group_IDs==group_i) ),
                                                upper_object_bounds = 
                                                  round( upper_object_bounds_percent * sum(group_IDs==group_i) )
                                              )
    }
     
    # join best LP result into 1 big table:
    join_LPs <- 
      lapply( run_LPs_inner,
              function(x){ x$clean_solution }
            ) %>% 
      do.call( rbind, . )
      
    return( join_LPs )
}  

```

try out my iterative LP function:
```{r}

test <- 
  run_iterative_LP( profit_matrix = sim_data,
                    max_groupsize = 100,         
                    lower_object_bounds_percent = rep(0.01, ncol(sim_data) ),         
                    upper_object_bounds_percent = rep(0.2, ncol(sim_data) )
                  )

# look at the mix of object assignments: 
colSums( test ) / nrow(test)

# total profit for this assignment:
sum( test * sim_data )

```

define swapping algorithm:
```{r}
# current_best_solution <- test 
# 
#

run_swapping_alg <- 
  function( profit_matrix,
            current_best_solution,
            lower_object_bounds_percent,         # min number of each object type that can be assigned
            upper_object_bounds_percent,          # max number of each object type that can be assigned
            n_iterations = 1000                        # number of swaps to perform
          ){
 
    # print current profit:
    current_profit <- sum(profit_matrix*current_best_solution)
    print( paste0( "total profit: ", current_profit ) )
    
    for( iteration_i in 1:n_iterations ){
      
      #####################################
      ## randomly change someones object ##  
      #####################################
      
      random_row <- sample( c(1, rep(0,ncol(profit_matrix)-1)) ) 
      proposed_new_solution <- current_best_solution
      proposed_new_solution[ sample(1:nrow(profit_matrix), size=1), ] <- random_row
      profit_under_proposed <- sum( profit_matrix*proposed_new_solution )
      if( profit_under_proposed > current_profit ){
        
        if( 
            # lower bound constraint satisfied:
            sum(   
                  ( colSums(proposed_new_solution)/nrow(proposed_new_solution) >= 
                    lower_object_bounds_percent 
                  ) == FALSE
            ) == 0 &
            # upper bound constraint satisfied:
            sum(   
                  ( colSums(proposed_new_solution)/nrow(proposed_new_solution) <= 
                    upper_object_bounds_percent 
                  ) == FALSE
            ) == 0 
        ){
          current_best_solution <- proposed_new_solution
          current_profit <- profit_under_proposed
          print( paste0( "total profit: ", current_profit ) )
        }
        
      }  
      
      
      #####################################
      ## swap 2 people's objects         ##  
      #####################################
      person1_id <- sample( 1:nrow(current_best_solution), size=1 )
      person2_id <- sample( 1:nrow(current_best_solution), size=1 )
      
      person1_row <- current_best_solution[ person1_id, ]
      person2_row <- current_best_solution[ person2_id, ]
        
      if( person1_id==person2_id ){ person1_id <- 1
                                    person2_id <- 2  
                                  }
      
      proposed_new_solution <- current_best_solution
      proposed_new_solution[ person1_id,  ] <- person2_row
      proposed_new_solution[ person2_id,  ] <- person1_row
      profit_under_proposed <- sum( profit_matrix*proposed_new_solution )
      if( profit_under_proposed > current_profit ){
        
        if( 
            # lower bound constraint satisfied:
            sum(   
                  ( colSums(proposed_new_solution)/nrow(proposed_new_solution) >= 
                    lower_object_bounds_percent 
                  ) == FALSE
            ) == 0 &
            # upper bound constraint satisfied:
            sum(   
                  ( colSums(proposed_new_solution)/nrow(proposed_new_solution) <= 
                    upper_object_bounds_percent 
                  ) == FALSE
            ) == 0 
        ){
          current_best_solution <- proposed_new_solution
          current_profit <- profit_under_proposed
          print( paste0( "total profit: ", current_profit ) )
        }
        
      }  
      
    }
   
    return( current_best_solution )
     
  }

```

```{r}
test2 <- run_swapping_alg( profit_matrix = sim_data,
                           current_best_solution = test,
                           lower_object_bounds_percent = rep(0.01, ncol(sim_data) ),
                           upper_object_bounds_percent = rep(0.2, ncol(sim_data) ),
                           n_iterations = 100
                        )

# check mix after swapping alg:
colSums( test2 ) / nrow(test2) 

# profit improvement from swapping:
sum( test2 * sim_data ) -
sum( test * sim_data )
```



# Let's try it out:
```{r}
tfg_profit_matrix <- 
  readRDS( "C:/Users/Joe/Desktop/local_object_allocation_dev/profit_matrix.rds" )

dim( tfg_profit_matrix )

head(tfg_profit_matrix)

#clean_tfg_profit_matrix <- round(tfg_profit_matrix, 2)
```

```{r}
tfg_run_lp_step <-  
  run_iterative_LP( profit_matrix = tfg_profit_matrix, # matrix of profit of each object for each person
            max_groupsize = 1000,         # max size of each subgroup on which LP is performed   (approximate)
            lower_object_bounds_percent = rep(0.01, ncol(tfg_profit_matrix)),
            upper_object_bounds_percent = rep(0.25, ncol(tfg_profit_matrix))
          )

colSums(tfg_run_lp_step) / nrow(tfg_run_lp_step)

saveRDS( tfg_run_lp_step,
          "C:/Users/Joe/Desktop/local_object_allocation_dev/run_lp_step_interim_solution.rds"
       )
```

```{r}

test <- tfg_run_lp_step

test <- 
  run_swapping_alg( profit_matrix = tfg_profit_matrix,
                    current_best_solution = test,
                    lower_object_bounds_percent = rep(0.01, ncol(tfg_profit_matrix)),
                    upper_object_bounds_percent = rep(0.25, ncol(tfg_profit_matrix)),
                    n_iterations = 1e5L                 
                )

colSums(test) / nrow(test)

# total profit at previous step: 
sum( tfg_run_lp_step * tfg_profit_matrix )

# total profit now:
sum( test * tfg_profit_matrix )
```










