---
title: "OLS_interrogation"
author: "Joe"
date: "11 January 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(plotly)
library(splines)
library(GGally)     # for ggpairs()
```

https://www.uv.es/uriel/3%20Multiple%20linear%20regression%20estimation%20and%20properties.pdf

http://home.cc.umanitoba.ca/~godwinrt/4042/material/part3.pdf

Elements of 

# General Regression Theory
We assume that our observed data are a partial view of a (possibly infinitely) complex system whose workings we cannot fully understand. However, given that we can see some outputs from this system, we can see that not all observed combinations of the values of the variables generated by the system are equally likely - we assume that these combinations occur with predictable frequencies. That is to say, we assume that $X$ (our predictor variables) and $Y$ (our response variable) follow a random joint probability distribution, and we make models which try to predict what the values of $Y$ are most likely to be given some observed values of $X$.

For example, suppose that we have a sample of customer data. We have observed a whole lot of customer characteristics $X$ and each customer's spend $Y$ in some given month. For example, we could have observed $X=(\text{age, gender, frequency of transacting, favourite brand, ...etc.})$ and want to predict their $Y=\text{spend}$. The actual spend of each customer is completely deterministic - if we had all of the variables which resulted in their spend decision, and were capable of comprehending the complex interactions between the variables which led to the spend decision (the full workings of the customer's brain would be included in these variables), then we could theoretically perfectly predict any customers spend. 

Feasibly, this is never possible. However, suppose that the variables in our dataset are some small subset of the actual variables which deterministically determine their spend decision. To us, the values of our observed variables are occurring seemingly at random, although some values occur more often than others and there are clearly patterns between the variables. For example, we might notice that of all of the young female customers that we see, most tend to shop very frequently. This we can represent using a probability distribution: a random draw from the distribution is more likely to yield a young frequently transacting female than it would a young infrequently transacting female.  

Because there are variables that we cannot see, we observe the system as a probability distribution. Our sample data (the data that we can see) is drawn from this distribution. We don't know what the form of the distribution is, but we can use our sample data to make inferences about it. The value of any one particular variable we cannot predict, although we can approximately quantify how likely it is to take on any particular value based on repeated observations of it (in this way it is random to us). Similarly, we can approximately quantify the relationships between the variables that we can see (e.g. when variable $X_1$ is high and variable $X_2$ is low, $X_3$ is likely to be high).

So here is structure we assume of our data-generating process:

$X \in \mathbb{R}^p$ is a random input vector. For example, one random draw from a 5-dimensional $X \in \mathbb{R}^5$ might yield $X=(2,-4,0,999,12)$

$Y \in \mathbb{R}$ is a random-valued output (response) variable. One random draw from $Y$ could be $Y=3.14159$

$X$ and $Y$ are distributed according to some joint distribution $Pr(X,Y)$. This means that some combinations of $X$ and $Y$ are more likely to arise from random samples from distribution $Pr(X,Y)$ than others.   

Each observation in our observed data (a single realisation of $X$ and $Y$) is a random draw from this distribution $Pr(X,Y)$

Building/fitting a regression model means choosing a function $f(X)$ which takes in a value of $X$ and outputs a prediction $f(X)=\hat Y$ for $Y$.

For any particular regression function $\hat Y = f(X)$ that we choose, we can use a loss function to quantify how well $f(X)$ predicts $Y$. There are many choices of loss function, each assigning different penalties to different sorts of prediction error. This loss function can be used to choose an optimal regression function $f(X)$.  
The most common choice for the loss function is **squared error loss**:

$$L\Big(Y, \space f(X)\Big) \quad = \quad \Big( Y - f(X) \Big)^2$$

For a function $f(X)$ that we've chosen, the expected value of our loss function is:

$$E\Bigg[L\Big( Y, \space f(X) \Big)\Bigg] \quad = \quad \int \Big(Y - f(X)\Big)^2 Pr(X,Y) \space dx dy $$

This is the average value of the loss function over all possible datasets drawn from $Pr(X,Y)$. 

This is also called the **Expected Squared Prediction Error** ($EPE$)

The $EPE$ can be rewritten like this:

$$\begin{array}{lcl} E\Bigg[\Big( Y - f(X) \Big)^2\Bigg] 
&=& E_X\Bigg[ \space E_{Y|X}\bigg[ \big(Y-f(X)\big)^2 \space | X \bigg] \Bigg]\\
\end{array}$$

We choose our regression function $f(X)$ to be the function $f()$ which minimises the $EPE$.

The $EPE$ is minimised by 

$$f(x) \quad = \quad E(Y|X=x)$$

..making $f(x) \quad = \quad E(Y|X=x)$ the best regression function by the least squares criterion.

Thus, the prediction for a given point $x$ resulting in the lowest average (mean) squared error is the mean of the $y$ values corresponding to those $x$.   

Here is a (contrived) concrete example:

Suppose that our true data generating process (which we would never know in reality) can only ever produce the following 15 states of $X=(animal,age)$ and $Y=speed$ (with the probabilities shown):

```{r true_process_dbn}
set.seed(6047)
true_process_dbn <- 
  data_frame( 
              x1_animal = c( rep("dog",6), rep("pig",9) ),
              x2_age = c( 1,2,2,2,3,3,      1,1,1,1,1,2,2,2,2)
            ) %>% 
  mutate( y_speed = 1 + (x1_animal=="dog")*5 + abs( round(rnorm(n(),3,1)) ) - x2_age ) %>% 
  mutate( probability = sample( 1:100, size=15, replace=FALSE ) ) %>% 
  mutate( probability = probability / sum(probability) )

true_process_dbn

```

We don't know the true relationship between animal, age and speed or any other variables that we haven't observed which might affect these, but we can see that the specific state which we observe most from this system is:

```{r most_often_state}
true_process_dbn %>% arrange( desc(probability) ) %>% slice(1)
```

In any sample, we will likely see more pigs than dogs and the pigs will tend to be younger and slower than the dogs. 

Suppose that our regression function is
$$ f(X) =  
    \begin{cases}  
      4 - \text{age} & \text{if animal = 'pig'} \\
      9 - \text{age} & \text{if animal = 'dog'} \\
   \end{cases}
$$

Then, our expected prediction error would be:

$$\begin{array}{lcl} E\Bigg[L\Big( Y, \space f(X) \Big)\Bigg] &=& \int \Big(Y - f(X)\Big)^2 Pr(X,Y) \space dx dy \\
&=& \sum_{i=1}^{15} \Big(y_i - f(x_i)\Big)^2 \space Pr(x_i,y_i) \\
&=& (9-8)^2 \times 0.0848 \space + \space (6-7)^2 \times 0.0950 \space + \space ... \space  + \space (1-2)^2 \times 0.0556 \\ 
&=& 0.6929825 \\
\end{array}$$

```{r showdem}
true_process_dbn %>% 
  mutate( fx = 9 - x2_age - (x1_animal=="pig")*5 ) %>% 
  mutate( loss_times_prob = (y_speed-fx)^2 * probability ) %>% 
  summarise( sum(loss_times_prob) ) %>% 
  unlist()
```

We can show that this expected prediction error (average loss over many samples) is true by sampling data from our true process and calculating our average sample error:  

```{r show_epe_is_true}

sample_n( tbl = true_process_dbn,
          size = 1e7L,
          replace = TRUE,
          weight = true_process_dbn$probability
        ) %>% 
  mutate( fx = 9 - x2_age - (x1_animal=="pig")*5 ) %>% 
  mutate( Lx = ( y_speed - fx )^2  ) %>% 
  summarise( mean(Lx) ) %>% 
  unlist()
  
```

The marginal distribution $Pr(Y|X)$ is:

```{r y_given_X}
y_given_x_dbn <- 
    true_process_dbn %>% 
      group_by( x1_animal, x2_age, y_speed ) %>%
      summarise( probability = sum(probability) ) %>% 
      group_by( x1_animal, x2_age ) %>% 
      mutate( x_probsum = sum(probability) ) %>% 
      mutate( probability = probability/x_probsum ) %>% 
      select( -x_probsum )

y_given_x_dbn

# within each x1/x2 combination, y_speed should sum to 1:
y_given_x_dbn %>% group_by( x1_animal, x2_age ) %>% summarise(sum(probability))
```
For example, if we have observed a pig of age 2, then it will have speed 3 with probability 0.187 (18.7% of age 2 pigs will have speed 3) 

We can verify this again by simulation using the true process:

```{r unbelievably_showdem}
sample_n( tbl = true_process_dbn,
          size = 1e7L,
          replace = TRUE,
          weight = true_process_dbn$probability
        ) %>% 
  summarise( sum(x1_animal=="pig" & x2_age==2 & y_speed==3) / 
             sum(x1_animal=="pig" & x2_age==2 )
            )
```

The marginal distribution $Pr(X)$ is:
```{r p_x_margarine}
X_dbn <- 
  true_process_dbn %>% 
      group_by( x1_animal, x2_age ) %>%
      summarise( probability = sum(probability) )

X_dbn
```
So, $E_{Y|X}\bigg[\Big(Y-f(X)\Big)^2 |X \bigg]$ for each $X$ is:

```{r Exp_y_given_X}
epe_y_given_x <- 
  y_given_x_dbn %>%
      mutate( fx = 9 - x2_age - (x1_animal=="pig")*5 ) %>% 
      mutate( Lx = ( y_speed - fx )^2  ) %>% 
      group_by( x1_animal, x2_age ) %>% 
      summarise( expected_loss_given_X = sum(Lx*probability) ) 

epe_y_given_x
```

..and then $E_X\Bigg[ \space E_{Y|X}\bigg[\Big(Y-f(X)\Big)^2 |X \bigg] \space \Bigg]$ is:

```{r showdem_wholting}
epe_y_given_x %>% 
  left_join( X_dbn ) %>% 
  ungroup() %>% 
  summarise( sum(expected_loss_given_X*probability) ) %>% 
  unlist()

```

So, for this distribution of data, the best regression function f(X) by the least squares criterion is:

```{r best_fx}
true_process_dbn %>% 
  group_by( x1_animal, x2_age ) %>% 
  summarise( `f(X)` = mean(y_speed) )
```

# Linear Model Concept
The linear model assumes that the regression function $f(X)=E\Big[Y|X \Big]$ has the form (or a form very well approximated by):

$$f(X) \quad = \quad E\Big[Y|X\Big] \quad \approx \quad  X\beta \quad = \quad \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3 + ...$$

The coefficients $\beta$ which minimise the expected squared error loss $E\Bigg[\Big(Y - f(X)\Big)^2\Bigg]=E\Bigg[\Big(Y - X\beta\Big)^2\Bigg]$ are $\beta \quad = \quad \Bigg(E\Big[XX^T\Big]\Bigg)^{-1} E\Big[XY\Big]$

For example, suppose that our data are generated by the following distribution function:

```{r true_data_generating_process}
generate_sample <- function(n=5000){
  x1 <- runif( n, -3, 2 )
  x2 <- runif( n, -2, 5 )
  y <- -5 + 2*x1 - 6*x2 + runif(n, -5, 5)

return( data_frame(x1=x1, x2=x2, y=y) )
}
```

For example, one dataset drawn from this distribution is:
```{r show_dataset}
generate_sample(n=10)
```
For this distribution:

$$f(X) \quad = \quad E\Big[Y|X\Big] \quad = \quad -5 + 2x_1 - 6 x_2 $$

and 

$$\beta = \Bigg(E\Big[XX^T\Big]\Bigg)^{-1} E\Big[XY\Big] = E\Bigg[\begin{pmatrix}1\\X_1\\X_2\end{pmatrix}\begin{pmatrix}1&X_1&X_2\end{pmatrix} \Bigg]^{-1} \space E\Bigg[\begin{pmatrix}1\\X_1\\X_2\end{pmatrix} Y \Bigg]$$ is:

If we generate a sample of size 1,000,000 from this distribution and calculate the sample expectation (mean) value of this quantity, we get $\beta = $  

```{r theoretical_beta}
test <- generate_sample( n = 1e6L )

first_part <- list()
second_part <- list()

for( i in 1:nrow(test) ){
  
  x1 <- test$x1[i] 
  x2 <- test$x2[i] 
  y <- test$y[i]

  first_part[[i]] <- rbind(1,x1,x2)%*%cbind(1,x1,x2)  
  second_part[[i]] <- rbind(1,x1,x2)%*%y
}

beta <- 
    solve( apply(simplify2array(first_part), c(1,2), mean) ) %*%
          apply(simplify2array(second_part), c(1,2), mean) %>% 
      round(digits=1)

rownames(beta) <- paste( "beta", 0:2  )
beta
```

Having assumed that the form of the regression function $f(X)=E[Y|X]$ is linear ($X\beta$) for the true distribution of our data $Pr(X,Y)$, we use our actual observed dataset to estimate the values of these unobservable 'true' $\beta$ values by choosing the estimates $\hat\beta$ which minimise 

$$RSS \quad = \quad (y-X\hat\beta)^T(y-X\hat\beta)$$

over all of our observed datapoints (our sample). 

If $X^TX$ is not singular for our sample data, then this has the unique solution:

$$\hat \beta \quad = \quad (X^TX)^{-1}X^Ty$$

If the observations in our dataset are random independent draws from the true population distribution $Pr(X,Y)$ of our data, then this is a good estimator of $\beta$.  


# Questions I want to answer:

Does the response need to be normally distributed?
What happens if Y is linearly related to some of the features, but non-linearly to others?
What is the effect of correlated predictors?
What are the assumptions of OLS? - What happens when I violate each specific one? 
                                 - What happens if I violate a combination of them?
                                 - How do I check if I have violated them?    
Does Y need to be linearly related to each predictor?
What is the Gauss Markov Theorem?
What is BLUE? When is the OLS estimate BLUE?
What are the assumptions of GLMs? - What happens when I violate each specific one? 
                                  - What happens if I violate a combination of them?
                                  - How do I check if I have violated them?  
Can I visualise lm fits in 3d in plot_ly?
What systems other than OLS for estimating the coefficients? How do these compare? When do I use them?
Are there situations in which the OLS model satisfies all of the assumptions but is useless?
When should I use a GLM rather than a LM? 

# Validation Plots
There are a few standard validation plots which are commonly used to signal whether the linear models assumptions are being grossly violated. I've written a function which creates a few of these plots: 

```{r validation_plot_functions}
validation_plots <- function( model_object,
                              original_data
){
  par(mfrow=c(3,3),pty="s", mar=c(4,1.1,3.5,1.1))
  plot( original_data$y ~ fitted.values(model_object), ylab="observed y", xlab="fitted y", 
        main = "I. observed y vs. fitted y")
  abline( 0, 1, lty=2, col=2 )
  plot( residuals(model_object) ~ fitted.values(model_object), ylab="residuals", xlab="fitted y", 
        main = "II. residuals vs. fitted y")
  abline( h=0, lty=2, col=2 )
  plot( residuals(model_object) ~ original_data$x1, ylab="residuals", xlab="x1",
        main = "III. residuals vs. x1")
  abline( h=0, lty=2, col=2 )
  plot( residuals(model_object) ~ original_data$x2, ylab="residuals", xlab="x2",
        main = "IV. residuals vs. x2")
  abline( h=0, lty=2, col=2 )
  if( exists("model_object$x3") ){ 
      plot( residuals(model_object) ~ original_data$x3, ylab="residuals", xlab="x3",
        main = "IV. residuals vs. x3")
      abline( h=0, lty=2, col=2 )
    }
  
  # Kolmogorov Smirnoff test for normality:
  ks_p.value <- 
     ks.test( residuals(model_object), "pnorm", 
              mean=mean(residuals(model_object)), 
              sd=sd(residuals(model_object))
             )$p.value
  
  # Shapiro Wilks test from normality:
  shapiro_wilks_p.value <- shapiro.test(residuals(model_object))$p.value
  
  hist( residuals(model_object), breaks=20, 
        main = paste0( "V. residuals distribution \n", "ks p: ", round(ks_p.value,3),
                       ", shapiro p: ", round(shapiro_wilks_p.value,3)
                       ), 
        freq=FALSE, xlab="residual" )
  abline( v=mean(residuals(model_object)), lty=2, col=2 )
  text( x=mean(residuals(model_object)),
        y=0,
        labels="mean",
        col=2,
        pos=3
        )
  curve( dnorm(x,mean=mean(residuals(model_object)), sd=sd(residuals(model_object))),add=TRUE,col=3 )
  
  qqnorm( residuals(model_object), main="VI. Normal QQ Plot of Residuals" )
  qqline( residuals(model_object), col=2, lwd=2 )
  
}

```

# Assumption 1: Linearity
The linear model assumes that 

 $$f(X) \quad = \quad E[Y|X] \quad\approx \quad X^T\beta \quad = \quad \beta_0 + \beta_1 x_1 + \beta_2 x_2 + ... + \beta_p x_p$$

Specifically, the relationship between response $Y$ and the independent variables $X$ is both linear and additive. 

Each explanatory variable $x$ must be related linearly to the response $y$ in order for the model to fit adequately - this can be checked by making a scatterplot of $x$ against response $y$ for every $x$ variable (as is done in the example below)   

Here is data with 2 predictors which are well-suited by a linear model:

```{r perfect_lm}
simdata <- 
  data_frame( x1 = runif(n=100, -100, 100),
              x2 = runif(n=100, -100, 100),
              errors = rnorm(n=100, 0, 120) 
            ) %>% 
    mutate( y = 5*x1 - 6*x2 + errors )

pairs(simdata[,c("y","x1","x2")], pch=16, cex=0.5)

fit_lm <- lm( y~x1+x2, data=simdata)

simdata$fitted_val <- fitted.values(fit_lm) 

plot_ly( data = simdata, 
         x = ~x1, 
         y = ~x2, 
         z = ~y
  ) %>% 
  add_markers(size=1) %>% 
  add_trace( data = simdata, 
             x = ~x1, 
             y = ~x2, 
             z = ~fitted_val,
             type = "mesh3d",
             alphahull = 0,
             opacity = 0.2
  )

```

The validation plots for this model look like this:

```{r validation_plots_perfect_model}
validation_plots( model_object = fit_lm, original_data = simdata )
```

These validation plots indicate how validation plots will look when the model very satisfactorily satisfies the assumptions of linearity and additivity between $y$ and each $x$ variable.   

Plot I: the points are symmetrically distributed on either side of the $45^{\circ}$ line with even spread (variance), indicating that there is no range of response values on which the model is underperforming or very variable.

plot II: the points are symmetrically distributed on either side of the horizontal line at 0 with even spread/variance. This indicates that there is no range of fitted values for which the model systematically underperforms, or gives more varied predictions. 

plot III and IV: the points are symmetrically distributed on either side of the horizontal line at 0 with even spread/variance. This indicates that the model does not systematically underperform, or give more variable predictions, for any range of predictor variable values.   

Now we look at a situation where the linearity assumption is violated: there is a linear relationship between response $y$ and the first predictor $x_1$, but the relationship between $y$ and $x_2$ is very non-linear. This example shows that the model provides a bad fit to the data if any one $x$ variable is not linearly related to response $y$.

```{r non_linear_x2_}
simdata <- 
  data_frame( x1 = runif(n=100, -100, 100),
              x2 = runif(n=100, -100, 100),
              errors = rnorm(n=100, 0, 50) 
            ) %>% 
    mutate( y = 40*x1 - x2^2 + errors )

pairs(simdata[,c("y","x1","x2")], pch=16, cex=0.5)

fit_lm <- lm( y~x1+x2, data=simdata)

simdata$fitted_val <- fitted.values(fit_lm) 

plot_ly( data = simdata, 
         x = ~x1, 
         y = ~x2, 
         z = ~y
  ) %>% 
  add_markers(size=1) %>% 
  add_trace( data = simdata, 
             x = ~x1, 
             y = ~x2, 
             z = ~fitted_val,
             type = "mesh3d",
             alphahull = 0,
             opacity = 0.2
  )

```

The linear model is clearly fitting the data very poorly. This would have been picked up in our validation plots:

```{r validation_plots}
validation_plots( model_object = fit_lm, original_data = simdata )
```

Plot I: The points seem fairly symmetrical about the line, but not with equal variance. It seems that fitted values are more variable where they are higher (less negative) than the observed values, meaning that when the model

Plot II: Again, the points seem fairly symmetrical about the line, meaning that the model gives as many positive errors as it does negative, and that the model is not underperforming for specific ranges of y. However, the residuals are definitely more variable when they are negative.   

plot IV: There is a clear pattern here. The non-linear relationship between $y$ and $x_2$ is causing a predictable pattern in the residuals - the model systematically overpredicts for certain ranges of $x_2$, and underpredicts for other ranges of $x_2$. 

plots V&VI: The errors are strongly non-normal, indicating that there is signal in the errors which has been missed by the model. It looks like the model sometimes predicts too high and sometimes too low, but rarely predicting correctly.

Non-linear relationships can still be modeled using a linear model by introducing new created variables.

For example, we could make our linear model

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_2^2$$

..which gives this fit to the data:

```{r polynomial_fit}
fit_lm <- lm( y~x1+ x2 + I(x2^2), data=simdata)

simdata$fitted_val <- fitted.values(fit_lm) 

plot_ly( data = simdata, 
         x = ~x1, 
         y = ~x2, 
         z = ~y
  ) %>% 
  add_markers(size=1) %>% 
  add_trace( data = simdata, 
             x = ~x1, 
             y = ~x2, 
             z = ~fitted_val,
             type = "mesh3d",
             opacity = 0.2
  )

```

The validation plots for this polynomial model: 
```{r validation_plots_polynomial_model}
validation_plots( model_object = fit_lm, original_data = simdata %>% mutate(x3 = x1*x2) )
```

The model is clearly not methodically underperforming for certain ranges of our response or predictors. A danger of fitting polynomial models is they are poor at predicting outside of the range of the data on which they were trained. i.e. we cannot trust predictions made using $x$ values outside of the range of the $x$ values which are in the data which we used to train (choose the parameters of) the model. 

A more sophisticated way to fit polynomial relationships within the linear model framework is to use basis splines (I use natural splines here). Splines fit $y$ as a polynomial function of an $x$ variable, but enforcing some some continuity restrictions: 

```{r spline_example}

fit_lm <- lm( y ~ x1 + ns(x2, df=5), data=simdata)
simdata$fitted_val <- fitted.values(fit_lm) 

plot_ly( data = simdata, 
         x = ~x1, 
         y = ~x2, 
         z = ~y
  ) %>% 
  add_markers(size=1) %>% 
  add_trace( data = simdata, 
             x = ~x1, 
             y = ~x2, 
             z = ~fitted_val,
             type = "mesh3d",
             opacity = 0.2
  )

```

Many different non-linear relationships between $y$ and a predictor $x$ can be modeled in this way, with the potential pitfall that we can accidentally overfit to our data (fit our model to random noise in the training data rather than to the meaningful relationships in it).

An alternative to engineering new $x$ variables for dealing with non-linear relationships between $y$ and one or more $X$ variables is to transform the response $y$ variable instead. For example, if we have the following data (where response $y$ is very non-linearly related to the explanatory $x$ variables):

```{r show_transformed_y}
simdata <- 
  data_frame( x1 = runif(n=100, 0, 5),
              x2 = runif(n=100, 0, 5),
              errors = rnorm(n=100, 0, 5) 
            ) %>% 
    mutate( y = exp(x1) + exp(x2) + errors ) %>% 
    mutate( y = abs(y)+1 )
pairs(simdata[,c("y","x1","x2")], pch=16, cex=0.5)
```

then we can instead regress response $log(y)$ against $x$, which makes the relationships much more linear:

```{r try_log_y}
simdata$log_y <- log(simdata$y)
pairs(simdata[,c("log_y","x1","x2")], pch=16, cex=0.5)
```

Here are the fitted models $y=\beta_0+\beta_1 x_1+\beta_2 x_2 + \epsilon$ and $log(y)=\beta_0+\beta_1 x_1+\beta_2 x_2 + \epsilon$ plotted on top of the data:

```{r plot_them_bitch}
# straight up model:
fit_lm <- lm( y ~ x1 + x2, data=simdata)
simdata$fitted_val <- fitted.values(fit_lm) 
plot_ly( data = simdata, 
         x = ~x1, 
         y = ~x2, 
         z = ~y
  ) %>% 
  add_markers(size=1) %>% 
  add_trace( data = simdata, 
             x = ~x1, 
             y = ~x2, 
             z = ~fitted_val,
             type = "mesh3d",
             opacity = 0.2
  )

# with log(y):
fit_lm <- lm( log(y) ~ x1 + x2, data=simdata)
simdata$fitted_val <- fitted.values(fit_lm) 
plot_ly( data = simdata, 
         x = ~x1, 
         y = ~x2, 
         z = ~y
  ) %>% 
  add_markers(size=1) %>% 
  add_trace( data = simdata, 
             x = ~x1, 
             y = ~x2, 
             z = ~exp(fitted_val),
             type = "mesh3d",
             opacity = 0.2
  )

plot_ly( data = simdata, 
         x = ~x1, 
         y = ~x2, 
         z = ~log(y)
  ) %>% 
  add_markers(size=1) %>% 
  add_trace( data = simdata, 
             x = ~x1, 
             y = ~x2, 
             z = ~fitted_val,
             type = "mesh3d",
             opacity = 0.2
  )

```
Here are the validation plots for the log(y) fit:
```{r validate_loggy}
validation_plots( model_object = fit_lm, original_data = simdata %>% mutate(y = log(y)) )
```

**Plot I:** Points are somewhat symmetrically distributed on either side of the diagonal line. Predictions (fitted values) generally agree in order of magnitude with observed $y$ (i.e. the model predicts correctly that y is small or y is large), but predictions are more variable for small values of y (small predictions from this model are less trustworthy).

**Plot II:** Again, we can see that small fitted values are less trustworthy (more likely to make a large error). The even spread of points around 0 indicates that the model is not systematically overpredicting or underpredicting. 

**Plots III and IV:** The model seems to be least trustworthy (most variable in it's predictions) when predicting from small values of $x_1$ and $x_2$. For the largest values of $x_1$ and $x_2$, the model seems to be producing larger positive residuals, meaning that it is underpredicting $y$ in this range.  


# Assumption 2: Additivity
This is the assumption that the effect of each independent variable $x$ on the response is not affected by changing levels of the other independent variables $X$ in the dataset. 

This is a strong assumption.

The model is $y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + ... + \beta_p x_p + \epsilon$

$\beta_j$ is the change in $y$ for a one unit increase in $x_j$, keeping all of the other independent variables fixed (i.e. having controlled for the other variables). In other words, $\beta_j$ is the marginal change in $y$ caused by a 1 unit increase in variable $x_j$ having  

For example, suppose that we have the following data:
```{r controlling_example}
simdata <- 
  data_frame( x1 = (rbeta(n=100, shape1=69, shape2=3) - 0.5)*50 ,
              x2 = (rbeta(n=100, shape1=3, shape2=69) - 0.5)*100 ,
              errors = rnorm(n=100, 0, 10) 
            ) %>%
  mutate( y = 100 - 2*x1 + x2 + x1*x2 + errors )
  
pairs(simdata[,c("y","x1","x2")], pch=16, cex=0.5)
```

If we fit the model $y = \beta_1 x_1 + \epsilon$, we estimate the effect of variable $x_1$ on $y$ to be

```{r z3ro}
fit_lm <- lm( y ~ x1 - 1, data = simdata )
coef(fit_lm)
```

Here is plot of the fitted model line to the data:
```{r on3}
plot( y ~ x1, data = simdata, las=2)
abline(fit_lm, col=2)

plot_ly( data = simdata, 
         x = ~x1, 
         y = ~x2, 
         z = ~y
  ) %>% 
  add_markers(size=1) %>% 
  add_trace( data = simdata, 
             x = ~x1, 
             y = ~x2, 
             z = ~fitted.values(fit_lm),
             type = "mesh3d",
             opacity = 0.2
  )
```

If we allow for the fact $y$ could have some baseline level, on top of which we want to measure the additive effect of $x_1$, then this corresponds to the model $y = \beta_0 + \beta_1 x_1 + \epsilon$:  

```{r tw0}
fit_lm <- lm( y ~ x1, data=simdata )
coef(fit_lm)
```

Where the baseline level is estimated to be `r fit_lm$coef["(Intercept)"]` and the additive effect of $x_1$ on top of basline is `r fit_lm$coef["x1"]`. Here is a plot of this model:

```{r 7hr33}
plot( y ~ x1, data = simdata, las=2)
abline(fit_lm, col=2)

plot_ly( data = simdata, 
         x = ~x1, 
         y = ~x2, 
         z = ~y
  ) %>% 
  add_markers(size=1) %>% 
  add_trace( data = simdata, 
             x = ~x1, 
             y = ~x2, 
             z = ~fitted.values(fit_lm),
             type = "mesh3d",
             opacity = 0.2
  )
```

If we now also include variable $x_2$ in the model, fitting $y = \beta_0 +\beta_1 x_1 + \beta_2 x_2 + \epsilon$, . We get the following estimates for $\beta_0, \beta_1 and \beta_2$:

```{r f0ur}
fit_lm <- lm( y ~ x1 + x2, data=simdata )
coef(fit_lm)
```

Notice the change in the estimates of the intercept and $x_1$ coefficients ($\hat\beta_0$ and $\hat\beta_1$) from the previous model fit. Now that we have additional variable $x_2$ in the model, our coefficient estimates are now for the additive effects of the baseline level of $y$ and $x_1$ **in the presence of** $x_2$. Some of the change in $y$ that we previously attributed to the intercept and variable $x_1$ (in the absence of any better information) we can now better attribute to variable $x_2$.

Here is a plot of the fit of this model:

```{r f0ur.pn7.f1v3}
plot_ly( data = simdata, 
         x = ~x1, 
         y = ~x2, 
         z = ~y
  ) %>% 
  add_markers(size=1) %>% 
  add_trace( data = simdata, 
             x = ~x1, 
             y = ~x2, 
             z = ~fitted.values(fit_lm),
             type = "mesh3d",
             opacity = 0.2
  )
```

If we include an interaction term in our model, fitting 

$$y = \beta_0 +\beta_1 x_1 + \beta_2 x_2 + \beta_3 x_1\times x_2 + \epsilon$$

we get updated estimates of the additive effects of the intercept, $x_1$ and $x_2$. We also get an estimate of the interaction effect between variables $x_1$ and $x_2$:

```{r fiv3}
fit_lm <- lm( y ~ x1 + x2 + x1:x2, data=simdata )
coef(fit_lm)
```

The coefficients have changed from the previous model estimates because we can now attribute some of the change in $y$ to the fact that the effect of variable $x_1$ on $y$ is possibly dependent on the value/level of $x_2$. The previous linear model found the best estimates for the additive effects of $x_1$ and $x_2$ that it could under the assumption that the **interaction** effect did not exist.     

Here is a plot of the fit of this model:
```{r fiv3_p0in7_3nd}
plot_ly( data = simdata, 
         x = ~x1, 
         y = ~x2, 
         z = ~y
  ) %>% 
  add_markers(size=1) %>% 
  add_trace( data = simdata, 
             x = ~x1, 
             y = ~x2, 
             z = ~fitted.values(fit_lm),
             type = "mesh3d",
             opacity = 0.2
  )
```


SPAAAAAAAAAAAAAAAACE
SPAAAAAAAAAAAAAAAACE
SPAAAAAAAAAAAAAAAACE
NEW IDEA

For example, suppose that the effect of variable $x_1$ on $y$ is linked to the value of variable $x_2$:

```{r additivity_violation}
simdata <- 
  data_frame( x1 = rexp(n=100, rate=0.5),
              x2 = runif(n=100, -100, 100),
              errors = rnorm(n=100, 0, 50) 
            ) %>% 
    mutate( y = 5 + 16*x1 -3*x2 + 9*x1*x2 + errors )
pairs(simdata[,c("y","x1","x2")], pch=16, cex=0.5)
```

In this case, the assumption of additivity provides a pretty bad model fit: 

```{r related_x1_x2, echo=FALSE}
fit_lm <- lm( y ~ x1 + x2, data=simdata)
simdata$fitted_val <- fitted.values(fit_lm) 
plot_ly( data = simdata, 
         x = ~x1, 
         y = ~x2, 
         z = ~y
  ) %>% 
  add_markers(size=1) %>% 
  add_trace( data = simdata, 
             x = ~x1, 
             y = ~x2, 
             z = ~fitted_val,
             type = "mesh3d",
             opacity = 0.2
  )

```

Scatterplots of the explanatory variables $x$ coloured by the response variable $y$ show that high values of $x_1$ combined with high values of $x_2$ seem to correspond to high values of response $y$ - even higher than we'd expect from the sum of the individual effects of $x_1$ on $y$ and $x_2$ on $y$.

```{r ggpairs_is_SHIT}
library(RColorBrewer)
cols <- brewer.pal(4, "Reds")
pal <- colorRampPalette(cols)
simdata$order <- findInterval(simdata$y, sort(simdata$y))

pairs( simdata %>% select(x1,x2,y),
       col=pal(nrow(simdata))[simdata$order]  
     )

```

We can sometimes fix this by including interaction terms between the x variables in the linear model, which assumes that the interaction effect is itself linear.  

$y=\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_1 x_2 + \epsilon$

```{r show_coef}
fit_lm <- lm( y ~ x1 + x2 + x1*x2, data=simdata)
simdata$fitted_val <- fitted.values(fit_lm) 
coef(fit_lm)
plot_ly( data = simdata, 
         x = ~x1, 
         y = ~x2, 
         z = ~y
  ) %>% 
  add_markers(size=1) %>% 
  add_trace( data = simdata, 
             x = ~x1, 
             y = ~x2, 
             z = ~fitted_val,
             type = "mesh3d",
             opacity = 0.2
  )
```

If the interaction effect on response $y$ between $x_1$ and $x_2$ is itself not linear, then including the linear interaction term won't help much:

```{r non_linear_interaction}
simdata <- 
  data_frame( x1 = runif(n=100, 0, 100),
              x2 = runif(n=100, 0, 500),
              errors = rnorm(n=100, 0, 100) 
            ) %>% 
    mutate( y = 5 + 16*x1 -0.3*x2 + 0.4*x1*x2 - 100*sqrt(x1*x2)  + errors )
pairs(simdata[,c("y","x1","x2")], pch=16, cex=0.5)

fit_lm <- lm( y ~ x1 + x2 + x1*x2, data=simdata)
simdata$fitted_val <- fitted.values(fit_lm) 
plot_ly( data = simdata, 
         x = ~x1, 
         y = ~x2, 
         z = ~y
  ) %>% 
  add_markers(size=1) %>% 
  add_trace( data = simdata, 
             x = ~x1, 
             y = ~x2, 
             z = ~fitted_val,
             type = "mesh3d",
             opacity = 0.2
  )
```

In which case, more complicated interaction terms can help: (IS THIS A GOOD IDEA?)

```{r complicated_non_linearity}
simdata$higher_interact_x1_x2 = (simdata$x1*simdata$x2)^2
fit_lm <- lm( y ~ x1 + x2 + x1*x2 + higher_interact_x1_x2, data=simdata)
simdata$fitted_val <- fitted.values(fit_lm) 
plot_ly( data = simdata, 
         x = ~x1, 
         y = ~x2, 
         z = ~y
  ) %>% 
  add_markers(size=1) %>% 
  add_trace( data = simdata, 
             x = ~x1, 
             y = ~x2, 
             z = ~fitted_val,
             type = "mesh3d",
             opacity = 0.2
  )
```
# Assumption 3: Errors are independently and normally distributed with mean 0 

The OLS estimates for the coefficients $\hat \beta$ are given by the solution to

$$\hat \beta \quad = \quad (X^TX)^{-1}X^Ty$$

Where $X$ and $y$ are the variables in our observed dataset, which we consider to be drawn from some unknown probability distribution $Pr(X,Y)$ 

These are the estimators $\hat \beta$ which minimise 

$$RSS \quad = \quad (y-X\hat\beta)^T(y-X\hat\beta) \quad = \quad \sum_{i=1}^n \Big( y-{x_i}^T  \hat\beta \Big)^2$$

over our observed sample data i.e. the estimators which minimise the sum of squared errors over the data that we have. 

If we assume that our optimal (mimises the $EPE$) regression function $f(X) = E\Big[Y|X\Big]$ is truly linear and that deviations of $Y|X$ around it's expected value are additive and Gaussian (normal) with mean 0 and constant variance, 

$$\begin{array}{lcl} E[Y|X] &=& X\beta + \epsilon \\
\epsilon &\sim& N\Big(0, \space \sigma^2 \Big) \\ \end{array}$$

then the distribution of $\hat\beta$ over different samples from the true data-generating process/distribution $Pr(X,Y)$ is:

$$\hat\beta \quad \sim \quad N\Bigg(\beta, \space \Big(X^TX \Big)^{-1}\sigma^2\Bigg) \\ $$

we can show this by simulation: 
suppose that we have a sample of size n=5000 and our $X$ data is:

```{r X_data}
simdata <- data_frame( x1 = sample( -100:100, size=5000, replace = TRUE),
                       x2 = rbeta( n=5000, shape1 = 2, shape2 = 12)*100,
                       x3 = rnorm( n=5000, mean=30, sd=30), 
                       x4 = rgamma(5000, shape=1.3, rate=0.1)*2 
                     )
head(simdata)
```

If we generate 1000 random datasets of size n=5000 each with the same $X$ values (those shown above), and Y values randomly generated according to 

$$\begin{array}{lcl} Y &=& X\beta + \epsilon \\
&=& -500 + 3x_1 -2 x_2 + 6 x_3 - x_4 -2 x_1 x_2 + \epsilon\\  
\end{array}$$

Where each $\epsilon$ value is a random value drawn from the distribution $N\Big(0,2000^2\Big)$,

then we can calculate $\hat\beta$ for each dataset, in order to see the distribution of $\hat\beta$ over different datasets drawn from $Pr(X,Y|X)$: 

```{r generate_y}
X <- cbind(1, simdata$x1, simdata$x2, simdata$x3, simdata$x4, simdata$x1*simdata$x2)
head(X)
true_beta <- matrix( c(-500, 3, -2, 6, -1, -2), ncol=1) 
rownames(true_beta) <- paste("b", 0:5, sep="")
true_beta

beta_hats <- 
    lapply( 1:1e5L,
            function(x) {  e <- rnorm(5000, mean=0, sd=2000) 
                           y <- X%*%true_beta + e
                           beta_hat <- solve( t(X)%*%X ) %*% t(X) %*% y
                           return( t(beta_hat) )
            }
          ) %>% 
    purrr::reduce( .f=rbind, .x=. )

colnames(beta_hats) <- paste("b", 0:5, sep="")
```

The true values for $\beta$ (the ones used to generate $y$) were:
```{r show_beta}
true_beta
```

This means that the distribution of $\hat\beta$ should be normal (Gaussian) with mean vector
```{r show_beta_again, echo=FALSE}
true_beta
```
e.g. the mean of the distribution of $\hat\beta_3$ should be `r true_beta[4,]`

and the variance covariance structure of $\hat \beta$ should be:
```{r var_covar}
covar <- solve( t(X)%*%X ) * (2000^2)
rownames(covar) <- paste("b", 0:5, sep="")
colnames(covar) <- paste("b", 0:5, sep="")
covar
```

The distibutions of  $\hat\beta$:

```{r beta_0}
hist( beta_hats[,"b0"], freq=FALSE, breaks = 50, main="distribution of hat[beta0] across samples" )
curve( dnorm(x, 
             mean = true_beta["b0",],
             sd = sqrt( covar["b0","b0"] )
            ),
       add = TRUE,
       col = "red"
      )

hist( beta_hats[,"b0"], freq=FALSE, breaks = 50, main="distribution of hat[beta0] across samples" )
curve( dnorm(x, 
             mean = true_beta["b0",],
             sd = sqrt( covar["b0","b0"] )
            ),
       add = TRUE,
       col = "red"
      )

hist( beta_hats[,"b1"], freq=FALSE, breaks = 50, main="distribution of hat[beta1] across samples" )
curve( dnorm(x, 
             mean = true_beta["b1",],
             sd = sqrt( covar["b1","b1"] )
            ),
       add = TRUE,
       col = "red"
      )

hist( beta_hats[,"b2"], freq=FALSE, breaks = 50, main="distribution of hat[beta2] across samples" )
curve( dnorm(x, 
             mean = true_beta["b2",],
             sd = sqrt( covar["b2","b2"] )
            ),
       add = TRUE,
       col = "red"
      )

hist( beta_hats[,"b3"], freq=FALSE, breaks = 50, main="distribution of hat[beta3] across samples" )
curve( dnorm(x, 
             mean = true_beta["b3",],
             sd = sqrt( covar["b3","b3"] )
            ),
       add = TRUE,
       col = "red"
      )

hist( beta_hats[,"b4"], freq=FALSE, breaks = 50, main="distribution of hat[beta4] across samples" )
curve( dnorm(x, 
             mean = true_beta["b4",],
             sd = sqrt( covar["b4","b4"] )
            ),
       add = TRUE,
       col = "red"
      )

hist( beta_hats[,"b5"], freq=FALSE, breaks = 50, main="distribution of hat[beta5] across samples" )
curve( dnorm(x, 
             mean = true_beta["b5",],
             sd = sqrt( covar["b5","b5"] )
            ),
       add = TRUE,
       col = "red"
      )
```


If we are happy to accept these assumptions, then we can use these distributional result to derive confidence intervals and hypothesis tests for our coefficients $\hat \beta$. 

Now, let us suppose that our dataset is from a distribution where the assumption that the regression function is linear ($f(X)=E[Y|X] = X\beta+\epsilon$) is an unreasonable assumption. 

Say that our true data distribution $Pr(X,Y)$ is:

```{r simdata_vba}
generate_errors <- function(n=1000){
  
    data_frame( e1 = rnorm(n, -40, 5),
                e2 = 0, #1/rnorm(n, 40, 5),
                e3 = sample( c(-100, -50, -20, 40), size=n, replace=TRUE ) 
              ) %>% 
    mutate( runif01_a = sample(c(0,1), replace=TRUE, size=n() ),
            runif01_b = sample(c(0,1), replace=TRUE, size=n() )
          ) %>% 
    mutate( error = runif01_a*e1 + runif01_b*e2 + e3 ) %>% 
    pull( error )
}

# n <- 5000
# x1 <- sample( -50:150, size=n, replace = TRUE)
# x2 <- rbeta( n=n, shape1 = 12, shape2 = 4)*100
# X_dm <- cbind(1,x1,x2) 

generate_sample <- function(n){ 
  
  x1 <- sample( -50:150, size=n, replace = TRUE)
  x2 <- rbeta( n=n, shape1 = 12, shape2 = 4)*100
  e <-generate_errors(n=n)
  y <- x1^2 - x1*x2 + 9*x2 + e - 0.3*e^2 
  return( data_frame(x1,x2,y) )
}

simdata <- 
  generate_sample( n=1000 )

```

So, a randomly drawn dataset from this population/true data-generating process looks like this:

```{r example1_bac}

simdata <- generate_sample( n=1000 )

plot_ly( data = simdata,
         x = ~x1,
         y = ~x2,
         z = ~y 
        ) %>% 
  add_markers( size=2 )

```

Here are a few more examples of sample datasets randomly drawn from this true data-generating process $Pr(X,Y)$, each time with a linear-model fit to the sample data shown:

```{r more_examples_of_badfits}
simdata <- generate_sample( n=1000 )
fit_lm <- lm( data=simdata, y~x1+x2)

plot_ly( data = simdata,
         x = ~x1,
         y = ~x2,
         z = ~y 
        ) %>% 
  add_markers( size=2 ) %>% 
  add_trace( data = simdata, 
             x = ~x1,
             y = ~x2,
             z = ~fitted.values(fit_lm),
             type = "mesh3d",
             opacity = 0.2
            )

simdata <- generate_sample( n=1000 )
fit_lm <- lm( data=simdata, y~x1+x2)

plot_ly( data = simdata,
         x = ~x1,
         y = ~x2,
         z = ~y 
        ) %>% 
  add_markers( size=2 ) %>% 
  add_trace( data = simdata, 
             x = ~x1,
             y = ~x2,
             z = ~fitted.values(fit_lm),
             type = "mesh3d",
             opacity = 0.2
            )

simdata <- generate_sample( n=1000 )
fit_lm <- lm( data=simdata, y~x1+x2)

plot_ly( data = simdata,
         x = ~x1,
         y = ~x2,
         z = ~y 
        ) %>% 
  add_markers( size=2 ) %>% 
  add_trace( data = simdata, 
             x = ~x1,
             y = ~x2,
             z = ~fitted.values(fit_lm),
             type = "mesh3d",
             opacity = 0.2
            )

```

We can get an approximation to the true function $f(X)=E[Y|X]$ via simulation by generating a very large dataset from the true data distribution and using sample statistics from it:

```{r approximate_eygx}
# generate a sample of size 1 million:
simdata <- generate_sample( n=1e6L ) 

approx_expected_Y_given_X <- 
    simdata %>% 
        # round each X variable to the nearest 10:
        mutate( x1 = round(x1,-1), x2=round(x2,-1) ) %>% 
        
        # get the average (mean) value of Y for each (X1,X2) combation:  
        group_by( x1, x2 ) %>% 
        summarise( expected_y_given_x = mean(y) ) %>%
        ungroup()

# get a few random points to plot around the function E[Y|X]:
simdata_extra <- generate_sample( n=500 )

plot_ly( data = simdata_extra,
         x = ~x1,
         y = ~x2,
         z = ~y
       ) %>% 
  add_markers( size=1 ) %>% 
  add_trace( data = approx_expected_Y_given_X, 
             x = ~x1,
             y = ~x2,
             z = ~expected_y_given_x,
             type = "mesh3d",
             opacity = 0.2
            )

```

Clearly, the function $f(X)=E[Y|X]$ is not linear 

Here is the best fitting $f(X)=X\beta$ superimposed over the true function $f(X)=E[Y|X]$:

```{r show_lack_of_fit}

best_fitting_lm <- lm( y~x1+x2, data=simdata)
X_grid <- expand.grid( x1 = seq( min(simdata$x1), max(simdata$x1), 1 ),
                       x2 = seq( min(simdata$x2), max(simdata$x2), 1 )
                      )
predict_X_grid <- predict( best_fitting_lm, newdata=X_grid )

lm_to_plot <- data_frame( x1 = X_grid$x1,
                          x2 = X_grid$x2,
                          y_hat_lm = predict_X_grid  
                        )

plot_ly( data = simdata_extra,
         x = ~x1,
         y = ~x2,
         z = ~y
       ) %>% 
  add_markers( size=1 ) %>% 
  add_trace( data = approx_expected_Y_given_X, 
             x = ~x1,
             y = ~x2,
             z = ~expected_y_given_x,
             type = "mesh3d",
             opacity = 0.2
            ) %>% 
  add_trace( data = lm_to_plot, 
             x = ~x1,
             y = ~x2,
             z = ~y_hat_lm,
             type = "mesh3d",
             opacity = 0.2
            )
```



Now suppose that we have assumed that the regression function $f(X)=E\Big[Y|X\Big]$ of the true data generating process is linear and additive (has form $f(X)=E\Big[Y|X\Big]=X\beta$) but it is in fact neither linear nor additive (we are still using that very non-normal error distribution too):

```{r not_actually_additive}
n <- 1e6L
x1 <- sample( -50:150, size=n, replace = TRUE)
x2 <- rbeta( n=n, shape1 = 12, shape2 = 4)*100
X_dm <- cbind(1,x1,x2) 

generate_sample <- function(n){ 
  e <-generate_errors(n=n)
  y <- x1^2 - x1*x2 + 9*x2 + e - 0.3*e^2 
  return( data_frame(x1,x2,y) )
}

simdata <- 
  generate_sample( n=n )

# simdata %>% 
#     group_by( x1, x2 ) %>% 
#     summarise( expected_y_given_x = mean(y) )

```
$$\beta = \Bigg(E\Big[XX^T\Big]\Bigg)^{-1} E\Big[XY\Big] = E\Bigg[\begin{pmatrix}1\\X_1\\X_2\end{pmatrix}\begin{pmatrix}1&X_1&X_2\end{pmatrix} \Bigg]^{-1} \space E\Bigg[\begin{pmatrix}1\\X_1\\X_2\end{pmatrix} Y \Bigg]$$ is:

```{r theoretical_beta_non_linear_dbn}

true_beta <- 
  coef( lm( y ~ x1+x2, data=simdata ) )
  
names( true_beta ) <- paste( "beta_", 0:2, sep="" )
true_beta
```

```{r plotLeeMorgan}

simdata_to_plot <- 
  simdata %>% 
    mutate( x1 = round(x1,-1), x2=round(x2,-1) ) %>% 
    group_by( x1, x2 ) %>% 
    summarise( expected_y_given_x = mean(y) ) %>%
    ungroup()
  
simdata_to_plot$fitted_plane <- true_beta["beta_0"] +
                                true_beta["beta_1"]*simdata_to_plot$x1 + 
                                true_beta["beta_2"]*simdata_to_plot$x2
  
plot_ly( data = simdata_to_plot,
         x = ~x1,
         y = ~x2,
         z = ~expected_y_given_x
  ) %>%
  add_trace( type="mesh3d" ) %>%
add_trace( data = simdata_to_plot,
           x = ~x1,
           y = ~x2,
           z = ~fitted_plane,
           type = "mesh3d",
           alphahull = 0,
           opacity = 0.2
)

```



```{r 1000_sims_fX_assumption_is_wrong}
generate_sample <- function(n){ 
  e <-generate_errors(n=n)
  y <- x1^2 - x1*x2 + 9*x2 + e - 0.3*e^2 
  return( data_frame(x1,x2,y) )
}

n <- 200
store_coefs <- 
  lapply( 1:1000,
          function(x) { randomsample <- generate_sample(n=n)
                        X <- cbind(1,randomsample$x1, randomsample$x2)
                        beta_hat <- solve( t(X)%*%X ) %*% t(X) %*% randomsample$y 
                            
                           #coef( lm(y ~ x1 + x2, data=randomsample) )
                        return( data_frame( b0=beta_hat[1],
                                            b1=beta_hat[2],
                                            b2=beta_hat[3]
                                          )
                              )
          }
  ) %>% 
  do.call( bind_rows, .)

hist( store_coefs$b0, breaks=30 )
abline( v=mean(store_coefs$b0), col=2 )

hist( store_coefs$b1, breaks=30 )
abline( v=mean(store_coefs$b1), col=2 )

hist( store_coefs$b2, breaks=30 )
abline( v=mean(store_coefs$b2), col=2 )

colMeans(store_coefs)
```
```{r quick_validation_plotz}

quick_sample <-  
  
validation_plots
```



If we do not want to accept these assumptions, we can still use a linear model to estimate the $\hat\beta$ coefficients, but we will have to find other ways to test the significance of our coefficients. 

Suppose that we have a regression function $$f(X)$$ which is linear, but with very non-normal errors:

$$f(X) \quad = \quad E\Big[Y|X\Big] \quad = \quad X\beta + \epsilon \\ \epsilon \quad \sim \quad exp(0.1)$$

Suppose that our observed $X$ data is:
```{r genX}
n <- 1000
x1 <- sample( -100:100, size=n, replace = TRUE)
x2 <- rbeta( n=n, shape1 = 2, shape2 = 12)*100
x3 <- rnorm( n=n, mean=30, sd=30)
X_dm <- cbind(1,x1,x2,x3) 
data_frame( x1,x2,x3)
```

The distribution of our errors looks like this:

```{r error_dbn}
hist( rexp(1e4L,rate=0.1), breaks=100 )
abline( v = 1/0.1, col=2 )
```

We can see that the errors do not have a mean of 0 (they have a mean of 10) and are definitely not normally distributed.

Then, our data-generating process/distribution (conditioned on X) is:

```{r dbn_given_X}
generate_sample <- function(n=1000){ 
  e <- rexp(n=n, rate=0.1)
  y <- 12 + 2*x1 - 3*x2 - x3 + e  
  return( data_frame(x1,x2,x3,y) )
}
```
Now, we can simulate datasets (each of size n=1000) from this distribution, calculate $\hat\beta$ for each dataset, and see the distribution of the $\hat\beta$s

```{r 1000_sims}
n <- 1000
store_coefs <- 
  lapply( 1:1e4L,
          function(x) { randomsample <- generate_sample(n=n)
                        beta_hat <- solve(t(X_dm)%*%X_dm) %*% t(X_dm) %*% randomsample$y 
                        return( data_frame( b0=beta_hat[1],
                                            b1=beta_hat[2],
                                            b2=beta_hat[3],
                                            b3=beta_hat[4]
                                          )
                              )
          }
  ) %>% 
  do.call( bind_rows, .)

hist( store_coefs$b0, breaks=30 )
abline( v=mean(store_coefs$b0), col=2 )

hist( store_coefs$b1, breaks=30 )
abline( v=mean(store_coefs$b1), col=2 )

hist( store_coefs$b2, breaks=30 )
abline( v=mean(store_coefs$b2), col=2 )

hist( store_coefs$b3, breaks=30 )
abline( v=mean(store_coefs$b3), col=2 )

colMeans(store_coefs)
```

Here is the same $X$ data with the following extreme error distribution:

```{r mad_errors}
generate_errors <- function(n=1000){
  
    data_frame( e1 = rnorm(n, -40, 5),
                e2 = 0, #1/rnorm(n, 40, 5),
                e3 = sample( c(-100, -50, -20, 40), size=n, replace=TRUE ) 
              ) %>% 
    mutate( runif01_a = sample(c(0,1), replace=TRUE, size=n() ),
            runif01_b = sample(c(0,1), replace=TRUE, size=n() )
          ) %>% 
    mutate( error = runif01_a*e1 + runif01_b*e2 + e3 ) %>% 
    pull( error )
}

hist( generate_errors(n=5000), breaks=50 )
```

We can see that our coefficient estimates $\hat\beta$ are still symetrically-distributed and each is centred on the true $\beta$ value that it is estimating (has mean $\beta$), with the exception of the intercept $\hat\beta_0$.  

Now, we can simulate datasets (each of size n=1000) using an even more extreme error distribution, calculate $\hat\beta$ for each dataset, and see the distribution of the $\hat\beta$s

```{r 1000_sims_insane_errors}
generate_sample <- function(n=1000){ 
  e <-generate_errors(n=n)
  y <- 12 + 2*x1 - 3*x2 - x3 + e  
  return( data_frame(x1,x2,x3,y) )
}

n <- 1000
store_coefs <- 
  lapply( 1:1e4L,
          function(x) { randomsample <- generate_sample(n=n)
                        beta_hat <- solve(t(X_dm)%*%X_dm) %*% t(X_dm) %*% randomsample$y 
                        return( data_frame( b0=beta_hat[1],
                                            b1=beta_hat[2],
                                            b2=beta_hat[3],
                                            b3=beta_hat[4]
                                          )
                              )
          }
  ) %>% 
  do.call( bind_rows, .)

hist( store_coefs$b0, breaks=30 )
abline( v=mean(store_coefs$b0), col=2 )

hist( store_coefs$b1, breaks=30 )
abline( v=mean(store_coefs$b1), col=2 )

hist( store_coefs$b2, breaks=30 )
abline( v=mean(store_coefs$b2), col=2 )

hist( store_coefs$b3, breaks=30 )
abline( v=mean(store_coefs$b3), col=2 )

colMeans(store_coefs)
```
The parameter distributions still look very normal, and the distribution means are at the true $\beta$ values.  




BLABLABLA


In order to make inferences (construct confidence intervals etc.) for our coefficients $\hat\beta$ and model predictions, we derive some sampling properties and sampling distribution of the coefficients $\hat\beta$:

$$\begin{array}{lcl} E\Big[\hat\beta | X \Big] &=& E\Big[(X^TX)^{-1}X^Ty \space | X \Big] \\
&=& (X^TX)^{-1} X^T \space E\Big[y|X\Big] \\
&=& (X^TX)^{-1}X^T \space E\Big[\big(X\beta + \epsilon \big)|X \Big] \hspace{5mm} \text{explain why we can do this} \\
&=& (X^TX)^{-1}X^T \space \big(X\beta + E\Big[\epsilon | X \Big] \big) \hspace{5mm} \text{linearity of expectation (even for dependent random variables!)} \\ 
&=& \beta + 0 \hspace{5mm} \text{assuming that errors have mean 0} \\
&=& \beta \\
\end{array}$$



Additionally, the assumption of normal independent errors can be used to create confidence intervals for the coefficients $\beta_0, \space \beta_2, \beta_3, ... $, to create confidence intervals for the predictions made using the model, and in order to test whether coefficients are significantly different from zero (i.e. to create hypothesis tests that assess the significance of each coefficient). 

If the distribution of the residuals is significantly non-normal or if the errors are dependent, then this will give misleading confidence intervals and hypothesis tests (leading to false conclusions) and the coefficient estimates may be biased.   

```{r perfect_data}
simdata <- 
  data_frame( x1 = rchisq(n=100, df=8),
              x2 = rbeta(n=100, shape1=5, shape2=2, ncp=4)*20,
              errors = rnorm(n=100, 0, 10) 
            ) %>% 
    mutate( y = 5*x1 - 6*x2 + errors )

pairs(simdata[,c("y","x1","x2")], pch=16, cex=0.5)

fit_lm <- lm( y~x1+x2, data=simdata)

simdata$fitted_val <- fitted.values(fit_lm) 

plot_ly( data = simdata, 
         x = ~x1, 
         y = ~x2, 
         z = ~y
  ) %>% 
  add_markers(size=1) %>% 
  add_trace( data = simdata, 
             x = ~x1, 
             y = ~x2, 
             z = ~fitted_val,
             type = "mesh3d",
             alphahull = 0,
             opacity = 0.2
  )

```

The validation plots for this model:

```{r validation_plots_heyo}
validation_plots( model_object = fit_lm, original_data = simdata )
```



# NOT A REQUIRED ASSUMPTION: y must be normally distributed
Yay I am happy to debunk this:) 

# Validating the Linearity and Additivity Assumptions
http://people.duke.edu/~rnau/testing.htm

TO JOE: ADD A WHOLE OF SINGLE VIOLATIONS AND SEE WHAT THEY DO TO EACH PLOT!

1 Correlation plots between independent variables

2 Fitted vs. Residuals

3 hm


```{r hm_check_for_non_linearity}
# plot( simdata$y ~ fitted.values(fit_lm), 
#       main = "Sample Data Observed Values vs. Linear Model Fitted Values",
#       ylab = "observed",
#       xlab = "fitted"  
#     )

plot( fitted.values(fit_lm) ~ residuals(fit_lm), 
      main = "Fitted Values vs. Residuals",
      ylab = "Fitted Value",
      xlab = "residuals"  
    )
```

```{r fitted_vs_residuals_examples}

##### true linear relationships #####
simdata <- 
  data_frame( x1 = runif(n=100, -100, 100),
              x2 = runif(n=100, -100, 100),
              x3 = runif(n=100, -100, 100),
              errors = rnorm(n=100, 0, 50) 
            ) %>% 
    mutate( y = -5*x1 + 7*x2 -6*x3 + errors )
pairs(simdata[,c("y","x1","x2","x3")], pch=16, cex=0.5)
fit_lm <- lm( y~x1+x2+x3, data=simdata)
plot( fitted.values(fit_lm) ~ residuals(fit_lm), 
      main = "LINEAR RELATIONSHIPS: Fitted Values vs. Residuals",
      ylab = "Fitted Value",
      xlab = "residuals"  
    )

###### non-linearity example 1 ##### 
simdata <- 
  data_frame( x1 = runif(n=100, -100, 100),
              x2 = runif(n=100, -100, 100),
              x3 = runif(n=100, -100, 100),
              errors = rnorm(n=100, 0, 50) 
            ) %>% 
    mutate( y = 5*x1 - x2^3/200 +6*x3 + errors )
pairs(simdata[,c("y","x1","x2","x3")], pch=16, cex=0.5)
fit_lm <- lm( y~x1+x2+x3, data=simdata)
plot( fitted.values(fit_lm) ~ residuals(fit_lm), 
      main = "LINEAR RELATIONSHIPS: Fitted Values vs. Residuals",
      ylab = "Fitted Value",
      xlab = "residuals"  
    )

###### non-linearity example 1 ##### 
simdata <- 
  data_frame( x1 = runif(n=100, -100, 100),
              x2 = runif(n=100, -100, 100),
              x3 = runif(n=100, -100, 100),
              errors = rnorm(n=100, 0, 50) 
            ) %>% 
    mutate( y = 5*x1 - x2^3/200 +6*x3 + errors )
pairs(simdata[,c("y","x1","x2","x3")], pch=16, cex=0.5)
fit_lm <- lm( y~x1+x2+x3, data=simdata)
plot( fitted.values(fit_lm) ~ residuals(fit_lm), 
      main = "LINEAR RELATIONSHIPS: Fitted Values vs. Residuals",
      ylab = "Fitted Value",
      xlab = "residuals"  
    )

```

# Correlated Predictors

# No Outliers

# The errors are independent
There must be no autocorrelation or dependence amongst any of errors. 

Suppose that we have autocorrelation in our errors: 

```{r autocorrelated_errors}
simdata <- 
  data_frame( x1 = runif(n=1000, -100, 100),
              x2 = runif(n=1000, -100, 100),
              errors = sort( rnorm(n=1000, 0, 50) ) 
            ) %>% 
    mutate( y = 40*x1 - x2^2 + errors )

plot( simdata$errors, pch=16, cex=0.5 )
hist( simdata$errors, breaks=20 )

```

# The errors have mean 0

# The errors have constant variance (homoscedasticity of errors)
Specifically, the errors must not become more variable at specific values of reponse $Y$ nor any of the predictors $X$.

# The predictors must be uncorrelated



