---
title: "Partial Correlation"
output:
  html_document:
    df_print: paged
  html_notebook:
    code_folding: hide
---
```{r setup}
library(dplyr)
```
Partial correlation measures the degree of association between 2 continuous variables, with the effect of a set of controlling variables removed.  

$$\space$$

The partial correlation between continuous variable $X$ and continuous variable $Y$, controlling for variable(s) $\mathbf{Z}=\{Z_1,Z_2,Z_3...\}$ (categorical or continuous) is measured by $\text{Cor}\Big[e_X,e_Y\Big]$ - the correlation between $e_X$ and $e_Y$ - where 

$$ e_X \quad \text{is the residuals from the linear regression } \quad \hat{X} = \beta_0+\beta_1Z_1+\beta_2Z_2+...+e_X $$

and

$$e_Y \quad \text{is the residuals from the linear regression} \quad \hat{Y} = \beta_0+\beta_1Z_1+\beta_2Z_2+...+e_Y$$

$$\space$$

$e_X$ is the part of $X$ that is not (linearly) explained by the variables $Z_1,Z_2,Z_3...$ and $e_Y$ is the part of $Y$ that can not be (linearly) explained by the variables $Z_1,Z_2,Z_3...$.   

$$\space$$

Considering the following dataset on exam anxiety:

```{r doit}
exam_anxiety_dat <- 
  data_frame(
    hours_spent_revising = c(4,	11,	27,	53,	4,	22,	16,	21,	25,	18,	18,	16,	13,	18,	98,	1,	14,	29,	4,	23,	14,	12,	22,	84,	23,	26,	24,	72,	37,	10,	3,	36,	43,	19,	12,	9,	72,	10,	12,	30,	15,	8,	34,	22,	21,	27,	6,	18,	8,	19,	0,	52,	38,	19,	23,	11,	27,	17,	13,	42,	4,	8,	6,	11,	7,	15,	4,	28,	22,	29,	2,	16,	59,	10,	13,	8,	5,	2,	38,	4,	10,	6,	68,	8,	1,	14,	42,	13,	1,	3,	5,	12,	19,	2,	19,	11,	15,	23,	13,	14,	1,	9,	20 ),
    exam_score = 
      c(40,	65,	80,	80,	40,	70,	20,	55,	50,	40,	45,	85,	70,	50,	95,	70,	95,	95,	50,	60,	80,	75,	85,	90,	30,	60,	75,	75,	27,	20,	75,	90,	60,	30,	80,	10,	85,	7,	5,	85,	20,	45,	60,	70,	50,	25,	50,	40,	80,	50,	35,	80,	50,	49,	75,	25,	65,	80,	50,	70,	40,	80,	10,	20,	40,	40,	70,	52,	50,	60,	80,	60,	65,	15,	85,	20,	80,	100,	100,	80,	10,	70,	100,	70,	70,	65,	75,	85,	30,	5,	10,	90,	70,	20,	85,	35,	30,	70,	55,	75,	2,	40,	50	
      ),
    anxiety_before_exam = 
      c(86.298,	88.716,	70.178,	61.312,	89.522,	60.506,	81.462,	75.82,	69.372,	82.268,	79.044,	80.656,	70.178,	75.014,	34.714,	95.164,	75.82,	79.044,	91.134,	64.536,	80.656,	77.432,	65.342,	0.0560000000000116,71.79,	81.462,	63.73,	27.46,	73.402,	89.522,	89.522,	75.014,	43.58,	82.268,	79.044,	79.044,	37.132,	81.462,	83.074,	50.834,	82.268,	78.238,	72.596,	74.208,	75.82,	70.984,	97.582,	67.76,	75.014,	73.402,	93.552,	58.894,	53.252,	84.686,	89.522,	71.79,	82.268,	69.372,	62.118,	68.566,	93.552,	84.686,	82.268,	81.462,	82.268,	91.134,	91.94,	86.298,	72.596,	63.73,	63.73,	71.79,	57.282,	84.686,	84.686,	77.432,	82.268,	10,50.834,	87.91,	83.88,	84.686,	20.206,	87.104,	83.88,	67.76,	95.97,	62.118,	84.686,	92.746,	84.686,	83.074,	73.402,	87.91,	71.79,	86.298,	84.686,	75.82,	70.984,	78.238,	82.268,	79.044,	91.134	
      ),
    gender = 
c("Male","Female","Male","Male","Male","Female","Female","Female","Female","Female","Male","Male","Male","Female","Male","Male","Male","Female","Female","Male","Male","Male","Female","Female","Female","Female","Male","Female","Female","Male","Female","Female","Male","Male","Male","Female","Male","Male","Female","Male","Male","Female","Male","Female","Female","Male","Male","Male","Male","Female","Female","Female","Female","Male","Female","Female","Male","Male","Male","Female","Male","Female","Male","Female","Male","Male","Female","Female","Male","Female","Male","Female","Male","Female","Male","Female","Female","Male","Female","Male","Male","Female","Female","Male","Female","Male","Female","Female","Male","Male","Female","Female","Male","Female","Male","Male","Female","Male","Female","Female","Male","Male","Female"
)      
  )
exam_anxiety_dat
```

Pearson's correlation between **hours spent revising** and **exam anxiety** is `r cor(exam_anxiety_dat$hours_spent_revising,exam_anxiety_dat$anxiety_before_exam)`

```{r straight_up_correlation}
cor( exam_anxiety_dat$hours_spent_revising, exam_anxiety_dat$anxiety_before_exam)
plot( x = exam_anxiety_dat$hours_spent_revising,
      y = exam_anxiety_dat$anxiety_before_exam,
      xlab = "hours spent revising",
      ylab = "anxiety before exam"
    )
abline( lm(anxiety_before_exam~hours_spent_revising, data=exam_anxiety_dat), col=2 )
```

However, both **anxiety before exam** and **hours spent revising** are linearly associated with **exam score**:

```{r quickplots, fig.width=10}
par(mfrow=c(1,2), pty="s")
plot( x = exam_anxiety_dat$exam_score, y = exam_anxiety_dat$anxiety_before_exam,
      xlab = "exam score", ylab = "anxiety before exam",
      main = "students doing well in the exam \n had less anxiety before the exam")
abline( lm(anxiety_before_exam~exam_score, data=exam_anxiety_dat), col=3 )

plot( x = exam_anxiety_dat$exam_score, y = exam_anxiety_dat$hours_spent_revising,
      xlab = "exam score", ylab = "hours spent revising",
      main = "students who revised more did \n better in the exam")
abline( lm(hours_spent_revising~exam_score, data=exam_anxiety_dat), col=4 )
```

The partial correlation between **hours spent revising** and **exam anxiety**, controlling for the effect of **exam score** is `r cor( 
  residuals( lm(hours_spent_revising~exam_score, data=exam_anxiety_dat) )
  ,
  residuals( lm(anxiety_before_exam~exam_score, data=exam_anxiety_dat) )
)`, indicating that the negative relationship between these 2 variables is slightly less strong than previously measured (part of the apparent negative relationship between arose from each's relationship to **exam score**) 

```{r controlit, fig.height=8}
cor( 
  residuals( lm(hours_spent_revising~exam_score, data=exam_anxiety_dat) )
  ,
  residuals( lm(anxiety_before_exam~exam_score, data=exam_anxiety_dat) )
)

plot( x = residuals( lm(hours_spent_revising~exam_score, data=exam_anxiety_dat) ),
      y = residuals( lm(anxiety_before_exam~exam_score, data=exam_anxiety_dat) ),
      xlab = bquote("residuals from linear model: "~hat(hours_spent_revising)~" = "~beta[0]~"+"~beta[1]*"exam_score"~"+"~e[hours_spent_revising]),
      ylab = bquote("residuals from linear model: "~hat(anxiety_before_exam)~" = "~beta[0]~"+"~beta[1]*"exam_score"~"+"~e[anxiety_before_exam]),
      main="Partial correlation between 'Hours Spent Revising' and 'Exam Anxiety' \n controlling for 'Exam Score'"
    )
abline( lm(y~x, data=data_frame(x=residuals( lm(hours_spent_revising~exam_score, data=exam_anxiety_dat) ),
                                y=residuals( lm(anxiety_before_exam~exam_score, data=exam_anxiety_dat) )
                              )
          ),
        col = 5
      )
```

$$\space$$

Similarly, we could calculate the partial correlation between **hours spent revising** and **anxiety before exam**, controlling for the effect of **exam score** and **gender**:

```{r dodododo}
cor( 
  residuals( lm(hours_spent_revising~exam_score + factor(gender), data=exam_anxiety_dat) )
  ,
  residuals( lm(anxiety_before_exam~exam_score + factor(gender), data=exam_anxiety_dat) )
)
```