---
title: "Evaluating Uplift Models: Qini Curve & Qini Coefficient"
author: "Joseph Bolton"
output: html_notebook
---

```{r setup}
library(tidyverse)
#install.packages("tools4uplift")
library(tools4uplift)
```

https://humboldt-wi.github.io/blog/research/theses/uplift_modeling_blogpost/

https://dms.umontreal.ca/~murua/research/UpliftRegression.pdf

Generate an experiment with 400 people: 200 treated and 200 in the control group: 
```{r}
simdata <- tibble(        id = 1:400,
                   treatment = c( rep(1,200), rep(0,200) )   
                 )
```

generate a single attribute $X_1$ for each person:
```{r}
simdata$X1 <- runif(   n = nrow(simdata), 
                     min = 0,
                     max = 100  
                   ) %>% 
              round(., digits=2)
ggplot( data = simdata,
        aes(      x = id,
                  y = X1, 
             colour = as.factor(treatment)
           )
       ) +
  geom_point() 
```

Simulate some true uplift signal:

```{r}
sigmoid_ftn <- function(x) exp(x) / (exp(x)+1)

tibble( x = 0:100 ) %>% 
  mutate( y_prob = 
     map_dbl( .x = x,
              .f = function(x1) 0.8 * ( -0.25 + sigmoid_ftn( -0.0005*(x1-30)^2 ) )                              )
        ) %>% 
  ggplot( data = .,
          aes( x = x,
               y = y_prob,
               group = 1
             )
        ) +
  geom_line() +
  labs(    title = "true uplift signal",
        subtitle = "(additional probability added by treatment)",
               x = bquote( x[1] ),
               y = "additional probability added"
      )

set.seed(6969)

simdata <- 
  simdata %>% 
    mutate( base_y_prob = 0.3 ) %>%          # base probability of responding 
    mutate( added_treat_prob = 
              map_dbl( .x = X1,
              .f = function(x1) 0.8 * ( -0.25 + sigmoid_ftn( -0.0005*(x1-30)^2 ) )                                      )
          ) %>% 
    mutate( added_treat_prob = added_treat_prob*treatment ) %>%    
    mutate( y_prob_final = base_y_prob + added_treat_prob ) %>% 
    mutate( y = map_dbl( .x = y_prob_final,
                         .f = function(prob)
                                            { 
                                              sample(           0:1, 
                                                         size = 1,
                                                         prob = c(1-prob, prob)
                                                    )
                                            } 
                      )
          )

simdata
```

```{r}
ggplot( data = simdata,
        aes(      x = X1,
                  y = y, 
             colour = as.factor(treatment)
           )
       ) +
  geom_jitter( height = 0.3 ) 
```






```{r}
data("SimUplift")
SimUplift
```

code that follows is adapted from https://dms.umontreal.ca/~murua/research/UpliftRegression.pdf

```{r}
set.seed(1988)

split_data <- 
  tools4uplift::SplitUplift(  data = SimUplift,
                                 p = 0.7, 
                             group = c("treat", "visit")
                            )

train <- split_data[[1]]
valid <- split_data[[2]]
```

```{r}
#Hmisc::describe(train)
```

fit 2-model to training data:
```{r}
base.tm <- 
 tools4uplift::DualUplift(       data = train,
                                treat = "treat",
                              outcome = "y",
                           predictors = colnames( train[, 3:22] )
                         )

# baseline model for control group:
base.tm[[1]]

# baseline model for treatment group:
base.tm[[2]]
```


Predict the uplift on the **validation** set using the 2-model:
(also return Qini Curve and Uplift Barplot)
```{r}
base.tm.valid <- 
  tools4uplift::DualPredict( data = valid,
                             treat = "treat",
                             outcome = "y",
                             model = base.tm,
                             nb.group = 5,     # number of groups 
                             plotit = FALSE   # don't return QiniCurve and QiniBarPlot
                           )
```

```{r}
# the validation set, with 'uplift_prediction' (for each individual) column added:
base.tm.valid[[1]]

# Qini value:
base.tm.valid[[2]]
```

get other uplift model assessment metrics:

metric        | description
--------------|--------------------------
cum_per       | $\phi$ = cumulative % of customers (ordered by predicted_uplift descending)
T_Y1          | Number of customers with y==1 and treat==1
T_n           | Number of customers with treat==1
C_Y1          | Number of customers with y==1 and treat==0
C_n           | Number of customers with treat==0
incremental_Y1| $h(\phi)=\text{T_Y1}-\frac{\text{T_n}}{\text{C_n}}\cdot\text{C_Y1}$
inc_uplift    | $g(\phi)=\frac{h(\phi)}{\sum_i T_i}$
uplift        | $\text{E[y|treat==1]-E[y|treat==0]}$ in this group of customers
Qini Curve    | Plots y=cum_per against x=inc_uplift

```{r}
base.tm.perf <- 
  tools4uplift::QiniTable(       data = base.tm.valid[[1]],
                                treat = "treat",
                              outcome = "y",   
                           prediction = "uplift_prediction",
                             nb.group = 5
                         )

base.tm.perf
tools4uplift::QiniCurve( base.tm.perf )
points(x=20, y=8.595229, pch=16)
tools4uplift::QiniBarPlot( base.tm.perf )
tools4uplift::QiniArea( base.tm.perf )

```

```{r}
valid_treat_ppl <- base.tm.valid[[1]] %>% filter(treat==1) %>% select(y, uplift_prediction)
valid_control_ppl <- base.tm.valid[[1]] %>% filter(treat==0) %>% select(y, uplift_prediction)

treat_0.2_grp <- valid_treat_ppl %>% top_n( 22, wt=uplift_prediction ) 
control_0.2_grp <- valid_control_ppl %>% top_n( 38, wt=uplift_prediction ) 

table( treat_0.2_grp$y )
table( control_0.2_grp$y )

# uplift:
mean(treat_0.2_grp$y)-mean(control_0.2_grp$y)

# Qini Plot:
par(mfrow=c(1,2))
tools4uplift::QiniCurve( base.tm.perf )
plot( x = c(0, base.tm.perf$cum_per),
      y = c(0, base.tm.perf$inc_uplift),
      type = "l"
    )
lines( x = c(0, 1),
       y = c(0, base.tm.perf$inc_uplift[5]) 
     )

# incremental uplift h(phi) in top 20%:
sum(treat_0.2_grp$y) - (22/38) * sum(control_0.2_grp$y)
base.tm.perf_0.2 <- base.tm.perf[1,]
h_phi <- 
  base.tm.perf_0.2$T_Y1 - 
  (base.tm.perf_0.2$T_n/base.tm.perf_0.2$C_n) * base.tm.perf_0.2$C_Y1
h_phi

# relative incremental uplift g(phi):
h_phi / 139
```

