---
title: "Rawtext product recommendation engine"
output:
  html_document:
    df_print: paged
---

```{r setup}
library(tidyverse)
library(reticulate)
reticulate::use_python('c:/users/joe/miniconda3')

library(plotly)
library(knitr)
library(reshape2)
```

```{python}
import pandas as pd
# nltk.download('stopwords')
from nltk.corpus import stopwords
from sklearn.metrics.pairwise import cosine_similarity, linear_kernel
from sklearn.feature_extraction.text import TfidfVectorizer, CountVectorizer
from sklearn import preprocessing
import numpy as np
from scipy import sparse
from scipy.sparse import csr_matrix
import random
```

# Content-based filtering recommender in R (theory)

still to do:

* does normalising the item vectors (rows) make any difference to cosine distance?

thoughts:

* need to down-weight product attributes that are very common 

* Do I need to deal with products that have more attributes?

* note that this method will very likely want to give customers products that they've just bought before.

```{r}
product_data <- 
  tibble( product_id = paste("product", 1:4, sep="_"), 
          product_desc = c( "red, green, green, green",
                           " red, blue, green",
                           " blue, blue, blue",
                           "red, red, blue"
                           )
        ) %>% 
  mutate(   tf_red = c( 1, 1, 0, 2),
           tf_blue = c( 0, 1, 2, 1 ),
          tf_green = c( 3, 1, 0, 0 ),
        ) %>% 
  mutate( red_idf = log( 4 / 3 ),
          blue_idf = log( 4 / 3 ),
          green_idf = log( 4 / 2 )
        ) %>% 
  mutate( red_tf_idf = tf_red * red_idf,
          blue_tf_idf = tf_blue * blue_idf,
          green_tf_idf = tf_green * green_idf
        )

product_data %>% kable()

```

plot the products in product space:
```{r fig.width=10, fig.height=10}

product1_vector <- product_data[1,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist()

product2_vector <- product_data[2,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist()

product3_vector <- product_data[3,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist()

product4_vector <- product_data[4,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist()

plot_ly( data = product_data,
            x = ~red_tf_idf,
            y = ~blue_tf_idf,
            z = ~green_tf_idf
       ) %>% 
  add_markers() %>% 
  add_trace( type = "scatter3d",
             x = c( 0, product1_vector[1] ), 
             y = c( 0, product1_vector[2] ),
             z = c( 0, product1_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            ) %>% 
    add_trace( type = "scatter3d",
             x = c( 0, product2_vector[1] ), 
             y = c( 0, product2_vector[2] ),
             z = c( 0, product2_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            ) %>% 
    add_trace( type = "scatter3d",
             x = c( 0, product3_vector[1] ), 
             y = c( 0, product3_vector[2] ),
             z = c( 0, product3_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            )  %>% 
    add_trace( type = "scatter3d",
               x = c( 0, product4_vector[1] ), 
               y = c( 0, product4_vector[2] ),
               z = c( 0, product4_vector[3] ),
               mode = "lines",
               line = list(width = 4, color="blue")
          )  %>% 
  layout(scene = list( xaxis = list(title = 'red'),
                       yaxis = list(title = 'blue'),
                       zaxis = list(title = 'green')
                    )
       )
```

Here is the product table for reference:
```{r}
product_data %>% 
  select( product_id, red_tf_idf, blue_tf_idf, green_tf_idf )
```

Suppose that customer $A$ buys four of product_3 and two of product_4. We average the 5 product vectors that customer $A$ has bought, giving customer $A$ an item profile of: 

```{r}
customerA_profile <- 
( 4*product3_vector + 2*product4_vector ) / 6 

names( customerA_profile ) <- c( "red", "blue", "green" )

customerA_profile
```

We can now plot this customer in the product space:
```{r fig.width=10, fig.height=10}
plot_ly( data = product_data,
            x = ~red_tf_idf,
            y = ~blue_tf_idf,
            z = ~green_tf_idf
       ) %>% 
  add_markers() %>% 
  add_trace( type = "scatter3d",
             x = c( 0, product1_vector[1] ), 
             y = c( 0, product1_vector[2] ),
             z = c( 0, product1_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            ) %>% 
    add_trace( type = "scatter3d",
             x = c( 0, product2_vector[1] ), 
             y = c( 0, product2_vector[2] ),
             z = c( 0, product2_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            ) %>% 
    add_trace( type = "scatter3d",
             x = c( 0, product3_vector[1] ), 
             y = c( 0, product3_vector[2] ),
             z = c( 0, product3_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            )  %>% 
    add_trace( type = "scatter3d",
             x = c( 0, product4_vector[1] ), 
             y = c( 0, product4_vector[2] ),
             z = c( 0, product4_vector[3] ),
             mode = "lines",
             line = list(width = 4, color="blue")
            )  %>% 
  layout(scene = list( xaxis = list(title = 'red'),
                       yaxis = list(title = 'blue'),
                       zaxis = list(title = 'green')
                    )
       ) %>% 
  add_trace( type = "scatter3d",
             x = c( 0, customerA_profile[1] ), 
             y = c( 0, customerA_profile[2] ),
             z = c( 0, customerA_profile[3] ),
             mode = "lines",
             line = list(width = 4, color="red")
            ) %>% 
    add_trace( type = "scatter3d",
             x = c( 0, customerA_profile[1] ), 
             y = c( 0, customerA_profile[2] ),
             z = c( 0, customerA_profile[3] ),
             mode = "points"
            )
```

Here is a reminder of what customer $A$ bought:
```{r}
product_data %>% 
  select( product_id, product_desc, red_tf_idf, blue_tf_idf, green_tf_idf ) %>% 
  bind_cols( ., tibble(n_bought=c(0, 0, 4, 2) ) ) %>% 
  select( product_id, product_desc, n_bought, everything() )
```

The angle $\theta$ between any 2 product vectors indicates their similarity. $cos(\theta)$ will therefore range as follows:

$\theta$   |   $cos(\theta)$
-----------|------------------
$0^\circ$  | 1
$30^\circ$ | 0.866
$45^\circ$ | 0.707
$60^\circ$ | 0.5
$90^\circ$ | 0

The cosine of the angle between 2 vectors $\overset{\rightarrow{}}{u}$ and $\overset{\rightarrow{}}{u}$ can be calculated using

$$cos\Big(\theta_{uv}\Big) \quad=\quad \displaystyle\frac{u \cdot v}{|u| \space |v|}$$

Here are the angles between the product vectors in our example:

```{r}
cosine_distmat <- matrix( nrow = 4, ncol = 4 )

for( i in 1:4 ){
  for( j in 1:4 ){
    
    product_i <- product_data[i,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist() 
    product_j <- product_data[j,] %>% 
                      select(red_tf_idf, blue_tf_idf, green_tf_idf) %>% 
                      unlist() 
    
    cos_theta <- ( product_i %*% product_j ) / 
                 ( sqrt( sum(product_i^2) ) * sqrt( sum(product_j^2) ) )
    
    cosine_distmat[i,j] <- cos_theta
  }
}

rownames(cosine_distmat) <- colnames(cosine_distmat) <- paste("product", 1:4, sep="")

cosine_distmat %>% round(2)
```

Here are the cosine distances between the customer and each product:
```{r}
cbind( product = 1:4,
       cos_dist_to_customerA = 
       c( 
          ( product1_vector %*% customerA_profile ) / 
            ( sqrt( sum(product1_vector^2) ) * sqrt( sum(customerA_profile^2) ) )
          ,
          ( product2_vector %*% customerA_profile ) / 
            ( sqrt( sum(product2_vector^2) ) * sqrt( sum(customerA_profile^2) ) )
          ,
          ( product3_vector %*% customerA_profile ) / 
            ( sqrt( sum(product3_vector^2) ) * sqrt( sum(customerA_profile^2) ) )
          ,
          ( product4_vector %*% customerA_profile ) / 
            ( sqrt( sum(product4_vector^2) ) * sqrt( sum(customerA_profile^2) ) )
       )
     )
```

So, if we needed to offer this customer a new product (choosing from either product_1 or product_2 which they haven't bought yet), we'd offer them product_2.

# Higher-dimensional example
```{r}

generate_random_prod_ftn <- function(){
  
  product_vector <- 
        tibble( formal=0, casual=0, sports=0,
                shoes=0, pants=0, shirt=0, tie=0, bag=0, socks=0,
                black=0, red=0, green=0, blue=0, white=0,
                stripes=0, bold=0, leopard_print=0, spots=0, shiny=0
              )
  
  choose_use <- sample( 1:3, size=1 )
  if( choose_use == 1 ){ product_vector$formal <- 1 } else 
  if( choose_use == 2 ){ product_vector$casual <- 1 } else
                       { product_vector$sports <- 1 }
  
  choose_garment <- sample( 1:6, size=1 )
  if( choose_garment == 1 ){ product_vector$shoes <- 1 } else 
  if( choose_garment == 2 ){ product_vector$pants <- 1 } else
  if( choose_garment == 3 ){ product_vector$shirt <- 1 } else
  if( choose_garment == 4 ){ product_vector$tie <- 1 } else
  if( choose_garment == 5 ){ product_vector$bag <- 1 } else
                           { product_vector$socks <- 1 }
  
  choose_colour <- sample( 1:5, size=1 )
  if( choose_colour == 1 ){ product_vector$black <- 1 } else 
  if( choose_colour == 2 ){ product_vector$red <- 1 } else
  if( choose_colour == 3 ){ product_vector$green <- 1 } else
  if( choose_colour == 4 ){ product_vector$blue <- 1 } else
  { product_vector$white <- 1 }
  
  choose_style <- sample( 1:5, size=1 )
  if( choose_style == 1 ){ product_vector$stripes <- 1 } else 
  if( choose_style == 2 ){ product_vector$bold <- 1 } else
  if( choose_style == 3 ){ product_vector$leopard_print <- 1 } else
  if( choose_style == 4 ){ product_vector$spots <- 1 } else
  { product_vector$shiny <- 1 }

  return( product_vector )
}

set.seed(1925)
full_product_list <- 
  bind_rows(
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn(),
    generate_random_prod_ftn()
  ) %>% 
  mutate( product_ID = row_number() 
) %>% 
  select( product_ID, everything() )
 
kable( full_product_list )
```

get item/product profiles (tf-idf) for full product list:
```{r}
product_colsums <- full_product_list %>% select(-product_ID) %>% colSums()
product_IDFs <- log( nrow(full_product_list) / product_colsums ) 

cbind( count = t( t(product_colsums) ), 
       IDF = t( t(product_IDFs) ) 
     )

product_IDFs_mat <- matrix( rep(product_IDFs, nrow(full_product_list)),
                            byrow = TRUE,
                            nrow = nrow(full_product_list)  
                          )
  
product_profiles <- full_product_list * cbind( 1, product_IDFs_mat )
```

Customer A buys the following:
```{r}
# customer A likes formal and sports stuff
# and he like shoes and socks a lot
# and he like black and white a lot
# and he likes shiny leopard print things

customerA_product_probs <- 
  full_product_list %>% 
    mutate( prob_this_product = 
              formal*10 + casual*1 + sports*10 +
              shoes*10 + pants*1 + shirt*1 + tie*1 + bag*4 + socks*10 +
              black*10 + red*2 + green*2 + blue*2 + white*10 +
              stripes*1 + bold*3 + leopard_print*10 + spots*2 + shiny*10
          ) %>% 
  pull( prob_this_product )

set.seed(6969420)
customerA_products_chosen <- 
  full_product_list[sample(1:nrow(full_product_list),prob=customerA_product_probs,size=5, replace=TRUE),]

customerA_products_chosen

customer_profile_vector <- 
  customerA_products_chosen %>% 
    select( product_ID ) %>% 
    left_join( product_profiles ) %>% 
    select( -product_ID ) %>% 
    summarise_all( mean ) %>% 
    unlist()

cosine_distances <- rep( NA, nrow(full_product_list) )
names(cosine_distances) <- paste("product", 1:nrow(full_product_list), sep="")

for( prod_i in 1:nrow(full_product_list) ){
  
  product_vector <- product_profiles[prod_i,-1] %>% unlist()
  
  cosine_distances[prod_i] <- 
                 c(t(product_vector) %*% customer_profile_vector) / 
                 ( sqrt( sum(product_vector^2) ) * sqrt( sum(customer_profile_vector^2) ) )
}
  
```

So, to recap, the customer bought:

```{r}
customerA_products_chosen
```

and his top matched products are:
( the ones he's already bought are obviously on top )

```{r}
bind_cols( full_product_list,
           tibble( cosine_dist = cosine_distances )
) %>% 
  arrange( desc(cosine_dist) ) %>% 
  kable()
```

If we give each product vector unit length (i.e. standardise the vectors to give them length 1), this does not affect the recommendations (since the angle between vectors does not change):

```{r}
standardised_customer_profile_vector <- 
  customer_profile_vector /
  sqrt( sum(customer_profile_vector^2) )

# give each row (product) euclidean length 1:
standardised_product_profiles <- 
  cbind( product_ID = product_profiles$product_ID
         ,
         product_profiles[,-1] /
         matrix( sqrt(rep(rowSums(product_profiles[,-1]^2), ncol(product_profiles[,-1]))),
                 byrow = FALSE,
                 ncol = ncol(product_profiles[,-1])
               )
       )

cosine_distances <- rep( NA, nrow(full_product_list) )
names(cosine_distances) <- paste("product", 1:nrow(full_product_list), sep="")

for( prod_i in 1:nrow(full_product_list) ){
  
  product_vector <- standardised_product_profiles[prod_i,-1] %>% unlist()
  
  cosine_distances[prod_i] <- 
                 c( t(product_vector) %*% standardised_customer_profile_vector ) / 
                 ( sqrt( sum(product_vector^2) ) * sqrt( sum(standardised_customer_profile_vector^2) ) )
}

bind_cols( full_product_list,
           tibble( cosine_dist = cosine_distances )
) %>% 
  arrange( desc(cosine_dist) ) %>% 
  kable()
```

# TF-IDF
```{r}
bind_rows( tibble( n = 10, df = 0:10 ) %>% mutate(idf = log( (1+n)/(1+df) ) +1 )
           ,
           tibble( n = 100, df = 0:100 ) %>% mutate(idf = log( (1+n)/(1+df) ) +1 )
           ,
           tibble( n = 1000, df = 0:1000 ) %>% mutate(idf = log( (1+n)/(1+df) ) +1 )
           ,
           tibble( n = 1e4L, df = 0:1e4L ) %>% mutate(idf = log( (1+n)/(1+df) ) +1 )
           ,
           tibble( n = 1e5L, df = 0:1e5L ) %>% mutate(idf = log( (1+n)/(1+df) ) +1 )
         ) %>% 
  ggplot( data = .,
          aes( x = df,
               y = idf
              )  
        ) +
  geom_line() +
  facet_wrap( ~n, scales = "free" ) +
  labs( title = "IDF values for different document sizes")
```

# Calculating TF-IDF in python:

TF-IDF calc for a given word is:

$$\begin{array}{lcl}
\underset{\text{Term Frequency - Inverse Document Frequency}}{\underbrace{\text{TF - IDF}}} 
&=& 
\text{TF} \quad \times \quad \underset{\text{IDF}}{\underbrace{\Bigg( log\Big\{ \displaystyle\frac{1+n}{1+DF} \Big\} +1 \Bigg)}} \\
n &=& \text{total number of documents (products)} \\
DF &=& \text{number of documents (products) containing this word} \\
\end{array}$$

the resulting TF-IDF vector (row) $\overset{\rightarrow{}}{v}$ is then normalised:

$$\begin{array}{lcl}
\text{L1 norm}: \quad \quad \overset{\rightarrow{}}{v}_{norm} &=& x \\
\text{L2 norm}: \quad \quad x &=& x \\
\end{array}$$

this information comes directly from the documentation https://scikit-learn.org/stable/modules/feature_extraction.html#text-feature-extraction

create dummy dataset:
```{python}
dummy_data = [ [1,'big big black boots'], [2,'womens running pants stripey'], [3,'black mens pants'], [4,'stripey mens socks blue and red'], [5,'red socks']  ]
dummy_data = pd.DataFrame( dummy_data, columns=['product_id','product_words'] )
dummy_data
```

```{python}
init_CountVectorizer = CountVectorizer( analyzer='word',
                                        strip_accents = 'unicode',
                                        lowercase = True,
                                        binary = True       # counts above 1 are counted as 1
                                      )
count_matrix = init_CountVectorizer.fit_transform( dummy_data['product_words'] )

# print( count_matrix )
# print('')

tf = pd.DataFrame( count_matrix.toarray() ) 
tf.columns = [init_CountVectorizer.get_feature_names()]

joined_df = pd.concat([dummy_data, tf], axis=1)

joined_df

```

Let's have a look at the matrix of word counts (in R) that we've created in python:

```{r}
py$joined_df
```

TF-IDF calc (for the scikit learn python package) for a given word is:

$$\begin{array}{lcl}
\underset{\text{Term Frequency - Inverse Document Frequency}}{\underbrace{\text{TF - IDF}}} 
&=& 
\text{TF} \quad \times \quad \underset{\text{IDF}}{\underbrace{\Bigg( log\Big\{ \displaystyle\frac{1+n}{1+DF} \Big\} +1 \Bigg)}} \\
n &=& \text{total number of documents (products)} \\
DF &=& \text{number of documents (products) containing this word} \\
\end{array}$$

the resulting TF-IDF vector (row) $\overset{\rightarrow{}}{v}$ is then normalised:

$$\begin{array}{lcl}
\text{L1 norm}: \quad \quad \overset{\rightarrow{}}{v}_{norm} &=& x \\
\text{L2 norm}: \quad \quad x &=& x \\
\end{array}$$

this information comes directly from the documentation https://scikit-learn.org/stable/modules/feature_extraction.html#text-feature-extraction

```{python}
word_document_counts = joined_df.drop(['product_id', 'product_words'], axis=1)
row_of_counts = word_document_counts.sum(axis=0)
row_of_counts = pd.DataFrame( row_of_counts ).transpose()
single_row_of_counts = row_of_counts

for i in range(joined_df.shape[0]-1):
    row_of_counts = row_of_counts.append(single_row_of_counts) 

n = joined_df.shape[0]     # 'n' is the number of documents/products
    
row_of_counts 

idf = np.log( (1+n) / (1+row_of_counts) ) +1
```

```{r}
py$row_of_counts
py$tf
py$idf
```

my manual tf-idf calc:

```{python}
tf_idf_not_normalised = pd.DataFrame( tf.values*idf.values, columns = tf.columns, index=tf.index )
#tf_idf_not_normalised
tf_idf_normalised = pd.DataFrame( preprocessing.normalize(tf_idf_not_normalised, norm='l2') )
tf_idf_normalised.columns = [init_CountVectorizer.get_feature_names()]
#tf_idf_normalised
```

```{r}
py$tf_idf_not_normalised
py$tf_idf_normalised
```

and here is how we do the tf-idf automatically in python:
```{python}
init_tf_idf_vectorizer = TfidfVectorizer( strip_accents='ascii',   # do punctuation cleaning before doing tf-idf
                                          lowercase=True,        # convert all to lowercase before doing tf-idf 
                                          analyzer='word',
                                          binary=True,           # forces tf to be either 0 or 1                   
                                          decode_error='strict',    
                                          input='content',
                                          max_df=1.0,
                                          max_features=None,      
                                          min_df=1,
                                          norm='l2',         # each row (vector) scaled to have (l2) length 1 
                                                             # Vectors are normalized to unit length before they are 
                                                             #   used for similarity calculation, making cosine similarity
                                                             #   and dot-product equivalent (Levy et al. 2015)
                                          preprocessor=None,
                                          smooth_idf=True,   # add 1 to every df number (prevents div. by 0)
                                          stop_words=None, 
                                          sublinear_tf=False,   # i.e. don't replace [tf] with [1 + log(tf)]
                                          token_pattern='(?u)\\b\\w\\w+\\b',  # what constitutes a token (word)
                                          tokenizer=None,
                                          use_idf=True,
                                          vocabulary=None
                                        )


tf_idf_matrix = init_tf_idf_vectorizer.fit_transform( dummy_data['product_words'] )
tf_idf_matrix = pd.DataFrame( tf_idf_matrix.toarray() ) 
#tf_idf_matrix

tf_idf_matrix.columns = [init_tf_idf_vectorizer.get_feature_names()]
#tf_idf_matrix

joined_df = pd.concat([dummy_data, tf_idf_matrix], axis=1)
```

```{r}
py$joined_df
```

```{python}
init_vectorizer = TfidfVectorizer()

count_matrix = init_vectorizer.fit_transform( r.tf_idf_test['some_text'] )


#pd.DataFrame( count_matrix.toarray() ) 
```


# The full rawtext personalisation process in Python

first, I create more dummy datasets in different messy formats:

```{r}

# EXAMPLE PRODUCTS:
# 
# black formal pants
# red and blue swimwear (synthetic)
# brown formal shoes (leather)
# black formal shoes (synthetic)
# black running shorts
# casual blue shoes stripes
# red sport pants running
# casual white running shoes 
# formal white shirt
# running shirt
# black and white casual shirt

dirty_attribute_data1 <- 
  bind_rows( tibble(product_id = 1,
                    attribute = "colour",
                    attribute_desc = "black"
                    ),
             tibble(product_id = 1,
                    attribute = "use",
                    attribute_desc = "formal"
                    ),
             
             tibble(product_id = 2,
                    attribute = "colour",
                    attribute_desc = "red"
                    ),
             tibble(product_id = 2,
                    attribute = "color",
                    attribute_desc = "red"
                    ),
             tibble(product_id = 2,
                    attribute = "color",
                    attribute_desc = "blue"
                    ),
             tibble(product_id = 2,
                    attribute = "material",
                    attribute_desc = "synthetic"
                    ),
             
             tibble(product_id = 3,
                    attribute = "colour",
                    attribute_desc = "brown"
                    ),
             tibble(product_id = 3,
                    attribute = "material",
                    attribute_desc = "leather"
                    ),
             
             tibble(product_id = 4,
                    attribute = "colour",
                    attribute_desc = "black"
                  ),
             tibble(product_id = 4,
                    attribute = "use",
                    attribute_desc = "running"
                  ),
             
             tibble(product_id = 5,
                    attribute = "use",
                    attribute_desc = "fashion"
                  ),
             tibble(product_id = 5,
                    attribute = "colour",
                    attribute_desc = "blue"
                  ),
             
            tibble( product_id = 6,
                    attribute = "colour",
                    attribute_desc = "red"
                  ),
            tibble( product_id = 6,
                    attribute = "use",
                    attribute_desc = "running marathon"
                  ),
            tibble( product_id = 6,
                    attribute = "material",
                    attribute_desc = "cotton"
                  ),
            
             tibble(product_id = 7,
                    attribute = "colour",
                    attribute_desc = "brown"
                    ),
             tibble(product_id = 7,
                    attribute = "use",
                    attribute_desc = "formal"
                    ),
             tibble( product_id = 7,
                    attribute = "material",
                    attribute_desc = "cotton"
                  ),
            
             tibble( product_id = 9,
                     attribute = "use",
                     attribute_desc = "sleep"
                    ),
          
             tibble( product_id = 9,
                    attribute = "material",
                    attribute_desc = "cotton"
                  ),
            
             tibble( product_id = 10,
                     attribute = "use",
                     attribute_desc = "sleep"
                    ),
            
             tibble( product_id = 10,
                     attribute = "material",
                     attribute_desc = "cotton"
                  )
            )

dirty_attribute_data2 <- 
  bind_rows( 
              tibble( product_id = 1, product_desc="smart suit pants", category="pants", department="formal", style="retro" ),
              tibble( product_id = 2, product_desc="colourful MENS swim shorts", category="swimming", department="sport", style="casual" ),
              tibble( product_id = 3, product_desc="light brown dinner shoes", category="footwear", department="formal", style="young" ),
              tibble( product_id = 4, product_desc="black house-brand running shorts", category="shorts", department="sport", style="sport" ),
              tibble( product_id = 5, product_desc="trendy blue sneakers", category="footwear", department="casual", style="young" ),
              tibble( product_id = 6, product_desc="athletic red sports pants", category="pants", department="sport", style="retro" ),
              tibble( product_id = 7, product_desc="brown suit trousers", category="pants", department="formal", style="slimfit" ),
              tibble( product_id = 8, product_desc="very small red swimsuit", category="swimming", department="sport", style="slimfit" ),
              tibble( product_id = 9, product_desc="blue pajama top", category="sleepwear", department="home", style="comfort" ),
              tibble( product_id = 10, product_desc="red pajama pants", category="sleepwear", department="home", style="comfort" ),
              tibble( product_id = 11, product_desc="furry slippers", category="shoes", department="home", style="comfort" )
              
  )
  
dirty_attribute_data1
dirty_attribute_data2

```

clean the first (long) dataset using python:

```{python}
dirty_attribute_data1_py = r.dirty_attribute_data1.drop(["attribute"], axis=1)

# remove duplicate rows from dirty_attribute_data1:
dirty_attribute_data1_py = dirty_attribute_data1_py.drop_duplicates()

# for the long data, paste all of the words into a single column, separated by space:
bagofwords_attribute_data1 = dirty_attribute_data1_py.groupby("product_id")["attribute_desc"].apply(" ".join)

# convert back to a pandas dataframe:
bagofwords_attribute_data1 = bagofwords_attribute_data1.to_frame()

# rename the attribute_desc column
bagofwords_attribute_data1.rename( columns={'attribute_desc':'data1_words'}, inplace=True)

```

```{r}
py$bagofwords_attribute_data1
```

clean the second (wide) dataset using python:

```{python}
# read in the product data from the R tibble:
dirty_attribute_data2 = r.dirty_attribute_data2

# paste all of the columns into 
dirty_attribute_data2['data2_words'] = dirty_attribute_data2[['product_desc',
                                                              'category',
                                                              'department',
                                                              'style'
                                                              ]].apply(lambda x: ' '.join(x), axis=1)

# only keep the product ID and the bag of product words:
bagofwords_attribute_data2 = dirty_attribute_data2[['product_id', 'data2_words']]
```

```{r}
py$bagofwords_attribute_data2
```

Join the 2 datasets in python to get one bag of words for each product:

```{python}
all_product_WordBags_raw = pd.merge(bagofwords_attribute_data1, bagofwords_attribute_data2, on='product_id', how='outer')

all_product_WordBags_raw

# turn the NaN (i.e. no match) values into empty string:
all_product_WordBags_raw = all_product_WordBags_raw.replace(np.nan, '', regex=True)

# merge the 2 word columns:
all_product_WordBags_raw['bag_of_words'] = all_product_WordBags_raw[['data1_words', 'data2_words',]].apply(lambda x: ' '.join(x), axis=1)
all_product_WordBags_raw = all_product_WordBags_raw[['product_id', 'bag_of_words']]
```

Look at the product wordclouds/bags of words that we have created:

```{r}
py$all_product_WordBags_raw
```

Now, turn each product_id into a word vector, using tf-idf (in python):

specify the parameters of the word-vectorizer function TfidfVectorizer():
```{python}
init_tf_idf_vectorizer = TfidfVectorizer( strip_accents='ascii',   # do punctuation cleaning before doing tf-idf
                                          lowercase=True,        # convert all to lowercase before doing tf-idf 
                                          analyzer='word',
                                          binary=True,           # forces tf to be either 0 or 1                   
                                          decode_error='strict',    
                                          input='content',
                                          max_df=1.0,
                                          max_features=None,      
                                          min_df=1,
                                          norm='l2',         # each row (vector) scaled to have (l2) length 1 
                                                             # Vectors are normalized to unit length before they are 
                                                             #   used for similarity calculation, making cosine similarity
                                                             #   and dot-product equivalent (Levy et al. 2015)
                                          preprocessor=None,
                                          smooth_idf=True,   # add 1 to every df number (prevents div. by 0)
                                          stop_words=None, 
                                          sublinear_tf=False,   # i.e. don't replace [tf] with [1 + log(tf)]
                                          token_pattern='(?u)\\b\\w\\w+\\b',  # what constitutes a token (word)
                                          tokenizer=None,
                                          use_idf=True,
                                          vocabulary=None
                                        )

```

```{python}
raw_tf_idf_matrix = init_tf_idf_vectorizer.fit_transform( all_product_WordBags_raw['bag_of_words'] )
tf_idf_matrix = pd.DataFrame( raw_tf_idf_matrix.toarray() ) 
#tf_idf_matrix

tf_idf_matrix.columns = [init_tf_idf_vectorizer.get_feature_names()]
#tf_idf_matrix

joined_df = pd.concat([all_product_WordBags_raw, tf_idf_matrix], axis=1)
```

```{r}
py$raw_tf_idf_matrix        # this is stored as a sparse matrix
py$tf_idf_matrix
py$joined_df
```


```{python}
product_cosine_similarities = cosine_similarity(raw_tf_idf_matrix, raw_tf_idf_matrix) 
product_cosine_similarities_alt = linear_kernel(raw_tf_idf_matrix, raw_tf_idf_matrix)      # same as cosine similarity because we used L2 norm 
```

```{r}
py$product_cosine_similarities

py$product_cosine_similarities_alt
```

make a little recommender function, returning closest 10 products to each product:

```{python}
# #checking out iterrows():
# print( product_wordbags.head(n=5) )
# print("\n")
# for index, row in product_wordbags.head(n=5).iterrows():
#     print(index, "\n\n",  row)
#     print("\n")        

# this code mostly stolen from https://heartbeat.fritz.ai/recommender-systems-with-python-part-i-content-based-filtering-5df4940bd831

results = {}
for rownum_i, row_i in all_product_WordBags_raw.iterrows():     # iterate through the rows of the dataframe product_wordbags
    # 'rownum_i' is the row number, 'row_i' is the row itself 
    # i.e. we are iterating through each product
    
    # store indices of 5 or so products with smallest cosine distance to this product:
    # (this includes the product itself with distance 0)
    similar_indices = product_cosine_similarities[rownum_i].argsort()[:-7:-1]  
    # print(similar_indices)
    
    # store distance to each of the 100 or so closest products:
    similar_items = [ (product_cosine_similarities[rownum_i][i],
                       all_product_WordBags_raw['product_id'][i]
                      ) for i in similar_indices 
                    ] 
    # print(similar_items)
    
    # store this information in the dictionary 'results':
    results[row_i['product_id']] = similar_items[1:]  
```

So **results** is now a dictionary, storing for each product the 5 closest products, with their cosine similarity score:
```{python}
results
```

Define a python function which fetches the bag of words (text) for a given product_ key_id:
```{python}
def get_item_words(product_id):  
    return all_product_WordBags_raw.loc[all_product_WordBags_raw['product_id'] == product_id]['bag_of_words'].tolist()[0].split(' - ')[0]
```

test out the function:
```{python}
get_item_words(product_id=4)
```

```{r}
py$all_product_WordBags_raw %>% filter(product_id==4)
```

define a little recommender function, returning closest matching X products given a certain product key:
```{python}
def get_recommendations(product_id, num_recommendations):
    print("Recommending " + 
          str(num_recommendations) + 
          " products similar to   " +
          get_item_words(product_id)
         )
    print( "------" )
    
    # extract [num_recommendations] closest products to product [product_id], with their cosine similarity:  
    recs = results[product_id][:num_recommendations]  
    
    for rec in recs:
        print("Recommended: " + get_item_words(rec[1]) + " (score:" +      str(rec[0]) + ")")
```

test out this product closeness recommender function:
```{python}
get_recommendations(product_id=6, num_recommendations=3)
```

```{r}
round( py$product_cosine_similarities, 2)
py$all_product_WordBags_raw
```

Now, suppose that we have observed some customer purchases:
```{python}
data_for_df = [ [101, -1], [101, -1], [102, -1], [103, -1], [103, -1] ]   # create array of data for dataframe
customer_purchases = pd.DataFrame( data_for_df, columns=['customer_id','product_id'] )           # turn array of data into dataframe
customer_purchases['product_id'] = all_product_WordBags_raw['product_id'][ np.random.choice( all_product_WordBags_raw.shape[0],
                                                                                             customer_purchases.shape[0],    # number of samples (i.e. number of rows of customer purchases table)
                                                                                             replace = True
                                                                                           ) ].values
                                                     
customer_purchases
```

```{r}
head( py$raw_tf_idf_matrix )
```

We turn these customers into word/product vectors by averaging the TF-IDF vectors(rows) of all of the products that they have bought:
```{python}
# link all product_IDs to tf-idf vector:
product_ID_and_tfidf = pd.concat( [all_product_WordBags_raw['product_id'], pd.DataFrame(raw_tf_idf_matrix.todense()) ],
                                  axis=1
                                )
                                
product_ID_and_tfidf
```

For each customer, we average (take the element-wise mean) of the TFIDF-vectors of the products which they bought:

```{python}
# get tf-idf info for each customer purchase:
customer_product_profiles = customer_purchases.merge( product_ID_and_tfidf, 
                                                      on = 'product_id', 
                                                      how = 'left' ).groupby('customer_id').mean().drop('product_id',axis=1)
customer_product_profiles
```

```{r}
py$customer_product_profiles
```

For each customer vector, store product_IDs of closest 7 products and return this information:
```{python}
customer_product_distances = cosine_similarity(customer_product_profiles, raw_tf_idf_matrix.todense() )
```

```{r}
py$customer_product_distances
```

```{python}
customer_closest_products = {}  # create an empty dictionary

for customer_i in range(customer_product_profiles.shape[0]):        # for each customer
  
  close_product_indices = customer_product_distances[customer_i].argsort()[::-1][:7]     # return indices of 5 top matches, in descending order of cosine similarity

  customer_closest_products[customer_product_profiles.index.values[customer_i]] = all_product_WordBags_raw['bag_of_words'][close_product_indices]      

customer_closest_products
```

```{r}
py$customer_closest_products
```

So, customer 101 bought:
```{python}
extract_customer_101_purchases = customer_purchases[customer_purchases['customer_id']==101]
#extract_customer_101_purchases

customer_101_purchases_words = extract_customer_101_purchases.merge( all_product_WordBags_raw,
                                                                     on = 'product_id', 
                                                                     how = 'left'
                                                                   )
```

```{r}
py$customer_101_purchases_words
```

and top recommendations for customer 101 are:
```{python}
customer_closest_products[101]
```

this is how to format the customer product recommendations as a single table, for example for export:
(taken top 5 products per customer)

```{python}
customer_close_products = []  # create an empty data frame

for customer_i in range(customer_product_profiles.shape[0]):        # for each customer
  close_product_indices = customer_product_distances[customer_i].argsort()[::-1]     # return indices of matches in descending order of cosine distance
  
  close_product_IDs = all_product_WordBags_raw['product_id'][close_product_indices]      
  close_product_cosine_similarity_values = customer_product_distances[customer_i][close_product_indices]
  
  product_IDs_actually_bought = customer_purchases['product_id'][customer_purchases['customer_id']==customer_product_profiles.index.values[customer_i]]
  
  for prod_ID in (close_product_IDs):
    if sum(prod_ID==product_IDs_actually_bought) == 0:     # if customer has not bought this product:
      prods_this_customer_dict = {}
      prods_this_customer_dict['customer_id'] = customer_product_profiles.index.values[customer_i]
      prods_this_customer_dict['product_id'] = prod_ID
      prods_this_customer_dict['cosine_similarity'] = 'NA'
      customer_close_products.append( prods_this_customer_dict )
    
print( customer_close_products )  

customer_close_products_df = pd.DataFrame( customer_close_products )  

print( customer_close_products_df )


```

```{python}
1 == product_IDs_actually_bought
```

```{r}
py$customer_close_products_df
```



# older code


alternative method of getting cosine distances between products:
```{python}
tf_idf_matrix_sparse_format = sparse.csr_matrix(tf_idf_matrix)

product_similarities_sparse = cosine_similarity(tf_idf_matrix_sparse_format, dense_output=False)

# pd.DataFrame(product_similarities_sparse.toarray())
# 
# print('pairwise sparse output:\n {}\n'.format(product_similarities_sparse))

mc = product_similarities_sparse.tocoo()

nice_similarity_matrix = pd.concat( [pd.DataFrame(mc.row), pd.DataFrame(mc.col), pd.DataFrame(mc.data)], axis=1, ignore_index=True )

nice_similarity_matrix['product1_id'] = np.asarray( joined_df['product_id'][np.asarray(nice_similarity_matrix[0])] )
nice_similarity_matrix['product2_id'] = np.asarray( joined_df['product_id'][np.asarray(nice_similarity_matrix[1])] )

#nice_similarity_matrix
nice_similarity_matrix = nice_similarity_matrix.drop([0,1], axis=1)
nice_similarity_matrix.rename( columns = {2:'cosine_similarity'}, inplace=True )

nice_similarity_matrix.sort_values('cosine_similarity', ascending=False)
```

```{r}
py$nice_similarity_matrix %>% 
  
  filter( product1_id != product2_id) %>% 
  
  # add product 1 bag of words: 
  left_join( py$joined_df %>% 
                select(product_id, bag_of_words) %>% 
                rename( product1_id = product_id,
                        product1_words = bag_of_words
                      )
            ) %>% 
  
  # add product 2 bag of words: 
  left_join( py$joined_df %>% 
                select(product_id, bag_of_words) %>% 
                rename( product2_id = product_id,
                        product2_words = bag_of_words
                      )
            ) %>% 
  select( -product1_id, -product2_id ) %>% 
  arrange( desc(cosine_similarity) ) %>% 
  kable()
  
```

simulate some random purchases for 3 customers:

```{python}
customer1 = tf_idf_matrix.sample(n=random.randint(1,5), replace=True)
customer1_products = joined_df.loc[customer1.index.values,:] 
customer1_products = customer1_products[['product_id', 'bag_of_words']]
customer1_vector = customer1.mean(axis=0)
customer1_vector_normalised = preprocessing.normalize( [customer1_vector], norm='l2')

customer2 = tf_idf_matrix.sample(n=random.randint(1,5), replace=True)
customer2_products = joined_df.loc[customer2.index.values,:] 
customer2_products = customer2_products[['product_id', 'bag_of_words']]
customer2_vector = customer2.mean(axis=0)
customer2_vector_normalised = preprocessing.normalize( [customer2_vector], norm='l2')

customer3 = tf_idf_matrix.sample(n=random.randint(1,5), replace=True)
customer3_products = joined_df.loc[customer3.index.values,:] 
customer3_products = customer3_products[['product_id', 'bag_of_words']]
customer3_vector = customer3.mean(axis=0)
customer3_vector_normalised = preprocessing.normalize( [customer3_vector], norm='l2')

```

So, customer 1 bought:
```{r}
py$customer1_products
```

and here are their product matches:
```{python}
customer1_product_matches_table = joined_df[['product_id','bag_of_words']]

customer_cosine_similar_vector = []

for row_i in range(tf_idf_matrix.shape[0]):      # iterate over rows of the product matrix
  # can use dot prod since normalised vectors:
  customer_cosine_similar_vector.append( np.dot(customer1_vector_normalised, tf_idf_matrix.loc[row_i,:])[0] )    

customer1_product_matches_table['cosine_similarity'] = customer_cosine_similar_vector
```

```{r}
py$customer1_product_matches_table %>% 
  arrange( desc(cosine_similarity) ) %>% 
  kable()
```

customer 2 purchases:
```{r}
py$customer2_products
```

customer 2 product matches:
```{python}
customer2_product_matches_table = joined_df[['product_id','bag_of_words']]

customer_cosine_similar_vector = []

for row_i in range(tf_idf_matrix.shape[0]):      # iterate over rows of the product matrix
  # can use dot prod since normalised vectors:
  customer_cosine_similar_vector.append( np.dot(customer2_vector_normalised, tf_idf_matrix.loc[row_i,:])[0] )    

customer2_product_matches_table['cosine_similarity'] = customer_cosine_similar_vector
```

```{r}
py$customer2_product_matches_table %>% 
  arrange( desc(cosine_similarity) ) %>% 
  kable()
```

customer 3 purchases:
```{r}
py$customer3_products
```

customer 2 product matches:
```{python}
customer3_product_matches_table = joined_df[['product_id','bag_of_words']]

customer_cosine_similar_vector = []

for row_i in range(tf_idf_matrix.shape[0]):      # iterate over rows of the product matrix
  # can use dot prod since normalised vectors:
  customer_cosine_similar_vector.append( np.dot(customer3_vector_normalised, tf_idf_matrix.loc[row_i,:])[0] )    

customer3_product_matches_table['cosine_similarity'] = customer_cosine_similar_vector
```

```{r}
py$customer3_product_matches_table %>% 
  arrange( desc(cosine_similarity) ) %>% 
  kable()
```






## still to do:

* remove stopwords

* lemmatize




# other useful text tools in R

```{r}
library(tidyverse)
library(tidytext)
library(janeaustenr)
library(stringr)
library(SnowballC) # for the function wordStem() for  stemming words

text <- c("Because I could not stop for Death -",
          "He kindly stopped for me -",
          "The Carriage held but just Ourselves -",
          "and Immortality"
        )

text_df <- tibble(line = 1:4, text = text)

text_df

# tokenize our text dataframe:
# a 'token' is a meaningful unit of text (e.g. a word)
text_df %>% 
  unnest_tokens( output = "word",     # name of output column
                 input = "text",      # name of column containing the text to be tokenized 
                 to_lower = TRUE,     # convert text to lowercase
                 strip_punct = TRUE  # remove punctuation (.,-/ etc.)
               )

  
raw_books <- austen_books()

raw_books %>% slice(500:510)

original_books <- 
  raw_books %>% 
    group_by(book) %>%   
      # within each book:
      mutate( # add a unique line number to each line:
              linenumber = row_number(),    
              
              # add a chapter column by getting the chapter number out of the raw text
              chapter = cumsum(str_detect(text, regex("^chapter [\\divxlc]",
                                                     ignore_case = TRUE
                                                     )
                                          )
                               )
              ) %>%
      ungroup()

# look at a random few lines from different books:
original_books %>% sample_n(10)  

# put into tidy text format:
tidy_books <- 
  original_books %>%
    unnest_tokens(input="text", output="word")

tidy_books %>%  head(20)

# get a list of stop-words (words not useful in analysis, like 'the', 'of', 'to' etc.)
data(stop_words)

nrow(stop_words)
head(stop_words, 20)

tidy_books_no_stopwords <-
  tidy_books %>% 
    anti_join(stop_words)
  
tidy_books %>% head(20)
tidy_books_no_stopwords %>% head(20)

# count most common words across all books:
tidy_books_no_stopwords %>% 
  count( word, sort=TRUE ) %>% 
  head(20)


# We use str_extract() here because the UTF-8 encoded texts from Project 
# Gutenberg have some examples of words with underscores around them to indicate 
# emphasis (like italics). The tokenizer treated these as words, but we don't want 
# to count "_any_" separately from "any" as we saw in our initial data 
# exploration before choosing to use str_extract()
# go search for this text in https://www.tidytextmining.com/tidytext.html

tibble( word = c("word", "_word_", "%w%o%r%d%") ) %>% 
  mutate( word_clean1 = str_extract(word, "[a-z']+"),
          word_clean2 = str_remove_all( word, "\\W+" )
        )



# checking out tf-idf:
example_dat <- tibble( document = paste("document", 1:5, sep="_"),
                       document_text = c( "black and big black boots and black",
                                          "black pants formal and with buttons",
                                          "BIG black and red pants with big buttons",
                                          "black hat with red and green stripes",
                                          "blue formal waistcoat"
                                        ) 
                       )

get_tf.idf <- 
example_dat %>% 
  unnest_tokens( input="document_text", output="word" ) %>% 
  group_by( document, word ) %>% 
  summarise( n_times_in_doc = n() ) %>% 
  ungroup() %>% 
  group_by( word ) %>% 
  mutate( n_docs_containing_word = n_distinct(document) ) %>% 
  group_by( document ) %>% 
  mutate( n_words_in_doc = sum(n_times_in_doc) ) %>% 
  ungroup() %>% 
  mutate( joe_tf = n_times_in_doc / n_words_in_doc ) %>% 
  ungroup() %>% 
  mutate( `n_docs/n_docs_containing_word` = 5 / n_docs_containing_word ) %>% 
  mutate( joe_idf = log(`n_docs/n_docs_containing_word`) ) %>% 
  mutate( joe_tf_idf = joe_tf * joe_idf ) %>% 
  bind_tf_idf( tbl =.,
               term = word,
               document = document,
               n = n_times_in_doc
  ) %>% 
  arrange( document, desc(tf_idf) ) 
# 
# # turn our tidy dataframe into a document-term matrix:
# example_dat %>% 
#   unnest_tokens( input="document_text", output="word" ) %>% 
#   group_by( document, word ) %>% 
#   summarise( n_times_in_doc = n() ) %>% 
#   ungroup() %>% 
#   cast_dtm( term = word,
#             document = document
#           )

# checking out stemming:
tibble( unstemmed_word = c("running", "run", "runs", "runned", "runner", "runners", "ran") ) %>% 
  mutate( stemmed_word = map_chr( .x = unstemmed_word,
                                .f = function(word){ wordStem( word, language="en" ) } 
                                )
  )


# check out document term matrix:
```



