---
title: "Latent Factor Matrix Decomposition Recommender"
output:
  html_document:
    df_print: paged
---

```{r setup}
library(tidyverse)
library(Matrix)
library(patchwork)
```

# Resources

1. For recommender theory: that awesome textbook that Dino gave me

2. Really nice high-level video on the theory: Lecture 55 — Latent Factor Recommender System | Stanford University: available at https://www.youtube.com/watch?v=E8aMcwmqsTg  

3. For plotting heatmaps in *ggplot2*: heatmapshttps://www.r-graph-gallery.com/79-levelplot-with-ggplot2.html

# Theory
Suppose that we have rating matrix $Y_{M\times N}$, which shows the ratings (ratings on 1-5 scale) that $M$ customers have given to a pool of $N$ items. For example, each item could be a movie and the matrix $Y_{M\times N}$ could show the ratings that various customers have given to various movies.  

The rating matrix $Y_{M\times N}$ is a sparse matrix (i.e. many entries in the matrix are missing). Entry $Y_{ij}$ of this matrix is the rating given to item $j$ by customer $i$. Here is an example of a rating matrix $Y_{10\times 6}$:

```{r echo=FALSE}
n_items <- 6
n_customers <- 10
percent_sparse <- 0.5

# assign some random 1-5 ratings for various customer/item combinations:
rating_matrix <- matrix( sample(1:5, size=n_items*n_customers, replace=TRUE),
                         ncol = n_items,
                         nrow = n_customers
                       )
rating_matrix <- 
  rating_matrix * 
  sample( 0:1, size=prod(dim(rating_matrix)), replace=TRUE, prob=c(percent_sparse,1-percent_sparse) )

colnames(rating_matrix) <- paste0("item", 1:ncol(rating_matrix), sep="" )
rownames(rating_matrix) <- paste0("customer", 1:nrow(rating_matrix), sep="" )

# make sure that each customer has rated at least 1 item:
for( row_i in 1:nrow(rating_matrix) ){
  if( sum(rating_matrix[row_i,])==0 ){ 
      rating_matrix[row_i,sample(1:ncol(rating_matrix),size=1)] <- 
        sample( 1:5, size=1)
    }
}

# make sure that each item has been rated at least once:
for( col_j in 1:ncol(rating_matrix) ){
  if( sum(rating_matrix[,col_j])==0 ){ 
      rating_matrix[sample(1:nrow(rating_matrix),size=1),col_j] <- 
        sample( 1:5, size=1)
    }
}

rating_matrix <- Matrix( rating_matrix, sparse=TRUE )

rating_matrix
```

We want to find matrices $U_{M\times L}$ and $V_{N\times L}$ where $UV^T$ provides a low-rank decomposition (approximation) of $Y$:

$$\hat Y \quad=\quad UV^T$$

$$i.e.\quad \hat y_{ij} \quad=\quad \mathbf{u}_i' \mathbf{v}_j$$

$\mathbf{u}_i'$ is the $i^{th}$ row of $U_{M\times L}$.

$\mathbf{v}_j'$ is the $j^{th}$ row of $V_{N\times L}$. 

$\mathbf{u}_i$ is a vector of length $L$.

$\mathbf{v}_j$ is a vector of length $L$.

We choose the dimensionality $L$ to be a much smaller number than either $M$ or $N$.  

$\mathcal{I}_j$ is the set of users who rated item $j$. 

$\mathcal{J}_i$ is the set of items that were rated by user $i$. 

What we are doing here is representing each customer (row $i$ of rating matrix $Y$) as an $L$-dimensional vector $u_i$, and each product (column $j$ of rating matrix $Y$) as an $L$-dimensional vector $v_j$. Then, the predicted rating for a particular customer/product combination $\{u_i,v_j\}$ is calculated as the dot product between those two vectors $\hat y_{ij}=u'\cdot v$.     

We can choose these vectors $u_1,u_2,...,u_M, v_1,v_2,...,v_N$ by minimising the sum of squared errors (squared differences between observed ratings $y_{ij}$ and predicted ratings $\hat y_{ij}=u'\cdot v$) over the observed ratings. This makes the objective function:

$$dothisjoe$$

Since 

To explore this concept, let's first create a rating matrix $Y$ filled with random information:

```{r}
n_items <- 6
n_customers <- 10
percent_sparse <- 0.5

# assign some random 1-5 ratings for various customer/item combinations:
rating_matrix <- matrix( sample(1:5, size=n_items*n_customers, replace=TRUE),
                         ncol = n_items,
                         nrow = n_customers
                       )
rating_matrix <- 
  rating_matrix * 
  sample( 0:1, size=prod(dim(rating_matrix)), replace=TRUE, prob=c(percent_sparse,1-percent_sparse) )

colnames(rating_matrix) <- paste0("item", 1:ncol(rating_matrix), sep="" )
rownames(rating_matrix) <- paste0("customer", 1:nrow(rating_matrix), sep="" )

# make sure that each customer has rated at least 1 item:
for( row_i in 1:nrow(rating_matrix) ){
  if( sum(rating_matrix[row_i,])==0 ){ 
      rating_matrix[row_i,sample(1:ncol(rating_matrix),size=1)] <- 
        sample( 1:5, size=1)
    }
}

# make sure that each item has been rated at least once:
for( col_j in 1:ncol(rating_matrix) ){
  if( sum(rating_matrix[,col_j])==0 ){ 
      rating_matrix[sample(1:nrow(rating_matrix),size=1),col_j] <- 
        sample( 1:5, size=1)
    }
}

rating_matrix <- Matrix( rating_matrix, sparse=TRUE )

rating_matrix
```

# Estimate latent factor matrices by alternating least squares

Create starting matrices $U$ and $V$ of *latent factor* vectors $\mathbf{u}_i$ and $\mathbf{v}_j$ (fill them with random values):

```{r}
# define dimension 
dim_L <- 3
dim_M <- nrow(rating_matrix)
dim_N <- ncol(rating_matrix)

U_matrix <- matrix( data = rnorm( n=dim_M * dim_L),
                    nrow = dim_M,
                    ncol = dim_L
                  )
colnames(U_matrix) <- paste("u", 1:ncol(U_matrix), sep="")

V_matrix <- matrix( data = rnorm( n=dim_N * dim_L),
                    nrow = dim_N,
                    ncol = dim_L
                  )
colnames(V_matrix) <- paste("v", 1:ncol(V_matrix), sep="")

U_matrix
V_matrix

U_matrix %*% t(V_matrix)
```

Estimate *U* and *V* by regularised alternating least squares:
(regularisation parameters $\lambda_1$ and $\lambda_2$ need to be chosen)

```{r}
lambda1 <- 0.1
lambda2 <- 0.1

lambda2_I <- diag( rep(lambda2, dim_L) )
lambda1_I <- diag( rep(lambda1, dim_L) )

for( iteration_i in 0:1000 ){

  # print iteration number and sum of squared errors every 50 iterations:
  if( iteration_i%%50 == 0 ){ 
        paste0( "iteration ", iteration_i ) %>% print()   
    
        # print the sum of square errors:
        # (over non-zero entries of Y)
        # sum(   [ y_{ij} - t(u_i)*v_j ]^2  )
        cat("new U: Sum of Squared Errors:")
      
        (
          sum(
               (
                 rating_matrix -
                 (  ( U_matrix %*% t(V_matrix) ) * (as.matrix(rating_matrix>0)*1)  )
               )^2
             ) +
          lambda1 * sum(U_matrix^2) +
          lambda2 * sum(V_matrix^2) 
        ) %>%
        print()
  }
  # keep [u_i]s constant and solve for each [v_j]:
  for( j in 1:dim_N ){   # for every row of V:
    V_matrix[j,] <-
      solve(
             lambda2_I +

              (
              lapply( which(rating_matrix[,j]!=0),   # all users who rated item [j]
                      function(row_i){ U_matrix[row_i,] %*% t(U_matrix[row_i,]) }
                    ) %>%
              Reduce( "+", . )
              )
      ) %*%

      (
        lapply( which(rating_matrix[,j]!=0),         # all users who rated item [j]
                function(row_i){ U_matrix[row_i, ] * rating_matrix[row_i, j] }
              ) %>%
        Reduce( "+", . )
      )
  }

  # keep [v_j]s constant and solve for each [u_i]:
  for( i in 1:dim_M ){   # for every row of U:

      U_matrix[i,] <-    # u_i
      solve(
             lambda1_I +

              (
              lapply( which(rating_matrix[i,]!=0),   # all items rated by user [i]
                      function(row_j){ V_matrix[row_j,] %*% t(V_matrix[row_j,]) }
                    ) %>%
              Reduce( "+", . )
              )
      ) %*%

      (
        lapply( which(rating_matrix[i,]!=0),   # all items rated by user [i]
                function(row_j){ V_matrix[row_j, ] * rating_matrix[i,row_j] }
              ) %>%
        Reduce( "+", . )
      )
  }

}
```


```{r fig.width=8, fig.height=6}
latent_decomp_matrix <- U_matrix %*% t(V_matrix)
rownames(latent_decomp_matrix) <- rownames(rating_matrix)
colnames(latent_decomp_matrix) <- colnames(rating_matrix)

rating_matrix
latent_decomp_matrix

rating_matrix_for_ggplot_heatmap <- 
  rating_matrix %>%
    as.matrix() %>% 
    as.data.frame() %>% 
    rownames_to_column( var="customer_id" ) %>%
    mutate( customer_id = as.numeric( gsub("customer", "", customer_id) ) ) %>% 
    pivot_longer(      cols = starts_with("item"),
                   names_to = "item_id",
                  values_to = "customer_rating"
                ) %>% 
    mutate( item_id = as.numeric( gsub("item", "", item_id) ) ) %>% 
    mutate( customer_rating = if_else(customer_rating==0, as.numeric(NA), customer_rating) ) %>% 
    mutate( customer_rating_text = if_else(customer_rating==0, " ", as.character(customer_rating))  )

lat_decomp_mat_for_ggplot_heatmap <- 
  latent_decomp_matrix %>%
    as.matrix() %>% 
    as.data.frame() %>% 
    rownames_to_column( var="customer_id" ) %>%
    mutate( customer_id = as.numeric( gsub("customer", "", customer_id) ) ) %>% 
    pivot_longer(      cols = starts_with("item"),
                   names_to = "item_id",
                  values_to = "predicted_customer_rating"
                ) %>% 
    mutate( item_id = as.numeric( gsub("item", "", item_id) ) )
  
bind_rows( 
           rating_matrix_for_ggplot_heatmap %>% mutate( source = "original rating matrix" )
           ,
           lat_decomp_mat_for_ggplot_heatmap %>% 
             mutate( source = "latent factor decomposition" ) %>%
             rename( customer_rating = predicted_customer_rating) %>% 
             mutate( customer_rating_text = as.character( round(customer_rating, digits=1) ) )
         ) %>% 
  ggplot( data = .,
          aes(     x = item_id,
                   y = customer_id,
                fill = customer_rating,
               label = customer_rating_text
             )
        ) +
  geom_tile() +
  geom_text() + 
  facet_wrap( ~source ) +
  scale_x_continuous( breaks = 1:n_items ) +
  scale_y_continuous( breaks = 1:n_customers ) +
  scale_fill_gradient2()

```

# Estimate latent factor matrices by stochastic gradient descent

```{r sgd}
lambda1 <- 0.1
lambda2 <- 0.1

calc_gradient_ij_ftn <- function( Umat, Vmat, Ymat, lamb1, lamb2, i, j ){

  list( 
        u_grad_row_i = 
            2 * (Ymat[i,j] - t(Umat[i,])%*%Vmat[j,]) %*% Vmat[j,] + ( (2*lamb1)/ sum(Ymat[i,]>0) )*Umat[i,],
        
        v_grad_row_j = 
            2 * (Ymat[i,j] - t(Umat[i,])%*%Vmat[j,]) %*% Umat[i,] + ( (2*lamb2)/ sum(Ymat[,j]>0) )*Vmat[j,]
      ) %>% 
    return(.)
  
} 

calc_SSE <- function( Umat, Vmat, Ymat, lamb1, lamb2 ){
          sum(
               (
                 Ymat -
                 (  ( Umat %*% t(Vmat) ) * (as.matrix(Ymat>0)*1)  )
               )^2
             ) +
          lamb1 * sum(Umat^2) +
          lamb2 * sum(Vmat^2) 
}

```

Set starting values for $U$ matrix and $V$ matrix:
```{r}
# define dimension 
dim_L <- 5
dim_M <- nrow(rating_matrix)
dim_N <- ncol(rating_matrix)

U_matrix <- matrix( data = rnorm( n=dim_M * dim_L),
                    nrow = dim_M,
                    ncol = dim_L
                  )
colnames(U_matrix) <- paste("u", 1:ncol(U_matrix), sep="")

V_matrix <- matrix( data = rnorm( n=dim_N * dim_L),
                    nrow = dim_N,
                    ncol = dim_L
                  )
colnames(V_matrix) <- paste("v", 1:ncol(V_matrix), sep="")

U_matrix
V_matrix
```

```{r}
learnrate_alpha <- 0.001
  
for( iteration_i in 0:1000 ){
  
  # print out SSE every 50 iterations:
  if( iteration_i%%50 == 0 ){ 
        paste0( "iteration ", iteration_i ) %>% print()   
  
  # print out the objective function value:
  cat("Sum of Squared Errors:  ")
  calc_SSE( Umat = U_matrix,
            Vmat = V_matrix,
            Ymat = rating_matrix,
            lamb1 = lambda1,
            lamb2 = lambda2
          ) %>% 
    cat()
  }
  
  # choose a random non-zero value from the rating matrix:
  rand_i <- sample( which(rowSums(rating_matrix)>0), size=1 )     # choose a random row which has one or more non-zero ratings
  rand_j <- sample( which(rating_matrix[rand_i,]>0), size=1 )

  # calculate gradient for U_i and V_j:
  calc_grads <- calc_gradient_ij_ftn( Umat = U_matrix,
                                      Vmat = V_matrix,
                                      Ymat = rating_matrix,
                                      lamb1 = lambda1,
                                      lamb2 = lambda2,
                                      i = rand_i,
                                      j = rand_j
                                    )
  
  # update u_i:
  U_matrix[rand_i,] <- U_matrix[rand_i,] + learnrate_alpha * calc_grads$u_grad_row_i
    
  # update v_i:
  V_matrix[rand_j,] <- V_matrix[rand_j,] + learnrate_alpha * calc_grads$v_grad_row_j
    
}


```

```{r}
latent_decomp_matrix <- U_matrix %*% t(V_matrix)
rownames(latent_decomp_matrix) <- rownames(rating_matrix)
colnames(latent_decomp_matrix) <- colnames(rating_matrix)

rating_matrix
latent_decomp_matrix

rating_matrix_for_ggplot_heatmap <- 
  rating_matrix %>%
    as.matrix() %>% 
    as.data.frame() %>% 
    rownames_to_column( var="customer_id" ) %>%
    mutate( customer_id = as.numeric( gsub("customer", "", customer_id) ) ) %>% 
    pivot_longer(      cols = starts_with("item"),
                   names_to = "item_id",
                  values_to = "customer_rating"
                ) %>% 
    mutate( item_id = as.numeric( gsub("item", "", item_id) ) ) %>% 
    mutate( customer_rating = if_else(customer_rating==0, as.numeric(NA), customer_rating) ) %>% 
    mutate( customer_rating_text = if_else(customer_rating==0, " ", as.character(customer_rating))  )

lat_decomp_mat_for_ggplot_heatmap <- 
  latent_decomp_matrix %>%
    as.matrix() %>% 
    as.data.frame() %>% 
    rownames_to_column( var="customer_id" ) %>%
    mutate( customer_id = as.numeric( gsub("customer", "", customer_id) ) ) %>% 
    pivot_longer(      cols = starts_with("item"),
                   names_to = "item_id",
                  values_to = "predicted_customer_rating"
                ) %>% 
    mutate( item_id = as.numeric( gsub("item", "", item_id) ) )
  
bind_rows( 
           rating_matrix_for_ggplot_heatmap %>% mutate( source = "original rating matrix" )
           ,
           lat_decomp_mat_for_ggplot_heatmap %>% 
             mutate( source = "latent factor decomposition" ) %>%
             rename( customer_rating = predicted_customer_rating) %>% 
             mutate( customer_rating_text = as.character( round(customer_rating, digits=1) ) )
         ) %>% 
  ggplot( data = .,
          aes(     x = item_id,
                   y = customer_id,
                fill = customer_rating,
               label = customer_rating_text
             )
        ) +
  geom_tile() +
  geom_text() + 
  facet_wrap( ~source ) +
  scale_x_continuous( breaks = 1:n_items ) +
  scale_y_continuous( breaks = 1:n_customers ) +
  scale_fill_gradient2()

```



