---
title: "SampleSize_Calc_for_YesNo_question"
author: "Joe"
date: "09 December 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(kableExtra)
```

We have a question, for which a proportion of the population would answer YES, and the remaining proportion would answer NO. 

For example, we might be interested in "Should Britain leave the EU?"

We call the proportion of YES people in the population $\pi$. For example, if our population was 60% YES people and 40% NO people, then we'd have $\pi = 0.6$. 

We obviously can't survey everyone in the entire population. Our goal is to conduct a survey - take a small subset of people from the population - and use this sample to try to work out what we think the population $\pi$ value is. So, we want to look at the proportion of YES people in our sample and use this to infer about the proportion of YES people in the population.

We'll call the size of our sample $n$, and the proportion of YES people in our sample $p$. So, if there are 100 people in our sample and 59 of them are yes people, we'd have $n=100$ and $p=0.59$.

Given that the following conditions are satisfied, then the number of YES people in our sample of size $n$ will follow a Binomial distribution:

* The population is large enough that every time we draw someone at random from it, the probability of that person being a YES person is $\pi$.

* Each individual in the population is equally likely to be chosen for the sample.

* Each observation is independent (i.e. someone answering YES in our sample does not affect the likelihood of anyone else in our sample saying NO or YES - the probability of each randomly drawn person is always $\pi$, regardless of what everyone else in the sample has answered)

* The population is at least 20 times larger than the sample. 

Call the number of people in our sample who answer YES $X$. For example, if we took a sample of $n=150$ people, and 12 answered YES, then we'd have 

$$x = 12$$
$$p = \frac{x}{n} = \frac{12}{150} = 0.08$$

Then, given that we've satisfied the assumptions above, we have the following probability for any chosen value $x$ of $X$:

$$Pr[X = x] \quad = \quad \binom{n}{x}\pi^x(1-\pi)^{n-x} \quad = \quad \frac{n!}{x!(n-x)!}\pi^x(1-\pi)^{n-x} \quad \text{... for x = 0,1,2,3,....}$$

For example, if we were taking many samples of size 5 (where probability of a YES is always 40%), we'd expect to see exactly 2 people answering YES in about 35% of the samples:

$$Pr[X=2] = \frac{5!}{2!\times 3!} \times 0.4^2 \times 0.6^3 = 0.3456$$

For example, if we simulate 10,000 different samples, we'd expect to see about 3,500 samples with exactly 2 YES answers:

```{r hellothere}
run_100 <- 
replicate( 10000,
           sample( c("Yes","No"), size=5, replace=TRUE, prob=c(0.4,0.6) )
         )

# count number of Yes answers:
table( apply( run_100, 2, function(x){ sum( x=="Yes" ) } ) )
```

If we take infinitely many samples (each of the same size $n$) and take the average value of $X$ across all of them, then we'd get the average value $E[X] = n\pi$. The variance of $X$ across all of the infinite samples would be $\sigma^2_X = n\pi(1-\pi)$. We'd get very close to these values by taking a lot of samples (we obviously don't have time to take infinitely many), where the approximation improves as we increase the number of samples. 

# Choosing a sample size ($n$)

What we want to work out is: how big a sample ($n$) do we need to have in order to be comfortable that our sample proportion of YES people $p$ is a good estimate of the population proportion of YES people $\pi$?

Given that we meet the following conditions:

* $n \pi \geq 10$

* $n \pi (1-\pi) \geq 10$

Then the distribution of $p$ across many samples is very well approximated by the distribution 

$$p \quad \sim \quad Normal \Big(\pi, \frac{\pi (1-\pi)}{n} \Big)$$

So, as long as we have a fairly large sample ($n$), and the proportion of YES people in the population is not too close to 0 or 1, we can use this normal distribution result to quantify how confident we can be in using our sample estimate $p$ to infer about the population proportion of YES people $\pi$.     

For example, suppose that in reality 40% of people in Britain want to leave the European Union.
Suppose that we survey 1000 people (i.e. take a sample of $n=1000$ people), asking these 1000 people if they want to leave the EU. In our sample, 434 of them say that they would like to leave the EU, which gives us a sample proportion of YES people $p=\frac{434}{1000}=0.434$

Because we know the population parameter $\pi=0.4$, we know that the distribution of 
$p$ is approximately

$$p \quad \sim \quad Normal \Big(0.4 \space ; \space \frac{0.4(0.6)}{1000}\Big)$$

So, we can say that in about 99% of samples, we'll see a $p$ bigger than 

```{r lower_bound}
qnorm( 0.005, mean=0.4, sd=sqrt(0.4*0.6/1000) )
```

and a $p$ lower than:

```{r uppabnd}
qnorm( 0.995, mean=0.4, sd=sqrt(0.4*0.6/1000) )
```

Let's see that this is true by simulating 100,000 samples (each of size $n=1000$) where the probability of drawing a YES person is always $\pi=0.4$:

```{r some_sim_in_R}
sim_100k_sample_proportions <- 
lapply( 1:100000,
  function(x) { 
    sum( 
        sample( c("Yes","No"), 
                replace=TRUE, 
                size=1000, 
                prob=c(0.4, 0.6) 
              ) == "Yes" 
                      
       ) / 1000
    }
) %>% 
  unlist()

# show the first 10 sample values of 'p': 
head(sim_100k_sample_proportions, 10)
```

And the percentage of samples in which $p$ falls inside our interval is:
```{r check_the_interval}
interval_lower <- qnorm( 0.005, mean=0.4, sd=sqrt(0.4*0.6/1000) )
interval_lower
interval_upper <- qnorm( 0.995, mean=0.4, sd=sqrt(0.4*0.6/1000) )
interval_upper
paste( 
    sum( sim_100k_sample_proportions >= interval_lower & 
         sim_100k_sample_proportions <= interval_upper 
       ) /
      length(sim_100k_sample_proportions) *100,
    "%"
)
```

Let's check that this also works if we have an extreme population proportion, like $\pi=0.03$: 

```{r some_sim_in_R2}
sim_100k_sample_proportions <- 
lapply( 1:100000,
  function(x) { 
    sum( 
        sample( c("Yes","No"), 
                replace=TRUE, 
                size=1000, 
                prob=c(0.03, 0.97) 
              ) == "Yes" 
                      
       ) / 1000
    }
) %>% 
  unlist()

# show the first 10 sample values of 'p': 
head(sim_100k_sample_proportions, 10)
```

And the percentage of samples in which $p$ falls inside our interval is:
```{r check_the_interval2}
interval_lower <- qnorm( 0.005, mean=0.03, sd=sqrt(0.4*0.6/1000) )
interval_lower
interval_upper <- qnorm( 0.995, mean=0.03, sd=sqrt(0.4*0.6/1000) )
interval_upper
paste( 
    sum( sim_100k_sample_proportions >= interval_lower & 
         sim_100k_sample_proportions <= interval_upper 
       ) /
      length(sim_100k_sample_proportions) *100,
    "%"
)
```

Here is a table of the 99% confidence intervals for sample proportion $p$ for different true population percentages $\pi\times100$ and different sample sizes $n$:

```{r table_of_99_percents}

expand.grid( true_proportion = seq(0.05,0.95,0.05),
             n = c(100,500,1000,2000,5000,10000)
             ) %>% 
  as_data_frame() %>% 
  mutate( sample_p_lwr0.99 = qnorm( 0.005, 
                                      mean=true_proportion, 
                                      sd=sqrt(true_proportion*(1-true_proportion)/n) 
                                    ),
          sample_p_upr0.99 = qnorm( 0.995, 
                                        mean=true_proportion, 
                                        sd=sqrt(true_proportion*(1-true_proportion)/n) 
                                    )
        ) %>% 
  # mutate( `sample p lwr 99%` = paste0( round(`sample p lwr 99%`*100,2), "%" ),
  #         `sample p upper 99%` = paste0( round(`sample p upper 99%`*100,2), "%" )  
  #       ) %>% 
  mutate( confints = paste0( "[",
                             round(sample_p_lwr0.99*100,2),
                             "% , ",
                             round(sample_p_upr0.99*100,2),
                             "%]"
                            )
         ) %>%
  select( true_proportion, n, confints ) %>% 
  mutate( true_percent = paste0( true_proportion*100, "%") ) %>% 
  mutate( n = paste0("n = ", n) ) %>% 
  spread( key = n, value = confints ) %>% 
  arrange( true_proportion ) %>% 
  select( true_percent, 
          "n = 100",
          "n = 500",
          "n = 1000",
          "n = 2000",
          "n = 5000",
          "n = 10000" ) %>% 
  kable(., "html") %>% 
  kable_styling("striped", full_width=FALSE)

```

We can get the same intervals using simulated data:

```{r sim_CI_0.99}

sim_sample_prop_confint <- function( n = 1000,
                                     n_sample = 500,
                                     true_proportion = 0.5,
                                     confidence = 0.99,
                                     lower_upper = 0    # 0 for lower bound, 1 for upper
                                    ){
  
  sample_prop_values <- 
      lapply( 1:n_sample,
              function(x){ data_frame( x = 
                                           sample( c(0,1), 
                                           size = n,
                                           replace = TRUE,
                                           prob = c(1-true_proportion, true_proportion)
                                                  )
                                       ) %>% 
                           summarise( proportion_yes = sum(x==1) / nrow(.) ) %>% 
                           pull(proportion_yes)
                         }  
            ) %>% 
    unlist()
  
  if( lower_upper == 0 ){ return(
    quantile( sample_prop_values, (1-confidence)/2 ) ) } else{
    return( quantile(sample_prop_values, 1-(1-confidence)/2 ) )  
    }
  
  # return( list( lower = quantile(sample_prop_values, (1-confidence)/2 ),
  #               upper = quantile(sample_prop_values, 1-(1-confidence)/2 )
  #             ) 
  #       )
}
  
# test the function:
sim_sample_prop_confint( n = 100,
                         n_sample = 100,
                         true_proportion = 0.3,
                         confidence = 0.99,
                         lower_upper = 1
                      )

# grid of parameter combinations to feed into function::
values_grid <- 
  expand.grid( true_proportion = seq(0.05,0.95,0.05),
               n = c(100,500,1000,2000,5000,10000)
             ) %>%
  as_data_frame() %>% 
  mutate( n_sample = 1e5L,
          confidence = 0.99,
          lower_upper = 0
        ) %>% 
  as.matrix()

head(values_grid)

cbind( values_grid, 
       lower0.99 = mapply( FUN = sim_sample_prop_confint,
                           values_grid[,"n"],
                           values_grid[,"n_sample"],
                           values_grid[,"true_proportion"],
                           values_grid[,"confidence"],
                           values_grid[,"lower_upper"]
                           ),
       upper0.99 = mapply( FUN = sim_sample_prop_confint,
                           values_grid[,"n"],
                           values_grid[,"n_sample"],
                           values_grid[,"true_proportion"],
                           values_grid[,"confidence"],
                           values_grid[,"lower_upper"]+1
                           )
     ) %>% 
  #head(10)
  as_data_frame() %>% 
  mutate( sample_p_lwr0.99 = lower0.99,
          sample_p_upr0.99 = upper0.99
        ) %>%
  # mutate( `sample p lwr 99%` = paste0( round(`sample p lwr 99%`*100,2), "%" ),
  #         `sample p upper 99%` = paste0( round(`sample p upper 99%`*100,2), "%" )  
  #       ) %>% 
  mutate( confints = paste0( "[",
                             round(sample_p_lwr0.99*100,2),
                             "% , ",
                             round(sample_p_upr0.99*100,2),
                             "%]"
                            )
         ) %>%
  select( true_proportion, n, confints ) %>% 
  mutate( true_percent = paste0( true_proportion*100, "%") ) %>% 
  mutate( n = paste0("n = ", n) ) %>% 
  spread( key = n, value = confints ) %>% 
  arrange( true_proportion ) %>% 
  select( true_percent, 
          "n = 100",
          "n = 500",
          "n = 1000",
          "n = 2000",
          "n = 5000",
          "n = 10000" ) %>% 
  kable(., "html") %>% 
  kable_styling("striped", full_width=FALSE)

```


# Confidence Interval for $\pi$

We can use results from the standard normal distribution to get a confidence interval for $\pi$:

Because $p$ is normally distributed, $\frac{p-\mu}{\sigma}$ will have a standard normal distribution ($Normal(0,1)$) - (where $\sigma=\sqrt{\frac{\pi(1-\pi)}{n}}$).

Because 99% of the density of a standard normal distribution lies between -2.575829 and 2.575829, we have that:

$$Pr\Big[ -2.575829 \leq \frac{p-\pi}{\sigma} \leq 2.575829 \Big] = 0.99$$

Re-arranging this, we can get a confidence interval for the population proportion $\pi$:

$$\begin{array}{lcl} Pr\Big[ -2.575829 \leq \frac{p-\pi}{\sigma} \leq 2.575829 \Big] &=& 0.99 \\
Pr\Big[ -2.575829\sigma \leq p-\pi \leq \sigma 2.575829 \Big] &=& 0.99 \\
Pr\Big[ -2.575829\sigma + p \geq \pi \geq \sigma 2.575829 + p \Big] &=& 0.99 \\
\end{array}$$

We don't know the variance $\sigma$ of the population parameter $\pi$, so the best that we can do is to estimate it as $\hat{\sigma} = \sqrt{\frac{p(1-p)}{n}}$ using our sample $p$

If we were to perform the same survey ($n=1000$ people in each survey) millions of times (this is a thought experiment), and in each of the samples we calculated a sample proportion $p$, and calculated the interval $\Big[p-2.575829\times \sqrt{\frac{p(1-p)}{1000}} \quad;\quad p-2.575829\times \sqrt{\frac{p(1-p)}{1000}} \Big]$, then in 99% of our samples, the true population proportion of YES people $\pi=0.4$ would lie within this interval. 

We can see this in R:

```{r label_the_options}
get_samples_100k <- 
 lapply( 1:1e5L,
        function(x){  
                      # draw random sample and calculate sample proportion 'p':    
                      sample_p <- 
                          sample( c("Yes","No"), 
                                  size=1000, 
                                  replace=TRUE,
                                  prob = c(0.4, 0.6)
                                ) %>% 
                          table(.) %>% 
                          prop.table(.) %>% 
                          .["Yes"]
                      
                      # calculate confidence interval and check that pi=0.4 lies within it:
                      interval_0.99_contains_0.4 <- 
                          0.4 >= sample_p - qnorm(0.995)*sqrt(sample_p*(1-sample_p)/1000) &
                          0.4 <= sample_p + qnorm(0.995)*sqrt(sample_p*(1-sample_p)/1000) 
                      
                      return(interval_0.99_contains_0.4)
          }  
        
      ) 

```

So, the proportion of our samples in which the sample 99% confidence interval contained $pi=0.4$ was:

```{r showmethechiknpi}
sum( unlist(get_samples_100k) ) / length(get_samples_100k)
```

Let's try this with a much more rare value of $\pi$, like $\pi=0.05$:

```{r more_risky}
get_samples_100k <- 
 lapply( 1:1e5L,
        function(x){  
                      # draw random sample and calculate sample proportion 'p':    
                      sample_p <- 
                          sample( c("Yes","No"), 
                                  size=1000, 
                                  replace=TRUE,
                                  prob = c(0.05, 0.95)
                                ) %>% 
                          table(.) %>% 
                          prop.table(.) %>% 
                          .["Yes"]
                      
                      # calculate confidence interval and check that pi=0.4 lies within it:
                      interval_0.99_contains_0.05 <- 
                          0.05 >= sample_p - qnorm(0.995)*sqrt(sample_p*(1-sample_p)/1000) &
                          0.05 <= sample_p + qnorm(0.995)*sqrt(sample_p*(1-sample_p)/1000) 
                      
                      return(interval_0.99_contains_0.05)
          }  
        
      ) 

```

So, the proportion of our samples in which the sample 99% confidence interval contained $\pi=0.05$ was:

```{r more_risky_the_result_the_movie}
sum( unlist(get_samples_100k) ) / length(get_samples_100k)
```




