---
title: "Shannon Information, Entropy, Mutual Information"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Shannon Information 
The *Shannon Information* $I(A)$ of an event $A$, which has probability $\theta$ of occurring i.e.

$$Pr \Big[A=a\Big] = \begin{cases} \theta \quad \text{for } a=1 \quad\text{i.e. event A occurs} \\ 1-\theta \quad \text{for } a=0 \quad\text{i.e. event A does not occur} \\ 0 \quad \text{otherwise} \end{cases}$$

..is defined as 

$$I(A) = log\Big(\displaystyle\frac{1}{\theta}\Big)=-log(\theta)$$ 

Here is a plot of this function for different probabilities $\theta$ and different log bases:

```{r fig.width=6, fig.height=6}
x_sequence <- seq(from=0, to=1, by=0.01)
plot( 
         x = x_sequence,
         y = -log(x_sequence, base=2),
      type = "l",
      xlab = bquote(theta~"(probability)"),
      ylab = bquote( -log[b](theta) ),
      main = "Shannon Information",
       col = "black"
    )
lines( 
        x = x_sequence,
        y = -log(x_sequence),
      col = "red"
    )
lines( 
        x = x_sequence,
        y = -log(x_sequence, base=10),
      col = "blue"
    )
legend(
                 "topright",
           col = c("black","red","blue"),
        legend = c("2","e","10"),
           lty = rep(1,3),
         title = "log base (b)" 
)
```

# Shannon Entropy 
For a random variable $X$ with probability distribution $p(x) = Pr\Big[X=x\Big]$, the *Shannon Entropy* is defined:

$$\begin{array}{lcl}
H(X) &=& E\Big[ I(X)\Big] \\
&=& -\displaystyle\sum_x p(x) \cdot log\Big[p(x)\Big] \\
\end{array}$$

The *Shannon Entropy* is the expected *Shannon Information* (average *Shannon Information*), averaged over the possible outcomes of $X$.

Here are the entropies of a few different random variables:

```{r fig.height=8, fig.width=12}

distribution_examples <- 
  list( 
          c( 0.5, 0.5 ),
          c( 0.6, 0.4 ),
          c( 0.1, 0.9 ),
          c( 0.4, 0.3, 0.3 ),
          c( 0.6, 0.2, 0.2 ),
          c( 0.6, 0.3, 0.1 ),
          c( 0.3, 0.3, 0.2, 0.1, 0.1),
          c( 0.6, 0.1, 0.1, 0.1, 0.1),
          c( 0.99, 0.0025, 0.0025, 0.0025, 0.0025)
    )

bind_rows(
            tibble( 
              distribution = lapply( distribution_examples,
                                     function(x) paste(x, collapse=" ; ") 
                                   ) %>% unlist(),
              log_base = "log_base=2",
              shannon_entropy = lapply( distribution_examples,
                                        function(x) -sum( x * log(x, base=2) ) 
                                      ) %>% unlist()
              )
            ,
            tibble( 
              distribution = lapply( distribution_examples,
                                     function(x) paste(x, collapse=" ; ") 
                                   ) %>% unlist(),
              log_base = "log_base=e",
              shannon_entropy = lapply( distribution_examples,
                                        function(x) -sum( x * log(x, base=exp(1)) ) 
                                      ) %>% unlist()
              )
            ,
            tibble( 
              distribution = lapply( distribution_examples,
                                     function(x) paste(x, collapse=" ; ") 
                                   ) %>% unlist(),
              log_base = "log_base=10",
              shannon_entropy = lapply( distribution_examples,
                                        function(x) -sum( x * log(x, base=10) ) 
                                      ) %>% unlist()
              )
            
            
  ) %>% 
  
  ggplot( data = .,
          aes( x = distribution,
               y = shannon_entropy,
               label = round(shannon_entropy, digits=2),
               fill = log_base
             )
        ) + 
  geom_bar( stat = "identity", position="dodge" ) +
  geom_text() + 
  coord_flip() + 
  facet_wrap( ~log_base, ncol=1, scales="free_x" )

```

Notice that random variables with a larger number of possible outcomes have higher entropy, and also that more uniform outcome probabilities result in higher entropy.  

Notice also that changing the base of the log function used does not change the relative entropies of the different outcomes of the random variable, but only the relative sizes of the entropy numbers.  

**Entropy** $H(X)$ measures the variability of a random variabile $X$. 



# Mutual Information






