---
title: "Survival Analysis & Survival Models"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(tidyverse)
library(mixtools)         # for fitting mixture distributions
```

# A discrete example

$\space$ | $\space$
---------|----------------
T        | Age (band) at which death will occur. A random variable.
t        | A chosen value/outcome of random variable $T$ 
f(t)     | $Pr[T=t]$
$\space$ | Probability of dying while at age (band) $t$.
$\space$ | Alternatively, proportion of population dying at age (band) $t$
F(t)     | $Pr\Big[T<t \Big]$ 
$\space$ | Probability of dying prior to reaching age (band) $t$
$\space$ | Alternatively, proportion of population dying before reaching age (band) $t$
S(t)     | $Pr \Big[T\geq t \Big]$
$\space$ | Probability of surviving to age (band) $t$, or older
$\space$ | Alternatively, proportion of population surviving to at least age (band) $t$
$\space$ | $S(t)=1-F(t)$
h(t)     | $Pr\Big[T=t \hspace{3mm} \Bigl| \hspace{3mm} T \geq t \Big]  \quad=\quad \displaystyle\frac{f(t)}{S(t)}$
$\space$ | Probability of dying at age (band) $t$ given that you've reached this age
$\space$ | Alternatively, proportion of population reaching age (band) $t$ that die at age (band) $t$

```{r echo=FALSE}
set.seed(1987)

deathdata <-
  tibble( age_lwr = c( 0, 10, 20, 30, 40, 50  ),
          age_upr = c( 9, 19, 29, 39, 49, 59 ),
          t =             c("0-9",
                            "10-19",
                            "20-29",
                            "30-39",
                            "40-49",
                            "50+"
                           )
        ) %>%
  mutate( `f(t)` =
            c( 6, 4, 1, 2, 8, 10 )
          ,
          `f(t)` =
              `f(t)` / sum(`f(t)`)
        ) %>% 
  mutate( `F(t)` = cumsum(`f(t)`) - `f(t)`,
          `S(t)` = 1 - `F(t)`,
          `h(t)` = `f(t)`/`S(t)`
        )

deathdata %>% select(-age_lwr, -age_upr)

sum(deathdata$`f(t)`)
```

```{r echo=FALSE, fig.width=10, fig.height=4}
deathdata %>% 
  pivot_longer(     cols = 4:7,
                names_to = "metric",
                values_to = "value"
              ) %>% 
  mutate( metric = factor( metric, levels = c("f(t)","F(t)","S(t)","h(t)") ) ) %>% 
  ggplot( data = .,
          aes(     x = t,
                   y = value,
                fill = metric,
               group = metric
             )
        ) +
  geom_bar( stat="identity", width=0.3 ) +
  geom_line() +
  geom_point( size = 3 ) + 
  coord_cartesian( ylim = c(0,1) ) + 
  # geom_hline( yintercept = 1 ) + 
  facet_grid( .~metric, scales="free_y" ) +
  theme( legend.position = "none", axis.text.x=element_text(angle=90) )
```

# Continuous Survival Theory

Random variable $T$ is the time until **event**. Let's call the event **death**, in which case $T$ is **survival time**.

Let's consider $T$ the number of years from birth until death (i.e. length of life) for a single person.  

$T$ is a non-negative continuous random variable.

The distribution of $T$ is described by a density function $f(t)$. 

The cumulative distribution function $\quad F(t)=Pr\Big[T<t\Big]=\int_{0}^t f(x) dx \quad$ gives the probability that the **event** (death) has happened by time $t$.

The **survival function** $S(t)$ is $\quad S(t)=1-F(t)=Pr\Big[T\geq t\Big]=\int_{t}^\infty f(x) dx$.

$S(t)$ gives the probability that the **event** has not yet happened by time $t$ (the probability of surviving until time $t$).

For a person alive at time $t$: the probability of them dying in the next $\Delta t$ years is

$$\begin{array}{lcl}
Pr\Big[ \quad T \in [t, \space t+\Delta t)  \quad \Bigl| \quad T\geq t \quad \Big]
&=& \displaystyle\frac{Pr\Big[ \quad T \in [t, \space t+\Delta t)  \quad \cap \quad T\geq t \quad\Big]}{Pr\Big[\quad T \geq t \quad \Big]} \hspace{5mm} \text{axiom of probability} \\
&=& \displaystyle\frac{Pr\Big[ \quad T \in [t, \space t+\Delta t)  \quad \cap \quad T\geq t \quad\Big]}{Pr\Big[\quad T \geq t \quad \Big]} \hspace{5mm} \text{axiom of probability} \\
&\approx& x \\
\end{array}$$

The **hazard function** $h(t)$ is:

$$\begin{array}{lcl}
h(t) &=& \underset{dt\rightarrow{}0}{lim} \displaystyle\frac{Pr\Big[t \leq T \leq t+dt  \quad \Bigl| \quad T\geq t\Big]}{dt} \\
&=& \displaystyle\frac{f(t)}{S(t)} \\
&=& -\displaystyle\frac{d}{dt} \Bigg\{ log \Big[S(t)\Big] \Bigg\} \\
\end{array}$$


# A quick simulation illustrating these principles
Suppose that the distribution of **age at death** in our population is:

```{r echo=FALSE}
# create a population based on the discrete (age band) data, 
# and fit a weibull distribution to this simulated population: 
set.seed(1769)
sim_population <- 
  lapply( 1:nrow(deathdata),
          function(rownum){ rnorm(    n = round( deathdata$`f(t)`[rownum]*1000 ),
                                   mean =
                                     (deathdata$age_upr[rownum]+deathdata$age_lwr[rownum])/2,
                                     sd = 5
                                 ) %>% 
                              abs(.)
                          }
        ) %>% 
    reduce( c ) 

fit_mixture_dbn <- 
  mixtools::gammamixEM( x = sim_population, k=2 )

mixgamma_1_mix_proportion <- fit_mixture_dbn$lambda[1]
mixgamma_1_shape <- fit_mixture_dbn$gamma.pars[1,1]
mixgamma_1_scale <- fit_mixture_dbn$gamma.pars[2,1]

mixgamma_2_mix_proportion <- fit_mixture_dbn$lambda[2]
mixgamma_2_shape <- fit_mixture_dbn$gamma.pars[1,2]
mixgamma_2_scale <- fit_mixture_dbn$gamma.pars[2,2]

f_ <- function(t){ mixgamma_1_mix_proportion * 
                       dgamma(     x = t,
                               shape = mixgamma_1_shape,
                               scale = mixgamma_1_scale
                              ) +
                   mixgamma_2_mix_proportion * 
                       dgamma(     x = t,
                               shape = mixgamma_2_shape,
                               scale = mixgamma_2_scale
                              )
                }

F_ <- function(t){ mixgamma_1_mix_proportion * 
                       pgamma(     q = t,
                               shape = mixgamma_1_shape,
                               scale = mixgamma_1_scale
                              ) +
                   mixgamma_2_mix_proportion * 
                       pgamma(     q = t,
                               shape = mixgamma_2_shape,
                               scale = mixgamma_2_scale
                       )
}

S_ <- function(t){ 1 - F_(t) }

hist(      x = sim_population, 
      breaks = 40,
        xlab = "Age",
        main = "Distribution of Age at Death in Population",
        freq = FALSE,
        xlim = c(0, 80)
      
    )

lines(   x = 0:80,
         y = sapply( 0:80, function(t){ f_(t) } ),
       col = 2
 )

```

The fitted distribution (red line) is a mixture of Gamma distributions that has been fit to this data using the package **mixtools**. The fit isn't perfect, but is good enough for purposes of illustration. 

```{r echo=FALSE, height=7, width=7}
par( mfrow=c(4,1), mar=c(1.8,5,3,1) )
plot( x=1:55, y=sapply(1:55, function(t){ f_(t) }), xlab="t", ylab="f(t)", type="l", las=1, col=2)
plot( x=1:55, y=sapply(1:55, function(t){ F_(t) }), xlab="t", ylab="F(t)", type="l", las=1, col=2)
plot( x=1:55, y=sapply(1:55, function(t){ S_(t) }), xlab="t", ylab="S(t)", type="l", las=1, col=2)
plot( x=1:55, y=sapply(1:55, function(t){ f_(t)/S_(t) }), xlab="t", ylab="f(t) / S(t)", type="l", las=1, col=2, ylim=c(0, 0.05) )
```

# Resources
1. https://data.princeton.edu/wws509/notes/c7.pdf

2. https://cran.r-project.org/web/packages/mixtools/vignettes/mixtools.pdf

3. https://stats.stackexchange.com/questions/218947/intuition-behind-the-hazard-rate

4. https://en.wikipedia.org/wiki/Failure_rate