---
title: "MVA_biplots"
author: "Joe Bolton"
date: "17 March 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

Information from: *Biplots in Practice* by *Michael Greenacre* (available for free in it's entirety online)

# Biplots

A biplot is like a scatterplot, but for visualising a dataset with more than 2 dimensions.

Biplots have as many axes as there are variables. 

Like a scatterplot, points are still projected perpendicularly onto an axis in order to read off their value on that particular axis.

A biplot is necessarily a reduced-dimensional approximation to the original data - data in many dimensions is reduced to 2 dimensions.

For example, suppose that we have the following data, with 5 observations (5 rows) of 4 variables (4 columns):

```{r littleY, echo=FALSE}
Y <- matrix( c(  8,  2,  2,  -6,
                 5,  0,  3,  -4,
                -2, -3,  3,   1,
                 2,  3, -3,  -1,
                 4,  6, -6,  -2
              ),
             byrow=TRUE,
             ncol=4 )
Y
```

This table of data can be visualised in the following symmetric biplot:

(get the value of an observation/row on a variable/column by projecting it perpendicularly onto that variable/column vector)

(e.g. row 4 takes on the value 3 on column 2, the value -3 on column 3 and value 2 on column 1)

```{r get_hat_Y, fig.width=10, fig.height=10, echo=FALSE}
u <- svd(Y)$u
d <- svd(Y)$d
v <- svd(Y)$v

# u %*% diag( c(d[1:2],0,0) ) %*% t(v)


left_matrix <- u[,1:2]%*%diag(sqrt(d[1:2]))
right_matrix <- diag(sqrt(d[1:2]))%*%t(v[,1:2])

par( pty = "s" )
# draw the biplot:
plot( x = c(-3,3 ),
      y = c(-3,3), 
      type = "n",
      xlab = "",
      ylab = "",
      axes = FALSE
    )
abline( v=0, h=0)

# draw in observations (points):
points( x = left_matrix[,1],
        y = left_matrix[,2],
        pch = 16,
        col = 3
      )

# draw in point labels:
text( x = left_matrix[,1],
      y = left_matrix[,2],
      labels = paste("row", 1:nrow(left_matrix), sep=""),
      pos = 4,
      col = 3
    )

# draw in axes:
arrows( x0 = rep(0, ncol(right_matrix) ),
        y0 = rep(0, ncol(right_matrix) ),
        x1 = right_matrix[1,],
        y1 = right_matrix[2,],
        length = 0.1,
        col = 2
      )
# draw in vector labels:
text( x = right_matrix[1,],
      y = right_matrix[2,],
      labels = paste("col", 1:ncol(right_matrix), sep=""),
      pos = 3,
      col = 2
      )

# draw in tickmarks on the right_matrix vectors:
for( col_j in 1:ncol(right_matrix) ){
  
  vec <- right_matrix[,col_j]
  length_of_vec <- sqrt(sum(vec^2))
  unit_vec <- vec / length_of_vec
  x_tick_coords <- -8:8 * (unit_vec/length_of_vec)[1]   
  y_tick_coords <- -8:8 * (unit_vec/length_of_vec)[2]   
  points( x = x_tick_coords,
          y = y_tick_coords,
          pch = "|",
          cex = 0.5
        )
  text( x = x_tick_coords,
        y = y_tick_coords,
        labels = -8:8,
        pos = 1
      )
}

```

Notice also that the directions of the vectors (representing columns/variables) indicate the correlation relationships between these variables/columns:

```{r showthecorridge}
cor(Y)
```

# Geometry of the biplot
Biplots are defined as the decomposition of a target matrix $\mathbf{S}$ into the product of a *left matrix* and a *right matrix*:

$$\mathbf{S} \quad = \quad \mathbf{X}\mathbf{Y}^T$$

The target matrix $\mathbf{S}$ is the data matrix which we would like to represent in the biplot. 

e.g.

$$\begin{array}{lcl}
\underset{\text{target matrix}}{
\begin{bmatrix}
8  & 2 & 2 & -6 \\
5  & 0 & 3 & -4   \\
-2 &-3 & 3 &  1   \\
 2 & 3 &-3 & -1  \\
4  & 6 &-6 & -2 \\
\end{bmatrix} 
}
&=&
\underset{\text{left matrix}}{
\begin{bmatrix}
2 & 2\\
1 & 2\\
-1 & 1\\
1 & -1\\
2 & -2 \\
\end{bmatrix}
}
\underset{\text{right matrix}}{
\begin{bmatrix}
3 & 2 & -1 & -2\\
1 & -1 & 2 & -1 \\
\end{bmatrix}
} \\
\underset{\text{target matrix}}{
\begin{bmatrix}
\mathbf{x}_1^T\mathbf{y}_1&\mathbf{x}_1^T\mathbf{y}_2&\mathbf{x}_1^T\mathbf{y}_3&\mathbf{x}_1^T\mathbf{y}_4\\
\mathbf{x}_2^T\mathbf{y}_1&\mathbf{x}_2^T\mathbf{y}_2&\mathbf{x}_2^T\mathbf{y}_3&\mathbf{x}_2^T\mathbf{y}_4\\
\mathbf{x}_3^T\mathbf{y}_1&\mathbf{x}_3^T\mathbf{y}_2&\mathbf{x}_3^T\mathbf{y}_3&\mathbf{x}_1^T\mathbf{y}_4\\
\mathbf{x}_4^T\mathbf{y}_1&\mathbf{x}_4^T\mathbf{y}_2&\mathbf{x}_4^T\mathbf{y}_3&\mathbf{x}_4^T\mathbf{y}_4\\
\mathbf{x}_5^T\mathbf{y}_1&\mathbf{x}_5^T\mathbf{y}_2&\mathbf{x}_5^T\mathbf{y}_3&\mathbf{x}_5^T\mathbf{y}_4\\
\end{bmatrix}}
&=&
\underset{\text{left matrix}}{
\begin{bmatrix}
\mathbf{x}_1^T \\
\mathbf{x}_2^T \\
\mathbf{x}_3^T \\
\mathbf{x}_4^T \\
\mathbf{x}_5^T \\
\end{bmatrix}}
\underset{\text{right matrix}}{
\begin{bmatrix}
\mathbf{y}_1 & \mathbf{y}_2 & \mathbf{y}_3 & \mathbf{y}_4 
\end{bmatrix}}
\end{array}$$

We can plot this data, with the rows of the *left matrix* as points in the cartesian plane, and the columns of the *right matrix* as vectors in the same plane:

```{r do_it_bitch, fig.width=10, fig.height=5}
par( pty="s" )

left_matrix <- 
        matrix( c( 2,  2,
                   1,  2,
                  -1,  1,
                   1, -1,
                   2, -2
                  ),
                byrow=TRUE,
                ncol=2) 
right_matrix <- 
         matrix( c(3,  2, -1, -2,
                   1, -1,  2, -1
                   ),
                 byrow=TRUE,
                 nrow=2
               )
left_matrix%*%right_matrix

# make a blank plot:
plot( x = c(-3, 4, 0, 0), 
      y = c( 0, 0, -3, 3),
      type = "n",
      axes = FALSE,
      xlab = "", ylab=""
    )
abline( h=0, v=0, lty=2 )
points( x=0, y=0, cex=2 )

for( row_i in 1:nrow(left_matrix) ){
  
  points( x = left_matrix[row_i, 1],
          y = left_matrix[row_i, 2],
          pch = 16
        )
  text( x = left_matrix[row_i, 1],
        y = left_matrix[row_i, 2],
        labels = paste0("x",row_i),
        col = 2,
        pos = 4
      )
}

for( col_j in 1:ncol(right_matrix) ){
  
  arrows( x0 = 0, y0 = 0,
          x1 = right_matrix[1, col_j],
          y1 = right_matrix[2, col_j],
          length = 0.1      # arrowhead length
        )
  points( x = right_matrix[1, col_j],
          y = right_matrix[2, col_j],
          pch=16
        )
  text( x = right_matrix[1, col_j],
        y = right_matrix[2, col_j],
        labels = paste0("y", col_j),
        pos = 3
      )
  
}

axis( side = 1, at = -5:5 )
axis( side = 2, at = -5:5 )

left_matrix
right_matrix

```

We wish to project vector $\mathbf{x}_1$ onto vector $\mathbf{y}_1$. 

The length of $\mathbf{x}_1$ is

$$||\mathbf{x}_1|| \quad = \quad \sqrt{(2-0)^2+(2-0)^2} \quad = \quad \sqrt{8}$$

Using the result $cos\theta = \frac{\mathbf{u} \cdot \mathbf{v}}{||\mathbf{u}||\space\space||\mathbf{v}||}$, the angle formed by $\mathbf{x}_1$ and $\mathbf{y}_1$ is

$$cos^{-1}\Bigg[\frac{\mathbf{x}_1 \cdot \mathbf{y}_1}{||\mathbf{x}_1||\space\space||\mathbf{y}_1||}\Bigg]\quad = \quad 0.4636476 \text{ radians}$$
```{r showquicks}
x1 <- left_matrix[1,]
y1 <- right_matrix[,1]
acos( c(t(x1)%*%y1) / (sqrt(sum(x1^2))*sqrt(sum(y1^2)) ) )
```

This means that the projection of $\mathbf{x}_1$ onto $\mathbf{y}_1$ has length 

$$||\mathbf{x}_1||\space \Bigg(\frac{\mathbf{x}_1^T\mathbf{y}_1}{||\mathbf{x}_1||\space||\mathbf{y}_1||}\Bigg) \quad = \quad ||\mathbf{x}_1||\space cos\Big(\theta\Big) \quad = \quad \sqrt{8}\space cos\Big(0.4636476\Big) \quad = \quad 2.529822 $$

and lies along the vector $\mathbf{y}_1$

So, the co-ordinate of the point at which the projection from $\mathbf{x}_1$ hits the vector $\mathbf{y}_1$  is $||\mathbf{x}_1||\space cos\Big(\theta\Big)$ times unit-length $\mathbf{y}_1$:

$$\sqrt{8}\space cos\Big(0.4636476\Big) \space \frac{\mathbf{y}_1}{||\mathbf{y}_1||} \quad = \quad 2.529822 \begin{bmatrix}\frac{3}{\sqrt{3^2+1^2}}\\ \frac{1}{\sqrt{3^2+1^2}} \end{bmatrix} \quad = \quad \begin{bmatrix}2.4 \\ 0.8\end{bmatrix}$$

```{r dotheshitshow, echo=FALSE}
the_proj <- c(2.4, 0.8)
  
par( pty="s" )
# make a blank plot:
plot( x = c(-3, 4, 0, 0), 
      y = c( 0, 0, -3, 3),
      type = "n",
      axes = FALSE,
      xlab = "", ylab=""
    )
abline( h=0, v=0, lty=2 )
points( x=0, y=0, cex=2 )

for( row_i in 1:nrow(left_matrix) ){
  
  points( x = left_matrix[row_i, 1],
          y = left_matrix[row_i, 2],
          pch = 16
        )
  text( x = left_matrix[row_i, 1],
        y = left_matrix[row_i, 2],
        labels = paste0("x",row_i),
        col = 2,
        pos = 4
      )
}

for( col_j in 1:ncol(right_matrix) ){
  
  arrows( x0 = 0, y0 = 0,
          x1 = right_matrix[1, col_j],
          y1 = right_matrix[2, col_j],
          length = 0.1      # arrowhead length
        )
  points( x = right_matrix[1, col_j],
          y = right_matrix[2, col_j],
          pch=16
        )
  text( x = right_matrix[1, col_j],
        y = right_matrix[2, col_j],
        labels = paste0("y", col_j),
        pos = 3
      )
  
}

axis( side = 1, at = -5:5 )
axis( side = 2, at = -5:5 )

segments( x0 = 0, y0 = 0,
          x1 = the_proj[1],
          y1 = the_proj[2],
          lwd=2,
          col=3
        )

segments( x0 = x1[1], y0 = x1[2],
          x1 = the_proj[1],
          y1 = the_proj[2],
          lty=3
        )

segments( x0 = 0, y0 = 0,
          x1 = x1[1], y1 = x1[2],
          lty = 3
          )

text( x=1, y=0.6,
      labels = expression(theta)
      )

points( x=the_proj[1], y=the_proj[2], pch=16 )
text( x=the_proj[1], y=the_proj[2], labels="(2.4, 0.8)", pos=1, cex=0.7)
```

So, the co-ordinate of the point (2.4, 0.8) at which the projection from $\mathbf{x}_1$ hits $\mathbf{y}_1$ we found using:

$$\begin{array}{lcl} \text{point co-ordinate} &=& \underset{||\mathbf{x}_1||cos(\theta)}{(\text{length of projection on } \mathbf{y}_1)} \times \underset{\frac{\mathbf{y_1}}{||\mathbf{y}_1||}}{(\text{unit vector in direction of } \mathbf{y}_1)} \\
&=& ||\mathbf{x}_1||\space \Bigg(\frac{\mathbf{x}_1^T\mathbf{y}_1}{||\mathbf{x}_1||\space||\mathbf{y}_1||}\Bigg) \frac{\mathbf{y}_1}{||\mathbf{y}_1||}\\
&=& \begin{bmatrix}2.4\\0.8\end{bmatrix}\end{array}$$

If we take the length of our projection vector (the vector joining $(0,0)$ to $(2.4,0.8)$ ) and multiply it by the length of $\mathbf{y}_1$, then we will get to the value $\mathbf{x}_1^T\mathbf{y}_1$ in the target matrix:

$$\text{length of projection} \times \text{length of }\mathbf{y}_1 \quad = \quad \sqrt{8}\space cos\Big(0.4636476\Big) \times \sqrt{10} \quad = \quad \sqrt{2.4^2+0.8^2} \times \sqrt{10}\quad = \quad 8$$

This is the standard dot product result:

$$\mathbf{x}^T\mathbf{y} \quad = \quad ||\mathbf{x}|| \space cos\Big(\theta\Big) \space ||\mathbf{y}|| $$

This means that if simply make the tickmarks on the $\mathbf{y}_1$ vector $\frac{1}{||\mathbf{y}_1||}$ units apart rather 1 unit apart, then projections of $\mathbf{x}_i$ onto $\mathbf{y}_1$ will give values from the **target matrix**. 

So, if the tickmarks along $\mathbf{y}_1$ are of length $\frac{1}{\sqrt{10}}=0.3162278$, then the length of the projection of $\mathbf{x}_1$ onto $\mathbf{y}_1$ being 2.529822 is in fact  $2.529822\times\frac{1}{1/\sqrt{10}}=8$ tickmarks/units along $\mathbf{y}_1$ (which corresponds to the value of row 1 and column 1 in the original target matrix).

Here are all of the $\mathbf{X}$ projections onto $\mathbf{y}_1$ - the vector algebra stays exactly the same: all that is changed is the spacing of the tickmarks on $\mathbf{y}_1$. 

```{r plot_more}
par( pty="s" )
# make a blank plot:
plot( x = c(-3, 4, 0, 0), 
      y = c( 0, 0, -3, 3),
      type = "n",
      axes = FALSE,
      xlab = "", ylab=""
    )
abline( h=0, v=0, lty=2 )
points( x=0, y=0, cex=2 )

for( row_i in 1:nrow(left_matrix) ){
  
  points( x = left_matrix[row_i, 1],
          y = left_matrix[row_i, 2],
          pch = 16
        )
  text( x = left_matrix[row_i, 1],
        y = left_matrix[row_i, 2],
        labels = paste0("x",row_i),
        col = 2,
        pos = 4
      )
}

arrows( x0 = 0, y0 = 0,
          x1 = right_matrix[1,1], y1=right_matrix[2,1] 
        )

arrows( x0 = 0, y0 = 0,
        x1 = -right_matrix[1,1], y1=-right_matrix[2,1],
        length = 0.1     # arrowhead length
      )

text( x = right_matrix[1,1],
      y = right_matrix[2,1],
      pos = 3,
      labels = "y1"
    )
 
axis( side = 1, at = -5:5 )
axis( side = 2, at = -5:5 )

points( x = -8:8 * 3/sqrt(10)*(1/1/sqrt(10)),
        y = -8:8 * 1/sqrt(10)*(1/1/sqrt(10)),
        pch = "|",
        cex = 0.5
        )

text( x = -8:8 * 3/sqrt(10)*(1/1/sqrt(10)),
      y = -8:8 * 1/sqrt(10)*(1/1/sqrt(10)),
      labels = -8:8,
      pos = 1,
      cex = 0.5
     )

# draw all of the projection lines:
for( row_i in 1:nrow(left_matrix)){

  x_i <- left_matrix[row_i,]
  
  projection_point <- c( t(x_i)%*%y1 ) * y1 / sum(y1^2) 
  
  segments( x0 = left_matrix[row_i,1],
            y0 = left_matrix[row_i,2],
            x1 = projection_point[1],
            y1 = projection_point[2],
            col = 2,
            lty = 2
          )
}

```

Here is the original target matrix for reference:

```{r theoriginaltarget, echo=FALSE}
rownames(Y) <- paste("x", 1:5, sep="" )
colnames(Y) <- paste("y", 1:4, sep="" )
Y
```

In this example, both the *left matrix* (providing points to plot) and the *right matrix* (providing vectors to plot) were 2-dimensional. This allowed us to make the biplot on the 2-dimensional plane. 

In practice, we will not be able to find a decomposition of the *target matrix* into two 2-dimensional matrices like this - the number of dimensions needed in the *left matrix* and the *right matrix* is in fact the *rank* of the *target matrix*. In order to produce a 2-dimensional biplot, we will need to reduce the dimensionality of the target matrix. 

# Reduced Dimension/Reduced Rank biplots using SVD
Suppose that we have a data matrix $\mathbf{Y}$ with $n$ observations on $p$ variables that we wish to visualise in a biplot. 

For example, suppose that we want to make a biplot for the following dataset $\mathbf{Y}$:

```{r Yme}
set.seed(13)
tempmat <- matrix( runif( 16, -5,5 ), ncol=4 )
Y <- MASS::mvrnorm( n = 20, 
                    mu = sample( -20:20, size=4 ),
                    Sigma = t(tempmat)%*%tempmat
                  ) %>% 
  scale()
Y

car::scatterplotMatrix(Y, col=1, regLine=list(col=4, lwd=1), smooth=FALSE)
```

Notice that I have standardised the data (to have mean 0 and variance 1). 

This matrix has rank 

```{r ranklieutentnat}
qr(Y)$rank
```

Meaning that we must find a 2-rank approximation of it in order to construct a biplot of it. 

Suppose that our $n\times p$ matrix has rank $r$ (in our case $20\times4$ and r=`r qr(Y)$rank`). 

In order to make a 2-dimensional biplot, we need to find a matrix $\hat{\mathbf{Y}}$ with the same dimension ($n\times p$=$20\times 4$) as $\mathbf{Y}$, but of rank $r=2$.

The best approximation $\hat{\mathbf{Y}}$, where 'best' is measured using the sum of squared element-wise differences between $\hat{\mathbf{Y}}$ and $\mathbf{Y}$, is achieved by using the Singular Value Decomposition (SVD) of $\mathbf{Y}$. The SVD of $Y$ is:

$$\begin{array}{lcl} \mathbf{Y} &=& \mathbf{U}\mathbf{\Lambda}\mathbf{V}^T\\ \end{array}$$
```{r svd_in_R}
u <- svd(Y)$u
d <- svd(Y)$d
v <- svd(Y)$v

Y

# which is the same as:
u %*% diag(d) %*% t(v)
```

For the rank 2 approximation, we replace $\Lambda$:

```{r lambstewda}
diag(d)
```

with 

```{r hmm}
diag(d)[1:2,1:2]
```

and we also only use the first 2 columns of $\mathbf{U}$ and the first 2 columns of $\mathbf{V}$. 

So, where the SVD of our data matrix $\mathbf{Y}$ was 

$$\mathbf{Y} \quad = \quad
\underset{\mathbf{U}}{\begin{bmatrix}*&*&.&.\\*&*&.&.\\*&*&.&.\\*&*&.&.\\*&*&.&.\\*&*&.&.\\*&*&.&.\\*&*&.&.\\\end{bmatrix}}\underset{\Lambda}{\begin{bmatrix}\lambda_1&0&0&0\\
                                    0&\lambda_2&0&0\\
                                    0&0&\lambda_3&0\\
                                    0&0&0&\lambda_4\\
                                    \end{bmatrix}}
\underset{\mathbf{V}^T}{\begin{bmatrix}*&*&*&*\\*&*&*&*\\.&.&.&.\\.&.&.&.\end{bmatrix}}$$

the rank-2 approximation of $\mathbf{Y}$ is then:

$$\mathbf{\hat{Y}} \quad = \quad \begin{bmatrix}*&*\\*&*\\*&*\\*&*\\*&*\\*&*\\*&*\\*&*\\\end{bmatrix}
\begin{bmatrix}\lambda_1&0\\
                0&\lambda_2
                \end{bmatrix}
\begin{bmatrix}*&*&*&*\\*&*&*&*\end{bmatrix}$$

which is the same as 

$$\mathbf{\hat{Y}} \quad = \quad
\begin{bmatrix}*&*&.&.\\*&*&.&.\\*&*&.&.\\*&*&.&.\\*&*&.&.\\*&*&.&.\\*&*&.&.\\*&*&.&.\\\end{bmatrix}\begin{bmatrix}\lambda_1&0&0&0\\
                                    0&\lambda_2&0&0\\
                                    0&0&0&0\\
                                    0&0&0&0\\
                                    \end{bmatrix}
\begin{bmatrix}*&*&*&*\\*&*&*&*\\.&.&.&.\\.&.&.&.\end{bmatrix}$$

```{r quickcheck}
u %*% diag( c(d[1:2],0,0) ) %*% t(v)
u[,1:2] %*% diag( d[1:2] ) %*% t(v[,1:2])
```

However, the first form can be written like this:

$$\mathbf{\hat{Y}} \quad = \quad \begin{bmatrix}*&*\\*&*\\*&*\\*&*\\*&*\\*&*\\*&*\\*&*\\\end{bmatrix}
\begin{bmatrix}\sqrt{\lambda_1}&0\\
                0&\sqrt{\lambda_2}
                \end{bmatrix}
\begin{bmatrix}\sqrt{\lambda_1}&0\\
                0&\sqrt{\lambda_2}
                \end{bmatrix}
\begin{bmatrix}*&*&*&*\\*&*&*&*\end{bmatrix}$$

which gives us the *left matrix* and *right matrix* for our biplot:

$$\mathbf{\hat{Y}} \quad = \quad \underset{\text{left matrix}}{\Bigg( \begin{bmatrix}*&*\\*&*\\*&*\\*&*\\*&*\\*&*\\*&*\\*&*\\\end{bmatrix}
\begin{bmatrix}\sqrt{\lambda_1}&0\\
                0&\sqrt{\lambda_2}
                \end{bmatrix}\Bigg)}
                \quad \quad \quad  
                \underset{\text{right matrix}}{
                \Bigg(
\begin{bmatrix}\sqrt{\lambda_1}&0\\
                0&\sqrt{\lambda_2}
                \end{bmatrix}
\begin{bmatrix}*&*&*&*\\*&*&*&*\end{bmatrix}
\Bigg)}$$

This is called a *symmetric biplot*, since the matrix of singular values is spread equally between the *left matrix* and the *right matrix*

Here is this biplot plotted manually in R: 

```{r do_the_whole_bipot, fig.width=10, fig.height=10}

left_matrix <- u[,1:2]%*%diag(sqrt(d[1:2]))
right_matrix <- diag(sqrt(d[1:2]))%*%t(v[,1:2])

par( pty = "s" )
# draw the biplot:
plot( x = c(-2,2 ),
      y = c(-2,2), 
      type = "n",
      xlab = "",
      ylab = "",
      axes = FALSE
    )

# draw in observations/points (left matrix) as text:
text( x = left_matrix[,1],
      y = left_matrix[,2],
      labels = 1:nrow(left_matrix),
      #pos = 4,
      col = 2
    )

# draw in vectors (right matrix):
arrows( x0 = rep(0, ncol(right_matrix) ),
        y0 = rep(0, ncol(right_matrix) ),
        x1 = right_matrix[1,],
        y1 = right_matrix[2,],
        length = 0.1,
        col = 3:6
      )
# draw in vector labels:
text( x = right_matrix[1,],
      y = right_matrix[2,],
      labels = paste("column ", 1:ncol(right_matrix), sep=""),
      pos = 3,
      col = 3:6
      )

# draw in tickmarks on the right_matrix vectors:
for( col_j in 1:ncol(right_matrix) ){
  
  vec <- right_matrix[,col_j]
  length_of_vec <- sqrt(sum(vec^2))
  unit_vec <- vec / length_of_vec
  x_tick_coords <- seq(-5,5,0.5) * (unit_vec/length_of_vec)[1]   
  y_tick_coords <- seq(-5,5,0.5) * (unit_vec/length_of_vec)[2]   
  points( x = x_tick_coords,
          y = y_tick_coords,
          pch = "|",
          cex = 0.5,
          col = 2+col_j
        )
  text( x = x_tick_coords,
        y = y_tick_coords,
        labels = seq(-5,5,0.5),
        pos = 1,
        cex = 0.5,
        col = 2+col_j
      )
}

```

Here is the original dataset (target matrix) again, for reference:

```{r lasttimeonemoretimecelebrate}
Y
```

and here is the same plot using the base R function biplot():

```{r andinR, fig.width=8, fig.height=8}
biplot( prcomp( Y ) )
```














