---
title: "Monotonic transformations of a Random Variable"
author: "Joseph Bolton"
output:
  html_document:
    df_print: paged
  html_notebook:
    code: hide
---

Some really good notes on this can be found at: 

http://www2.econ.iastate.edu/classes/econ671/hallam/documents/Transformations.pdf

# A simple example of the *Method of Transformations/Change of Variable Method*

Suppose that we have a continuous univariate random variable $X$ with an exponential distribution:

$$p_X(x) \quad=\quad \lambda e^{-\lambda x} \hspace{10mm} x \geq 0 \quad , \quad \lambda > 0 $$

```{r}
x_seq <- seq(0,100,0.01)
plot( x = x_seq,
      y = dexp( x_seq, rate = 0.1 ),
      type = "l",
      ylab = "f(x)",
      xlab = "x",
      main = bquote("exponential distribution function of X, with "*lambda*"=0.1"),
      lwd = 3,
      col = "blue"
    )
lines( x = x_seq,
       y = 0.1 * exp(-0.1*x_seq),
       col = "red"
     )
```

Suppose that we want to find the distribution of the random variable $Y$: a differentiable and monotonic (either increasing or decreasing) function of $X$, such as:

$$Y \quad=\quad \displaystyle\frac{1}{\sqrt{X}}$$

```{r}
x_seq <- seq(0,5,0.01)
y <- 1 / sqrt(x_seq)
plot( y ~ x_seq, 
      type = "l",
      xlab = "x",
      main = "Y=g(X)"
    )

```

The *change of variable* technique says that the distribution of $Y=g(X)$ is:

$$\begin{array}{lcl}
f_Y(y) &=& f_X\Big(g^{-1}(y)\Big) \quad\cdot\quad \Biggl| \displaystyle\frac{\partial \space \big[ g^{-1}(y) \big]}{\partial \space y}\Biggr| \\
&=& f_X\Big(\frac{1}{y^2}\Big) \quad \cdot \quad \Biggl| \displaystyle\frac{\partial \space \big[ \frac{1}{y^2} \space \big]}{\partial \space y} \Biggr| \\
&=& \lambda e^{-\lambda \Big(\frac{1}{y^2} \Big)} \quad \cdot \quad \Biggl| \displaystyle\frac{-2}{y^3} \Biggr| \\
&=& \displaystyle\frac{2\lambda e^{-\lambda \Big(\frac{1}{y^2} \Big)}}{y^3} \\ 
\end{array}$$

Where $g^{-1}(y)$ was calculated like this: 

$$\begin{array}{lcl}
Y &=& \displaystyle\frac{1}{\sqrt{X}} \\
\therefore \sqrt{X} &=& \displaystyle\frac{1}{Y} \\
\therefore X &=& \displaystyle\frac{1}{Y^2} \\
\therefore g^{-1}(y) &=& \displaystyle\frac{1}{y^2} \\
\end{array}$$

We can see that this calculation was correct by doing a quick simulation: 

```{r}
# simulate 100,000 values from X with parameter lambda=10: 
lambda <- 10
x <- rexp( 1e5L, rate = lambda)

# transform each simulated X into Y = 1/sqrt(X):
y <- 1 / sqrt(x)

# plot the expirical distribution of the simulated Y values using a histogram:
hist( y, breaks=1000, freq=FALSE, xlim = c(0,50) )

# draw the theoretical distribution curve on top of the empirical distribution:
y_seq <- seq(0, 100, 0.01)
lines( x = y_seq,
       y = (2*lambda*exp(-lambda/y_seq^2))/y_seq^3,
       col = 2
     )
```

