---
title: "What the Bayesian lm means"
author: "fuck yeah"
output:
  html_document:
    df_print: paged
---

```{r }
library(tidyverse)
#library(epade)
library(knitr)
```

prior distribution of $\beta_0$:

$$\begin{array}{lcl} 
Pr[\beta_0 = -1] &=& 0.4 \\
Pr[\beta_0 = 1] &=& 0.6 \\
\end{array}$$

```{r echo=FALSE}
tibble( beta0 = c(-1, 1),
        prob = c(0.4, 0.6) 
      ) %>% 
  ggplot( data = .,
          aes( x = beta0,
               y = prob 
              )
          ) +
  geom_bar( stat = "identity" ) +
  labs( title = bquote( "distribution of"~beta[0]) )
```

prior distribution of $\beta_1$:

$$\begin{array}{lcl} 
Pr[\beta_1 = -1] &=& 0.3 \\
Pr[\beta_1 = 0] &=& 0.5 \\
Pr[\beta_1 = 1] &=& 0.2 \\
\end{array}$$

```{r echo=FALSE}
tibble( beta1 = c(-1, 0, 1) %>% as.factor(),
        prob = c(0.3, 0.5, 0.2) 
      ) %>% 
  ggplot( data = .,
          aes( x = beta1,
               y = prob 
              )
          ) +
  geom_bar( stat = "identity" ) +
  labs( title = bquote( "distribution of"~beta[1]) )
```

error distribution:

```{r echo=FALSE}
tibble( e = -5:5,
        prob = c(1,1,2,5,8,15,8,5,2,1,1)/sum( c(1,1,2,5,8,15,8,5,2,1,1) )
      ) %>% 
  ggplot( data = .,
          aes( x = e,
               y = prob 
              )
          ) +
  geom_bar( stat = "identity" ) +
  geom_text( aes( x = e,
                  y = prob,
                  label = round(prob, digits=3),
                  vjust = -0.3
                )
           ) +
  theme_bw() + 
  labs( title = bquote("distribution of the errors"~epsilon) )
```


we have that:

$$y \quad=\quad \beta_0 \quad + \quad \beta_1 x \quad + \quad \epsilon$$

our observed data is:
```{r echo=FALSE}
set.seed(5213)
x <- c(-2, 1, 2)         # sample( -2:2, size=3, replace=TRUE ) 
b0 <- -1
b1 <- 1
e <- c(0,0,1)

# sample( x = c(-2, -1, 0, 1, 2),
#              prob = c(0.05, 0.2, 0.5, 0.2, 0.05),
#              size = 3,
#              replace = TRUE
#            )

y <- b0 + b1*x + e
observed_data <- tibble(x,y) %>% arrange(x,y) %>% filter( abs(y)<4 )

observed_data

plot( y ~ x )
```

So, we can see that in the scenario that we have specified: 

$$Pr\Big[y = -3 \space \Bigl|\space \beta_0=-1, \beta_1=0, x=-1\Big] \quad=\quad 0.01$$

Since we know the system which created our data, we can make probability statements about each $y_i|\beta_0,\beta_1,x_i$.

Since we assume that every row of our data is an independent draw from our data-creation process, we can make probability statements about our whole dataset:

$$\begin{array}{lcl}
Pr\Big[ y=\begin{bmatrix}-3\\-2\\-1\\-1\\-1\\1\\2\\2\end{bmatrix} \space \Bigl| \space \beta_0=-1, \beta_1=0, x = \begin{bmatrix}x\\x\\x\\x\end{bmatrix} \Big] &=& 
Pr\Big[y=-3|\beta_0=-1, \beta_1=0, x=-2\Big] \times \\ 
&\space& Pr\Big[y=-2|\beta_0=-1, \beta_1=0, x=-1\Big] \times \\
&\space& . \\
&\space& . \\
&\space& . \\
&\space& \times Pr\Big[y=2|\beta_0=-1, \beta_1=0, x=2\Big]  \\
\end{array}$$


$$\begin{array}{lcl}
Pr(\beta_0, \beta_1 | y, x) &\propto& Pr(y|\beta_0,\beta_1,x)\times Pr(\beta_0,\beta_1) \\
                            &=& Pr(y|\beta_0,\beta_1,x)\times Pr(\beta_0)\times Pr(\beta_1) \\
                            &=& \begin{cases}
                                      \times \quad\quad \text{for } \beta_0=0, \space \beta_1=-2 \\
                                      \times \quad\quad \text{for } \beta_0=1, \space \beta_1=-2 \\
                                      \times \quad\quad \text{for } \beta_0=5, \space \beta_1=-2 \\
                                      \times \quad\quad \text{for } \beta_0=0, \space \beta_1=1 \\
                                      \times \quad\quad \text{for } \beta_0=1, \space \beta_1=1 \\
                                      \times \quad\quad \text{for } \beta_0=5, \space \beta_1=1 \\
                                      \end{cases}
\end{array}$$

```{r}

likelihood_y_i <- function( y_i, x_i, b0, b1 ){
  
  tibble( error = -5:5,
          prob = c(1,1,2,5,8,15,8,5,2,1,1)/sum( c(1,1,2,5,8,15,8,5,2,1,1) )
        ) %>% 
    mutate( y = b0 + b1*x_i + error ) %>% 
    filter( y == y_i ) %>% 
    summarise( prob = sum(prob) ) %>% 
    pull( prob )
  
}

theoretical_probs <- 
  expand.grid( 
                 b0 = c(-1, 1),
                 b1 = c(-1, 0, 1)
              ) %>% 
      as_tibble() %>% 
    
      # add b0 prior probs:
      left_join( tibble( b0 = c(-1, 1),
                         b0_prob = c(0.4, 0.6) 
                       ),
                 by = "b0"
                ) %>%
      # add b1 prior probs:
      left_join( tibble( b1 = c(-1, 0, 1),
                         b1_prob = c(0.3, 0.5, 0.2) 
                       ),
                 by = "b1"
                ) %>% 
    
      mutate( prior_prob = b0_prob * b1_prob ) %>% 
    
      #select( -b0_prob, -b1_prob ) %>% 
    
      mutate( likelihood = map2_dbl( .x = b0,
                                     .y = b1,
                                     .f = function(beta0, beta1){
                                       
                                       observed_data %>% 
                                          mutate( likelihood_y_i = pmap_dbl( list(X = x,
                                                                                  Y = y,
                                                                                  B0 = beta0,
                                                                                  B1 = beta1
                                                                                  ),
                                                                             .f = function(X,Y,B0,B1){
                                                                               likelihood_y_i( y_i=Y,
                                                                                               x_i=X,
                                                                                               b0=B0,
                                                                                               b1=B1 ) 
                                                                                 }
                                                                            )
                                          ) %>% 
                                         summarise( likelihood_y = prod(likelihood_y_i) ) %>% 
                                         pull( likelihood_y )
                                    }
                                  )
            ) %>% 
    mutate( unscaled_prob = prior_prob * likelihood ) %>% 
    mutate( scaled_prob = unscaled_prob / sum(unscaled_prob) ) 
    
theoretical_probs
```

```{r}

conditional_match <- tibble( b0=NA, b1=NA ) %>% slice(0)

for( i in 1:1e7L ){
  
  if( i%%1e4L == 0 ){ print(i) }
  
  b0 <- sample( c(-1, 1), prob = c(0.4, 0.6), size=1 )
  b1 <- sample( c(-1,0, 1), prob = c(0.3, 0.5, 0.2), size=1 ) 

  e <- sample( -5:5,
               prob = c(1,1,2,5,8,15,8,5,2,1,1)/sum( c(1,1,2,5,8,15,8,5,2,1,1) ),
               size = nrow(observed_data),
               replace = TRUE
             )
  
  y <- b0 + observed_data$x*b1 + e
    
  if( all.equal(y, observed_data$y) == TRUE ){
    conditional_match <- bind_rows( conditional_match, tibble(b0=b0, b1=b1) )
  }
}

conditional_match %>% 
  group_by( b0, b1) %>% 
  tally() %>% 
  ungroup() %>% 
  mutate( simulated_prob = n/sum(n) ) %>% 
  left_join( theoretical_probs %>% 
                select( b0, b1, scaled_prob )   
           ) %>% 
  rename( theoretical_prob = scaled_prob )

```